### 2021-10-15 draft.Rmd
initial commit

---
title: "Extended early adulthoods"
author: "Kevin Kiley & Nicolas Restrepo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The last few decades have seen dramatic changes in the expectations surrounding the transition from adolescence to adulthood. The trajectories between these two stages have become more varied, blurring the institutional boundaries that used to demarcate them. Individuals are more likely to return to higher education after graduating, and they tend to switch professional tracks more often, instead of being locked-in in a trade. The period between leaving the parental nest and "settling down" has extended. The primary example here is the increasing median age at first marriage, especially for women. This "age of independence" - characterized by exploration and a degree of detachment from social institutions - appears to be becoming longer and more diverse.

This stage of the life-course has gained increasing attention from sociologists because work - from varied sub-fields - highlights the pivotal role that it might play in processes of social change. Young adults nowadays are more likely to be away from familial supervision, which can make them prone to break with social mores. Their relative distance from certain social institutions - like home ownership - means that they can risk more and, therefore, they tend to inhabit the front lines of social movements. Additionally, recent work on attitudinal change suggests people above 25 rarely change their minds. Young adulthood then might also represent the tail-end of our formative windows, where we are still open to revisiting our habits and beliefs.

This paper seeks to examine this latter relationship. The expanding duration of young adulthood provides an important opportunity to study how belief-updating unfolds across an individual's lifetime. The finding that a period of exploration during one's early years is followed by relative stability is recurrent in many fields. A question that is interesting for sociologists is whether the length of these formative windows varies across socio-historical contexts. The evidence that the life-course and its stages are socially patterned would make that conclusion intuitive but there has been little empirical work that tries to adjudicate this question. However, if formative windows are dependent on social contexts - and not entirely maturational - then expanded periods of young adulthood should also result in higher rates of belief updating among people in that age group in more recent cohorts. In this paper, we test exactly this proposition.

Finding traces of a relationship between extended periods of young adulthood and higher rates of updating would provide important insights for understanding change, within an agent's lifetime and at the aggregate level. It would, for instance, constitute novel evidence that individuals' likelihood to update their beliefs is not just ontogenetic but rather depends on the social contexts in which they are situated. If this is the case, then sociologists are well-positioned to contribute to ongoing interdisciplinary work about formative windows and learning schedules. Furthermore, if longer young adulthoods result in extended periods of exploration, then there is another mechanism through which this "age of independence" affects social change: by letting individuals remain open to novel information for longer.

### 2023-02-06 draft.Rmd
modifies script to add what Kevin wrote and uploads initial data

@@
 author: "Kevin Kiley & Nicolas Restrepo"
-output: html_document
+fontsize: 12pt
+output:
+  pdf_document: default
+  html_document: default
+header-includes:
+  - \usepackage{setspace}
+  - \doublespacing
+  - \usepackage{lineno}
+  - \linenumbers
+  - \def\linenumberfont{\normalfont\tiny\sffamily}
+bibliography: efw_bib.bib
+abstract: "Young adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of attitude change.
+This probability has often been attributed to the social conditions of adolescence.
+Over the past 70 years, changes in the timing of major life course events such as marriage have transformed and extended this life stage.
+We leverage these changes to explore whether these life course changes, as well as broader social changes, have affected how likely people are to make durable changes of attitudes.
+We find strikingly little difference in the rates of any kind of stable opinions and opinion change among young people in different eras.
+People under 30 are similarly different from people over 30 in all eras.
+The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates."
 ---
@@
 
-The last few decades have seen dramatic changes in the expectations surrounding the transition from adolescence to adulthood.
-The trajectories between these two stages have become more varied, blurring the institutional boundaries that used to demarcate them.
-Individuals are more likely to return to higher education after graduating, and they tend to switch professional tracks more often, instead of being locked-in in a trade.
-The period between leaving the parental nest and "settling down" has extended.
-The primary example here is the increasing median age at first marriage, especially for women.
-This "age of independence" - characterized by exploration and a degree of detachment from social institutions - appears to be becoming longer and more diverse.
+Adolescence and young adulthood are typically identified in social science literature as “impressionable years” – a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples’ core ideas and attitudes are formed [].
+Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other life stages, with opinions becoming more stable once these life stages have passed [].
+This pattern, in which young people are open to attitude change and become resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1962; @mannheiumYEAR].
 
-This stage of the life-course has gained increasing attention from sociologists because work - from varied sub-fields - highlights the pivotal role that it might play in processes of social change.
-Young adults nowadays are more likely to be away from familial supervision, which can make them prone to break with social mores.
-Their relative distance from certain social institutions - like home ownership - means that they can risk more and, therefore, they tend to inhabit the front lines of social movements.
-Additionally, recent work on attitudinal change suggests people above 25 rarely change their minds.
-Young adulthood then might also represent the tail-end of our formative windows, where we are still open to revisiting our habits and beliefs.
+However, why people in these life stages appear to show more frequent opinion change relative to older people is unclear.
+Work in biology, psychology, and cultural evolution shows that peoples’ cognition changes in ways that might make them more resistant to new information as they age [].
+Explanations in social sciences, in contrast, tend to highlight the unique social features of these life stages – high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing “new experiences” – that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information [].
+While both mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.
 
-This paper seeks to examine this latter relationship.
-The expanding duration of young adulthood provides an important opportunity to study how belief-updating unfolds across an individual's lifetime.
-The finding that a period of exploration during one's early years is followed by relative stability is recurrent in many fields.
-A question that is interesting for sociologists is whether the length of these formative windows varies across socio-historical contexts.
-The evidence that the life-course and its stages are socially patterned would make that conclusion intuitive but there has been little empirical work that tries to adjudicate this question.
-However, if formative windows are dependent on social contexts - and not entirely maturational - then expanded periods of young adulthood should also result in higher rates of belief updating among people in that age group in more recent cohorts.
-In this paper, we test exactly this proposition.
+Clarifying the mechanism that drives the frequency of opinion change during this formative period is important for understanding the broader process of social change and cultural evolution.
+If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader process of macro-social change [].
+Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2020; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than in older eras, diminishing the importance of “cohorts” for social change.
+If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social change in the same way.
 
-Finding traces of a relationship between extended periods of young adulthood and higher rates of updating would provide important insights for understanding change, within an agent's lifetime and at the aggregate level.
-It would, for instance, constitute novel evidence that individuals' likelihood to update their beliefs is not just ontogenetic but rather depends on the social contexts in which they are situated.
-If this is the case, then sociologists are well-positioned to contribute to ongoing interdisciplinary work about formative windows and learning schedules.
-Furthermore, if longer young adulthoods result in extended periods of exploration, then there is another mechanism through which this "age of independence" affects social change: by letting individuals remain open to novel information for longer.
+In this paper, we seek to further elaborate this debate by exploring how changes in the broader social structure and changes in the timing of life course events since the 1950s have affected rates of change in political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population.
+Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one’s parents’ home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann; @bruckner2005].
+Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2007].
+This suggests both a lengthening and a intensification of this stage of the life course in ways that should increase the rates at which people make changes of opinion.
+
+We take advantage of this change over time to explore whether young adults of different cohorts demonstrate more or less change in general political opinions over time relative to other members of the population.
+We compare rates of opinion change in comparable general political attitudes across four eras – 1956-1960; 1972-1976; 1992-1996; and 2006-2010 – and between young people and old people in these eras to evaluate a range of competing hypotheses about what accounts for the divergence in attitude behaviors over the life course.
+
+Broadly, we find that the probability of political opinion change in people under 30 have not increased as the “early adulthood” window has lengthened and the conditions of this life stage have changed.
+Across eras, young people make political opinion changes at remarkably similar rates, with social transitions such as marriage, having children, and completing education playing little role in the rates at which these opinion changes happen.
+While these findings do not provide conclusive evidence of one mechanism over the other, they confine the range of potential theories of attitude development and the nature of political opinions.
+In particular, they suggest that there is something relatively timeless – potentially biological and cognitive factors – that accounts for increased opinion variance in young people.
+
+## Impressionable Years / Opinion Change and Stability
+
+The idea that the early stages of the life course are especially influential in the process of attitudinal formation is dominant across contemporary accounts of socialization (Dinas 2014; @guhin2021).
+The argument – at times implicit – is that the circumstances in which people are brought up leave an enduring and traceable mark on their worldviews (Gerber and Green 1998).
+In the study of political socialization, this perspective enjoys a longstanding history.
+“Generational effects” are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups (Mannheim 1970; Bartels and Jackman 2014).
+However, the notion of “impressionable years” is more general.
+The burgeoning literature on cohort effects rests on the assumption that our early years play key role in shaping not just political opinions, but peoples’ cognition on many dimensions (Elder 2018).
+Their early years will leave a mark that will be predictive of their life outcomes, well beyond those impressionable years have passed.
+Thus, the notion of a formative period – at the early stages of the life-course – underpins diverse and longstanding lines of research.
+
+The “impressionable years” hypothesis – that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age – has often been tested against other theories of opinion change over the life course.
+The most common are an aging stability or increasing persistence hypothesis that argues that openness to change gradually declines over time and a life stages hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visserkrosnick1998].
+While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnickalwin1989; @alwinkrosnick1991; @searsfunk1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality [@wagner2019].
+More generally, recent research has found that, with a couple notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S.
+adults [@kileyvaisey], and Vaisey and Lizardo [-@vaiseylizardo2016] find that cohort explains much more variance than period for most items in the General Social Survey, suggesting a strong impressionable years pattern.
+
+In recent years, researchers have attempted to use new approaches to quantify “openness” to social influence across the life course, especially around politics [@bartelsjackmanYEAR; @ghitzagelman].
+This work has produced remarkably similar patterns with openness peaking from a person’s mid-teens through age 25 and declining quickly after that, lending more credence to the “impressionable years” hypothesis.
+However, this work tends to assume a constant age-based rate of openness across cohorts – that 20-year-olds in 1950 are open to influence at the same rate as 20-year-olds in 2010 – which has not been tested.
+
+## Theoretical Models
+
+Why is it research consistently finds the period of time between 14 and 25 to be distinctly “impressionable years?” At the outset, it is important to note that changing one’s mind involves at least two distinct processes that can be difficult to disentangle: exposure to new ideas and openness to change conditional on exposure to new ideas.
+For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change.
+Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one’s mind as a function of exposure does not, producing the same pattern.
+It could also be the case that both change as a function of age.
+Any patterns we observe should be interpreted as coming from changes in one or both of these processes.
+In this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.
+
+### Biological Mechanisms
+
+The first potential mechanism for changes in attitude stability over time is biological: that the development of the of human brain as people age makes it more difficult for people to make changes of opinions after some point in their life.
+This approach tends to suggest that human cognition works similar to Bayesian updating (Bartels and Jackman 2014; Sherratt and Morand-Ferron 2018; Gopnik 2020).
+In this framework, humans weight information equally, whether they encounter it during adolescence or middle age.
+However, given that by the time people reach adulthood they have already encountered a lot of information, each new piece will have less of an impact (Bartels and Jackman 2014).
+The distinctive thing about our early years is simply that they happen early.
+They prove so influential because the incremental effect of new information diminishes as we age.
+
+It is also possible that there is something unique about human cognition during these periods: that agents ascribe more weight to the information acquired within them.
+Maturational arguments about the increased plasticity during our early years are consistent with this position (Mata and von Helversen 2015).
+Our brains exhibit more malleability during childhood and adolescence (Mata and von Helversen 2015).
+This is consistent with the fact that these early formative periods have been observed across species and contexts.
+It is not only human children that exhibit a heightened capacity to adapt.
+In a study of social learning among birds, for instance, Aplin et al (2017) show that younger individuals – unlike their older counterparts – were consistently able to retool their habits in light of new information.
+
+Though the Bayesian updating explanation is simple and elegant, research across different disciplines has produced evidence that supports the existence of these periods of heightened sensitivity.
+In light of empirical evidence about these sensitive windows, there has been extensive modeling work about their evolution.
+This line of research seeks to understand why periods of heightened sensitivity would emerge and what contexts would facilitate their emergence (Fawcett and Frankenhuis 2015).
+The central argument is that the emergence and timing of these sensitive windows depends on the interaction between the access to information, the reliability of information, and the stability of the environment (Frankenhuis and Walasek 2020).
+Early formative periods are likely to evolve in specific configurations: when environments are unstable, and information is plentiful yet unreliable (Frankenhuis and Walasek 2020).
+In these circumstances, it makes sense for agents to explore early on, to find something that works and then to stick to strategies that have worked well in the past.
+Access to different types of information might vary across time and this might explain the emergence of sensitive periods at later stages of the life-course (Walasek, Frankenhuis, and Panchanathan 2021).
+For example, the importance of musical tastes and fashion choices becomes important around adolescence, and, at this stage, we are awash with – at best contradictory – information about what one should or should not like, should or should not wear.
+Sensitive periods around the development of these tastes might peak around this period, when there are more signals and when the payoffs are more salient.
+
+The thrust of the biological arguments is that, through millennia of evolution, humans have developed a cognitive architecture that makes them relatively open changing dispositions at younger ages and relatively resistant at older ages.
+However, while biological mechanisms can shape how individuals respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change.
+For that, we must turn attention to the structuring of the social world.
+
+### Life Course // Social Role Mechanisms:
+
+Humans’ early years might be distinct in terms of cognitive plasticity, but they are also socially distinct: they are characterized by sustained social learning and punctuated with constant changes of social contexts.
+Here we briefly consider some of the ways that the “impressionable years” differ in ways that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.
+
+First, people in this age group tend to move through social contexts much more quickly than people in other age groups.
+In the 11-year window from ages 14 to 25, in which people seem to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year in these contexts potentially exposing people to new classmates, teachers, and other role models, peers, and sources of information.
+People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons [is this true?].
+Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups.
+
+Second, people in the “impressionable years” window spend their time in different social contexts than people in middle age.
+Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55.
+These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions.
+For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, gender egalitarianism, and racial inequality [@campbell2016; wodtke2018].
+Social networks also differ in this time period.
+Previous work has found that young peoples’ networks tend to be larger and include more friends than kin, for example [@marsden1987].
+It could be the case that larger “weak tie” networks expose younger people to more new information that drives opinion change [@granovetter1975; burt2005].
+
+Third, these are periods in which exploration and self-discovery are expected and encouraged; something that should not be taken for granted given social variation in norms around when maturity is reached [@cunningham2012; @elder1994].
+
+### Updating conditional on exposure
+
+While social conditions might expose young people to new information at higher rates, it might also make it easier for them to act on these new sources of information in ways that facilitate opinion change.
+
+Previous work on stability of attitudes over the life course suggests that people tend to value stability in figures of authority.
+As people move into roles with more authority in mid-life, both in family and occupational contexts, they feel this expectation from others and act accordingly, closing themselves off to updating beliefs [@eaton2009].
+This potentially explains another phenomenon with some evidence behind it: that people appear to open up to changing their minds again as they enter post-retirement [@danigelis].
+At this point, peoples’ authority diminishes, potentially opening them up to change later in life.
+
+Similarly, mid-life is associated with more homogenous networks [@marsden1985; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@vissermirible].
+If young people exist in less dense, more heterogeneous social networks, they might it easier to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks [], or close off parts of their lives to people who might not be supportive of new attitudes.
+
+Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the “impressionable years ” are principally an effect of how we have chosen to structure the life course, or the interaction between these life course stages and biological mechanisms.
+
+## Changing Life Course
+
+Contemporary historical trends offer the opportunity to disentangle the role of socio-cultural and  biological factors in explaining why we observe “impressionable years.” In the U.S., the last decades have seen considerable changes around adolescence and young adulthood (Buchmann 1989; Rosenfeld 2007; Rosenfeld and Kim 2005).
+The expansion of higher education and increased possibilities for mobility are among the many social processes that have led to changing expectations for the trajectories that young adults should follow (Rosenfeld 2007).
+In general, as Buchmann  (1989) notes, life trajectories have become less predictable.
+The “script” of young adulthood has become less rigid: it might involve prolonged higher education, moving to a different city, or dating without a stable romantic partner.
+The routes that people are taking into adulthood are increasingly diverse.
+For the purposes of this paper, there are two accompanying trends that are particularly important: the elongation and intensification of young adulthood.
+
+### Timing
+
+The period between adolescence and “settling down” seems to be getting longer, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts.
+Buchmann (1989) notes that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood.
+In the U.S., the median age at first marriage has consistently risen in the last decades and so has the median age at which families are having their first child (Rosenfeld 2007).
+We also see higher rates of non-marital childbearing across all demographic groups [@cherlin].
+Thus, the markers of the transition into adulthood seem to be occurring later in the life-course and potentially in different orders.
+Therefore, the definition of what counts as our “impressionable years” might be extending.
+This, in turn, could have an effect in how long exploration lasts throughout the life-course.
+If this is the case, we should expect to see more attitudinal updating among individuals in age-ranges that, in previous cohorts, had displayed mainly stability.
+In other words, people who were born say in the mid 20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still exhibit attitudinal updating when they reach that age.
+Adolescents are taking longer to move through the rituals of adulthood, and this might in turn elongate the period in which exploration and self-discovery are expected.
+
+### Changing Impressionable Years
+
+Another second trend relates to just how much exploration young adults can engage in during these “impressionable years.” In his work, Rosenfeld (2007) explores the rapid rise in same-sex and interracial marriages in the U.S.
+His argument is that one of the processes underlying these changes is that young adults are increasingly escaping the controlling gaze of their parents.
+When young people live with – and/or financially dependent on – their parents, the latter have more say about their children’s choice of partner.
+The increased independence and mobility that young adults have enjoyed since the mid 20th century means that they have been able to engage in courtship outside the gaze of their parents, thus making choices that would not have been sanctioned by them.
+Rosenfeld’s (2007) work highlights an important insight: exploration might not be only a product of how sensitive we are to novel information but also how – and whether – we can act upon such information.
+The social institutions where we spend our impressionable years might shape how much exploration we can engage in.
+An increase in independence then might entail more exploration.
+This assertion in turn engenders a prediction: we should see more attitudinal updating among contemporary young adults than their older counterparts, who tended to be closer to the grip of familial control, even if they have met similar life course milestones.
+The contemporary social organization of young adulthood, which has been developing for the last few decades, might give rise to more meaningful attitudinal changes by providing more leeway to deviate from established social norms.
+
+This approach suggests that it is not the timing of life events that is important for rates of change, but the broader social conditions of early adulthood that might make people more malleable in recent cohorts even net of changes.
+In other words, unmarried, childless 25-year-old men and women in 1956 might have experienced a set of social constraints that unmarried, childless 25-year-old men and women in 2008 do not experience, leaving the latter much more open to developing new opinions than their predecessors.
+
+## Hypotheses
+
+Sociological work then highlights that there are principled reasons to believe that length and intensity of formative periods might be affected by social factors, and yet this question remains unexplored.
+The importance of youth remains the backbone of sociological theories of socialization and there has been work trying to outline the cognitive assumptions that would lead to marked influence of our early years.
+While the evidence for these formative periods has accrued, there has been no research on the social variability of these periods.
+Any evidence of differences in the rate of attitudinal updating among people of similar ages across different social contexts would suggest that sensitive periods are not only maturational or a result of “tallying up” information.
+In turn, finding similar patterns of attitudinal updating across widely different social contexts would lend further credence to the idea that our early years are “impressionable” either because we weight information more heavily during these periods or because we are cementing our opinions, regardless of the social contexts we happen to find ourselves in.
+
+The argument so far implies three hypotheses that we will test in our study.
+Given historical changes in the meaning of adolescence, we hypothesize that we will see differences in the rates of attitudinal updating that the same age-groups across different cohorts will exhibit.
+The first hypothesis then is non-directional:
+
+H1: There will be differences in the patterns of attitudinal updating from individuals in the same age groups but in different cohorts.
+
+The second hypothesis is based on Buchmann’s work and it concerns the fact that young adults are increasingly postponing the rituals associated with the transition into adulthood.
+Thus, we posit that more recent cohorts should display more exploration well into their twenties than previous cohorts.
+
+H2: Young adults in recent cohorts should exhibit active attitudinal updating later than their counterparts of similar ages from older cohorts.
+
+The last hypothesis in turn relates to Rosenfeld’s ideas.
+Young adults in recent cohorts have enjoyed more independence and, therefore, might have had more opportunities to substantially change their beliefs and habits.
+We hypothesize then that:
+
+H3: Young adults in recent cohorts should exhibit higher rates of attitudinal updating than people of a similar age in previous generations.
+
+Finally, it could be case that an explanation for overall change in rates of the population are explained by factors that we have not explicitly considered here.
+It could be the case that rates of opinion change are tied to things like racial and ethnic identity, religious identity, or geographic region, all of which have changed significantly in the same window.
+If we identify broad shifts in attitude change that are not explained by , this suggests .
+
+## Kevin's Hypotheses
+
+Kevin’s Hypotheses:
+-	Broadly speaking the “impressionable years” hypothesis would suggest that: People under 30 will demonstrate higher rates of attitude change than people over 30 across all cohorts.
+-	Both the “extended adolescence” and the “intensification of young adulthood” theories would both suggest that people under 30 will demonstrate higher rates of attitude change (relative to people over 30) in more recent panels than people in earlier panels.
+This could be because people 25-30 make changes at higher rates than they were in the past or because people 18-25 and 25-30 make changes of attitudes at higher rates.
+-	Adjudicating these might require separating people 18-25 from 25-3X. But this seems tough since the numbers are too small in most of the panels to separate these. But I might try.
+-	The “timing of life course” events hypothesis would seem to suggest that: “Controlling” for marital status and having children [and maybe other stuff] will reduce the difference between people under 30 and people over 30.
+o	In other words, net of age, we should see that marriage and having children makes people more stable opinion holders (maybe also something about education).
+So I’m going to try to include these variables in some analyses.
+-	The “intensity of early adulthood” argument would seem to suggest that: Controlling for differences in life course stages, younger people in more recent panels will still demonstrate higher rates of attitude change than younger people in older cohorts.
+
+## Kevin’s Other Thoughts
+-	There’s the possibility that people over 30 have increased their likelihood of making a change.
+First, the social environment is significantly different: people are likely exposed to new information at higher rates.
+Second, deinstitutionalization of the life course and increases in divorce rates, etc.
+might mean that older people are more free to make opinion changes than they were in the past too.
+-	Similarly, there’s the “life stages” hypothesis (its not that early life is distinctly open, it’s that middle age is distinctly closed) that has some support.
+It might be worth either holding out or including a separate coefficient for people > 60 or something.
+-	There’s a possibility that we see differences in rates of change between cohorts that are not explained by life course stuff.
+One thing that I’ve been thinking is like a “population composition” thing.
+Black and Hispanic people are less stable opinion holders than white people on a number of issues (something I’ve been thinking about in my spare time).
+People who attend religious services are also more stable opinion holder than other people.
+Changes in this stuff might be worth considering.
+Or maybe it’s part of the discussion of the results.
+
+## Data
+
+To test the hypotheses outlined above, we require data on within-individual change in comparable attitudes over a stable window of time across eras.
+To do this, we focus on general political beliefs, including views on the role of government, political ideology, and partisan identification, that have been asked in panel surveys.
+This is in part a practical matter, as these are the kinds of questions that have been included in panel studies that date back to the mid-1950s, giving us the largest window of cohorts to explore.
+However, these kinds of questions have frequently been used to identify the “impressionable years” [].
+
+We draw on data from four different survey panels: the 1956-60, 1972-76, and 1992-96 American National Election Study panels , and the 2006-10 General Social Survey Panels.
+All four surveys are nationally representative samples of non-institutionalized adults in the United States.
+All four panels survey respondents three times over a four-year window, making the durations comparable.
+
+### Political Opinions
+
+We sought to identify questions that tapped general political sentiments that were echoed in at least one other survey.
+We avoided questions that asked about specific policies or political figures, since they were often not replicated across panels.
+We also focused principally on questions asking about the role of the federal government in domestic affairs, rather than international concerns.
+
+No questions are directly comparable across each of the panels.
+For this reason, we include a broad range of questions, in hopes of establishing general patterns.
+Table XXX outlines the questions used in this analysis, their specific wording, and response options.
+
+### Life Course Events
+
+To evaluate changes in the life course, and their relationship to attitude stability, we focus on a set of events that are traditionally accepted as marking the transition to adulthood.
+
+While question wording differs slightly across surveys, all surveys include a measure of whether people are married, whether they have any children, and .
+
+For example, while marriage rates have declined , rates of cohabitation (romantic partners living together without being married) have increased.
+
+## Analytical Strategy
+
+It is not obvious what it means for a person to change their opinion in a survey, and debates about how to measure “change” in opinions often focuses on differences in the approach to measurement [@converse; @achen1975; ].
+Rather than adopt a single approach to quantify the rates of change across questions, we use a variety of methods with different assumptions about what “change” entails and attempt to triangulate the findings to see if patterns are robust across these assumptions.
+
+The first approach to measuring change in political opinions is to assume that all change is meaningful and measure the total “distance” people move across a scale during the time period in question.
+
+A second approach is to a prioi define what kinds of opinion behavior count as “change” and define what counts as “stable opinions” and compare their prevalence across age groups and panels.
+For example, we might count any response pattern in which people stay on the same side of a midpoint (“agree”-“strongly agree”-“agree”) as a stable opinion and any response option that crosses the midpoint (“agree”-“disagree”-“disagree”) as representing a change of opinion.
+
+A third and related approach is to classify people who change their responses within a specified range, such as changing less than two scale points (e.g., “agree”-“neither”-“agree”), as expressing stability while counting other respondents as making changes of opinion.

### 2023-02-09 Analyses/build_long_dataframe.R
begins script to build the dataset necessary to test hypotheses on different was of conceptualizing change

@@ -1,0 +1,591 @@
+# Packages ----
+library(tidyverse)
+library(patchwork)
+
+# Load Data ----
+load("~/Documents/extended_formative_windows/Data/clean_anes5.Rdata")
+load("~/Documents/extended_formative_windows/Data/clean_anes7.Rdata")
+load("~/Documents/extended_formative_windows/Data/clean_anes9.Rdata")
+load("~/Documents/extended_formative_windows/Data/clean_g6.Rdata")
+
+# PARTYID - ANES56 ----
+# Strict
+a5_pid_s <- clean_anes5 %>%
+  mutate(id = 1:nrow(.)) %>%
+  select(id, age_1, contains("party")) %>%
+  filter(
+    partyid_1 != 7,
+    partyid_1 != 8,
+    partyid_1 != 9,
+    partyid_2 != 7,
+    partyid_2 != 8,
+    partyid_2 != 9,
+    partyid_3 != 7,
+    partyid_3 != 8,
+    partyid_3 != 9
+  ) %>%
+  mutate(
+    change_w2 = abs(partyid_2 - partyid_1),
+    change_w3 = abs(partyid_3 - partyid_2),
+    total_change_s = change_w2 + change_w3, 
+    any_change_s = if_else(total_change_s >= 1, 1, 0), 
+    crossed_midpoint_s = case_when(
+      partyid_1 < 3 & partyid_2 > 3 ~ 1,
+      partyid_1 > 3 & partyid_2 < 3 ~ 1,
+      partyid_2 < 3 & partyid_3 > 3 ~ 1,
+      partyid_2 > 3 & partyid_3 < 3 ~ 1,
+      partyid_1 < 3 & partyid_3 > 3 ~ 1,
+      partyid_1 > 3 & partyid_3 < 3 ~ 1,
+      TRUE ~ 0
+    ), 
+    two_points_s = if_else(
+      abs(partyid_3-partyid_1) >= 2, 1,0
+    )
+  ) %>% 
+  select(id, age_1, total_change_s, any_change_s, crossed_midpoint_s, two_points_s) %>% 
+  mutate(variable = "partyid")
+
+# Recodes 7s and 8s into 3s to be less strict 
+a5_pid_l <- clean_anes5 %>%
+  mutate(id = 1:nrow(.)) %>%
+  select(id, age_1, contains("party")) %>%
+  filter(
+    partyid_1 != 9,
+    partyid_2 != 9,
[…]

### 2023-02-09 Analyses/build_long_dataframe.R
Adds the ANES74 to the coded variables

@@ -587,5 +587,635 @@ a5_private_l <- clean_anes5 %>%
          two_points_l) %>%
   mutate(variable = "private")
 
-# PARTID - ANES74
+# PARTYID - ANES74 ----
 
+# This variable seems to be messed up here because we have a *vast* majority of NAs / DKs
+
+# Strict
+a7_pid_s <- clean_anes7 %>%
+  mutate(id = 1:nrow(.)) %>%
+  select(id, age_1, contains("party")) %>%
+  filter(
+    partyid7_1 != 7,
+    partyid7_1 != 8,
+    partyid7_1 != 9,
+    partyid7_2 != 7,
+    partyid7_2 != 8,
+    partyid7_2 != 9,
+    partyid7_3 != 7,
+    partyid7_3 != 8,
+    partyid7_3 != 9
+  ) %>%
+  mutate(
+    change_w2 = abs(partyid7_2 - partyid7_1),
+    change_w3 = abs(partyid7_3 - partyid7_2),
+    total_change_s = change_w2 + change_w3, 
+    any_change_s = if_else(total_change_s >= 1, 1, 0), 
+    crossed_midpoint_s = case_when(
+      partyid7_1 < 3 & partyid7_2 > 3 ~ 1,
+      partyid7_1 > 3 & partyid7_2 < 3 ~ 1,
+      partyid7_2 < 3 & partyid7_3 > 3 ~ 1,
+      partyid7_2 > 3 & partyid7_3 < 3 ~ 1,
+      partyid7_1 < 3 & partyid7_3 > 3 ~ 1,
+      partyid7_1 > 3 & partyid7_3 < 3 ~ 1,
+      TRUE ~ 0
+    ), 
+    two_points_s = if_else(
+      abs(partyid7_3-partyid7_1) >= 2, 1,0
+    )
+  ) %>% 
+  select(id, age_1, total_change_s, any_change_s, crossed_midpoint_s, two_points_s) %>% 
+  mutate(variable = "partyid")
+
+# Recodes 7s and 8s into 3s to be less strict 
+a7_pid_l <- clean_anes7 %>%
+  mutate(id = 1:nrow(.)) %>%
+  select(id, age_1, contains("party")) %>%
+  filter(
+    partyid7_1 != 9,
+    partyid7_2 != 9,
+    partyid7_3 != 9
+  ) %>%
+  mutate(
+    across(contains("party"), 
[…]

### 2023-02-15 Analyses/build_long_dataframe.R
Runs models, builds the report for different ways of measuring change

@@ -1,6 +1,8 @@
 # Packages ----
 library(tidyverse)
 library(patchwork)
+library(brms)
+parallel::detectCores()
 
 # Load Data ----
 load("~/Documents/extended_formative_windows/Data/clean_anes5.Rdata")
@@ -1996,14 +1998,16 @@ a5_complete_strict <- rbind(a5_bldschool_s,
                             a5_hlpblk_s, 
                             a5_integrate_s, 
                             a5_pid_s, 
-                            a5_private_s)
+                            a5_private_s) %>% 
+  mutate(young = if_else(age_1 <= 30, 1, 0))
 
 a5_complete_loose <- rbind(a5_bldschool_l, 
                             a5_govjobs_l, 
                             a5_hlpblk_l, 
                             a5_integrate_l, 
                             a5_pid_l, 
-                            a5_private_l)
+                            a5_private_l) %>% 
+  mutate(young = if_else(age_1 <= 30, 1, 0))
 
 # Build ANES74 datasets ----
 
@@ -2013,7 +2017,8 @@ a7_complete_strict <- rbind(a7_accused_s,
                             a7_jobguar_s, 
                             a7_libcon_s, 
                             a7_pid_s, 
-                            a7_unrest_s)
+                            a7_unrest_s) %>% 
+  mutate(young = if_else(age_1 <= 30, 1, 0))
 
 a7_complete_loose <- rbind(a7_accused_l, 
                             a7_busing_l, 
@@ -2021,7 +2026,8 @@ a7_complete_loose <- rbind(a7_accused_l,
                             a7_jobguar_l, 
                             a7_libcon_l, 
                             a7_pid_l, 
-                            a7_unrest_l)
+                            a7_unrest_l) %>% 
+  mutate(young = if_else(age_1 <= 30, 1, 0))
 
 # Build ANES94 datasets ----
 
@@ -2030,14 +2036,16 @@ a9_complete_strict <- rbind(a9_govins_s,
                             a9_jobguar_s, 
                             a9_libcon_s, 
                             a9_pid_s, 
-                            a9_spendserv_s)
+                            a9_spendserv_s) %>% 
+  mutate(young = if_else(age_1 <= 30, 1, 0))
 
 a9_complete_loose <- rbind(a9_govins_l, 
                             a9_hlpblk_l, 
                             a9_jobguar_l, 
[…]

### 2023-02-17 alternative_measures.R
expanded draft. pdf plots. cleaning file. more data.

@@ -1,0 +1,613 @@
+
+
+# Analyses of Clean Data
+
+
+load("~/Dropbox/formative_period/clean_anes5.Rdata")
+load("~/Dropbox/formative_period/clean_anes7.Rdata")
+load("~/Dropbox/formative_period/clean_anes9.Rdata")
+load("~/Dropbox/formative_period/clean_g6.Rdata")
+
+
+m1 <- function(.x) {
+  tidy(lm(total_dist ~ young, data = .x)) %>%
+    mutate(outcome = "1. Total Dist.")
+}
+
+m2 <- function(.x) {
+  tidy(lm(sd ~ young, data = .x)) %>%
+    mutate(outcome = "2. S.D.")
+}
+
+m3 <- function(.x) {
+  tidy(glm(stable_opinion ~ young, data = .x,
+           family = binomial(link = "logit"))) %>%
+    mutate(outcome = "3. Cross")
+}
+
+m4 <- function(.x) {
+  tidy(glm(stable_middle ~ young, 
+           data = .x,
+           family = binomial(link = "logit"))) %>%
+    mutate(outcome = "4. Switch")
+}
+
+m5 <- function(.x) {
+  tidy(glm(change_2 ~ young,
+           data = .x,
+           family = binomial(link = "logit"))) %>%
+    mutate(outcome = "5. Change > 2")
+}
+
+m6 <- function(.x) {
+  tidy(glm(durable_change ~ young,
+           data = .x,
+           family = binomial(link = "logit"))) %>%
+    mutate(outcome = "6. Cross+Stay")
+}
+
+
+
+canes5 <- clean_anes5 %>%
+  mutate(id = 1:nrow(clean_anes5)) %>%
+  select(-c(ed_2, ed_3, age_2, age_3, marital_2, marital_3, child_2, child_3,
+            sex_1:race_3)) %>%
+  zap_labels() %>%
+  mutate(ed_1 = factor(ed_1),
+         ed_1 = relevel(ed_1, ref = "hs"),
+         marital_1 = factor(marital_1),
+         marital_1 = relevel(marital_1, ref = "married")) %>%
+  mutate(across(c(helpblk5_1, helpblk5_3, integrate5_1, integrate5_3,
[…]

### 2023-02-17 draft.Rmd
more drafting. more code.

--- 2021-10-15 draft.Rmd
@@
 author: "Kevin Kiley & Nicolas Restrepo"
-output: html_document
+fontsize: 12pt
+output:
+  pdf_document: default
+  html_document: default
+header-includes:
+  - \usepackage{setspace}
+  - \doublespacing
+  - \usepackage{lineno}
+  - \linenumbers
+  - \def\linenumberfont{\normalfont\tiny\sffamily}
+bibliography: efw_bib.bib
+abstract: "Young adulthood is consistently identified as a “formative period” in the life course when people have heightened probability of attitude change.
+This probability has often been attributed to cognitive factors like brain plasticity as well as the social conditions of adolescence.
+Over the past 70 years, changes in the timing of major life course events such as marriage have transformed and extended this life stage.
+We leverage these changes to explore whether these life course changes, as well as broader social changes, have affected how likely people are to make durable changes of attitudes.
+We find strikingly little difference in the opinion behaviors among young people in different eras.
+People under 30 are similarly different from people over 30 in all eras, with slightly less stable attitudes.
+The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates."
+editor_options: 
+  markdown: 
+    wrap: sentence
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+
+library(tidyverse)
+library(patchwork)
+library(brms)
+theme_set(theme_bw())
+
 ```
@@
 
-The last few decades have seen dramatic changes in the expectations surrounding the transition from adolescence to adulthood.
-The trajectories between these two stages have become more varied, blurring the institutional boundaries that used to demarcate them.
-Individuals are more likely to return to higher education after graduating, and they tend to switch professional tracks more often, instead of being locked-in in a trade.
-The period between leaving the parental nest and "settling down" has extended.
-The primary example here is the increasing median age at first marriage, especially for women.
-This "age of independence" - characterized by exploration and a degree of detachment from social institutions - appears to be becoming longer and more diverse.
+Adolescence and young adulthood are typically identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples' core ideas and attitudes are formed [].
+Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other life stages, with opinions becoming more stable once these life stages have passed [].
+This pattern, in which young people are open to attitude change and become resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952; @vaisey2016].
 
-This stage of the life-course has gained increasing attention from sociologists because work - from varied sub-fields - highlights the pivotal role that it might play in processes of social change.
-Young adults nowadays are more likely to be away from familial supervision, which can make them prone to break with social mores.
-Their relative distance from certain social institutions - like home ownership - means that they can risk more and, therefore, they tend to inhabit the front lines of social movements.
-Additionally, recent work on attitudinal change suggests people above 25 rarely change their minds.
-Young adulthood then might also represent the tail-end of our formative windows, where we are still open to revisiting our habits and beliefs.
+However, why people in these life stages appear to show more frequent opinion change relative to older people is unclear.
+Work in biology, psychology, and cultural evolution shows that peoples' cognition changes in ways that might make them more resistant to new information as they age [].
+Explanations in social sciences, in contrast, tend to highlight the unique social features of these life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of this information [].
+While both mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.
 
-This paper seeks to examine this latter relationship.
-The expanding duration of young adulthood provides an important opportunity to study how belief-updating unfolds across an individual's lifetime.
-The finding that a period of exploration during one's early years is followed by relative stability is recurrent in many fields.
-A question that is interesting for sociologists is whether the length of these formative windows varies across socio-historical contexts.
-The evidence that the life-course and its stages are socially patterned would make that conclusion intuitive but there has been little empirical work that tries to adjudicate this question.
-However, if formative windows are dependent on social contexts - and not entirely maturational - then expanded periods of young adulthood should also result in higher rates of belief updating among people in that age group in more recent cohorts.
-In this paper, we test exactly this proposition.
+Clarifying the mechanism that drives the frequency of opinion change during this formative period is important for understanding the broader process of social change and cultural evolution.
+If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader process of social and cultural change [].
+Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2020; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than in previous eras, diminishing the importance of "cohorts" for social change.
+If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social and cultural change in the same way.
 
-Finding traces of a relationship between extended periods of young adulthood and higher rates of updating would provide important insights for understanding change, within an agent's lifetime and at the aggregate level.
-It would, for instance, constitute novel evidence that individuals' likelihood to update their beliefs is not just ontogenetic but rather depends on the social contexts in which they are situated.
-If this is the case, then sociologists are well-positioned to contribute to ongoing interdisciplinary work about formative windows and learning schedules.
-Furthermore, if longer young adulthoods result in extended periods of exploration, then there is another mechanism through which this "age of independence" affects social change: by letting individuals remain open to novel information for longer.
+Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005].
+Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2007].
+This suggests both a lengthening and a intensification of this stage of the life course in ways that should increase the rates at which people make changes of opinion.
+In this paper, we attempt to clarify the relative contributions of these different mechanisms by exploring how these changes have affected rates of change in political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population.
+We compare rates of opinion change in comparable general political attitudes across four eras -- 1956-1960; 1972-1976; 1992-1996; and 2006-2010 -- and between young people and old people in these eras to evaluate a range of competing hypotheses about what accounts for the divergence in attitude behaviors over the life course.
+
+Broadly, we find that the probability of political opinion change in people under 30, relative to the rest of the population, has not increased as the "early adulthood" window has lengthened and the conditions of this life stage have changed.
+Across eras, both young people and older people make political opinion changes at remarkably similar rates, with social transitions such as marriage, having children, and completing education playing little role in the rates at which these opinion changes happen.
+While these findings do not provide conclusive evidence of one mechanism over the other, they confine the range of potential theories of attitude development and the nature of political opinions.
+They either suggest that there is something relatively timeless -- potentially biological and cognitive factors -- that accounts for increased opinion variance in young people or that countervailing forces, some increasing stability in young people and some decreasing it, have produced aggregate stability over seven decades.
+
+## Impressionable Years / Opinion Change and Stability
+
+The idea that the early stages of the life course are especially influential in the process of attitudinal formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021].
+The argument -- at times implicit -- is that the circumstances in which people are brought up leave an enduring and traceable mark on their worldviews (Gerber and Green 1998).
+In the study of political socialization, this perspective enjoys a longstanding history.
+"Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014].
+However, the notion of "impressionable years" is more general.
+The literature on cohort effects rests on the assumption that our early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions (Elder 2018).
+Peoples' early years will leave a mark that will be predictive of their life outcomes, well beyond when those impressionable years have passed.
+Thus, the notion of a formative period -- at the early stages of the life-course -- underpins diverse and longstanding lines of research.
+
+The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course.
+The most common are an aging stability or increasing persistence hypothesis that argues that openness to change gradually declines over time and a life stages hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a].
+While different attitudes occasionally follow slightly different patterns, the impressionable years hypothesis or aging stability hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality [@wagner2019].
+More generally, recent research has found that, with a couple notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S.
+adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort explains much more variance than period for most items in the General Social Survey, suggesting a strong impressionable years pattern.
+
+In recent years, researchers have used new approaches to quantify "openness" to social influence across the life course, especially around politics [@bartels2014; @ghitza2022].
+This work has produced remarkably similar patterns, with openness peaking from a person's mid-teens through age 25 and declining quickly after that, lending more credence to the "impressionable years" hypothesis.
+However, this work tends to assume a constant age-based rate of openness across cohorts -- that 20-year-olds in 1950 are open to influence at the same rate as 20-year-olds in 2010 -- which has not been clearly tested.
+
+## Theoretical Models
+
+Why is it research consistently finds the period of time between 14 and 25 to be distinctly "impressionable years?" Changing one's mind involves at least two distinct processes that can be difficult to disentangle: exposure to new ideas and changing one's opinion conditional on exposure to new ideas.
+For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change.
+Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern.
+It could also be the case that both change as a function of age.
+In this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.
+
+### Biological Mechanisms
+
+The first potential mechanism for changes in attitude stability over the life course is biological: that the development of the of human brain as people age makes it more difficult for people to make changes of opinions after some point in their life.
+This approach tends to suggest that human cognition works similar to Bayesian updating (Bartels and Jackman 2014; Sherratt and Morand-Ferron 2018; Gopnik 2020).
+In this framework, humans weight information equally, whether they encounter it during adolescence or middle age.
+However, given that by the time people reach adulthood they have already encountered a lot of information, each new piece will have less of an impact (Bartels and Jackman 2014).
+In this thinking, the distinctive thing about our early years is simply that they happen early.
+They prove so influential because the incremental effect of new information diminishes as we age.
+
+It is also possible that there is something unique about human cognition during these periods: that agents ascribe more weight to the information acquired within them.
+Maturational arguments about the increased plasticity of the human brain during our early years are consistent with this position (Mata and von Helversen 2015).
+Our brains exhibit more malleability during childhood and adolescence (Mata and von Helversen 2015).
+This is consistent with the fact that these early formative periods have been observed across species and contexts.
+It is not only human children that exhibit a heightened capacity to adapt.
+In a study of social learning among birds, for instance, Aplin et al (2017) show that younger individuals -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.
+
+Though the Bayesian updating explanation is simple and elegant, research across different disciplines has produced evidence that supports the existence of periods of heightened sensitivity.
+In light of empirical evidence about these sensitive windows, there has been extensive modeling work to understand how they might have developed through evolution.
+This line of research seeks to understand why periods of heightened sensitivity would emerge and what contexts would facilitate their emergence (Fawcett and Frankenhuis 2015).
+The central argument is that the emergence and timing of these sensitive windows depends on the interaction between the access to information, the reliability of information, and the stability of the environment (Frankenhuis and Walasek 2020).
+Early formative periods are likely to evolve in specific configurations: when environments are unstable, and information is plentiful yet unreliable (Frankenhuis and Walasek 2020).
+In these circumstances, it makes sense for agents to explore early on, to find something that works and then to stick to strategies that have worked well in the past.
+Access to different types of information might vary across time and this might explain the emergence of sensitive periods at later stages of the life-course (Walasek, Frankenhuis, and Panchanathan 2021).
+For example, the importance of musical tastes and fashion choices becomes important around adolescence, and, at this stage, we are awash with -- at best contradictory -- information about what one should or should not like, should or should not wear.
+Sensitive periods around the development of these tastes might peak around this period, when there are more signals and when the payoffs are more salient.
+
+Regardless of the underlying reasons, the thrust of the biological arguments is that millennia of evolution have shaped human cognitive architecture in ways that makes humans relatively open to changing dispositions at younger ages and relatively resistant at older ages.
+However, while biological mechanisms can shape how individuals respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change in this window of time.
+For that, we must turn attention to the structuring of the social world.
+
+### Life Course // Social Role Mechanisms:
+
+Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct: they are characterized by sustained social learning and punctuated with constant changes of social contexts.
+Here we briefly consider some of the ways that the "impressionable years" differ in ways that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.
+
+First, people in this age group tend to move through social contexts much more quickly than people in other age groups.
+In the 11-year window from ages 14 to 25, in which people seem to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year in these contexts potentially exposing people to new classmates, teachers, role models, peers, and sources of information.
+People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons [@geist2008] .
+Assuming that new social contexts shape exposure to new information, and that exposure to new information affects the rate at which people change their opinions, this heightened contextual mobility might explain part of the difference in rates of opinion change between age groups.
+
+Second, people in the "impressionable years" window spend their time in different social contexts than people in middle age.
+Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55.
+These contexts, especially education, are explicity designed to expose people to new information in ways that can lead to updated opinions.
+For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, gender egalitarianism, and racial inequality [@campbell2016; @wodtke2018].
+Social networks also differ in this time period.
+Previous work has found that young peoples' networks tend to be larger and include more friends than kin, for example [@marsden1987].
+It could be the case that larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1975; @burt2005].
+
+Third, these are periods in which exploration and self-discovery are expected and encouraged; something that should not be taken for granted given social variation in norms around when maturity is reached [@cunningham2012; @elder1994].
+One reason young people might be exposed to new information at higher rates is that they are implicitly asked to do it by the broader culture, while older people are implicitly asked not to.
+
+For all these reasons, young people might encounter new information at higher rates than older people, which might account for higher rates of opinion updating during this window.
+And changes in the norms around geographic and social mobility during the life course might affect the rates at which these changes happen at different life stages.
+
+The social conditions of early adulthood might also influence the rate at which young people demonstrate attitude change by making it easier for young people to act on new sources of information in ways that facilitate opinion change.
+Previous work on stability of attitudes over the life course suggests that people tend to value stability in figures of authority.
+As people move into roles with more authority in mid-life, both in family and occupational contexts, they feel this expectation from others and act accordingly, closing themselves off to updating beliefs [@eaton2009].
+This potentially explains another phenomenon with some evidence behind it: that people appear to open up to changing their minds again as they enter post-retirement [@danigelis2007].
+At this point, peoples' authority diminishes, potentially opening them up to change later in life.
+
+Similarly, mid-life is associated with more homogeneous networks [@marsden1985; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@vissermirible].
+If young people exist in less dense, more heterogeneous social networks, they might it easier to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks [], or close off parts of their lives to people who might not be supportive of new attitudes.
+
+Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the "impressionable years" are principally an effect of how we have chosen to structure the life course, or the interaction between these life course stages and biological mechanisms, rather than simply a function of how the human brain works.
+
+## Changing Life Course
+
+Contemporary historical trends offer the opportunity to evaluate the role of socio-cultural and biological factors in explaining why we observe "impressionable years" in adolescence and early adulthood.
+In the U.S., the last seven decades have seen considerable changes around adolescence and young adulthood (Buchmann 1989; Rosenfeld 2007; Rosenfeld and Kim 2005).
+The expansion of higher education and increased possibilities for geographic mobility are among the many social processes that have led to changing expectations for the trajectories that young adults should follow (Rosenfeld 2007).
+In general, as Buchmann (1989) notes, life trajectories have become less predictable.
+The "script" of young adulthood has become less rigid: it might involve prolonged higher education (and movement between the workforce and education), moving to a different city, or dating without a stable romantic partner.
+The routes that people are taking into adulthood are increasingly diverse.
+For the purposes of this paper, there are two trends that are particularly important: the elongation of young adulthood and the intensification of young adulthood.
+
+### Timing
+
+The period between adolescence and "settling down" seems to be getting longer, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts.
+Buchmann (1989) notes that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood.
+In the U.S., the median age at first marriage has risen consistently since the middle of the 20th Century, from about 20 for women and 22 for men in the mid-1950s to 26.5 for women and 28.7 for men in 2011 [@manning2014].
+The median age of womens' first birth also increased a comparable amount, from 21.4 years in 1970 to 26.8 years in 2017 [@eickmeyer2017].
+We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2004].
+Thus, the markers of the transition into adulthood seem to be occurring later in the life-course and potentially in different orders.
+Therefore, the definition of what counts as our "impressionable years" might be extending.
+This, in turn, could have an effect in how long exploration lasts throughout the life-course.
+
+Adolescents are taking longer to move through the rituals of adulthood, and this might in turn elongate the period in which exploration and self-discovery are expected.
+If this is the case, we should expect to see more attitudinal updating among individuals in age-ranges that, in previous cohorts, had displayed mainly stability.
+In other words, people who were born say in the mid 20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still exhibit attitudinal updating when they reach that age.
+
+### Changing Impressionable Years
+
+A second trend relates to just how much exploration young adults can engage in during these "impressionable years." In his work, Rosenfeld (2007) explores the rapid rise in same-sex and interracial marriages in the U.S.
+His argument is that one of the processes underlying these changes is that young adults are increasingly escaping the controlling gaze of their parents.
+When young people live with -- or are financially dependent on -- their parents, the latter have more say about their children's choice of acquaintances, how they spend their time, and even the things they can believe.
+The increased independence and mobility that young adults have enjoyed since the mid 20th century means that they have been able to engage in courtship outside the gaze of their parents, thus making choices that would not have been sanctioned by them.
+Rosenfeld's (2007) work highlights an important insight: exploration might not be only a product of how sensitive we are to novel information but also how -- and whether -- we can act upon such information.
+The social institutions where we spend our impressionable years might shape how much exploration we can engage in.
+An increase in independence then might entail more exploration.
+
+This assertion in turn engenders a prediction: we should see more attitudinal updating among contemporary young adults than their older counterparts, who tended to be closer to the grip of familial control, even if they have met similar life course milestones.
+The contemporary social organization of young adulthood, which has been developing for the last few decades, might give rise to more meaningful attitudinal changes by providing more leeway to deviate from established social norms.
+
+This approach suggests that it is not the timing of life events that is important for rates of change, but the broader social conditions of early adulthood that might make people more malleable in recent cohorts even net of changes.
+In other words, unmarried, childless 25-year-old men and women in 1956 might have experienced a set of social constraints that unmarried, childless 25-year-old men and women in 2008 do not experience, leaving the latter much more open to developing new opinions than their predecessors.
+
+## Hypotheses
+
+Sociological work then suggests principled reasons to believe that the length and intensity of formative periods -- the ages at which we observe heightened rates of attitude change, and the difference between these ages and other ages -- have been affected by social factors over the past 70 years.
+We're principally interested in the age range from about 20, when many of these life stage transitions were completed in the 1950s, to about 30, the age at which many of these life stage transitions are completed now.
+We refer to people in this group as "young people" for simplicity.
+
+First, consistent with previous work, we expect young people to be more likely to make changes of attitudes than older individuals across all years.
+Biological arguments suggest that humans' cognition is still developing in this window, and, even in the mid-20th Century, young people were sill less likely to be married, less likely to have children, and most likely to still be enrolled in education.
+
+H1:
+
+Second, we expect that if social conditions are the principal drivers of rates of attitude change over the life course, then the sum of changes in the life course will produce higher rates of change in more recent cohorts.
+Both the extension of early adulthood and the intensification of early adulthood suggest that young people will make changes in attitudes at higher rates than they did in previous eras.[\^extension]
+
+H2: The rate at which young people make attitude changes, relative to older people, will have increased since the 1950s.
+
+Third, we can distinguish between the "elongation of young adulthood" and the "intensification of young adulthood" arguments by exploring the associaiton of life-course transitions and opinion change across cohorts.
+The elongation of young adulthood argument principally suggests that delaying the life course transitions is what would explain heightened rates of attitudinal updating in young people in recent cohorts.
+In contrast, the "intensification of young adulthood" argument suggests that attitudinal updating may be occurring at heightened rates in recent cohorts net of these life course transitions.
+
+H3a: Heightened rates of attitudinal updating will go away when we control for life course transitions.
+
+H3b: Heightened rates of attitudinal updating will persist when we control for life course transitions.
+
+In contrast, finding similar patterns of attitudinal updating across widely different social contexts would lend further credence to the idea that our early years are "impressionable" either because we weight information more heavily during these periods or because we are cementing our opinions, regardless of the social contexts we happen to find ourselves in.
+
+Finally, it could be case that an explanation for overall change in rates of the population are explained by factors that we have not explicitly considered here.
+It could be the case that rates of opinion change are tied to things like racial and ethnic identity, religious identity, or geographic region, all of which have changed significantly in the same window.
+If we identify broad shifts in attitude change that are not explained by age groups, or life course transitions, it is possible that further exploration of demographic changes is necessary.
+
+## Data
+
+To test the hypotheses outlined above, we require data on within-individual change in comparable attitudes over a stable window of time across eras within the same general population.
+To do this, we focus on general political beliefs, including views on the role of government, political ideology, and partisan identification, that have been asked in a variety of panel surveys.
+This is in part a practical matter, as these are the kinds of questions that have been included in panel studies that date back to the mid-1950s, giving us the largest window of cohorts to explore across comparable issues.
+However, these kinds of questions have frequently been used to identify the "impressionable years" [@alwin1991; @ghitza2022], so using these questions helps our results speak to these previous findings more directly.
+
+We draw on data from four different survey panels: the 1956-60, 1972-76, and 1992-96 American National Election Study panels, and the 2006-10 General Social Survey Panels.
+All four surveys are nationally representative samples of non-institutionalized adults in the United States.
+All four panels survey respondents three times over a four-year window, making the duration comparable.
+And all four panels include a variety of comparable questions about general sentiments toward government action, as well as measures of life course transitions.
+
+### Political Opinions
+
+We sought to identify questions that tapped general political sentiments that were echoed in at least one other survey.
+We avoided questions that asked about specific policies or political figures, such as the president, since they were often not replicated across panels.
+We also focused principally on questions asking about the role of the federal government in domestic affairs, rather than international concerns.
+Questions about international affairs were common in the early panels, but more recent panels do not contain comparable questions.
+
+The only question that is replicated across all the panels is a question about partisan identification on a seven-point scale.
+No other questions are directly comparable across all of the panels, though some are asked in comparable ways across several.
+For this reason, we include a broad range of questions in hopes of establishing general patterns.
+Table XXX outlines the questions used in this analysis, their specific wording, and response options.
+We discuss decisions regarding coding strategies below.
+
+### Life Course Events
+
+To evaluate changes in the life course, and their relationship to attitude stability, we focus on a set of events that are traditionally accepted as marking the transition to adulthood.
+First, we include a measure of respondents' ages.
+In our analyses below, we use an indicator for whether respondents were 30 years old or younger in the first wave of the survey.
+In the 1956-60 ANES panel, 205 of the 1153 respondents (18 percent) who completed all waves were under 30.
+In the 1972-76 panel, 339 respondents out of about 1300 respondents (26 percent) who completed all three waves were under 30.
+In the 1992-96 panel, 144 of the 713 respondents (20 percent) who responded in each wave were under 30.
+And in the GSS panel, 156 of the 840 respondents (19 percent) who responded in each wave were under 30.
+In all samples, the total number of respondents under 30 is relatively small, which might pose a problem for evaluating statistical differences.
+
+Other life course measures are less comparable across all panel surveys.
+While question wording differs slightly across surveys, all surveys include a measure of marital status.
+We recode this into indicators for either never being married, currently married, or previously married but not currently (divorced, separated, or widowed).
+The 1956-60 ANES panel did not distinguish between "single" and "never married" as more recent surveys do, so there are likely some cases that should fall into the "previously married" category that are coded as "never married." We code respondents based on their response in the first wave of the panel, ignoring transitions that happen during the course of the panel.
+
+The four panels were highly inconsistent in how they measured whether people had children.
+The 1956-60 panel of the ANES asked respondents how many children under 18 years old they had in the household.
+We coded individuals as having children in the first wave of the panel if they answered at least one to this question.
+The 1970s panel of the ANES asked respondents in the first wave if they had any school-aged children.
+In the second wave, they asked respondents if they ever had children, how many, and the ages of all children in the household.
+We code respondents as "having children in the first wave of the panel if they reported having at least one child older than one in the second wave of the panel.
+Because some of these questions focus on ever having children while others focus on children in the household, we caution readers from making any significant judgments about these coefficients.
+
+Finally, we include a measure of educational attainment.
+Rather than focus on whether a person was currently enrolled in an educational program, we code respondents by their highest level of education completed, either less than a high school degree, a high school degree, or at least a bachelor's degree.
+Even in more recent panels, the proportion of respondents enrolled in education is too small to generate a good estimate of its association with attitude change.
+
+## Analytic Strategy
+
+The broadest challenge for our analysis is that it is not obvious what it means for a person to demonstrate opinion change in a survey.
+Debates about how much "change" we see in opinions often focus on differences in the approach to measurement [@converse1965; @achen1975; @ansolabehere2008].
+The kinds of questions we seek to examine, general political dispositions, tend to be asked on five- or seven-point ordinal scales, either from strongly agree to strongly disagree, or from one policy position ("government should help minority groups") to another ("minority groups should help themselves").
+These kinds of scales support a wide range of different measures of opinion change and stability.
+Rather than adopt a single approach to quantify the rates of change across questions, we use a variety of approaches with different assumptions about what "change" entails and attempt to identify patterns that are robust across a range of assumptions.
+When these approaches disagree, having multiple measures allows us to more clearly evaluate what forms of change might be driving differences between groups.
+
+The first approach we pursue is to simply measure whether people change their opinions at all during the course of the panel.
+In this approach, we use an indicator variable if they change at all from their first position.
+This imposes an extremely strong definition of what it means to hold a "stable" opinion, and does not distinguish between changes of opinion that might be due to measurement error and changes that reflect substantive shifts in positions.
+
+The second approach to measuring change in political opinions is to assume that numbering of the ordinal scale of the response options is approximately correct and that all change is meaningful.
+Under these assumptions, we measure the total "distance" people move across a scale during the time period in question, adding the absolute change from wave 1 and 2 to the absolute change from wave 2 and 3.
+This approach does not distinguish directional change (ending with a new opinion) from reversion (changing opinion at time 2 and reverting at time 3).
+This approach also has the downside of not being comparable across questions that are on different scales.
+A one-unit change on a five-point scale might be more substantial than a one-point change on a seven-point scale.
+
+A third approach we use is to define a prioi what kinds of opinion behavior count as "change" and compare the prevalence of this kind of behavior across age groups and panels.
+In this case, we count any response pattern in which people stay on the same side of a midpoint ("agree"-"strongly agree"-"agree") as a stable opinion and any response option that crosses the midpoint ("agree"-"disagree"-"disagree") as representing a change of opinion.
+In this approach, we code people as "changing" if they cross the scale midpoint.[^1]
+Like the previous approach, this approach counts directional change the same as reversion.
+
+[^1]: Coding people as changing if they simply "touch" the midpoint ("agree"-"neither"-"agree") compared to crossing the midpoing ("agree"-"disagree"-"agree") produces substantively similar results.
+
+A fourth and related approach is to classify people who change their responses within a specified range of the scale, such as changing less than two scale points (e.g., "agree"-"neither"-"agree"), as expressing stability while counting other respondents as making changes of opinion.
+This allows people who are stably in the middle of the scale (somewhere between agreeing and neither agreeing nor disagreeing, for example) to count as holding a stable opinion, but as changing if they shift too far in either direction.
+In principle, this approach should account for measurement error better than other approaches, as "true" changes of opinion should, on average, be larger than those driven by measurement error.
+
+Fifth, we try to distinguish change that reflects moving from one side of the scale to the other from what might be "random" change.
+To do this, we classify people as making a durable change if they cross the scale midpoint once and only once over the course of the panel.[^2]
+This is potentially the most stringent definition of "true" opinion change.
+
+[^2]: This approach has a fundamental problem.
+    As the proportion of people answering the question at random increases, the proportion of people demonstrating this pattern will also increase.
+
+Sixth, in an attempt to deal with measurement error, we generate an issue scale of all general political questions in each year [@ansolabehere2008].
+In these scales, all domestic policy issues (excluding partisan identification) are recoded on a 1 to 7 scale so that higher values are associated with more conservative positions.
+Responses are then averaged to generate a single value for each respondent.
+We calculate the standard deviation of this scale as a measure of overall change, as well as the overall change from time 1 to time 3.
+These scales should not be assumed to be comparable across panels.
+Since some have more items than others, they are likely to have lower measurement error, resulting in lower standard deviations over time.
+However, we are able to compare young and old people on these metrics in each panel.
+
+For every definition of change, we regress the outcome measure on an indicator variable for whether people are aged 30 or less.
+For the continuous outcomes, we use ordinary least squares regression.
+For the dichotomous outcomes, we use logistic regression.
+This produces an extremely large number of models for each of many questions, and will likely produce statistically significant coefficients at the $p < .05$ level for a substantial proportion of coefficients simply by chance.
+Our goal is not to interpret any single coefficient but to interpret overall patterns if they emerge.
+Comparing logistic regression coefficients across models presents a number of challenges [@allison1999].
+Instead, we principally compare the predicted rates of opinion change generated by these models over time visually.
+
+### Strict and Loose Coding
+
+Many of the questions have response categories like "Don't Know" and "Haven't thought about it." These are usually coded towards the end of the scale, so taking their coding at face value would imply that moving from these answers to one end of the scale is "easier" (at least in terms of distance).
+We consider two ways of handling these response options.
+
+The first is to accept that we cannot categorize how far "now knowing" is from say "extremely liberal" and "extremely conservative," or whether these kinds of responses represent a qualitative shift from previous responses.
+Moreover, we might say that this kind of response is qualitatively not the same as being in the middle of the scale as an "moderate." Therefore, we get rid of these cases in all models, and only attempt to make inferences for the kinds of people who select other options in all waves of a survey.
+We call this approach "strict filtering."
+
+The second approach is to code these responses as falling at the middle of the opinion scale; that there is no difference between someone saying they "neither agree nor disagree," for example, and saying they "do not know" their position on an issue.
+Recoding these people allows us to have a much larger sample size and draw clearer distinctions between cases.
+There are substantial justifications for this decision for many of the questions we explore here.
+First, many Likert scales explicitly label the scale midpoint with "no opinion," which suggests that respondents might perceive it this way too, even if it is labeled something like, "neither agree nor disagree." Second, people who report "Don't know" in one wave tend to report an opinion near the scale midpoint in subsequent waves [@kinder2017].
+This suggests that they view these positions as substantively similar as well.
+
+Each of these approaches might be more or less valid for different questions, so we consider all models under each approach.
+
+## Results
+
+Figure XXX presents either the average expected change (for continuous outcomes) or the predicted probability of people over 30 making a "change" of opinion (for categorical outcomes) as defined by the various metrics we outline above.
+
+There are several takeaways from the figure.
+First, change as defined by most of the metrics we outline above is incredibly common.
+On almost all items, the majority of respondents change their response more than two scale points and cross the midpoint of the scale over the course of the survey window.
+At the same time, this tends not to manifest as "durable change" -- defined as starting on one side of the scale and ending on the other.
+This kind of opinion behavior is rare across panels, being observed in fewer than 10 percent of all cases for almost every question.
+
+Substantively, there is little difference across the panels in the average rates of opinion change.
+All panels tend to have a very similar level of each kind of change, with one or two questions showing more or less change in each panel.
+Partisan identification is a clear outlier in all panels.
+It is the least likely to show change from one side of the scale to the other and the least likely to show changes of greater than 2 points on the scale in almost every panel.
+This finding agrees with previous work in this domain suggesting that partisan identification is a substantively different kind of response than the other responses here [@alwin1991].
+Partisan identification is registered externally to an individual, has the clearest support from social institutions, and is the most likely to be reinforced in media representations.
+
+Self-placement on a liberal-conservative scale also shows a somewhat different pattern than other opinion questions.
+For most questions, the rates of change greater than 2 points and change across the scale midpoint are relatively similar.
+However, self-placement on the liberal-conservative scale shows high rates of people switching sides or crossing the midpoint over the course of the scale, but much lower rates (relatively) of people changing by more than two points during the course of the panel.
+This suggests that many respondents hover around the middle of the scale (identifying as moderate), while on other questions they spread responses more widely, vacillating between ends of the scale over time.
+
+These rates provide a baseline against which we can compare young people.
+We now turn to exploring the differences between young (under 30) and older respondents across the panels.
+Figure XXX plots predicted differences between young and old people generated from the regression models outlined above.
+These are presented in logodds for the dichotomous outcomes and scalar differences for the continuous (total change/standard deviation) outcomes.
+
+```{r}
+#Alterantive measures plot here.
+```
+
+In general, there are very few statistically significant coefficients in the plot, suggesting that young peoples' opinion behavior is not substantially different than that of older people for the issues examined here across all eras.
+This is not to say that young people do not have substantively different opinions (they likely do), just that they are not any more or less stable by the metrics we outlined above.
+There are only a handful of issues where we observe a consistent pattern across most measures of attitude change: partisan identification in most panels, the issue of busing in the 1970s, and general views on government spending in the 1990 panel.
+
+One concern might be that young people are changing at notably different rates than older respondents, but that the models explored here are under-powered and not able to identify true differences.
+However, it is hard to observe how narrower standard errors would produce a clearer pattern.
+For both standard deviation and total distance changed, coefficients for young people in all waves are equally likely to be positive and negative, suggesting no clear pattern over time.
+In all eras, across all measures, there are issues where young people are more stable and issues where young people are less stable.
+On the two issues that are asked across multiple panels -- partisan identification, asked in all panels, and position on a liberal to conservative scale, asked since the 1970s panel -- coefficients look remarkably stable across panels.
+
+Comparing coefficient estimates across logistic regression models can be misleading [@allison1999].
+We can further confirm the patterns observed in the above plot by plotting the predicted probability of old and young people making each kind of change in each panel, or the predicted difference in overall change.
+
+```{r}
+#Figure out a way to put the predicted probability plot in here.
+```
+
+Predicted probabilities reinforce previous statements: there is not a clear change over time in the difference between respondents under 30 and those over 30.
+It might be reasonable to suggest that there is a general tendency of young people to change opinions at higher rates than older people that might be detectable with larger samples, but the difference is probably on the order of less than 5 percentage points, with the exception of a few questions like partisan identification, where we see a substantial difference.
+This difference, to the extent that it exists, principally manifests as inconsistent change, rather than a switch from one side of the scale to the other.
+
+### Life Course Stages
+
+We have to now explored the relationship between age and attitude change and observed no clear pattern, but we suggested earlier that features of the life course, such as whether people are married or whether they graduated college, might matter as well.
+Because the samples of young people are already quite small, we focus on each life stage separately.
+
+Figure XXX plots the partial association between measures of marital status and change, net of the coefficient for young people.
+Contrary to expectations, there is no clear pattern in the effect of marital status on the probability of change, especially never having been married.
+Almost none of these coefficients show statistical significance, and are equally likely to be positive and negative, suggesting no substantial difference between married respondents and other respondents.
+Having been previously married appears to be associated with slightly higher rates of attitude change change over time, especially in more recent panels, where this group shows up in higher rates.
+
+```{r}
+#marital coef plot hereyeyeye
+```
+
+Figure XXX plots the coefficient estimates for the association of education level and various measures of change, with high school degree holders serving at the reference category.
+Broadly speaking, prior to since the 1970s, higher levels of education are associated with a decreased likelihood of attitude change on most measures.
+
+```{r}
+#ed coef plot herehey
+```
+
+Including either education or marital status in the model does not affect the overall pattern of change over time.
+These covariates have almost no effect on the covariate for "youth" in any model.
+This suggests that changes these life course factors are not affecting the relationship between youth and attitude change over time.
+
+## Discussion
+
+This paper set out to evaluate whether social and cultural changes in the length and intensity of young adulthood have produced changes in the rates at which young people hold stable political opinions or change their political opinions over time.
+Considering a wide range of definitions about what constitutes "opinion change," and comparing young and old people on these measures across general measures of political opinion in four panel data sets spanning 60 years, we find remarkably little change in opinion dynamics, either in the population as a whole or among young people specifically.
+
+Broadly, in line with previous work on the topic (much of which uses the same data), survey measures designed to tap general political dispositions produce highly inconsistent responses from the public [@converse1965; @zaller1992; @hout2016].
+On many of the questions explored here, more than 60 percent of the sample places themselves on either side of a scale over the three-wave, four-year panel.
+The outlier to this pattern is partisan identification, which demonstrates much less within-person variance than the other questions, a finding that is also in line with previous work [@hout2016; @alwin1991].
+We strongly caution other researchers from using partisan identification to draw inferences about political attitudes in general.
+
+In general, young people are more inconsistent opinion holders than older people, demonstrating more change or inconsistency over time across a variety of metrics than their older counterparts.
+However, this difference is not pronounced.
+When it does manifest, young people are probably about 5 percentage points more likely to demonstrate an inconsistent opinion than older people, and on many issues younger people and older people are indistinguishable in their opinion behavior.
+
+In addressing our central question, we find that there has not been a obvious, significant shift in the difference between younger and older respondents across the four panels.
+Young people have not diverged from older individuals in their likelihood of opinion instability or opinion change.
+People under 30 today make opinion changes at roughly the same rate as people under 30 in 1950 or 1970, despite being much less likely to be married, and much less likely to have a child.
+The overall takeaway must be that the extension and changes of early adulthood has not had profound effects on the rate at which young people update their beliefs.
+We suggest two explanations for this pattern.
+
+First, it might be the case that countervailing forces have produced aggregate stability in the differences between young and old people.
+While more young adults are putting off marriage and having children, which might make them more inconsistent opinion holders for longer periods of time, they are also attending higher education at higher rates, which might make them more consistent opinion holders.
+In more recent panels, having attended college is associated with more stable opinions, even controlling for age.
+This was not the case in the early panels.
+However, these factors -- rates of marriage and rates of education -- do not explain or obscure any differences over time.
+When we include these factors in a regression model with youth, the differences in opinion behavior between young people and old people stay the same.
+
+The alternative explanation is that changes in the social conditions around early adulthood have not had substantial effects on the rates at which young people make opinion changes because it is biological and cognitive processes, rather than the social conditions of young adulthood, that explain the heightened rate at which young people make opinion changes.
+The remarkable consistency of the patterns of opinion behavior despite monumental shifts in the nature of information access, political polarization, and the organization of work, as well as the entrance of women into the workforce in large numbers, the dramatic expansion of higher education, and dramatic changes in the demographic composition of the United States in the same time period suggests that these patterns might be rooted in something other than social structure.
+
+### Limitations
+
+There are several important limitations of our analysis.
+First, the high degree of opinion instability in general makes it difficult to distinguish "real" change from the kind of inconsistent change that seems to be captured in most of our measures.
+While we attempted to control for this by using composite scales of multiple items, these scales themselves still have relatively high inconsistency.
+
+Second, while all the questions generally touch on the role of the federal government in addressing different issues, the substantive topics asked in each panel are quite different.
+Early panels focus on questions about the government's role in managing public utilities, questions of the government's role in racial integration, and the federal role in local school construction.
+Later panels focus on questions of redistribution of wealth, and the government's role in the health care and the broader economy.
+Because we cannot hold constant the topics under investigation, and because the topics do not themselves represent a random sample of opinions, it is not clear whether the lack of change over time is due to this turnover in topics.
+It could be the case that the overall pattern we observe -- high rates of instability, with instability higher in young people -- is only applicable to highly salient issues in the general public.
+
+## Conclusions
+
+Despite significant shifts in the timing of life course events and the broad social structure of early adulthood from the mid-1950s to today, there is little change in the rates at which people between the ages of 18 and 30 undergo major changes of political opinions.
+The patterns observed in this analysis broadly suggest that there is something timeless -- such as how the human brain develops over time -- that explains broad patterns in attitude instability early in the life course.
+It could also suggest that the features that explain stability are still social in nature, but are rooted in factors that have not undergone significant changes since the 1950s, though it is unclear what those might be.

### 2023-02-21 Analyses/extended_abstract_asa23.Rmd
Changes some wording in the main draft. Plausible structure for the extended abstract

---
title: "Extended Abstract"
author: "Kevin Kiley & Nicolas Restrepo"
fontsize: 12pt
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \def\linenumberfont{\normalfont\tiny\sffamily}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
library(tidyverse)
library(brms)
theme_set(jtools::theme_apa())
```

## Motivation

Adolescence and young adulthood are typically identified in the social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when their core attitudes are formed []. Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other periods, with opinions becoming more stable once these stages have passed []. This pattern underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952; @vaisey2016].

However, why people in these life stages appear to show more frequent opinion change relative to older people is unclear. Work in biology, psychology, and cultural evolution shows that peoples' cognition changes in ways that might make them more resistant to new information as they age []. Explanations in sociology, in contrast, tend to highlight the unique social features of these life stages -- like high rates of mobility between social contexts, and participation in educational institutions -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of this information []. While both mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.

Clarifying the mechanism that drives the frequency of opinion change during this formative period is important for understanding the broader process of social change. If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader patterns of social change. Similarly, trends such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @brückner2005] that decrease distinctions between life course stages might make older people more open to change in more recent eras, diminishing the importance of "cohorts" for social change. If the mechanisms are principally developmental, rather than social, then these changes are unlikely to affect rates of social and cultural change in the same way.

Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @brückner2005].
Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2005]. This suggests both a lengthening and a intensification of this stage of the life course in ways that should increase the rates at which people update their attitudes. In this paper, we attempt to clarify the relative contributions of these different mechanisms by exploring how these social trends have affected rates of change in political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population.

## Hypotheses

Recent changes in the structuring of the life-course should help us adjudicate the relative contributions between social and developmental factors in rates of attitudinal updating among young adults. If their openness to information is purely developmental, then recent social changes should not affect rates of attitudinal change across young people from different periods (**H1**). However, if social structure plays a key role, we should be able to detect differential rates of change across these groups.

Sociological work outlines two specific trends that engender clear and testable predictions. The first is related to the elongation of young adulthood: people are increasingly postponing the rituals associated with "settling down". If social conditions affect the length of formative periods, then we should expect young people in more recent cohorts to exhibit higher rates of change (**H2**).

Nonetheless, work like that of Rosenfeld [] suggests that it is not only about the speed at which we move through the life-course, but also about the opportunities we have for exploration and self-expression. Contemporary social changes can be explained by the fact that people are more mobile and financially independent and, therefore, have more freedom to deviate from convention. This is an argument then about the *intensity* of formative periods. While this argument also predicts higher rates of absolute change among young people in more recent periods, it suggests that these differences will persists even when we account for differential timing of life events like marriage (**H3**).

## Data & Analytic Strategy

To test the hypotheses outlined above, we require data on within-individual change in comparable attitudes, over a stable window of time. We also need this data to capture different time periods within the same general population. To do this, we focus on general political beliefs, including views on the role of government, political ideology, and partisan identification, that have been asked in a variety of panel surveys. This is in part a practical matter, as these are the kinds of questions that have been included in panel studies that date back to the mid-1950s, giving us the largest window of cohorts to explore across comparable issues. However, these kinds of questions have frequently been used to identify the "impressionable years" [@alwin1991; @ghitza2022], so using these questions helps our results speak to these previous findings more directly.

We draw on data from four different survey panels: the 1956-60, 1972-76, and 1992-96 American National Election Study panels, and the 2006-10 General Social Survey Panels. All four surveys are nationally representative samples of non-institutionalized adults in the United States. All panels also survey respondents three times over a four-year window, making the duration comparable. They also include a variety of comparable questions about general sentiments toward government action, as well as measures of life course transitions.

The broadest challenge for our analysis is that it is not obvious what it means for a person to demonstrate opinion change in a survey. Thus, rather than adopt a single approach to quantify the rates of change across questions, we use a variety of approaches. We calculate whether participants changed at all across the three waves, we measure their absolute movement across the scales, whether they crossed the midpoint of the scale at any point, and whether their last poisition is at leasty two points away from their starting one. All these outcome variables presuppose different assumptions about what "change" entails and, thus, we deploy them all in an attempt to identify patterns that are robust across a range of assumptions.

For every definition of change, we regress the outcome measure on an indicator variable for whether people are aged 30 or less. For the continuous outcomes, we use ordinary least squares regression and, for the dichotomous outcomes, we use logistic regression. This produces an extremely large number of models for each of many questions, and will likely produce statistically significant coefficients at the $p < .05$ level for a substantial proportion of coefficients simply by chance.
Our goal is not to interpret any single coefficient but to interpret overall patterns if they emerge.

## Results

Figure 1 plots predicted differences between young and old people generated from the regression models outlined above. These are presented in log-odds for the dichotomous outcomes and scalar differences for the continuous (total change/standard deviation) outcomes. If rates of attitudinal updating among young people had been increasing across time, we would see the plotted coefficients becoming larger as panels become more recent. Here, it means that the brown coefficients should be smaller than the blue/green ones, and the darkest blue/green one -- which represents our most recent panel -- should be the largest. Instead, in general, there are very few coefficients whose confidence intervals do not overlap 0, suggesting that young peoples' opinion behavior is not substantially different than that of older people for the issues examined here across all eras.

```{r, out.height='200%'}
model_paths <- list.files(path="../Models", pattern=".rds$", all.files=TRUE,
                          full.names=TRUE, recursive = T)

all_coefficients <- data.frame()

for (i in 1:length(model_paths)) {
  path <- model_paths[[i]]

  model <- read_rds(path)

  young_variance <- coef(model)$variable[, , 'young'] %>%
    as.data.frame()

  name <- str_remove(path, "../Models/")

  young_variance <- young_variance %>%
    mutate(model = substr(name, 1, 3),
           change_type = str_extract(name, "any|crossed|total|twopoints"),
           variable = rownames(young_variance))

  all_coefficients <- rbind(all_coefficients, young_variance)
}

all_coefficients %>% 
  filter(model %in% c("a5s", 
                      "a7s", 
                      "a9s", 
                      "gss")) %>% 
  mutate(model = case_when(
    model == "a5s" ~ "ANES56", 
    model == "a7s" ~ "ANES74", 
    model == "a9s" ~ "ANES94", 
    model == "gss" ~ "GSS06"
  )) %>% 
  ggplot(aes(x = variable, y = Estimate, col = model)) +
  geom_pointrange(aes(ymin = `Q2.5`, 
                      ymax = `Q97.5`), 
                  position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  scale_color_brewer(palette="BrBG") +
  facet_wrap(~ change_type) + 
  theme(legend.position = "bottom") + 
  coord_flip() +
  labs(title = "Figure 1", 
       x = "", 
       y = "Predicted Difference")
```

This is not to say that young people do not have substantively different opinions (they likely do), just that they are not any more or less stable by the metrics we outlined above. There are only a handful of issues where we observe a consistent pattern across most measures of attitude change: partisan identification in most panels, the issue of busing in the 1970s, and general views on government spending in the 1990 panel.

## Conclusion

Despite significant shifts in the timing of life course events and the broad social structure of early adulthood from the mid-1950s to today, there is little change in the rates at which people between the ages of 18 and 30 undergo major changes of political opinions. The patterns observed in this analysis broadly suggest that there is something developmental  that explains broad patterns in attitude instability early in the life course. It could also suggest that the features that explain stability are still social in nature, but are rooted in factors that have not undergone significant changes since the 1950s, though it is unclear what those might be.

### 2023-02-21 draft.Rmd
Changes some wording in the main draft. Plausible structure for the extended abstract

@@
 Over the past 70 years, changes in the timing of major life course events such as marriage have transformed and extended this life stage.
-We leverage these changes to explore whether these life course changes, as well as broader social changes, have affected how likely people are to make durable changes of attitudes.
+We take advantage of these changes to explore whether they -- as well as broader social changes -- have affected how likely people are to make durable changes of attitudes.
 We find strikingly little difference in the opinion behaviors among young people in different eras.
@@
 
-Adolescence and young adulthood are typically identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples' core ideas and attitudes are formed [].
-Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other life stages, with opinions becoming more stable once these life stages have passed [].
+Adolescence and young adulthood are typically identified in the social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples' core ideas and attitudes are formed [].
+Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other periods, with opinions becoming more stable once these stages have passed [].
 This pattern, in which young people are open to attitude change and become resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952; @vaisey2016].
@@
 Clarifying the mechanism that drives the frequency of opinion change during this formative period is important for understanding the broader process of social change and cultural evolution.
-If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader process of social and cultural change [].
-Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2020; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than in previous eras, diminishing the importance of "cohorts" for social change.
+If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader process of social and cultural change.
+Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @brückner2005] that decrease distinctions between life course stages might make older people more open to change in more recent eras than in previous eras, diminishing the importance of "cohorts" for social change.
 If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social and cultural change in the same way.
 
-Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005].
-Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2007].
+Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @brückner2005].
+Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2005].
 This suggests both a lengthening and a intensification of this stage of the life course in ways that should increase the rates at which people make changes of opinion.
@@
 
-The idea that the early stages of the life course are especially influential in the process of attitudinal formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021].
+The idea that the early stages of the life course are especially influential in the process of attitudinal formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2020].
 The argument -- at times implicit -- is that the circumstances in which people are brought up leave an enduring and traceable mark on their worldviews (Gerber and Green 1998).
@@
 The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course.
-The most common are an aging stability or increasing persistence hypothesis that argues that openness to change gradually declines over time and a life stages hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a].
+The most common are an aging stability or increasing persistence hypothesis that argues that openness to change gradually declines over time and a life stages hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004].
 While different attitudes occasionally follow slightly different patterns, the impressionable years hypothesis or aging stability hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality [@wagner2019].
 More generally, recent research has found that, with a couple notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S.
-adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort explains much more variance than period for most items in the General Social Survey, suggesting a strong impressionable years pattern.
+adults.
+[@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort explains much more variance than period for most items in the General Social Survey, suggesting a strong impressionable years pattern.
 
@@
 Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55.
-These contexts, especially education, are explicity designed to expose people to new information in ways that can lead to updated opinions.
+These contexts, especially education, are explicitly designed to expose people to new information in ways that can lead to updated opinions.
 For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, gender egalitarianism, and racial inequality [@campbell2016; @wodtke2018].
@@
 Previous work has found that young peoples' networks tend to be larger and include more friends than kin, for example [@marsden1987].
-It could be the case that larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1975; @burt2005].
+It could be the case that larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].
 
@@
 
-Similarly, mid-life is associated with more homogeneous networks [@marsden1985; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@vissermirible].
-If young people exist in less dense, more heterogeneous social networks, they might it easier to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks [], or close off parts of their lives to people who might not be supportive of new attitudes.
+Similarly, mid-life is associated with more homogeneous networks [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004].
+If young people exist in less dense, more heterogeneous social networks, they might it easier to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks or close off parts of their lives to people who might not be supportive of new attitudes.
 
@@
 
-Third, we can distinguish between the "elongation of young adulthood" and the "intensification of young adulthood" arguments by exploring the associaiton of life-course transitions and opinion change across cohorts.
+Third, we can distinguish between the "elongation of young adulthood" and the "intensification of young adulthood" arguments by exploring the association of life-course transitions and opinion change across cohorts.
 The elongation of young adulthood argument principally suggests that delaying the life course transitions is what would explain heightened rates of attitudinal updating in young people in recent cohorts.
@@
 In the second wave, they asked respondents if they ever had children, how many, and the ages of all children in the household.
-We code respondents as "having children in the first wave of the panel if they reported having at least one child older than one in the second wave of the panel.
+We code respondents as "having children" in the first wave of the panel if they reported having at least one child older than one in the second wave of the panel.
 Because some of these questions focus on ever having children while others focus on children in the household, we caution readers from making any significant judgments about these coefficients.
@@
 The broadest challenge for our analysis is that it is not obvious what it means for a person to demonstrate opinion change in a survey.
-Debates about how much "change" we see in opinions often focus on differences in the approach to measurement [@converse1965; @achen1975; @ansolabehere2008].
+Debates about how much "change" we see in opinions often focus on differences in the approach to measurement [@converse1964; @achen1975; @ansolabehere2008].
 The kinds of questions we seek to examine, general political dispositions, tend to be asked on five- or seven-point ordinal scales, either from strongly agree to strongly disagree, or from one policy position ("government should help minority groups") to another ("minority groups should help themselves").
@@
 It could also suggest that the features that explain stability are still social in nature, but are rooted in factors that have not undergone significant changes since the 1950s, though it is unclear what those might be.
+
+## References

### 2023-08-11 Analyses/extended_abstract_asa23.Rmd
edits 811

@@
   - \def\linenumberfont{\normalfont\tiny\sffamily}
+bibliography: ./efw_bib.bib
+
 ---
@@
 
-Adolescence and young adulthood are typically identified in the social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when their core attitudes are formed [].
-Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other periods, with opinions becoming more stable once these stages have passed [].
-This pattern underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952; @vaisey2016].
+Adolescence and young adulthood are typically identified in the social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents; when their opinions are more malleable; and when their core dispositions, identities, and opinions are formed [@alwin1991].
+A broad body of research finds that rates of opinion change tend to peak during the window of ages 14-25, with opinions becoming more stable once people move into middle age [@sears1999; @bartels2014; @ghitza2022].
+And this recurring empirical pattern is central to broader dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952; @vaisey2016].
 
 However, why people in these life stages appear to show more frequent opinion change relative to older people is unclear.
-Work in biology, psychology, and cultural evolution shows that peoples' cognition changes in ways that might make them more resistant to new information as they age [].
-Explanations in sociology, in contrast, tend to highlight the unique social features of these life stages -- like high rates of mobility between social contexts, and participation in educational institutions -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of this information [].
-While both mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.
+Work in biology, psychology, and cultural evolution suggests that peoples' cognition changes in ways that might make them more resistant to new information as they age.
+Explanations in sociology, in contrast, tend to focus on the unique social features of these life stages -- including high rates of mobility between social contexts and participation in educational institutions -- that shape how people are exposed to new sources of information and their ability to update opinions in the face of this information [@visser2004a].
+While both mechanisms likely play a role in explaining the higher rates of opinion change in this life stage, the relative contributions of each are unclear.
 
-Clarifying the mechanism that drives the frequency of opinion change during this formative period is important for understanding the broader process of social change.
-If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader patterns of social change.
-Similarly, trends such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @brückner2005] that decrease distinctions between life course stages might make older people more open to change in more recent eras, diminishing the importance of "cohorts" for social change.
-If the mechanisms are principally developmental, rather than social, then these changes are unlikely to affect rates of social and cultural change in the same way.
+In this paper, we leverage changes in the timing, ubiquity, and standardization of life course transitions over the second half of the 20th Century [@buchmann1989; @bruckner2005] to adjudicate the relative contributions of developmental and social mechanisms to the heightened rates of opinion change we observe in young people.
+Since the 1950s, the median age of marriage and childbirth have each been pushed back by about 10 years for both men and women [@eickmeyer2017; @manning2014], rates of college-going have increased dramatically, and people make more transitions between workplaces and between the workforce and education [@bruckner2005].
+These shifts have led to a dramatically expanded window of participation in educational institutions, geographic and social mobility, and freedom from various social obligations and expectations, all of which are assumed to facilitate opinion change [@rosenfeld2009].
+If the heightened rates of opinion change we observe in young adults are principally due to the social structure of this life stage, then we should observe changes in the rates of opinion change in this group over time.
 
-Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @brückner2005].
-Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2005].
-This suggests both a lengthening and a intensification of this stage of the life course in ways that should increase the rates at which people update their attitudes.
-In this paper, we attempt to clarify the relative contributions of these different mechanisms by exploring how these social trends have affected rates of change in political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population.
-
-## Hypotheses
-
-Recent changes in the structuring of the life-course should help us adjudicate the relative contributions between social and developmental factors in rates of attitudinal updating among young adults.
-If their openness to information is purely developmental, then recent social changes should not affect rates of attitudinal change across young people from different periods (**H1**).
-However, if social structure plays a key role, we should be able to detect differential rates of change across these groups.
-
-Sociological work outlines two specific trends that engender clear and testable predictions.
-The first is related to the elongation of young adulthood: people are increasingly postponing the rituals associated with "settling down".
-If social conditions affect the length of formative periods, then we should expect young people in more recent cohorts to exhibit higher rates of change (**H2**).
-
-Nonetheless, work like that of Rosenfeld [] suggests that it is not only about the speed at which we move through the life-course, but also about the opportunities we have for exploration and self-expression.
-Contemporary social changes can be explained by the fact that people are more mobile and financially independent and, therefore, have more freedom to deviate from convention.
-This is an argument then about the *intensity* of formative periods.
-While this argument also predicts higher rates of absolute change among young people in more recent periods, it suggests that these differences will persists even when we account for differential timing of life events like marriage (**H3**).
+Drawing on four nationally representative panel surveys, dating from the mid-1950s through the early 2000s, we compare multiple different measures of opinion change in general political opinions and partisan identification of early adults (people under 30) and the rest of the population.
+Despite the significant changes in the standardization of the life course and the duration of early adulthood, we find remarkable consistency in opinion behaviors over time, both for young adults and for the rest of the population.
+Across multiple measures of opinion change, we find that people under 30 change their opinions at slightly higher rates than people over 30, but this gap remains stable across panels.
+The overall pattern of results suggests either that the heightened rates of attitude change in early adulthood are principally driven by relatively constant developmental factors -- how the brain reacts to new information as people age -- or that the dramatic changes in the social structuring of early adulthood have combined in a way to produce aggregate stability over six decades.
 
@@
 
-To test the hypotheses outlined above, we require data on within-individual change in comparable attitudes, over a stable window of time.
-We also need this data to capture different time periods within the same general population.
-To do this, we focus on general political beliefs, including views on the role of government, political ideology, and partisan identification, that have been asked in a variety of panel surveys.
-This is in part a practical matter, as these are the kinds of questions that have been included in panel studies that date back to the mid-1950s, giving us the largest window of cohorts to explore across comparable issues.
-However, these kinds of questions have frequently been used to identify the "impressionable years" [@alwin1991; @ghitza2022], so using these questions helps our results speak to these previous findings more directly.
-
-We draw on data from four different survey panels: the 1956-60, 1972-76, and 1992-96 American National Election Study panels, and the 2006-10 General Social Survey Panels.
+We draw on data from four different panel surveys of the American population: the 1956-60, 1972-76, and 1992-96 American National Election Study panels, and the 2006-10 General Social Survey panel.
 All four surveys are nationally representative samples of non-institutionalized adults in the United States.
-All panels also survey respondents three times over a four-year window, making the duration comparable.
-They also include a variety of comparable questions about general sentiments toward government action, as well as measures of life course transitions.
+All panels also interviewed survey respondents three times over a four-year window, making the duration of each survey window comparable.
+All four panels also include comparable questions about general sentiments toward government action, a measure of partisan identification, and measures of life course transitions such as marriage and childbearing.
 
@@
 Thus, rather than adopt a single approach to quantify the rates of change across questions, we use a variety of approaches.
-We calculate whether participants changed at all across the three waves, we measure their absolute movement across the scales, whether they crossed the midpoint of the scale at any point, and whether their last poisition is at leasty two points away from their starting one.
-All these outcome variables presuppose different assumptions about what "change" entails and, thus, we deploy them all in an attempt to identify patterns that are robust across a range of assumptions.
+We calculate whether participants changed at all across the three waves (), their absolute movement across the scales (), whether they crossed the midpoint of the scale at any point (), and whether their last position is at least two points away from their starting one ().
+All these outcome variables presuppose different assumptions about what "change" entails and, thus, we present them all in an attempt to identify patterns that are robust across a range of assumptions.
 
 For every definition of change, we regress the outcome measure on an indicator variable for whether people are aged 30 or less.
-For the continuous outcomes, we use ordinary least squares regression and, for the dichotomous outcomes, we use logistic regression.
+For the continuous outcomes, we use ordinary least squares regression and for the dichotomous outcomes we use logistic regression.
 This produces an extremely large number of models for each of many questions, and will likely produce statistically significant coefficients at the $p < .05$ level for a substantial proportion of coefficients simply by chance.
@@
 ## Results
+
+Figure 1 plots coefficient estimates generated from regression models of change on whether respondents are under 30.
+Each column represents a different measure of "change" outlined above.
+...
+If young people have become more likely to make
+
+```{r}
+
+```
 
@@
 
-```{r, out.height='200%'}
-model_paths <- list.files(path="../Models", pattern=".rds$", all.files=TRUE,
-                          full.names=TRUE, recursive = T)
-
-all_coefficients <- data.frame()
-
-for (i in 1:length(model_paths)) {
-  path <- model_paths[[i]]
-
-  model <- read_rds(path)
-
-  young_variance <- coef(model)$variable[, , 'young'] %>%
-    as.data.frame()
-
-  name <- str_remove(path, "../Models/")
-
-  young_variance <- young_variance %>%
-    mutate(model = substr(name, 1, 3),
-           change_type = str_extract(name, "any|crossed|total|twopoints"),
-           variable = rownames(young_variance))
-
-  all_coefficients <- rbind(all_coefficients, young_variance)
-}
-
-all_coefficients %>% 
-  filter(model %in% c("a5s", 
-                      "a7s", 
-                      "a9s", 
-                      "gss")) %>% 
-  mutate(model = case_when(
-    model == "a5s" ~ "ANES56", 
-    model == "a7s" ~ "ANES74", 
-    model == "a9s" ~ "ANES94", 
-    model == "gss" ~ "GSS06"
-  )) %>% 
-  ggplot(aes(x = variable, y = Estimate, col = model)) +
-  geom_pointrange(aes(ymin = `Q2.5`, 
-                      ymax = `Q97.5`), 
-                  position = position_dodge(width = 0.8)) +
-  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
-  scale_color_brewer(palette="BrBG") +
-  facet_wrap(~ change_type) + 
-  theme(legend.position = "bottom") + 
-  coord_flip() +
-  labs(title = "Figure 1", 
-       x = "", 
-       y = "Predicted Difference")
-```
-
 This is not to say that young people do not have substantively different opinions (they likely do), just that they are not any more or less stable by the metrics we outlined above.
@@
 
-## Conclusion
+This is a rather striking pattern that challenges the notion that
+
+## Discussion
 

### 2023-08-11 formative_period.Rmd
edits 811

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "7/06/2023"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: Young adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of attitude change. This probability has often been attributed to the social conditions of adolescence. Over the past 70 years, changes in the timing of major life course events such as marriage have transformed and extended this life stage. We leverage these changes to explore whether these life course changes, as well as broader social changes, have affected how likely people of different cohorts are to make durable changes of general political dispositions over a comparable time frame. We find strikingly little difference in the rates of any kind of stable opinions and opinion change among young people in different eras. People of all ages have similar attitude behaviors over time. The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates.

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Adolescence and young adulthood are typically identified in social science literature as “impressionable years” – a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples’ core ideas and attitudes are formed [@krosnick1989; @alwin1991]. Research regularly finds that people are more likely to change their attitudes and opinions in these life stages than other life stages, with opinions becoming more stable once these life stages have passed [@sears2013]. This pattern, in which young people are open to attitude change and become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952].

However, why people in this life stage appear to show more frequent opinion change relative to older people is unclear. Work in biology, psychology, and cultural evolution shows that peoples’ cognition changes in ways that might make them more resistant to new information as they age. Explanations in social sciences, in contrast, tend to highlight the unique social features of these life stages – high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing “new experiences” – that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information. While both mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.

Clarifying the mechanism that drives the frequency of opinion change during this formative period is important for understanding the broader process of social change and cultural evolution. If openness to change in early adulthood is principally a function of the social structure of life in that time period, changes in the timing and ubiquity of life course events like marriage and childbearing can have significant impacts on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than in older eras, diminishing the importance of “cohorts” for social change. If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social change in the same way.

In this paper, we seek to further elaborate this debate by exploring how changes in the broader social structure and changes in the timing of life course events since the 1950s have affected rates of change in political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population. Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one’s parents’ home, and having children, have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005]. Similarly, this stage of life has become more detached from family and other role obligations, giving people more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. This suggests both a lengthening and a intensification of this stage of the life course in ways that should increase the rates at which people make changes of opinion.

We take advantage of this change over time to explore whether young adults of different cohorts demonstrate more or less change in general political opinions over time relative to other members of the population. We compare rates of opinion change in comparable general political attitudes across four eras – 1956-1960; 1972-1976; 1992-1996; and 2006-2010 – and between young people and old people in these eras, to evaluate a range of competing hypotheses about what accounts for the divergence in attitude behaviors over the life course.

Broadly, we find that the probability of political opinion change in people under 30 have not increased as the “early adulthood” window has lengthened and the conditions of this life stage have changed. Across eras, young people make political opinion changes at remarkably similar rates, with social transitions such as marriage, having children, and completing education playing little role in the rates at which these opinion changes happen. While these findings do not provide conclusive evidence of one mechanism over the other, they confine the range of potential theories of attitude development and the nature of political opinions. In particular, they suggest that there is something relatively timeless – potentially biological and cognitive factors – that accounts for increased opinion variance in young people.

# Impressionable Years

The idea that the early stages of the life course are especially influential in the process of attitudinal formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The argument – at times implicit – is that the circumstances in which people are brought up leave an enduring and traceable mark on their worldviews [@gerber1998]. In the study of political socialization, this perspective enjoys a longstanding history. “Generational effects” are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014] However, the notion of “impressionable years” is more general. The burgeoning literature on cohort effects rests on the assumption that our early years play key role in shaping not just political opinions, but peoples’ cognition on many dimensions (Elder 2018). Their early years will leave a mark that will be predictive of their life outcomes, well beyond those impressionable years have passed. Thus, the notion of a formative period – at the early stages of the life-course – underpins diverse and longstanding lines of research.

The “impressionable years” hypothesis – that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age – has often been tested against other theories of opinion change over the life course. The most common are an aging stability or increasing persistence hypothesis that argues that openness to change gradually declines over time and a life stages hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a].  While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality [@wagner2019]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort explains much more variance than period for most items in the General Social Survey, suggesting a strong impressionable years pattern.

In recent years, researchers have attempted to use new approaches to quantify “openness” to social influence across the life course, especially around politics [@bartels2014; @ghitza2022]. This work has produced remarkably similar patterns with openness peaking from a person’s mid-teens through age 25 and declining quickly after that, lending more credence to the “impressionable years” hypothesis. However, this work tends to assume a constant age-based rate of openness across cohorts – that 20-year-olds in 1950 are open to influence at the same rate as 20-year-olds in 2010 – which has not been tested.

# Theoretical Models

Why is it research consistently finds the period of time between 14 and 25 to be distinctly “impressionable years?” At the outset, it is important to note that changing one’s mind involves at least two distinct processes that can be difficult to disentangle: exposure to new ideas and openness to change conditional on exposure to new ideas. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one’s mind as a function of exposure does not, producing the same pattern. It could also be the case that both change as a function of age. Any patterns we observe should be interpreted as coming from changes in one or both of these processes. In this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.

## Biological Mechanisms:

The first potential mechanism for changes in attitude stability over time is biological: that the development of the of human brain as people age makes it more difficult for people to make changes of opinions after some point in their life. This approach tends to suggest that human cognition works similar to Bayesian updating (Bartels and Jackman 2014; Sherratt and Morand-Ferron 2018; Gopnik 2020). In this framework, humans weight information equally, whether they encounter it during adolescence or middle age. However, given that by the time people reach adulthood they have already encountered a lot of information, each new piece will have less of an impact (Bartels and Jackman 2014). The distinctive thing about our early years is simply that they happen early. They prove so influential because the incremental effect of new information diminishes as we age.

It is also possible that there is something unique about human cognition during these periods: that agents ascribe more weight to the information acquired within them. Maturational arguments about the increased plasticity during our early years are consistent with this position (Mata and von Helversen 2015). Our brains exhibit more malleability during childhood and adolescence (Mata and von Helversen 2015). This is consistent with the fact that these early formative periods have been observed across species and contexts. It is not only human children that exhibit a heightened capacity to adapt. In a study of social learning among birds, for instance, Aplin et al (2017) show that younger individuals – unlike their older counterparts – were consistently able to retool their habits in light of new information.

Though the Bayesian updating explanation is simple and elegant, research across different disciplines has produced evidence that supports the existence of these periods of heightened sensitivity. In light of empirical evidence about these sensitive windows, there has been extensive modeling work about their evolution. This line of research seeks to understand why periods of heightened sensitivity would emerge and what contexts would facilitate their emergence (Fawcett and Frankenhuis 2015). The central argument is that the emergence and timing of these sensitive windows depends on the interaction between the access to information, the reliability of information, and the stability of the environment (Frankenhuis and Walasek 2020). Early formative periods are likely to evolve in specific configurations: when environments are unstable, and information is plentiful yet unreliable (Frankenhuis and Walasek 2020). In these circumstances, it makes sense for agents to explore early on, to find something that works and then to stick to strategies that have worked well in the past. Access to different types of information might vary across time and this might explain the emergence of sensitive periods at later stages of the life-course (Walasek, Frankenhuis, and Panchanathan 2021). For example, the importance of musical tastes and fashion choices becomes important around adolescence, and, at this stage, we are awash with – at best contradictory – information about what one should or should not like, should or should not wear. Sensitive periods around the development of these tastes might peak around this period, when there are more signals and when the payoffs are more salient.

The thrust of the biological arguments is that, through millennia of evolution, humans have developed a cognitive architecture that makes them relatively open changing dispositions at younger ages and relatively resistant at older ages. However, while biological mechanisms can shape how individuals respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change. For that, we must turn attention to the structuring of the social world.

## Life Course and Social Roles:

Humans’ early years might be distinct in terms of cognitive plasticity, but they are also socially distinct: they are characterized by sustained social learning and punctuated with constant changes of social contexts. Here we briefly consider some of the ways that the “impressionable years” differ in ways that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.

First, people in this age group tend to move through social contexts much more quickly than people in other age groups. In the 11-year window from ages 14 to 25, in which people seem to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year in these contexts potentially exposing people to new classmates, teachers, and other role models, peers, and sources of information. People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons [is this true?]. Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups.

Second, people in the “impressionable years” window spend their time in different social contexts than people in middle age. Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55. These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions. For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, gender egalitarianism, and racial inequality [@campbell2016; wodtke2018]. Social networks also differ in this time period. Previous work has found that young peoples’ networks tend to be larger and include more friends than kin, for example [@marsden1987].  It could be the case that larger “weak tie” networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].

Third, these are periods in which exploration and self-discovery are expected and encouraged; something that should not be taken for granted given social variation in norms around when maturity is reached [@cunningham2007; @elder1994].

## Updating Conditional on Exposure

While social conditions might expose young people to new information at higher rates, it might also make it easier for them to act on these new sources of information in ways that facilitate opinion change.

Previous work on stability of attitudes over the life course suggests that people tend to value stability in figures of authority. As people move into roles with more authority in mid-life, both in family and occupational contexts, they feel this expectation from others and act accordingly, closing themselves off to updating beliefs [@eaton2009]. This potentially explains another phenomenon with some evidence behind it: that people appear to open up to changing their minds again as they enter post-retirement [@danigelis2007]. At this point, peoples’ authority diminishes, potentially opening them up to change later in life.

Similarly, mid-life is associated with more homogenous networks [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004]. If young people exist in less dense, more heterogeneous social networks, they might it easier to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks, or close off parts of their lives to people who might not be supportive of new attitudes.

Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the “impressionable years ” are principally an effect of how we have chosen to structure the life course, or the interaction between these life course stages and biological mechanisms.

# Changing Life Course

Contemporary historical trends offer the opportunity to disentangle the role of socio-cultural and  biological factors in explaining why we observe “impressionable years.” In the U.S., the last decades have seen considerable changes around adolescence and young adulthood (Buchmann 1989; Rosenfeld 2007; Rosenfeld and Kim 2005). The expansion of higher education and increased possibilities for mobility are among the many social processes that have led to changing expectations for the trajectories that young adults should follow (Rosenfeld 2007). In general, as Buchmann  (1989) notes, life trajectories have become less predictable. The “script” of young adulthood has become less rigid: it might involve prolonged higher education, moving to a different city, or dating without a stable romantic partner. The routes that people are taking into adulthood are increasingly diverse. For the purposes of this paper, there are two accompanying trends that are particularly important: the elongation and intensification of young adulthood.

## Timing

The period between adolescence and “settling down” seems to be getting longer, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts. Buchmann (1989) notes that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood. In the U.S., the median age at first marriage has consistently risen in the last decades and so has the median age at which families are having their first child [@rosenfeld2009]. We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021]. Thus, the markers of the transition into adulthood seem to be occurring later in the life-course and potentially in different orders. Therefore, the definition of what counts as our “impressionable years” might be extending. This, in turn, could have an effect in how long exploration lasts throughout the life-course. If this is the case, we should expect to see more attitudinal updating among individuals in age-ranges that, in previous cohorts, had displayed mainly stability. In other words, people who were born say in the mid 20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still exhibit attitudinal updating when they reach that age. Adolescents are taking longer to move through the rituals of adulthood, and this might in turn elongate the period in which exploration and self-discovery are expected.

## Changing Impressionable Years:

Another second trend relates to just how much exploration young adults can engage in during these “impressionable years.” In his work, Rosenfeld (2007) explores the rapid rise in same-sex and interracial marriages in the U.S. His argument is that one of the processes underlying these changes is that young adults are increasingly escaping the controlling gaze of their parents. When young people live with – and/or financially dependent on – their parents, the latter have more say about their children’s choice of partner. The increased independence and mobility that young adults have enjoyed since the mid 20th century means that they have been able to engage in courtship outside the gaze of their parents, thus making choices that would not have been sanctioned by them. Rosenfeld’s (2007) work highlights an important insight: exploration might not be only a product of how sensitive we are to novel information but also how – and whether – we can act upon such information. The social institutions where we spend our impressionable years might shape how much exploration we can engage in. An increase in independence then might entail more exploration. This assertion in turn engenders a prediction: we should see more attitudinal updating among contemporary young adults than their older counterparts, who tended to be closer to the grip of familial control, even if they have met similar life course milestones. The contemporary social organization of young adulthood, which has been developing for the last few decades, might give rise to more meaningful attitudinal changes by providing more leeway to deviate from established social norms.

This approach suggests that it is not the timing of life events that is important for rates of change, but the broader social conditions of early adulthood that might make people more malleable in recent cohorts even net of changes. In other words, unmarried, childless 25-year-old men and women in 1956 might have experienced a set of social constraints that unmarried, childless 25-year-old men and women in 2008 do not experience, leaving the latter much more open to developing new opinions than their predecessors.

# Hypotheses

Sociological work then highlights that there are principled reasons to believe that length and intensity of formative periods might be affected by changes in the timing and structure of the life course, and yet this question remains unexplored. The distinctiveness of youth remains the backbone of sociological theories of socialization and there has been work trying to outline the cognitive assumptions that would lead to marked influence of our early years. While the evidence for these formative periods has accrued, there has been limited research on the social variability of these periods. Any evidence of differences in the rate of attitudinal updating among people of similar ages across different periods would suggest that sensitive periods are not only maturational or a result of “tallying up” information. In turn, finding similar patterns of attitudinal updating across widely different social contexts would lend further credence to the idea that our early years are “impressionable” either because we weight information more heavily during these periods or because we are cementing our opinions, regardless of the social contexts we happen to find ourselves in.

The argument so far implies three hypotheses that we will test in our study. Given historical changes in the meaning of adolescence, we hypothesize that we will see differences in the rates of attitudinal updating that the same age-groups across different cohorts will exhibit. The first hypothesis then is non-directional:

*H1: There will be differences in the patterns of attitudinal updating from individuals in the same age groups but in different cohorts.*

This does not make any specific predictions about the form of attitudinal updating that more recent cohorts of young people will exhibit, just that their opinion reports will be more variable than that of previous cohorts.

The second hypothesis is based on Buchmann’s work and it concerns the fact that young adults are increasingly postponing the rituals associated with the transition into adulthood. Thus, we posit that more recent cohorts should display more exploration well into their twenties than previous cohorts.

*H2: Young adults in recent cohorts should exhibit higher levels of attitudinal updating later than their counterparts of similar ages from older cohorts.*

We expect this to take the form of decreased reliability in their attitudes, as they make more transitive, non-durable changes in attitudes over time.

The last hypothesis in turn relates to Rosenfeld’s ideas. Young adults in recent cohorts have enjoyed more independence and, therefore, might have had more opportunities to substantially change their beliefs and habits. We hypothesize then that:

*H3: Young adults in recent cohorts should exhibit higher rates of attitudinal updating than people of a similar age in previous generations.*

Broadly speaking, we expect young people to be more likely to make "real" change in their attitudes, meaning their "true" values at time $t$ will be worse predictors of their true value at time $t+1$. Second, we expect young people to have more "noise" in their opinions, meaning they will have lower reliability on their opinions than other age groups as they vacillate more as a result of influences from the environment.

In particular, we focus on two age groups compared to the broader population. The first is respondents who are under 25 when they begin the survey. These respondents have typically displayed the most distinctive attitudinal patterns relative to the rest of the population, and this period has been considered the "impressionable years," at least of those that are documented in surveys [@alwin2007; @krosnick1989]. The second age group is respondents aged 26-33 when they start the survey. This group has seen the most dramatic shift in life course structuring over the past seven decades. To the extent that we see an elongation of impressionable years, we should see changes among this group.

# Data and Methods

Analyzing these questions requires data on attitude reports over time for members of different cohorts at different time points. To address this, we draw on data from seven panels administered since the 1950s. These include four major panels administered by the American National Election Study in 1956-58-60, 1972-74-76, 1992-94-96, and 2000-02-04. These were all administered over a four-year period, with waves two years apart and contain a diverse array of attitudinal questions. We also use the General Social Survey's rotating panels from 2006-10, 2008-12, 2010-2014. We also evaluate the 1980 Major Panel study from the American National Election Study. Rather than being administered three times over four years, the 1980 panel was administered three times over a single year. All seven surveys reflect nationally representative samples of U.S. households.

In all panels, we identified questions that tap several categories of political attitudes: (1) political efficacy and alienation; (2) attitudes toward social groups (such as labor unions, blacks, whites, the police, Democrats, liberals, etc.); (3) attitudes toward policy issues; (4) ideological identification on a liberal-conservative scale, and (5) partisan identification. For the 1950s, 1970s, and 1980, we broadly follow Krosnick and Alwin [-@krosnick1989] in selecting variables, with the addition of some questions about foreign policy that they omitted. In the 1990 panel, we follow Alwin's identification of questions [-@alwin2007]. For the 2000s ANES panel and the three GSS panels, we identify questions similar to those identified in the earlier panels.

This produces a set of 265 attitude measures asked on a variety of scales tapping a diverse array of topics. They include four- and five-point Likert scales, binary agree/disagree response scales, 100-point "feeling thermometers", and a range of other structures.

Most questions are asked in only one panel, but a small number of questions are replicated exactly across at least two panels (excluding questions duplicated across the GSS panels, where all questions are duplicated). There are theoretically relevant trade-offs from focusing on the same questions over time. By focusing on the same questions over time, we control for question content and structure. If attributes of questions changes over time, that might affect the estimates of stability or reliability. However, the questions asked in major surveys change over time as the dominant issues change. By focusing only on questions that are repeated across eras, we limit our attention only to issues that have been salient for an extended period of time. These might behave different than the "issues of the day" -- they might get settled earlier in life, be more resistant to change, or be reported more reliably than other kinds of questions.

To ensure that our findings are robust to both sets of issues, we conduct a separate analysis of questions that are repeated across multiple panels (excluding questions replicated just across the GSS). These include partisan identification (asked in all panels), political ideology on a seven-point liberal-conservative scale (asked in ## panels), questions about federal spending on different priorities (asked in the 1990 and 2000 ANES panels and the GSS panels), and questions about government efficacy and alienation from the political process (asked in the 1970s, 1990s, and 2000 ANES panels), as well as others.

## Methods of Analysis

We follow previous work that seeks to separate true change in attitude response from measurement error in the study of attitude change and stability over the life cycle. To do this, we follow Alwin and Krosnick's approach of using just-identified simplex models that specify two structural equations for a set of three over-time measures of a given variable, $y_i$:

$$ y_{t} = \tau_t + \epsilon_t$$

$$ \tau_t = \beta_{t,t-i} \tau_{t-1} + \zeta_t $$
The first equation represents a set of measurement assumptions indicating that (1) over-time measures are assumed to be $\tau$-equivalent, except for true attitude change, and (2) that measurement error is random. It decomposes the variance in reports into two components, one due to the latent attitude ($tau_t$) and the other due to random measurement error ($\epsilon_t$). The second equation specifies a causal process involving attitude change over time. $\beta_{t,t-i}$ is the stability of the latent attitude during the period from $t-1$ to $t$. $\zeta_t$ represents sources of change over time in the latent attitude. The model assumes that the system of attitudes is in a dynamic equilibrium described by a lag-1 or Markovian process in which the distribution of a "true" attitude at time $t$ is dependent on the distribution at time $t-1$ and not on the distribution of the variable at previous times.

As presented above, the model is unidentified. All strategies for estimating these quantities require the lag-1 assumption regarding the nature of "true" attitude change, as well as an assumption about measurement error. An approach proposed by Heise [@heise1969] assumes equal reliabilities over occasions of measurement. Another approach is to assume constant error variances rather than constant reliabilities [@wiley1970]. We follow the Heise approach and constrain reliabilities to be constant across waves, rather than constraining error variance to be the same. However, extensive work suggests that both models produce similar estimates of the standardized parameters [@alwin2007; @alwin1991].

We divide each sample into seven age categories: 18-25, 26-33, 34-41, 42-49, 50-57, 58-65, 66-83, assessed at the initial survey year. We analyze these estimates in two ways. First, we explore the stability of the latent attitudes, $\beta_21$ and $\beta_32$. These estimates are non-independent, so we examine them as multiple indicators of the stability of the attitude in that panel. We take these reliabilities as a measure of how much durable change in attitudes occurs in an issue during the panel, with higher values indicating less change. Second, we examine a single reliability estimate generated for each group for each attitude measure. We take this as a combined measure of actual measurement error and transitive, non-durable change that respondents make during the course of the panel. For each age group, we average across all questions in a given panel to generate an average estimate of reliability and an average estimate of stability for each age group in each panel.

We compare these average estimates of reliability and stability for age groups across panels to evaluate our hypotheses.

## Question Fixed Effects

For questions that are asked in multiple panels, we also use estimates of stability and reliability as outcome variables in a regression on time with fixed effects for questions. We run these models separately for each age group. These models let us allow for an average trajectory of reliability and stability for age groups over time, controlling for questions, to evaluate whether respondents in the same age groups have become more or less stable reporters of the same opinions over time. We also control for attributes of panels that might affect reliability or stability estimates -- such as the shortened wave duration in the 1980 panel or the mode of administration -- by including panel fixed effects in the model as well.

# Results

## Stability and Reliability

Before digging into the time trends, I first evaluate the average estimates for reliability and stability for each panel, and for each age group to establish a baseline of comparison. Table 1 presents the average reliability and stability for each panel. Across panels, the 1956 panel shows the lowest average reliability of all the panels.[^improvement] The 1980 panel, with the shortest time between waves, has notably higher reliability than other panels. This has been established in previous work [@krosnick1989]. Having less time between waves increases the reliability of responses, though the exact mechanism for this has not been explained. The remaining panels all have comparable reliabilities, with many within .02 of each other.

[^improvement]: This is potentially due to improvements in question-asking since the ANES first administered this survey. Many questions in this panel have been critiqued as difficult for average Americans to answer.

Average stability estimates are generally comparable across panels, ranging withing a couple points of each other. Of note here is that differences between the GSS panels are quite large, with the 2008-10-12 panel showing the highest label of stability while the 2010-12-14 panel showing the lowest, even though these panels overlap considerably.

Table XXX presents the average reliability and stability for age groups. Here we see results broadly consistent with the "impressionable years" hypothesis. First, the 18-25 year old group has notably lower stability than other groups, suggesting higher rates of change in their "true" attitudes over time. Stability tends to increase up through middle age (42-29) before leveling off. Second, reliability is also substantially lower for the youngest age group and increases consistently as groups get larger. For both measures, the largest jump is between the youngest age group (18-25) and the one following it (26-33), supporting the impressionable years model over an increasing persistence model.

While these results suggest that life-course patterns that have been established in other work using earlier panels hold true in more recent panels, they do not say anything about whether changes in the life course have extended or intensified the "impressionable years" by decreasing stability and reliability in the youngest group (18-25) over time or extending the values seen in that group up the age ladder (to respondents in the 26-33 group). For that we must examine age groups separately by panel.

## Over-Time Trends

Table XXX summarizes the average reliability and stability estimated for each of the age groups for each of the panels analyzed. The "formative period" or "impressionable years" hypothesis and previous work [@alwin2007; @alwin1989] suggests that we should observe low reliabilities for the youngest cohorts and large and approximately equal in the older groups. It is important to note that some age-group/panel combinations have very small sample sizes, which can lead to large sampling errors and produce problematic estimates of reliability. Because of this, we remove some reliability and stability estimates that are outside the bounds of plausibility or were not able to be estimated.

Because information in the table is difficult to process, Figure ... plots the average reliability estimates by year. The figure and table show the same overall age trend as the aggregate data. Young people are the least reliable respondents in most panels, suggesting high rates of non-persistnet change. Broadly, across panels, we see a general positive relationship between age and stability up through middle age, but a decrease in stability following that. This is consistent with some work suggesting that people "open up" for change later in life after a period of increased stability in middle age [@danigelis2007].

While respondents in other age groups have a generally positive trend in reliability, meaning they make less non-persistent change in more recent panels, the same is not true of the youngest age group. Members of this group have a flat age trend. At the same time, the same is not true of the second-youngest age group (26-33), which is where we might have expected to see the most significant shift over time. While this group tends to have the second-lowest average reliability over time, the trend for this age group is highly consistent with that of the rest of the population, displaying higher levels of reliability over time.

Turning to stability, we see a similar pattern. Figure XXX plots the average stability for each age group over time. Young people are consistently less stable opinion holders than the rest of the population. Stability generally increases with age, with a decline after middle age (sometimes getting as low as the 18-25 group. The gap between the 18-25 group and the 26-33 group is often the largest gap. However, we do not see a strong time effect among stability. Groups are all slightly more stable in more recent panels than in the early panels, but not enough to say there has been a significant shift in stability since the 1950s.

## Question Fixed Effects

The preceding analysis has assumed that panels are generally comparable, and that average estimates of reliability and stability are comparable across panels as measures of reliability and stability toward "salient" issues for that period. We now turn toward exploring the reliability and stability for questions that are asked in multiple panels. Table XXX presents regression coefficient estimates for model regressing stability estimates on time and fixed effects for question and panel, with an interaction between time and age group. The table shows that there is no discernible time trend for the stability estimates of 18-25 year olds, nor is their time trend statistically significantly different from the time trends for the other age groups.

Table XXX presents similar coefficient estimates for the model regressing reliability on time and fixed effects for question, with an interaction between time and age group. The table also shows no meaningful time trend for respondents 18-25. The coefficient for the interaction between time and the 58-65 age group is statistically significant, suggesting people in this group became slightly more stable over time compared to 18-25 year olds, but it is difficult to make a clear judgment about the overall trend, and there are no other statistically significant interaction terms.

# Discussion

Theoretical models that explain the heightened malleability of attitudes during early adulthood by referencing the unique social conditions of that life phase would posit that significant changes in the timing of life course events should produce heightened rates of attitude change during early adulthood (18-25) or should extend rates of attitude updating later into life (ages 26-33). As Americans have pushed back marriage, childbirth, and entering the workforce, we should expect these to have some effect on the sources of attitude change, such as meeting new significant others, getting exposed to new kinds of information, or delaying social positions that expect stability. We tested that claim by looking at the stability and reliability of political attitude reports among different cohorts as they moved through the same age groups.

The results from the preceding analyses suggest that the changing life course has had minimal impact on rates of attitude change in the population. Despite significant shifts in age of first marriage and age of first child birth, and a much higher proportion of the population attending higher education, attitude behavior as measured by reliability and attitude stability are remarkably similar over time. Across panels, we see consistent evidence that ages 18-25 have decreased attitude reliability and decreased attitude stability relative to the rest of the population, a pattern consistent with the "impressionable years" hypothesis that has broad support in the sociological literature. However, there is consistent change over time in either reliability or stability estimates for people in this age group. They are different from the rest of the population, but similarly different over time. It does not seem to be the case that the emergence of the "Age of Independence" since the middle of the 20th Century has increased attitude change among people in this group.

Similarly, as life course transitions have been pushed back since the middle of the 20th Century, we expected to see decreased reliability and stability extend into the next age group (ages 26-33). However we see no clear evidence of changes in either measure. These peoples' attitude behavior is consistently closer to the next oldest group (34-41) than the 18 to 25 age group. While social life for people in this group tends to resemble the younger group, their attitude behavior tends to resemble older groups.

Controlling for changes in the kinds of questions asked using fixed effects models, we see no significant time trends for either stability or reliability.

Overall, our findings suggest that the substantial shifts in the timing of life course transitions, increased rates of college-going, and other major social shifts that have occurred over the past 60 years has not affect attitudinal behavior in ways suggested by a "social" explanation for the "impressionable years" hypothesis would suggest. Instead, it seems that most of what makes the "impressionable" years impressionable has to do with age itself. This points toward biological or cognitive mechanisms about how people respond to new information over time as the key source of "impressionable years".

While we have focused on the time trend as the major question in our analysis, our results speak to a couple other domains of sociological inquiry. Our results reinforce findings that middle-aged respondents appear to have the most reliable and stable attitudes than members of other age groups. Respondents 50-65 consistently demonstrated the highest average reliability, with respondents over 65 displaying slightly lower levels of reliability.

## Limitations

We face several limitations in trying to make broader claims. First, many of our age-group estimates of stability and reliability are based on relatively small samples and as a result are not highly reliabile. This is particularly true for the youngest and oldest age groups, which tend to be the smallest in each panel. We have not accounted for that imprecision in our analysis here.

Second, while we look at more than 250 questions in panels spanning 60 years, we do not actually have a large number of distinct time points. In total, we only have eight estimates of "average stability" and "average reliability" for each age group. Four of our eight panels fall within the same decade (and three of these panels overlap). Again, given the large amount of unreliability in our estimates for groups, we cannot rule out the possibility that clearer trends would emerge with more data. It is not clear how to improve this limitation, as we can not go back in time and generate new panels.

Second, the high diversity of questions comes at the cost of not having many questions asked in multiple panels. Outside of the GSS panels (which span eight years total), most questions examined here are only asked in one panel. Of those that are asked in multiple panels, most are asked only twice, which gives us limited power for understanding change in reliability and stability net of change in question wording. It also forces us to assume that estimates of reliability and stability are comparable across panels as estimates of general reliability and stability of "political attitudes" at that time -- a claim that might not be valid.

# Conclusion

This paper set out to test whether the unique rates of attitude change found among young adults are due to the social features of this life stage or due to features of how human brains develop and process information over time. We leverage delays in the timing of life course transitions and the emergence of an "age of independence" over the past 70 years to explore whether attitude behavior shifted in response to these changes. We expected that we should see more durable attitude change in early adulthood and an extension of the "impressionable years" pattern up the age ladder.

We found no evidence of either process, broadly suggesting that changes in the timing of life course transitions, as well as broader social and cultural changes around early adulthood have not broadly affected attitude behavior among young adults. The pattern of results generally suggests that what makes early adulthood unique when it comes to attitude development is not its social conditions, but how humans process information as they age. These years are not formative because they are unfettered, they are formative because they are early.

# References

### 2023-08-11 clean_datasets.R
edits 811

@@ -112,24 +112,24 @@ clean_anes5 <- anes5 %>%
                          age_1), 
          age_1 = ifelse(age_1 %in% c(98,99) & age_3 < 98, age_3-4,
                          age_1)) %>%
-  mutate(helpblk5_1 = V560044, helpblk5_2 = V580329,
-         helpblk5_3 = V600628,
-         integrate5_1 = V560074, integrate5_2 = V580333,
-         integrate5_3 = V600636,
-         govjob5_1 = V560032, govjob5_2 = V580321,
-         govjob5_3 = V600620,
-         fgnaid5_1 = V560041, fgnaid5_2 = V580327,
-         fgnaid5_3 = V600626,
-         bldschls5_1 = V560053, bldschls5_2 = V580325,
-         bldschls5_3 = V600624,
-         fghtcomm5_1 = V560056, fghtcomm5_2 = V580331,
-         fghtcomm5_3 = V600630,
-         stayhome5_1 = V560035, stayhome5_2 = V580323,
-         stayhome5_3 = V600622,
-         prvtpwr5_1 = V560059, prvtpwr5_2 = V580319,
-         prvtpwr5_3 = V600618,
-         partyid5_1 = V560088, partyid5_2 = V580360,
-         partyid5_3 = V600835) %>%
+  mutate(helpblk_1 = V560044, helpblk_2 = V580329,
+         helpblk_3 = V600628,
+         integrate_1 = V560074, integrate_2 = V580333,
+         integrate_3 = V600636,
+         govjob_1 = V560032, govjob_2 = V580321,
+         govjob_3 = V600620,
+         fgnaid_1 = V560041, fgnaid_2 = V580327,
+         fgnaid_3 = V600626,
+         bldschls_1 = V560053, bldschls_2 = V580325,
+         bldschls_3 = V600624,
+         fghtcomm_1 = V560056, fghtcomm_2 = V580331,
+         fghtcomm_3 = V600630,
+         stayhome_1 = V560035, stayhome_2 = V580323,
+         stayhome_3 = V600622,
+         prvtpwr_1 = V560059, prvtpwr_2 = V580319,
+         prvtpwr_3 = V600618,
+         partyid_1 = V560088, partyid_2 = V580360,
+         partyid_3 = V600835) %>%
   select(-c(V560044,V580329,V600628,
[…]

### 2023-08-11 clean_panels.R

@@ -1,0 +1,1313 @@
+#Cleaning for Reliability Models
+library(tidyverse)
+library(haven)
+library(broom)
+
+
+anes5 <- read_dta("~/Dropbox/data/anes/anes5660/anes_mergedfile_1956to1960.dta") 
+
+clean_anes5 <- anes5 %>% 
+  select(#Negroes fair treatment
+    V560044,V580329,V600628, 
+    #deseg
+    V560074,V580333,V600636,
+    #Job guarantee
+    V560032,V580321,V600620,
+    #Help poor countries
+    V560041,V580327,V600626,
+    #Build schools
+    V560053,V580325,V600624,
+    #Fight communism
+    V560056,V580331,V600630,
+    #Leave utilities to prvt mkt
+    V560059,V580319,V600618,
+    #US better off stay home
+    V560035,V580323,V600622,
+    #Partyid
+    V560088,V580360,V600835,
+    #closerjob
+    V560034, V580322, V600621,
+    #closeraid
+    V560043, V580328, V600627,
+    #closertreat
+    V560046, V580330, V600629,
+    #closerbuild
+    V560055, V580326, V600625,
+    #closerfghtcomm
+    V560058, V580332, V600631,
+    #closerprvtpwr
+    V560061, V580320, V600619,
+    #closerdeseg
+    V560076, V580336, V600639,
+    #Age
+    V560295, V580472, V600688, 
+    #Marital status
+    V560177, V580474, V600690,
+    #Education
+    V560181, V580478, V600694,
+    #Children in household
+    V560178, V580475, V600691,
+    #sex
+    V560171, V580468, V600684,
+    #race
+    V560172, V580469, V600685,
+    #survey weight
+    V600569) %>%
+  mutate(weight = V600569,
+         age_1 = V560295,
+         age_2 = V580472,
+         age_3 = V600688,
+         marital_1 = V560177, marital_2 = V580474,
+         marital_3 = V600690,
+         marital_1 = ifelse(V560177 == 9, NA, V560177),
+         marital_1 = ifelse(V560177 %in% c(2), "single/nm", marital_1),
+         marital_1 = ifelse(V560177 %in% c(1), "married", marital_1),
+         marital_1 = ifelse(V560177 %in% c(3,4,5,7), "other", marital_1),
[…]

### 2023-08-11 correct_rels.R

@@ -1,0 +1,342 @@
+library(tidyverse)
+library(lavaan)
+library(haven)
+
+source("~/Dropbox/extended_formative_windows/clean_panels.R")
+
+
+
+
+
+heise_rel_con <- '
+
+y1 ~~ v1*y1
+y2 ~~ v2*y2
+y3 ~~ v3*y3
+
+y1 ~~ c12*y2
+y1 ~~ c13*y3
+y2 ~~ c23*y3
+
+r12 := c12/(sqrt(v1)*sqrt(v2))
+r13 := c13/(sqrt(v1)*sqrt(v3))
+r23 := c23/(sqrt(v2)*sqrt(v3))
+
+stab12 := r13/r23
+stab23 := r13/r12
+
+rel := (c12*c23)/(c13*v2)
+'
+
+
+heise_rel_fun <- function(x, n_vals) {
+  n_vals <- x$n_vals[1]
+  if (n_vals == 2) {
+    m1 <- sem(heise_rel_2, x,
+              ordered = c("y1", "y2", "y3"))
+    t1 <- tidy(m1) %>% mutate(model = "heise_rel_2")
+  } else if (n_vals == 3) {
+    m1 <- sem(heise_rel_3, x,
+              ordered = c("y1", "y2", "y3"))
+    t1 <- tidy(m1) %>% mutate(model = "heise_rel_3")
+  } else if (n_vals == 4) {
+    m1 <- sem(heise_rel_4, x,
+              ordered = c("y1", "y2", "y3"))
+    t1 <- tidy(m1) %>% mutate(model = "heise_rel_4")
+  } else if (n_vals == 5) {
+    m1 <- sem(heise_rel_5, x,
+              ordered = c("y1", "y2", "y3"))
+    t1 <- tidy(m1) %>% mutate(model = "heise_rel_5")
+  } else if (n_vals == 6) {
+    m1 <- sem(heise_rel_6, x,
+              ordered = c("y1", "y2", "y3"))
+    t1 <- tidy(m1) %>% mutate(model = "heise_rel_6")
+  } else if (n_vals == 7) {
+    m1 <- sem(heise_rel_7, x,
+              ordered = c("y1", "y2", "y3"))
+    t1 <- tidy(m1) %>% mutate(model = "heise_rel_7")
+  } else {
+    m1 <- sem(heise_rel_con, x)
+    t1 <- tidy(m1) %>% mutate(model = "heise_rel_con")
+  }
+  return(t1)
+}
+
+
[…]

### 2023-10-05 clean_panels.R
added data

@@ -1311,3 +1311,16 @@ cg10 <- clean_g10 %>%
   separate(name, into = c("var", "wave")) %>%
   mutate(wave = paste0("y", wave)) %>%
   spread(wave, value) 
+
+save(cg10, file = "/Dropbox/extended_formative_windows/clean_data/cg10.Rdata")
+save(cg8, file = "/Dropbox/extended_formative_windows/clean_data/cg8.Rdata")
+save(cg6, file = "/Dropbox/extended_formative_windows/clean_data/cg6.Rdata")
+save(canes5, file = "/Dropbox/extended_formative_windows/clean_data/canes5.Rdata")
+save(canes7, file = "/Dropbox/extended_formative_windows/clean_data/canes7.Rdata")
+save(canes8, file = "/Dropbox/extended_formative_windows/clean_data/canes8.Rdata")
+save(canes90, file = "/Dropbox/extended_formative_windows/clean_data/canes90.Rdata")
+save(canes9, file = "/Dropbox/extended_formative_windows/clean_data/canes9.Rdata")
+save(canes0, file = "/Dropbox/extended_formative_windows/clean_data/canes0.Rdata")
+
+
+

### 2023-10-05 correct_rels.R

@@ -2,7 +2,7 @@ library(tidyverse)
 library(lavaan)
 library(haven)
 
-source("~/Dropbox/extended_formative_windows/clean_panels.R")
+source("~/Dropbox/formative_period/clean_panels.R")
 
 
 
@@ -104,21 +104,30 @@ simplex_reliabilities <- bind_rows(canes5 %>% mutate(df = "anes5"),
 
 
 simplex_full <- bind_rows(canes5 %>% mutate(df = "anes5"),
-                          canes7 %>% mutate(df = "anes7"),
-                          canes9 %>% mutate(df = "anes9"),
-                          canes90 %>% mutate(df = "anes90"),
-                          cg6 %>% mutate(df = "gss6"),
-                          canes8 %>% mutate(df = "anes8"),
-                          canes0 %>% mutate(df = "anes0"),
-                          cg8 %>% mutate(df = "gss8"),
-                          cg10 %>% mutate(df = "gss10")) %>% 
+                                   canes7 %>% mutate(df = "anes7"),
+                                   canes9 %>% mutate(df = "anes9"),
+                                   cg6 %>% mutate(df = "gss6"),
+                                   canes8 %>% mutate(df = "anes8"),
+                                   cg8 %>% mutate(df = "gss8"),
+                                   cg10 %>% mutate(df = "gss10")) %>% 
   filter(age_1 < 83) %>%
   filter(!is.na(y1), !is.na(y2), !is.na(y3)) %>%
   group_by(var, df) %>%
-  mutate(n_vals = (max(y1, na.rm = TRUE)-min(y1, na.rm = TRUE))+1) %>%
   nest() %>%
   mutate(measure = map(data, heise_rel_fun)) %>%
   unnest(measure)
+
+
+nix <- canes9 %>% 
+  filter(var == "abort") %>% 
+  filter(!is.na(y1), !is.na(y2), !is.na(y3)) %>%
+  mutate(group = ifelse(age_1 < 26, "18-25",
+                        ifelse(age_1 >= 26 & age_1 < 34, "26-33",
+                               ifelse(age_1 >= 34 & age_1 < 42, "34-41",
+                                      ifelse(age_1 >= 42 & age_1 < 50, "42-49",
+                                             ifelse(age_1 >= 50 & age_1 < 58, "50-57",
+                                                    ifelse(age_1 >= 58 & age_1 < 66, "58-65",
+                                                           "66-83")))))))
 
[…]

### 2023-10-05 extended_abstract_asa23.Rmd

---
title: "Extended Abstract: Extended Early Adulthood and the Unchanging Probability of Opinion Change"
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
fontsize: 12pt
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{setspace}
  - \usepackage{lineno}
  - \def\linenumberfont{\normalfont\tiny\sffamily}
bibliography: efw_bib.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
library(tidyverse)
library(brms)
library(patchwork)
theme_set(jtools::theme_apa())
```

## Motivation

Adolescence and young adulthood are typically identified in the social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents; when their attitudes are more malleable; and when their core dispositions, identities, and opinions are formed [@alwin1991]. A broad body of research finds that rates of opinion change tend to peak during the ages of 14 to 25, with opinions becoming more stable once people move into middle age [@sears1999; @bartels2014; @ghitza2022]. And this recurring empirical pattern is central to broader dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952; @vaisey2016].

However, why people in these life stages appear to show more frequent opinion change relative to older people is unclear. Work in biology, psychology, and cultural evolution suggests that peoples' cognition changes in ways that might make them more resistant to new information as they age. Explanations in sociology, in contrast, tend to focus on the unique social features of these life stages -- including high rates of mobility between social contexts and participation in educational institutions -- that shape how people are exposed to new sources of information and their ability to update opinions in the face of this information [@visser2004]. While both mechanisms likely play a role in explaining the higher rates of opinion change in this life stage, the relative contributions of each are unclear.

In this paper, we leverage changes in the timing, ubiquity, and standardization of life course transitions over the second half of the 20th Century [@buchmann1989; @brückner2005] to adjudicate the relative contributions of developmental and social mechanisms to the heightened rates of opinion change we observe in young people. Since the 1950s, the median age of marriage and childbirth have each been pushed back by about 10 years for both men and women [@eickmeyer2017; @manning2014], rates of college-going have increased dramatically, and people make more transitions between workplaces and between the workforce and education [@brückner2005]. These shifts have led to a dramatically expanded window of participation in educational institutions, geographic and social mobility, and freedom from various social obligations and expectations, all of which are assumed to facilitate opinion change [@rosenfeld2005]. If the heightened rates of opinion change we observe in young adults are principally due to the social structure of this life stage, then we should observe changes in the rates of opinion change in this group over time.

## Data & Analytic Strategy

Drawing on four nationally representative panel surveys, dating from the mid-1950s through the early 2000s, we compare multiple different measures of opinion change in general political opinions and partisan identification of early adults (people under 30) and the rest of the population. We draw on four different panel surveys of the American adult population: the 1956-60, 1972-76, and 1992-96 American National Election Study panels, and the 2006-10 General Social Survey panel. All four surveys are nationally representative samples of non-institutionalized adults in the United States. All panels also interviewed survey respondents three times over a four-year window, making the duration of each survey window comparable. All four panels also include comparable questions about general sentiments toward government action, a measure of partisan identification, and measures of life course transitions such as marriage and childbearing.

We focus on questions about general political dispositions that were asked in all three waves of each panel, with a total of 29 questions, with between 6 and 8 questions in each panel. These questions tend to focus on the national government's role in the economy, the government's obligation to ensure equal opportunity or outcomes, and the government's role in helping minorities, as well as self-placement on a liberal to conservative scale. We also include a seven-point measure of partisan identification that is asked similarly in each panel. No other questions are directly comparable across all of the panels, though some questions are asked in comparable ways across at least two panels. Rather than adopt a single approach to quantify the rates of change across questions, we use a variety of measures of "change." We calculate the absolute difference between respondents' opinions at time 3 and their response at time 1 ("Total Distance"), the standard deviation of responses at all three waves ("S.D."), and construct indicators for whether they cross or select the scale midpoint at any point over the three waves ("Cross"), whether they finish on a different side of the scale from where they started ("Cross+Stay"), and whether they change more than two positions on the five- or seven-point scale that they are measured on ("Change \>2"). Each of these measures has different strengths and weaknesses, though none perfectly captures an error-free measure of change.

For each definition of change, we regress that outcome on an indicator variable for whether people are aged 30 or less. For the continuous outcomes, we use ordinary least squares regression and for the dichotomous outcomes we use logistic regression. In later analyses, we include covariates for marital status, whether the respondent has children, and their highest level of education to evaluate whether these life-course transitions account for (or obscure) gaps between age groups over time. This approach produces an extremely large number of models, and will likely produce statistically significant coefficients at the $p < .05$ level for a substantial proportion of coefficients simply by chance. Our goal is not to interpret any single coefficient but to interpret overall patterns if they emerge. To that end, we principally focus on comparing plots of many models rather than evaluating specific questions.

## Results

Figure 1 plots the predicted opinion change (for scalar outcomes: total distance and standard deviation) or predicted probability of opinion change (all other measures) for each of these questions, for respondents aged 30 and older (red dots) and for respondents under 30 (blue dots). Each column represents a different measure of "change" as outlined above, and the vertical axis is separated into the different panels, with questions from the oldest panel (the 1956-60 ANES panel) at the top and questions from the most recent panel (the 2006-10 GSS) at the bottom. If early adults have become more likely to make opinion changes over time relative to people over 30, we should see points diverge as we move down the figure.

```{r, include=TRUE, fig.height=8, fig.align="center", fig.cap=c("Predicted change distance or probability of change"), echo=FALSE}
knitr::include_graphics("./predicted_change_values.pdf")
```

While there are idiosyncratic questions in each panel, broadly speaking, there are a number of consistent patterns across the panels. First, there are high rates of opinion instability in the population, with the proportion of respondents crossing or selecting the scale midpoint ("Cross") often exceeding 80 percent for both early adults and the rest of the population. At the same time, little of this change is durable, with the proportion of respondents ending the survey on a different side of the scale ("Cross+Stay") often being less than 10 percent of the population. In other words, respondents are highly inconsistent in reporting their general political beliefs and tend to report opinions near the scale mid-point. This pattern is consistent with several decades of work on the stability of Americans' political opinions [@converse1964; @zaller1992]

Second, the population as a whole does not appear to become either more or less consistent on these general political issues over time. While it is possible that variation in which questions are included in each panel obscure an overall trend, there is no obvious shift in stability or instability over time in the measures we looked at. An average of 57 percent of respondents over 30 switched sides of the scale to each question in the 1950s, and about 60 percent of respondents over 30 switched sides of the scale to each question in the 2000s. That is remarkable stability despite significant shifts in the partisan landscape, the issues at the center of American politics, and the composition of the population.

Third, there is no consistent age-based pattern of opinion stability, especially over time. On most questions, in most panels, the rates of opinion change between early adults (under 30) and the rest of the population are statistically indistinguishable. This is not simply an issue of statistical power. Across all issues, young respondents are about as likely to have lower rates of opinion change than older people as higher rates of opinion change, and these differences are frequently only one or two percentage points. When we create composite scales from all issue questions in a panel by rescaling and averaging all questions so that higher values indicate more conservative responses, an approach designed to minimize measurement error in issue positions [@ansolabehere2008], we find no statistically significant difference between people under 30 and people 30 and older in any of the panels (not shown). This is somewhat surprising, given the general expectation that young people would be more likely to change opinions.

At the same time, there are some observable differences between young and old respondents on a small number of questions that do reinforce the general expectation. On partisan identification, the one question included in all panels, respondents under 30 are consistently more likely than older respondents to change sides of the scale. But we do not see a clear trend over time in this difference. This question also shows lower overall rates of inconsistency across both young people and old people, suggesting it behaves quite differently from the other kinds of opinions analyzed here.

None of this should imply that young people do not have substantively different opinions (they likely do, but that is beyond the scope of this analysis). But what we can say is that they are not any more or less stable by the metrics we outlined above. There are only a handful of issues where we observe a consistent pattern across most measures of attitude change: partisan identification in most panels, the issue of busing in the 1970s, and general views on government spending in the 1990 panel.

Further analyses (not included here) explore whether life-course transitions obscure or exacerbate age-based differences. We find little relationship between marital status and the likelihood of opinion change in any panel. While higher levels of education are associated with less change over time in the later panels, we find that including it in a model with age does not affect the age coefficient in any meaningful way.

## Discussion

Despite significant shifts in the timing of life course events and the broad social structure of early adulthood from the mid-1950s to today, there is little change in the rates at which people between the ages of 18 and 30 undergo major changes of political opinions. We suggest four possible explanations for this aggregate stability. First, and we believe most plausibly, the patterns observed in this analysis broadly suggest that there is something developmental that explains broad patterns in attitude instability early in the life course. Second, it is possible that shifts in social structure push in opposite directions to produce aggregate stability both in the rate at which respondents over 30 make changes of opinion and the difference between younger and older respondents. However, these shifts would have to be unrelated to life-course transitions like completing education and getting married, which are unrelated to the difference between young and old respondents. Third, it is possible that the features that explain stability are still social in nature, but are rooted in factors that have not undergone significant changes since the 1950s, though it is unclear what those might be. Finally, it might be the case that issue turnover across the panels explains the lack of a trend. It could be the case that panels select questions that produce high amounts of opinion instability in the general population and that obscure differences between young and old people. However, because we observe no clear trend in the one question that is asked in all panels, partisan identification, we think this is unlikely.

## References

### 2023-10-05 manuscript/formative_period.Rmd

---
title: Extended Early Adulthood and Attitude Change
author:
  - Kevin Kiley, North Carolina State University
  - Nicolas Restrepo, University of California, Davis
date: "7/06/2023"
output: 
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-02.docx
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    toc: false
    number_sections: true
bibliography: ["kileybib.bib"]
header-includes:
   - \usepackage{setspace}
   - \setlength{\parindent}{4em}
   - \setlength{\parskip}{0em}
   - \usepackage[markers,nolists]{endfloat}
   - \usepackage{graphicx}
mainfont: Minion Pro
fontsize: 12pt
abstract: Young adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of changing their personal culture. This probability has often been attributed to the social conditions of adolescence. Over the past 70 years, changes in the timing of major life course events such as marriage and childbirth have transformed and extended this life stage. We leverage these changes to explore whether life course changes, as well as broader social changes, have affected how likely people of different cohorts are to make durable changes of general political dispositions over a comparable time frame. We find strikingly little difference in the rates of any kind of stable opinions and opinion change among young people in different eras. People of all ages have similar attitude behaviors over time. The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Adolescence and young adulthood are typically identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples' core ideas and attitudes are formed [@krosnick1989; @alwin1991]. Research regularly finds that people are more likely to change their preferences, attitudes, and opinions in these life stages than other life stages, with opinions and dispositions becoming more stable once these life stages have passed [@sears2013; @kiley2020]. This pattern, in which young people are open to cultural change and become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952].

However, why people in this life stage appear to show more frequent opinion change relative to older people is unclear. Explanations in the social sciences tend to highlight the unique social features of these life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information. Work in biology, psychology, and cultural evolution, in contrast, shows that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION]. While both mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.

Clarifying the relative importance of different mechanisms that drive the frequency of opinion change during this formative period is important for understanding the mutually reinforcing processes of social change and cultural evolution. If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time. If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social change in the same way.

In this paper, we explore how changes in the broader social structure and changes in the timing of life course events since the 1950s have affected the stability and reliability of political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population. Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005]. Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. This suggests both a lengthening of and an intensification of this stage of the life course in ways that potentially affect the rates at which people undergo cultural change.

We take advantage of this change over time to explore whether young adults of different cohorts demonstrate more or less change in their political opinions over time relative to other members of the population. We decompose over-time responses into reliability (a lack of random, non-persisting fluctuations) and stability (true consistency in latent dispositions net of these random fluctuations) for a range of political attitudes in nine nationally representative panel data sets spanning from the 1950s through the 2010s. We separately estimate these parameters for a comparable set of age groups in each panel to evaluate whether response reliability or stability changed as new cohorts moved through the same age ranges under different social conditions.

Consistent with the "impressionable years" hypothesis, we find that the youngest respondents in each panel (ages 18-25) are consistently the least stable opinion reporters, suggesting they are more likely than other respondents to make durable changes in their views. They are also among the least reliable reporters, suggesting considerable short-term but non-persisting change in their views. But we find no consistent trajectory in either measure over time, among members of this age group or other members of the population. Notably, respondents aged 26 to 33 continue to look more like older age groups than the youngest age group in their rates of opinion stability and reliability, providing little evidence of "lengthening" of this formative window.

We believe these findings have important implications for our understanding of why early adulthood is a significant formative period. While these findings do not provide conclusive evidence of one mechanism over the other, they confine the range of potential theories about the relationship between life course stages and attitude change. They suggest that there is something relatively timeless -- potentially biological and cognitive factors -- that accounts for increased opinion variance in young people, and that, barring significant evidence of counter-currents that stabilize people, changes in the social conditions of early adulthood have not increased the degree to which young people make meaningful changes of opinion.

# Impressionable Years

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on their personal culture -- worldviews, attitudes, dispositions, behaviors, and preferences [@gerber1998]. In the study of political socialization, this perspective enjoys a longstanding history. "Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than just political preferences. The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed (Elder 2018). Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.

These models of social imprinting have implications for the kinds of patterns we should observe in peoples' attitudes over time, and suggest specific forms that opinion change should take. The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality traits [@wagner2019]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort -- the year of a person's birth - explains much more variance than period -- the year of observation -- for most items in the General Social Survey, suggesting a strong impressionable years pattern.

In recent years, researchers have attempted to use new approaches to quantify "openness" to social influence across the life course, especially around politics [@bartels2014; @ghitza2022]. This work has produced remarkably similar patterns with openness peaking from a person's mid-teens through age 25 and declining quickly after that, lending more credence to the "impressionable years" hypothesis [a pattern that seems to hold for music tastes as well @way2019]. However, this work tends to assume a constant age-based rate of openness across cohorts -- that 20-year-olds in 1950 are open to influence at the same rate as 20-year-olds in 2010 -- which has not been tested.

# Theoretical Models

Why is it research consistently finds the period of time between 14 and 25 to be distinctly "impressionable years?" At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern. It could also be the case that both change as a function of age. Because these mechanisms have not been disentangled, in this section we outline theoretical processes that have been suggested to contribute to one or both of these processes.

## Biological Mechanisms:

The first potential mechanism for changes in attitude stability over time is rooted in the biology of the human brain and science of how people process information. This approach tends to suggest that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally, whether they encounter it during adolescence or middle age, and are exposed to information at a (more or less) constant rate (Bartels and Jackman 2014; Sherratt and Morand-Ferron 2018; Gopnik 2020). However, because people have already encountered a lot of information by the time they reach adulthood, each new piece has less of an impact on their overall dispositions [@bartels2014]. The distinctive thing about our early years, then, is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.

It is also possible that there is something unique about human cognition during these periods making people ascribe more weight to the information acquired within them than to information acquired at other points. Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life (Mata and von Helversen 2015). Supporting this model is evidence that early formative periods have been observed not just in people, but across species. In a study of social learning among birds, for instance, Aplin et al (2017) show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.

In light of empirical evidence about these sensitive windows across species, there has been modeling work about why periods of heightened sensitivity would emerge and what contexts would facilitate their emergence (Fawcett and Frankenhuis 2015). The central argument from this work is that that the emergence and timing of these sensitive windows depends on the interaction between the access to information, the reliability of information, and the stability of the environment, and that early formative periods are likely to evolve when environments are unstable and information is plentiful yet unreliable (Frankenhuis and Walasek 2020). In these circumstances, it makes sense for agents to explore early on, to find something that works, and then to stick to strategies that have worked well in the past. Access to different types of information might vary across time and this might explain the emergence of sensitive periods at later stages of the life-course (Walasek, Frankenhuis, and Panchanathan 2021). For example, the importance of musical tastes and fashion choices becomes important around adolescence, and, at this stage, we are awash with at best contradictory information about what one should or should not like, or should or should not wear. Sensitive periods around the development of these tastes might peak around this period, when there are more signals and when the payoffs are more salient.

The thrust of the biological arguments is that, irrespective of the social conditions they inhabit, millennia of evolution have led to the development of a cognitive architecture that makes humans relatively open changing dispositions at younger ages and relatively resistant at older ages. However, while biological mechanisms can shape how individuals respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change. For that, we must turn attention to the structuring of the social world.

## Life Course and Social Roles:

Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct, characterized by sustained social learning and constant changes of social contexts. Here we briefly consider some of the ways that the "impressionable years" differ in ways that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.

First, people in this age group tend to move across social contexts at a much higher rates than people in other age groups. In the 11-year window from ages 14 to 25, in which people tend to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year potentially exposing people to new classmates, teachers and other role models, and sources of information. People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons. Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups.

Second, people in the "impressionable years" window spend their time in different social contexts than people in middle age. Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55. These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions. For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, and racial inequality [@campbell2016; @wodtke2018; @edelmann2023]. Social networks also differ in this time period. Young peoples' networks tend to be larger and include more friends than kin, for example [@marsden1987]. It could be the case that these larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].

Third, these are periods are also culturally distinct, with exploration and self-discovery expected and encouraged. People in this "early adulthood" phase are not expected to be stable adults, something should not be taken for granted given social variation in norms around when maturity is reached [@cunningham2007; @elder1994].

## Updating Conditional on Exposure

While the distinct social and cultural conditions of adolescence and early adulthood might expose young people to new information at higher rates than older people, the impact of this information on peoples' attitudes is contingent on their ability to incorporate it into their cultural repertoire. The social conditions of different life stages might make it easier or more difficult for people to act on these new sources of information in ways that facilitate opinion change.

Social networks not only shape peoples' exposure to new information, but they also affect the degree to which people adopt new beliefs and attitudes. People are much less likely to defect from their traditional beliefs and adopt new beliefs if they are embedded in dense networks [@stark1980]. Consistent with this, mid-life is associated with denser, more homogenous networks [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004]. If young people exist in less dense, more heterogeneous social networks, it might be easier for them to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks. Similarly, if information flows less freely through weak-tie networks, these structures might allow respondents to limit their exposure to or close off parts of their lives to people who might not be supportive of new attitudes.

Previous work on opinion stability over the life course suggests that people tend to value stability in figures holding social power. As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information [@eaton2009]. This potentially explains another phenomenon: that people appear to open up to changing their minds again as they enter later life and retirement, when these expectations presumably diminish [@danigelis2007].

# Changing Life Course

Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the "impressionable years" are principally an effect of how we have chosen to structure the life course, or the interaction between these life course stages and biological mechanisms. Because of this, contemporary historical trends offer a window to disentangle the role of socio-cultural and biological factors in explaining why we observe "impressionable years."

In the United States, the last seven decades have seen considerable social and cultural changes around adolescence and young adulthood [@buchmann1989; @rosenfeld2007; @rosenfeld2005]. The expansion of higher education and increased possibilities for mobility are among the many social processes that have led to changing expectations for the trajectories that young adults should follow [@rosenfeld2007]. These changes have made life course trajectories less predictable, and the "script" of young adulthood has become less rigid: it might involve prolonged higher education, moving to a different city, or dating without a stable romantic partner [@buchmann1989]. For the purposes of this paper, there are two related trends that are particularly important: the elongation of young adulthood and the intensification of this life stage.

## Timing

First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts. Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood. In the United States, both the median age at first marriage and median age of first child birth has consistently risen since the mid-20th Century [@rosenfeld2009]. We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021]. Thus, the markers of the transition into adulthood seem to be occurring later in the life-course and potentially in different orders.

If the distinctive nature of "impressionable years" is principally driven by their unique social circumstances, then these shifts have the potential to extend these social circumstances into later ages. This, in turn, could have an effect in how long exploration, and therefore durable change, lasts throughout the life-course. If this is the case, we should expect to see lower levels of attitude stability (and therefore more change) among age groups that, for earlier cohorts, displayed mainly stability. In other words, people who were born say in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age. Specifically, we expect that people aged 25-33, who in earlier cohorts displayed opinion patterns similar to middle-aged respondents [@krosnic1991; @alwin1991], should more closely resemble the youngest adults in more recent years. This produces our first hypothesis:

**Hypothesis 1: Rates of attitude stability will be lower among respondents aged 25-33 in recent surveys than in earlier surveys.**

## Changing Impressionable Years:

A second trend in early adulthood relates to just how much exploration young adults can engage in during these "impressionable years." In his work, Rosenfeld [-@rosenfeld2007] argues that one of the processes underlying the rapid rise in same-sex and interracial marriages in the United States is that young adults are increasingly escaping the controlling gaze of their parents. When young people live with, or are financially dependent on, their parents, the latter have more say about their children's choice of partner. Social and cultural changes that have increased independence from parental authority and mobility mean they have been able to engage in courtship outside the gaze of their parents, making choices that would not have been sanctioned by them.

Rosenfeld's work suggests that the social conditions where we spend our impressionable years might shape how much exploration we can engage in. An increase in independence might entail more exploration, and greater ability to update attitudes as a result of this exploration. This assertion in turn engenders a prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.

This approach suggests that it is not the timing of life events that is important for rates of change, but the broader social conditions of early adulthood that might make people more malleable in recent cohorts. In other words, unmarried, childless 25-year-old men and women in 1956 might have experienced a set of social constraints that unmarried, childless 25-year-old men and women in 2008 do not experience, leaving the latter much more open to developing new opinions than their predecessors. The "intensification" of impressionable years suggests two hypotheses.

*Hypothesis 2: People ages 18 to 25 will demonstrate lower levels of reliability in more recent years than in earlier years.*

Reliability reflects not just the degree to which people make errors , but the baseline level of variance around a fixed attitude position that people express.

*Hypothesis 3: People aged 18 to 25 will demonstrate lower levels of stability in recent years than they did in earlier years.*

## Biological and Cognitive Factors

We have framed our hypotheses as expectations derived from the framework suggesting that social conditions play a large role in explaining the uniquely high rates of attitude change among early adults and adolescents. However, to the extent that we do not see evidence that these social changes have not contributed, but that young adults maintain a similar pattern of lower rates of attitude stability that older adults over time, we take this as evidence that social circumstances play a relatively small role in explaining change in favor of biological and cognitive factors.

# Analytical Strategy

## Data and Methods

Analyzing these questions requires data on attitude reports over time for members of different cohorts at different time points, presumably coming from the same population. While ideally we would be able to observe different cohorts responding to the same broad range of questions as they pass through the same age periods, we do not know of such a data source.

Instead, we focus on a range of questions tapping a similar domain -- national politics in the United States. We draw on data from seven different panels administered since the 1950s. These include four major panels administered by the American National Election Study in 1956-58-60, 1972-74-76, 1992-94-96, and 2000-02-04. These were all administered over a four-year period, with waves two years apart and contain a diverse array of attitudinal questions. We also use the General Social Survey's rotating panels from 2006-08-10, 2008-10-12, 2010-12-14. We also include the 1980 Major Panel study from the American National Election Study, a four-wave panel spanning about eight months. Finally, we incldue the the 1990-91-92 Merged File, a three-wave panel spanning two years. All seven surveys reflect nationally representative samples of U.S. households.

In all panels, we identified questions that tap several categories of political attitudes: (1) political efficacy and alienation; (2) attitudes toward social groups (such as labor unions, blacks, whites, the police, Democrats, liberals, etc.); (3) attitudes toward both domestic and policy issues; (4) ideological identification on a liberal-conservative scale; and (5) partisan identification. For the 1956-60, 1972-76, and 1980 panels, we broadly follow Krosnick and Alwin's selection of variables [-@krosnick1989]. In the 1992-96 panel, we follow Alwin's identification of questions [-@alwin2007]. For the 1990-92 panel, the 2000-04 panel, and the three GSS panels, we identify questions similar to those identified in the earlier panels.

This produces a set of 265 panel-attitude measures asked on a variety of scales tapping a diverse array of topics related to national government and politics in the United States. They include four- and five-point Likert scales, binary agree/disagree response scales, 100-point "feeling thermometers", and a range of other structures.

Most questions are asked in only one panel, but a small number of questions are replicated exactly across at least two panels (excluding questions duplicated across the GSS panels, where all questions are duplicated). There are theoretically relevant trade-offs from focusing only on the same questions over time. By focusing on the same questions over time, we control for question content and structure. If attributes of questions changes over time, that might affect the estimates of stability or reliability. However, the questions asked in major surveys change over time as the dominant issues change. By focusing only on questions that are repeated across eras, we limit our attention only to issues that have been salient for an extended period of time. These might behave different than the "issues of the day" -- they might get settled earlier in life, be more resistant to change, or be reported more reliably than other kinds of questions.

By focusing on all questions tapping the domains outlined above, the average level of stability and reliability reflect change in "important" issues -- including issues that remain important for a long time and issues that only maintain salience for a few years.

To ensure that our findings are robust to both sets of issues, we conduct a separate analysis of questions that are repeated across multiple panels (excluding questions replicated just across the GSS). These include partisan identification (asked in all panels), political ideology on a seven-point liberal-conservative scale (asked in \## panels), questions about federal spending on different priorities (asked in the 1990 and 2000 ANES panels and the GSS panels), and questions about government efficacy and alienation from the political process (asked in the 1970s, 1990s, and 2000 ANES panels), as well as others.

Weighting schemes are different across panels. When available, I use weights that account for differential selection. The 1972-76, 1980, and

## Methods of Analysis

There are a number of potential ways to measure change, including linear trajectory models, latent growth curve models, and sequence analysis. We follow previous work that seeks to separate true change in attitude response from measurement error in the study of attitude change and stability over the life cycle. To do this, we follow Alwin and Krosnick's approach of using just-identified simplex models that specify two structural equations for a set of three over-time measures of a given variable, $y_i$:

$$ y_{t} = \tau_t + \epsilon_t$$

$$ \tau_t = \beta_{t,t-i} \tau_{t-1} + \zeta_t $$ The first equation represents a set of measurement assumptions indicating that (1) over-time measures are assumed to be $\tau$-equivalent, except for true attitude change, and (2) that measurement error is random. It decomposes the variance in reports into two components, one due to the latent attitude ($tau_t$) and the other due to random measurement error ($\epsilon_t$). The second equation specifies a causal process involving attitude change over time. $\beta_{t,t-i}$ is the stability of the latent attitude during the period from $t-1$ to $t$. $\zeta_t$ represents sources of change over time in the latent attitude. The model assumes that the system of attitudes is in a dynamic equilibrium described by a lag-1 or Markovian process in which the distribution of a "true" attitude at time $t$ is dependent on the distribution at time $t-1$ and not on the distribution of the variable at previous times.

As presented above, the model is unidentified. All strategies for estimating these quantities require the lag-1 assumption regarding the nature of "true" attitude change, as well as an assumption about measurement error. An approach proposed by Heise [@heise1969] assumes equal reliabilities over occasions of measurement. Another approach is to assume constant error variances rather than constant reliabilities [@wiley1970]. We follow the Heise approach and constrain reliabilities to be constant across waves, rather than constraining error variance to be the same. However, extensive work suggests that both models produce similar estimates of the standardized parameters [@alwin2007; @alwin1991].

We divide each sample into seven age categories: 18-25, 26-33, 34-41, 42-49, 50-57, 58-65, 66-83, assessed at the initial survey year. We analyze these estimates in two ways. First, we explore the stability of the latent attitudes, $\beta_21$ and $\beta_32$. These estimates are non-independent, so we examine them as multiple indicators of the stability of the attitude in that panel. We take these measures of stability as a measure of how much durable change in attitudes occurs in an issue during the panel, with higher values indicating less change. Second, we examine a single reliability estimate generated for each group for each attitude measure. We take this as a combined measure of actual measurement error and transitive, non-durable change that respondents make during the course of the panel. For each age group, we average across all questions in a given panel to generate an average estimate of reliability and an average estimate of stability for each age group in each panel.

We compare these average estimates of reliability and stability for age groups across panels to evaluate our hypotheses.

## Question Fixed Effects

For questions that are asked in multiple panels, we also use estimates of stability and reliability as outcome variables in a regression on time with fixed effects for questions. We run these models separately for each age group. These models let us allow for an average trajectory of reliability and stability for age groups over time, controlling for questions, to evaluate whether respondents in the same age groups have become more or less stable reporters of the same opinions over time. We also control for attributes of panels that might affect reliability or stability estimates -- such as the shortened wave duration in the 1980 panel or the mode of administration -- by including panel fixed effects in the model as well.

# Results

## Life Course Changes

While not the central question of our analysis, we first highlight changes in three life-course measures -- proportion of respondents who are married, proportion of respondents who have at least one child, and proportion of respondents who have a college degree -- across the panels we include in our analysis to illustrate the substantial shifts social conditions of early adulthood. The measurement of children varies between panel, sometimes asking respondents how many children they have ever had and sometimes asking respondents how many children currently live in their home. These responses are often similar for people in younger age groups, but produces very different responses in respondents in older age groups who have children that no longer live with them. For that reason, we hold respondents 50 and older off of the figure.

In the early panel (1956-60), respondents in the lowest age group look very similar to respondent in the older groups on all major measures. However, they quickly diverge by the 1970s, displaying the lowest probability of being married and the lowest probability of having children by quite a larger margin. Over time, the next youngest agr group (26-33) starts to diverge toward the youngest age group, especially in terms of marriage.

These trends set the stage for what to expect in terms of reliability and stability. To the extent that these measures as proxies for life stages that influence attitude stability and reliability, then we should see the youngest age group look similar to the older age groups in the 1950s, and diverge by the 1970s, continuing to diverge over time. Respondents in the 26-33 age group should remain similar to older respondents for most panels, but begin to diverge by the 1990s and especially by the 2000s.

## Stability and Reliability

We first evaluate the average estimates for reliability and stability for each panel and for each age group to establish a baseline of comparison. Table 1 presents the average reliability and stability for each panel. Across panels, the 1956-58-60 panel shows the lowest average reliability of all the panels.[^badq] The 1980 panel, with the shortest time between waves (about 3 months between each wave), has notably higher reliability than other panels, a fact established in previous work [@krosnick1989]. Having less time between waves increases the reliability of responses. The remaining panels all have comparable average reliabilities, with many within .02 of each other.

[^badq]: This is potentially due to improvements in question-asking since the ANES first administered this survey. Many questions in this panel have been critiqued as difficult for average Americans to answer.

Average stability estimates are also generally comparable across panels, ranging withing a couple points of each other. Of note here is that differences between the GSS panels are quite large, with the 2008-10-12 panel showing the highest label of stability while the 2010-12-14 panel showing the lowest, even though these panels contain the exact same questions and overlap considerably in terms of time.

Table XXX presents the average reliability and stability for age groups. Here we see results broadly consistent with the "impressionable years" hypothesis. First, the 18-25 year old group has notably lower stability than other groups, suggesting higher rates of change in their "true" attitudes over time. Stability tends to increase up through middle age (42-29) before leveling off. Second, reliability is also substantially lower for the youngest age group and increases consistently as groups get larger. For both measures, the largest jump is between the youngest age group (18-25) and the one following it (26-33), supporting the impressionable years model over an increasing persistence model.

These results reinforce life-course patterns established in other work and extend it to more recent panels, but they do not say anything about whether changes in the life course have extended or intensified the "impressionable years" by decreasing stability and reliability in the youngest group (18-25) over time or extending the values seen in that group up the age ladder (to respondents in the 26-33 group). For that we must examine age groups separately by panel.

## Over-Time Trends

Table XXX summarizes the average reliability and stability estimated for each of the age groups for each of the panels analyzed. The "formative period" or "impressionable years" hypothesis and previous work [@alwin2007; @alwin1989] suggests that we should observe low stability for the youngest cohorts and large and approximately equal stability in the older groups. The "extended early adulthood" hypotheses suggests that, over time, people in the 26-33 age group should come to more closely resemble people in the 18-25 age group, meaning their average stability will decrease over time as people in this age group continue to make durable changes of culture. The "intensification" of early adulthood hypothesis suggests that 18-25 year olds (and possible 26-33 year olds) will move farther apart from the rest of the population, both in terms of stability and reliability, as they make more temporary changes (lower reliability) and durable changes (lower stability) over time.

It is important to note that some age-group/panel combinations have very small sample sizes, which can lead to large sampling errors and produce problematic estimates of reliability. Because of this, we remove some reliability and stability estimates that are outside the bounds of plausibility or were not able to be estimated. If a question produced an impalusible value for one of the age groups, we remove it for all age groups.

Because information in the table is difficult to process, and because some of the panels have much higher average levels of stability and reliability because of shorter duration, Figure ... plots the average reliability estimates by panel, with each panel pegged to year it begain (e.g., the 1972-74-76 panel is plotted at 1972). The figure and table show the same overall age trends as the aggregate data. Across panels, the youngest respondents (ages 18-25) are the least stable respondents. In eight of the nine panels, they have the lowest average stability, and in the remaining panel, they have the second-lowest. However, there is no clear age pattern beyond that. Respondents in the second-youngest age group (26-33) have the second-lowest stability in only two of the panels (1990 and 2008), but this does not seem to indicate a pattern. In other panels, both early and late, these respondents seem indistinguishable from other respondents, especially in the 2006-10 and 2010-14 GSS panels (the most recent).

There is no clear time trajectory for any age group when it comes to stability, even accounting for different average duration across panels. Groups do not, on average, appear to be getting more or less stable over time, with responses from the most recent GSS panels looking quite similar to responses from the 1972-76 ANES panel.

Turning next to the question of reliability, the youngest aged group (18-25) are typically among the least reliable respondents in each panel, suggesting they are making higher rates of non-durable change in their attitudes than other member of the population. However, the pattern is not as consistent as with stability, and there is not always a clear gap between the youngest and the rest of the population. Other groups -- especially the oldest (66-83) -- frequently show less reliability in each panel. In terms of reliability, the second-youngest group (26-33) are functionally indistinguishable from middle-aged respondents in each panel, and much closer to those respondents than the youngest, providing no evidence for the "extended formative period" hypothesis. Again, as with stability, these findings reflect previous work on reliability and show that patterns continue in more recent panels.

In terms of change over time, there appears to be a slight upward trend in reliability across the panels, even accounting for the two panels with different duration -- the 1980 and the 1990-91-92 ANES panels. However, aside from the general upward trend, there do not appear to be distinct time trends for any of the age groups.

## Question Fixed Effects

The preceding analysis has assumed that panels are generally comparable, and that average estimates of reliability and stability are comparable across panels as measures of reliability and stability toward "salient" issues for that period. We now turn toward exploring the reliability and stability for questions that are asked only in multiple panels. Table XXX presents regression coefficient estimates for model regressing stability estimates on time and fixed effects for question and panel, with an interaction between time and age group. The table shows that there is no discernible time trend for the stability estimates of 18-25 year olds, nor is their time trend statistically significantly different from the time trends for the other age groups.

Table XXX presents similar coefficient estimates for the model regressing reliability on time and fixed effects for question, with an interaction between time and age group. The table also shows no meaningful time trend for respondents 18-25. The coefficient for the interaction between time and the 58-65 age group is statistically significant, suggesting people in this group became slightly more stable over time compared to 18-25 year olds, but it is difficult to make a clear judgment about the overall trend, and there are no other statistically significant interaction terms.

# Discussion

Theoretical models that explain the heightened malleability of attitudes during early adulthood by referencing the unique social conditions of that life phase suggest that significant changes in the timing of life course events should produce heightened rates of attitude change during early adulthood (18-25) or should extend rates of attitude updating later into life (ages 26-33). As Americans have pushed back marriage, childbirth, and entering the workforce, we should expect these transitions to have some effect on the sources of attitude change, such as meeting new significant others, getting exposed to new kinds of information, or delaying social positions that expect stability. We tested that claim by looking at the stability and reliability of political attitude reports among different cohorts as they moved through the same age groups.

The results from the preceding analyses suggest that the changing life course has had minimal impact on rates of attitude change in the population. Despite significant shifts in age of first marriage and age of first child birth, and a much higher proportion of the population attending higher education, attitude behavior as measured by reliability and attitude stability are remarkably similar over time. Across panels, we see consistent evidence that ages 18 to 25 have lower attitude stability and potentially lower attitude reliability relative to the rest of the population, a pattern consistent with the "impressionable years" hypothesis that has broad support in the sociological literature. However, there is no consistent change over time in either reliability or stability estimates for people in this age group. They are different from the rest of the population, but similarly different over time. It does not seem to be the case that the emergence of the "Age of Independence" since the middle of the 20th Century has increased rates of attitude change among people in this group.

Similarly, as life course transitions have been pushed back since the middle of the 20th Century, we expected to see decreased reliability and stability extend into the next age group (ages 26-33). However we see no clear evidence of changes in either stability or reliability. Attitude behavior of people in this group is consistently closer to that of people in the next oldest group (34 to 41) than to people the 18 to 25 age group. While social life for people in this group increasingly resembles the younger group, their attitude behavior tends to resemble older groups.

It does not appear that the lack of change is attributable to changes in questions over time. We observe the same patterns in stability and reliability of attitudes whether we focus only on questions that are replicated across panels .

## Limitations

We face several limitations in trying to make broader claims. First, many of our age-group estimates of stability and reliability are based on relatively small samples and as a result are not highly reliabile. This is particularly true for the youngest and oldest age groups, which tend to be the smallest in each panel, meaning we have the least confidence in each individual group-question estimate for these groups. We have attempted to limit this problem by averaging as many items as possible for each panel, but the number of questions in each panel is highly variable.

Second, while we look at more than 250 questions in panels spanning 60 years, we do not actually have a large number of distinct time points. In total, we only have nine estimates of "average stability" and "average reliability" for each age group. Four of our nine panels fall within the same decade, and three of these panels overlap. Again, given the large amount of unreliability in our estimates for groups, we cannot rule out the possibility that clearer trends would emerge with more data. It is not clear how to improve this limitation, as we can not go back in time and generate new panels.

Relatedly, most the panels span two presidential election years, in which respondents might be exposed to more information about national politics than usual that might produce heightened attitude change. Because of that, our results might not speak to other kinds of questions related to religion, and the patterns we observe might not be applicable to all times.

Second, the high diversity of questions comes at the cost of not having many questions asked in multiple panels. Outside of the GSS panels (which span eight years total), most questions examined here are only asked in one panel. Of those that are asked in multiple panels, most are asked only twice, which gives us limited power for understanding change in reliability and stability net of change in question wording. It also forces us to assume that estimates of reliability and stability are comparable across panels as estimates of general reliability and stability of "political attitudes" at that time -- a claim that might not be valid.

# Conclusion

This paper set out to test whether the unique rates of attitude change found among young adults are due to the social features of this life stage or due to features of how human brains develop and process information over time. We leverage delays in the timing of life course transitions and the emergence of an "age of independence" over the past 70 years to explore whether attitude behavior shifted in response to these changes. We expected that we should see more durable attitude change in early adulthood and an extension of the "impressionable years" pattern up the age ladder.

We found no evidence of either process, broadly suggesting that changes in the timing of life course transitions, as well as broader social and cultural changes around early adulthood have not broadly affected attitude behavior among young adults. The pattern of results generally suggests that what makes early adulthood unique when it comes to attitude development is not its social conditions, but how humans process information as they age. These years are not formative because they are unfettered, they are formative because they are early.

# References

### 2024-03-29 code/clean_panels.R
kkeditold

@@ -1312,15 +1312,15 @@ cg10 <- clean_g10 %>%
   mutate(wave = paste0("y", wave)) %>%
   spread(wave, value) 
 
-save(cg10, file = "/Dropbox/extended_formative_windows/clean_data/cg10.Rdata")
-save(cg8, file = "/Dropbox/extended_formative_windows/clean_data/cg8.Rdata")
-save(cg6, file = "/Dropbox/extended_formative_windows/clean_data/cg6.Rdata")
-save(canes5, file = "/Dropbox/extended_formative_windows/clean_data/canes5.Rdata")
-save(canes7, file = "/Dropbox/extended_formative_windows/clean_data/canes7.Rdata")
-save(canes8, file = "/Dropbox/extended_formative_windows/clean_data/canes8.Rdata")
-save(canes90, file = "/Dropbox/extended_formative_windows/clean_data/canes90.Rdata")
-save(canes9, file = "/Dropbox/extended_formative_windows/clean_data/canes9.Rdata")
-save(canes0, file = "/Dropbox/extended_formative_windows/clean_data/canes0.Rdata")
+save(cg10, file = "~/Dropbox/extended_formative_windows/clean_data/cg10.Rdata")
+save(cg8, file = "~/Dropbox/extended_formative_windows/clean_data/cg8.Rdata")
+save(cg6, file = "~/Dropbox/extended_formative_windows/clean_data/cg6.Rdata")
+save(canes5, file = "~/Dropbox/extended_formative_windows/clean_data/canes5.Rdata")
+save(canes7, file = "~/Dropbox/extended_formative_windows/clean_data/canes7.Rdata")
+save(canes8, file = "~/Dropbox/extended_formative_windows/clean_data/canes8.Rdata")
+save(canes90, file = "~/Dropbox/extended_formative_windows/clean_data/canes90.Rdata")
+save(canes9, file = "~/Dropbox/extended_formative_windows/clean_data/canes9.Rdata")
+save(canes0, file = "~/Dropbox/extended_formative_windows/clean_data/canes0.Rdata")
 
 
 

### 2024-03-29 code/correct_rels.R
kkeditold

@@ -149,7 +149,7 @@ simplex_reliabilities %>% filter(label == "rel") %>%
        subtitle = "1980 reliabilities adjusted to be comparable with other panels")
 
 simplex_reliabilities %>% filter(label %in% c("stab12", "stab23")) %>% 
-  group_by(df, var) %>%
+  #group_by(df, var) %>%
   filter(max(estimate) < 1.4,
          min(estimate) > 0) %>%
   group_by(df, group) %>% summarise(mean = mean(estimate)) %>%

### 2024-03-29 manuscript/formative_period.Rmd
kkedits329

@@
 However, the notion of "impressionable years" is more general than just political preferences.
-The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed (Elder 2018).
+The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@elder2016].
 Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990].
@@
 
-These models of social imprinting have implications for the kinds of patterns we should observe in peoples' attitudes over time, and suggest specific forms that opinion change should take.
+These models of social imprinting have implications for the kinds of patterns we should observe in a population's attitudes over time and suggest specific forms that within-person opinion change should take over time.
 The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course.
@@
 
-## Biological Mechanisms:
+## Biological Mechanisms
 
 The first potential mechanism for changes in attitude stability over time is rooted in the biology of the human brain and science of how people process information.
-This approach tends to suggest that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally, whether they encounter it during adolescence or middle age, and are exposed to information at a (more or less) constant rate (Bartels and Jackman 2014; Sherratt and Morand-Ferron 2018; Gopnik 2020).
-However, because people have already encountered a lot of information by the time they reach adulthood, each new piece has less of an impact on their overall dispositions [@bartels2014].
-The distinctive thing about our early years, then, is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.
+This approach tends to suggest that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally, whether they encounter it during adolescence or middle age, and are exposed to information at a (more or less) constant rate [@bartels2014; @gopnik2020; @sherratt2018].
+However, because people have already encountered a lot of information by the time they reach adulthood, each new piece of information encountered later has less of an impact on their overall dispositions [@bartels2014].
+From this perspective, the distinctive thing about our early years is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.
 
 It is also possible that there is something unique about human cognition during these periods making people ascribe more weight to the information acquired within them than to information acquired at other points.
-Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life (Mata and von Helversen 2015).
+Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life [@mata2015].
 Supporting this model is evidence that early formative periods have been observed not just in people, but across species.
-In a study of social learning among birds, for instance, Aplin et al (2017) show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.
+In a study of social learning among birds, for instance, Aplin and colleagues [-@aplin2017] show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.
 
@@
 
-## Life Course and Social Roles:
+## Life Course and Social Roles
 
@@
 
-Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the "impressionable years" are principally an effect of how we have chosen to structure the life course, or the interaction between these life course stages and biological mechanisms.
-Because of this, contemporary historical trends offer a window to disentangle the role of socio-cultural and biological factors in explaining why we observe "impressionable years."
+Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the "impressionable years" are principally an effect of how societies choose to structure the life course or the interaction between these life course stages and biological mechanisms.
+Because of this, contemporary historical trends offer a window to disentangle the role of socio-cultural and biological or cognitive factors in explaining why we observe "impressionable years."
 
@@
 We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021].
-Thus, the markers of the transition into adulthood seem to be occurring later in the life-course and potentially in different orders.
+Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts.
 
 If the distinctive nature of "impressionable years" is principally driven by their unique social circumstances, then these shifts have the potential to extend these social circumstances into later ages.
-This, in turn, could have an effect in how long exploration, and therefore durable change, lasts throughout the life-course.
+This, in turn, could have an effect on how long exploration, and therefore openness to durable change, lasts throughout the life-course.
 If this is the case, we should expect to see lower levels of attitude stability (and therefore more change) among age groups that, for earlier cohorts, displayed mainly stability.
-In other words, people who were born say in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age.
-Specifically, we expect that people aged 25-33, who in earlier cohorts displayed opinion patterns similar to middle-aged respondents [@krosnic1991; @alwin1991], should more closely resemble the youngest adults in more recent years.
+In other words, people who were born in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age.
+Specifically, we expect the attitude behavior of people aged 25-33 to more closely resemble middle-aged respondents in earlier cohorts [@krosnic1991; @alwin1991] but younger respondents in more recent cohorts.
 This produces our first hypothesis:
@@
 
-## Changing Impressionable Years:
+## Intensifying Impressionable Years
 
 A second trend in early adulthood relates to just how much exploration young adults can engage in during these "impressionable years." In his work, Rosenfeld [-@rosenfeld2007] argues that one of the processes underlying the rapid rise in same-sex and interracial marriages in the United States is that young adults are increasingly escaping the controlling gaze of their parents.
-When young people live with, or are financially dependent on, their parents, the latter have more say about their children's choice of partner.
-Social and cultural changes that have increased independence from parental authority and mobility mean they have been able to engage in courtship outside the gaze of their parents, making choices that would not have been sanctioned by them.
+When a young person lives with, or is financially dependent on, his or her parents, the latter have more say about their child's choice of partner.
+Social and cultural changes that have increased geographic mobility and independence from parental authority mean people in this life stage have been able to engage in courtship outside the gaze of their parents, making choices that would not have been sanctioned by them.
 
-Rosenfeld's work suggests that the social conditions where we spend our impressionable years might shape how much exploration we can engage in.
+Rosenfeld's work suggests that the social conditions that shape where we spend our impressionable years might also shape how much exploration we can engage in.
 An increase in independence might entail more exploration, and greater ability to update attitudes as a result of this exploration.
@@
 
-Reliability reflects not just the degree to which people make errors , but the baseline level of variance around a fixed attitude position that people express.
+Reliability reflects not just the degree to which people make errors in their reporting of attitudes over time, but the baseline level of variance around a fixed attitude position that people express.
+In other words, we might ...
+if younger people bounce around
 

### 2024-04-08 manuscript/formative_period_kk.Rmd
new_approach

--- 2024-03-29 manuscript/formative_period.Rmd
@@
   - Nicolas Restrepo, University of California, Davis
-date: "7/06/2023"
+date: "`r format(Sys.time(), '%B %d, %Y')`"
 output: 
@@
 fontsize: 12pt
-abstract: Young adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of changing their personal culture.
+abstract: Early adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of changing their personal culture.
 This probability has often been attributed to the social conditions of adolescence.
@@
 We leverage these changes to explore whether life course changes, as well as broader social changes, have affected how likely people of different cohorts are to make durable changes of general political dispositions over a comparable time frame.
-We find strikingly little difference in the rates of any kind of stable opinions and opinion change among young people in different eras.
+We find strikingly little difference in the rates of opinion change among young people in different eras.
 People of all ages have similar attitude behaviors over time.
-The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates.
+The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates of change.
 ---
@@
 However, why people in this life stage appear to show more frequent opinion change relative to older people is unclear.
-Explanations in the social sciences tend to highlight the unique social features of these life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information.
+Explanations in the social sciences tend to highlight the unique social features of these early life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information [@alwin1991; @danigelis2007].
+Alternative explanations in the social sciences also focus on the unique features of middle age that tend to stabilize attitudes during this period, including higher levels of social authority [@eaton2009; @visser2004].
 Work in biology, psychology, and cultural evolution, in contrast, shows that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION].
-While both mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.
+While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.
 
@@
 If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change.
-Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time.
+Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016].
 If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social change in the same way.
 
-In this paper, we explore how changes in the broader social structure and changes in the timing of life course events since the 1950s have affected the stability and reliability of political attitudes over time, especially differences in rates of opinion change between young adults and the rest of the population.
+In this paper, we explore how rates of opinion change in different age groups have changed in response to broader social structure and changes in the timing of life course events since the 1950s.
 Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005].
@@
 We take advantage of this change over time to explore whether young adults of different cohorts demonstrate more or less change in their political opinions over time relative to other members of the population.
+
 We decompose over-time responses into reliability (a lack of random, non-persisting fluctuations) and stability (true consistency in latent dispositions net of these random fluctuations) for a range of political attitudes in nine nationally representative panel data sets spanning from the 1950s through the 2010s.
@@
 
+There are a broad range of approaches to measure the amount of change a group undergoes. We could compare the average slope of different age groups, but this assumes that , and would miss
+
+My starting point for our analysis if the Life Course Adaption Model (LCAM) formalized by Lersch [-@lersch2023].
+The LCAM is a mixed-effects growth curve model that formalizes several key insights from the literature on attitude change over time.
+First, it assumes that survey respondents have an underlying disposition shaped by early-life socialization.
+Second, it assumes that people can, but do not necessarily, change this disposition over time as a result of aging and general life-course experiences.
+Third, it assumes that people fluctuate around their underlying dispositions over time as a result of random shocks and measurement error.
+Fourth, it assumes respondents can draw on their prior experiences when adapting to new events.
+
+The model takes the form of:
+
+$$ c_{i,t}^* = \beta_0 + \beta_ia_{i,t} + u_{0,i} + u_{i,t}a_{i,t} + \epsilon_{i,t}$$
+
+where $c_{i,t}^*$ is the standardized survey response for respondent $i$ at time $t$, $\beta_0$ is the intercept with random term $u_{0,i}$ capturing early-life experiences.
+$\beta_1$ with random term $u_{i,t}a_{i,t}$ captures the individual-specific slope for age $a_{i,t}$.
+$\epsilon_{i,t}$ is random error around this individual-specific slope for person $i$ at time $t$.
+
+Because we are unconcerned with whether respondents change in the same direction or tend to change in different directions (which could indicate either convergence or polarization, depending on where respondents start).
+The central quantity of interest for our analysis is the absolute change in responses for a year, or $|\beta_ia_{i,t} + u_{i,t}a_{i,t}|$.
+Averaging this quantity across all respondents in a group produces an average absolute change in responses for a year for members of that group.
+
+As a second quantity of interest, we calculate the mean square residual for respondents in a particular group.
+Groups might differ both in their linear (directional) change, as well as whether they are consistent at respodning around that trajectory.
+High variance around a latent trajectory could indicate that respondents are either more error-prone in making responses or more subject to temporary considerations.
+To calculate this, we calculate:
+
+$$\frac{1}{n}\sum{(y_{i,t} - \hat{y_{i,t}})^2} $$
+
+Our central question deals with comparisons across age groups at one time and within age groups over time in the amount of change we observe.
+Because of this, we need a way to extract a single measure of change .
+
 There are a number of potential ways to measure change, including linear trajectory models, latent growth curve models, and sequence analysis.
@@
 However, extensive work suggests that both models produce similar estimates of the standardized parameters [@alwin2007; @alwin1991].
-
-We divide each sample into seven age categories: 18-25, 26-33, 34-41, 42-49, 50-57, 58-65, 66-83, assessed at the initial survey year.
-We analyze these estimates in two ways.
-First, we explore the stability of the latent attitudes, $\beta_21$ and $\beta_32$.
-These estimates are non-independent, so we examine them as multiple indicators of the stability of the attitude in that panel.
-We take these measures of stability as a measure of how much durable change in attitudes occurs in an issue during the panel, with higher values indicating less change.
-Second, we examine a single reliability estimate generated for each group for each attitude measure.
-We take this as a combined measure of actual measurement error and transitive, non-durable change that respondents make during the course of the panel.
-For each age group, we average across all questions in a given panel to generate an average estimate of reliability and an average estimate of stability for each age group in each panel.
-
-We compare these average estimates of reliability and stability for age groups across panels to evaluate our hypotheses.
-
-## Question Fixed Effects
-
-For questions that are asked in multiple panels, we also use estimates of stability and reliability as outcome variables in a regression on time with fixed effects for questions.
-We run these models separately for each age group.
-These models let us allow for an average trajectory of reliability and stability for age groups over time, controlling for questions, to evaluate whether respondents in the same age groups have become more or less stable reporters of the same opinions over time.
-We also control for attributes of panels that might affect reliability or stability estimates -- such as the shortened wave duration in the 1980 panel or the mode of administration -- by including panel fixed effects in the model as well.
 

### 2024-04-08 code/kkapproach/slope_analysis.R

@@ -1,0 +1,502 @@
+
+library(lme4)
+
+source("./code/clean_panels.R")
+
+canes5 <- canes5 %>% filter(var != "clsrprvtpwr")
+vars <- unique(canes5$var)
+group_meansc5 <- vector(mode = "list", length = length(vars))
+variancec5 <- vector(mode = "list", length = length(vars))
+for (i in 1:length(vars)) {
+  v <- vars[i]
+  df <- canes5 %>%
+    filter(var == v) %>%
+    pivot_longer(y1:y3) %>%
+    filter(!is.na(value), !is.na(age_1)) %>%
+    group_by(id) %>% filter(n() > 1) %>%
+    ungroup() %>%
+    mutate(year = as.numeric(gsub("y", "", name))) %>%
+    select(-name) %>%
+    mutate(value = (value - mean(value,na.rm=TRUE))/(sd(value,na.rm=TRUE))) %>%
+    mutate(year = (year - 1)*2) %>%
+    mutate(group = ifelse(age_1 <= 25, "18-25",
+                          ifelse(age_1 >= 26 & age_1 <= 33, "26-33",
+                                 "34+"))) 
+  #Estimate the effect of one year on standardized response.
+  m1 <- lmer(value ~ year + (1 + year|id), data = df)
+  #Extract individual year effect
+  cm1 <- coef(m1)
+  df$yhat <- predict(m1)
+  #Put them in data frame
+  effects <- data.frame(cm1$id) %>% 
+    select(-X.Intercept.) %>%
+    rownames_to_column(var = "id") %>%
+    mutate(id = as.numeric(id)) %>%
+    mutate(slope = year) %>% select(-year)
+  #Link to ids/groups
+  df <- df %>%
+    left_join(effects, by = c("id"="id"))
+  group_meansc5[[i]] <- df %>% 
+    group_by(id) %>% 
+    slice(1) %>%
+    group_by(group) %>%
+    summarise(n = n(), abs_mean_slope = mean(abs(slope),na.rm=TRUE)) %>%
+    mutate(var = v)
+  
+  variancec5[[i]] <- df %>%
+    mutate(resid = (value - yhat)^2) %>%
+    group_by(group) %>%
+    summarise(ssr = sum(resid), 
+              n_obs = n(),
+              n_id = length(unique(id))) %>%
+    mutate(mse = ssr/n_obs) %>%
+    mutate(var = v)
+  
+  
+}
+
[…]

### 2024-04-10 code/kkapproach/slope_analysis.R
kkapproach_410

@@ -222,8 +222,7 @@ bind_rows(variancec9) %>%
 
 
 
-canes8 <- canes8 %>% filter(var != "thermbaker",
-                            var != "thermbush")
+canes8 <- canes8
 vars <- unique(canes8$var)
 group_meansc8 <- vector(mode = "list", length = length(vars))
 variancec8 <- vector(mode = "list", length = length(vars))
@@ -239,7 +238,7 @@ for (i in 1:length(vars)) {
     mutate(year = as.numeric(gsub("y", "", name))) %>%
     select(-name) %>%
     mutate(value = (value - mean(value,na.rm=TRUE))/(sd(value,na.rm=TRUE))) %>%
-    mutate(year = (year - 1)*.5) %>%
+    mutate(year = (year - 1)*.375) %>%
     mutate(group = ifelse(age_1 <= 25, "18-25",
                           ifelse(age_1 >= 26 & age_1 <= 33, "26-33",
                                  "34+"))) 
@@ -369,134 +368,326 @@ bind_rows(variancec2k) %>%
   theme_bw()
 
 
+canes90 <- canes90 %>%
+  filter(var != "adjmoral",
+         var != "newstyles")
+vars <- unique(canes90$var)
+group_meansc90 <- vector(mode = "list", length = length(vars))
+variancec90 <- vector(mode = "list", length = length(vars))
+for (i in 1:length(vars)) {
+  v <- vars[i]
+  df <- canes90 %>%
+    filter(var == v) %>%
+    pivot_longer(y1:y3) %>%
+    filter(!is.na(value),
+           !is.na(age_1)) %>%
+    group_by(id) %>% filter(n() > 1) %>%
+    ungroup() %>%
+    mutate(year = as.numeric(gsub("y", "", name))) %>%
+    select(-name) %>%
+    mutate(value = (value - mean(value,na.rm=TRUE))/(sd(value,na.rm=TRUE))) %>%
+    mutate(year = (year - 1)) %>%
+    mutate(group = ifelse(age_1 <= 25, "18-25",
+                          ifelse(age_1 >= 26 & age_1 <= 33, "26-33",
+                                 "34+"))) 
+  #Estimate the effect of one year on standardized response.
+  m1 <- lmer(value ~ year + (1 + year|id), data = df)
+  #Extract individual year effect
+  cm1 <- coef(m1)
[…]

### 2024-06-05 code/clean_panels.R
kkedits605

@@ -47,6 +47,8 @@ clean_anes5 <- anes5 %>%
     V560181, V580478, V600694,
     #Children in household
     V560178, V580475, V600691,
+    #Occupational code
+    V560120,
     #sex
     V560171, V580468, V600684,
     #race
@@ -57,6 +59,7 @@ clean_anes5 <- anes5 %>%
          age_1 = V560295,
          age_2 = V580472,
          age_3 = V600688,
+         student_1 = ifelse(V560120 %in% c(92), 1, 0),
          marital_1 = V560177, marital_2 = V580474,
          marital_3 = V600690,
          marital_1 = ifelse(V560177 == 9, NA, V560177),
@@ -180,6 +183,8 @@ clean_anes7 <- anes7 %>%
     #age
     age_1 = V720294, age_2 = V742406, 
     age_3 = V763369,
+    #student status
+    student_1 = ifelse(V720306 == 8, 1, 0),
     #toofast
     toofast_1 = V720112, toofast_2 = V742264, 
     toofast_3 = V763213,
@@ -344,6 +349,12 @@ clean_anes8 <- anes80 %>%
          energypol_1=VMP0177,energypol_2=VMP2196,energypol_3=VMP3314,
          rationgas_1=VMP0179,rationgas_2=VMP2198,rationgas_3=VMP3316,
          nukeplant_1=VMP0184,nukeplant_2=VMP2203,nukeplant_3=VMP3321,
+         baby = ifelse(VMP0589 > 0 & VMP0589 != 9, 1, 0),
+         kid = ifelse(VMP0590 > 0 & VMP0590 != 9, 1, 0),
+         tween = ifelse(VMP0591 > 0 & VMP0591 != 9, 1, 0),
+         teen = ifelse(VMP0592 > 0 & VMP0592 != 9, 1, 0),
+         child_1 = ifelse(baby + kid + tween + teen > 0, 1, 0),
+         student_1 = ifelse(VMP0352 == 8, 1, 0),
          marital_1=recode(VMP0326, "1"="married", "2"="single/nm",
                           "3"="other", "4"="other", "5"="other",
                           "7"="married", "9"=NA_character_),
@@ -355,13 +366,25 @@ clean_anes8 <- anes80 %>%
                           "7"="married", "9"=NA_character_),
          ed_1=ifelse(VMP0327<12, "lessthan",
                      ifelse(VMP0327>=16, "ba", "hs"))) %>%
[…]

### 2024-06-05 code/kkapproach/model_kk.Rmd

---
title: "Modeling Change"
author: "Kevin Kiley"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Model

Our central questions are 1) whether respondents in the youngest age group (18-25) make more change in more recent panels than in earlier panels and 2) whether respondents in the next-youngest age group (26-34) become more similar to respondents in the youngest age group over time. To evaluate these hypotheses, we need a measure of the rate of change for respondents of different age groups over time in different era.

A central challenge for this analysis is that panels differ in the questions they ask, the number of times they ask participants to respond to different questions, and the duration between observations.

In the one-item, one-panel case (partisan identification in the 1972-76 panel, for example), we could model the absolute change from wave-to-wave as a function of the duration between observations.

$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1*d_i + \epsilon_i $$

In this equation, $d_{i}$ captures the duration of time between $t-1$ and $t$ in years. In the 1972-76 panel, respondents are asked their partisan identification three times: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart. If a person participates in all three waves, they have three observations of $|y_{i,t}-y_{i,t-1} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).

If respondents have a stable underlying partisan identification they report with error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$. In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise).  Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$). For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3, and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common.

Our theoretical question deals primarily with whether these coefficients ($\beta_0$ and $\beta_1$) change over time. To explore that, we can add an additional term for time and interact the duration term with time.

$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1d + \beta_2t + \beta_3(d*t) + \epsilon_i $$

In this formulation, $\beta_2$ captures the change in the baseline level of random change over time. A negative value of $\beta_2$ would suggest that people are becoming more consistent reporters of their true underlying opinion over time. $\beta_3$ captures the expected change in absolute change over time for a one-year change in duration. A positive $\beta_3$ value would suggest that people are more likely to undergo a "real" change in their opinion over time in later panels than earlier panels.

A challenge for our analysis is that while the panels we explore often have relatively large samples, most questions are only asked in a few panels. As such, we lack large numbers of any particular question, meaning our time trends for any individual question is likely to be unstable. One potential solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels. It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable.

Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty.

$$ |y_{i,j,t}-y_{i,j,t-1} | = \beta_{0j} + \beta_{1j}d + \beta_{2j}t + \beta_{3j}(d*t) + \epsilon_i $$

Where:

$$ \beta_{0j} = \gamma_0 + \upsilon_{0j} $$
$$ \beta_{1j} = \gamma_1 + \upsilon_{1j} $$
$$ \beta_{2j} = \gamma_2 + \upsilon_{2j} $$
$$ \beta_{3j} = \gamma_3 + \upsilon_{3j} $$

In this framework, the first-level model models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above. The second-level equations represent the items. Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for time, duration, and their interaction. These effects are assumed to be normally distributed around a main effect.

## Results

```{r}
long_data %>%
  group_by(name) %>%
  nest() %>%
  mutate(t = map(data, lm_func)) %>%
  unnest(t) %>%
  filter(term == "date") %>%
  unnest(data) %>%
  mutate(direction = ifelse(estimate < 0, "negative", "positive")) %>%
  ggplot(aes(x = date, y = a, color = direction)) + 
  facet_wrap(~reorder(name, estimate)) +
  geom_smooth(method = "lm") + 
  theme_bw() + 
  labs(x = "Year", y = "Absolute difference")
```

The figure above presents the expected absolute difference over time generated from separate regressions for each item only on date of observation. This model does not account for varying duration observed in different panels. Generally speaking, most items show a negative slope over time, suggesting that respondents are becoming more stable in their responses over time.

Notably, most of the "feeling thermometer" items show positive slopes over time, suggesting people are becoming more variable in how they feel toward other groups. In contrast, general policy positions show negative slopes over time.

### 2024-06-05 manuscript/formative_period_kk.Rmd

@@
 
-**Hypothesis 1: Rates of attitude stability will be lower among respondents aged 25-33 in recent surveys than in earlier surveys.**
+*Hypothesis 1: More recent cohorts will show higher rates of attitude change in the age range of 25 to 33 than cohorts born in earlier years.*
 
@@
 
-*Hypothesis 2: People ages 18 to 25 will demonstrate lower levels of reliability in more recent years than in earlier years.*
+*Hypothesis 2: More recent cohorts will show higher levels of residual variance in the 18 to 25 age range, net of their over-time trajectories, compared to previous generations in the same age range.*
 
-Reliability reflects not just the degree to which people make errors in their reporting of attitudes over time, but the baseline level of variance around a fixed attitude position that people express.
-In other words, we might ...
-if younger people bounce around
+Residual variance reflects not just the degree to which people make errors in their reporting of attitudes over time, but the baseline level of variance around a fixed attitude position that people express.
+In other words, we might expect younger people to bounce around their latent trajectory more if they engage in more exploration of attitudes over time, trying on different identities and ideas.
+
+Finally, the intensification of early adulthood hypothesis expects that the rates of durable attitude change will also increase for more recent cohorts.
+Freedom to explore outside of the watch of family members should result in more respondents making durable changes of opinion over time.
 
@@
 We have framed our hypotheses as expectations derived from the framework suggesting that social conditions play a large role in explaining the uniquely high rates of attitude change among early adults and adolescents.
-However, to the extent that we do not see evidence that these social changes have not contributed, but that young adults maintain a similar pattern of lower rates of attitude stability that older adults over time, we take this as evidence that social circumstances play a relatively small role in explaining change in favor of biological and cognitive factors.
+However, to the extent that we observe significant changes in the social context of early adulthood but no corresponding changes in rates of attitude change or response stability across cohorts, we take this as evidence that social circumstances play a relatively small role in explaining change in favor of biological and cognitive factors.
 
@@
 
-There are a broad range of approaches to measure the amount of change a group undergoes. We could compare the average slope of different age groups, but this assumes that , and would miss
-
-My starting point for our analysis if the Life Course Adaption Model (LCAM) formalized by Lersch [-@lersch2023].
+There are a range of approaches to measure the amount of change a group undergoes.
+The starting point for our analysis if the Life Course Adaption Model (LCAM) formalized by Lersch [-@lersch2023].
 The LCAM is a mixed-effects growth curve model that formalizes several key insights from the literature on attitude change over time.
@@
 
-Because we are unconcerned with whether respondents change in the same direction or tend to change in different directions (which could indicate either convergence or polarization, depending on where respondents start).
-The central quantity of interest for our analysis is the absolute change in responses for a year, or $|\beta_ia_{i,t} + u_{i,t}a_{i,t}|$.
+By combining the main effect for time and the individual-level random effects for time, we generate a person-specific slope estimate, or the expected change in that attitude for a one-year change.
+Because we are unconcerned with whether respondents change in the same direction or tend to change in different directions (which could indicate either convergence or polarization, depending on where respondents start), the central quantity of interest for our analysis is the absolute change in respondents' latent trajectories for a year, or $|\beta_ia_{i,t} + u_{i,t}a_{i,t}|$.
 Averaging this quantity across all respondents in a group produces an average absolute change in responses for a year for members of that group.
@@
 As a second quantity of interest, we calculate the mean square residual for respondents in a particular group.
-Groups might differ both in their linear (directional) change, as well as whether they are consistent at respodning around that trajectory.
+Groups might differ both in their linear (systematic) change, as well as whether they are consistent at responding around that trajectory.
 High variance around a latent trajectory could indicate that respondents are either more error-prone in making responses or more subject to temporary considerations.
-To calculate this, we calculate:
+To evaluate this, we calculate:
 
 $$\frac{1}{n}\sum{(y_{i,t} - \hat{y_{i,t}})^2} $$
-
-Our central question deals with comparisons across age groups at one time and within age groups over time in the amount of change we observe.
-Because of this, we need a way to extract a single measure of change .
-
-There are a number of potential ways to measure change, including linear trajectory models, latent growth curve models, and sequence analysis.
-We follow previous work that seeks to separate true change in attitude response from measurement error in the study of attitude change and stability over the life cycle.
-To do this, we follow Alwin and Krosnick's approach of using just-identified simplex models that specify two structural equations for a set of three over-time measures of a given variable, $y_i$:
-
-$$ y_{t} = \tau_t + \epsilon_t$$
-
-$$ \tau_t = \beta_{t,t-i} \tau_{t-1} + \zeta_t $$ The first equation represents a set of measurement assumptions indicating that (1) over-time measures are assumed to be $\tau$-equivalent, except for true attitude change, and (2) that measurement error is random.
-It decomposes the variance in reports into two components, one due to the latent attitude ($tau_t$) and the other due to random measurement error ($\epsilon_t$).
-The second equation specifies a causal process involving attitude change over time.
-$\beta_{t,t-i}$ is the stability of the latent attitude during the period from $t-1$ to $t$.
-$\zeta_t$ represents sources of change over time in the latent attitude.
-The model assumes that the system of attitudes is in a dynamic equilibrium described by a lag-1 or Markovian process in which the distribution of a "true" attitude at time $t$ is dependent on the distribution at time $t-1$ and not on the distribution of the variable at previous times.
-
-As presented above, the model is unidentified.
-All strategies for estimating these quantities require the lag-1 assumption regarding the nature of "true" attitude change, as well as an assumption about measurement error.
-An approach proposed by Heise [@heise1969] assumes equal reliabilities over occasions of measurement.
-Another approach is to assume constant error variances rather than constant reliabilities [@wiley1970].
-We follow the Heise approach and constrain reliabilities to be constant across waves, rather than constraining error variance to be the same.
-However, extensive work suggests that both models produce similar estimates of the standardized parameters [@alwin2007; @alwin1991].
 
@@
 
-## Stability and Reliability
+## Systematic Change
+
+We first evaluate the average estimates for absolute mean change for each age group.
+Figure 1 plots the average asbolute slope for each of the three age groups in each panel across all questions asked in that panel.
+Again, which questions are asked in each panel varies significantly.
+
+```{r}
+bind_rows(mean_val5, mean_val8, mean_val9, mean_val7,
+          mean_val2k, mean_val90, mean_val06,
+          mean_val08, mean_val10) %>%
+  mutate(year = recode(df, "anes2k"=2000, "anes5"=1956,
+                       "anes7"=1972, "anes9"=1992,
+                       "anes8"=1980, "anes90"=1990,
+                       "gss06"=2006, "gss08"=2008,
+                       "gss10"=2010)) %>%
+  filter(year != 1980, year != 1990) %>%
+  ggplot(aes(x = year, y = mean, fill = group)) + 
+  geom_point(size = 2, shape = 21, color = "black") + 
+  theme_bw() + 
+  labs(x = "Year", y = "Abs. Slope", fill = "Age Group")
+```
+
+Overall, we see few general patterns across age groups over time.
+There is little consistency in which age group displays the largest absolute change in each of the panels or the trajectory of rates of change across the panels.
+The 2006 and 2008 General Social Survey have the highest absolute change over time, but the 2010 panel, the most recent, is highly comparable to the other panels.
+While people in the 18 to 25 age range display higher rates of change than people in the 26-33 and 34 and older age group in the most two recent panels (the 2008 and 2010 GSS panels), suggesting some evidence of an intensification of early adulthood, they fall below people ages 34 and over in the panel begun only two years prior and overlapping (the 2006 GSS panel).
+
+In aggregating questions, we see no evidence for the "elongation" hypothesis that people in the 26 to 33 age range increasingly look like people in the 18 to 25 age range.
+People in the 26 to 33 age range are functionally identical to people over 34 in all panels, and they do not show any evidence of becoming more like people in the 18 to 25 age range over time.
+
+One challenge we noted at the outset is that the composition of questions in each panel changes over time, meaning our understanding of the general trajectories of age groups across cohorts might be biased by changes in question content, questions wording, or survey administration.
+To control for these potential sources of bias, we explore only questions that have been repeated almost verbatim in multiple panels.
+These are the seven-point scale of partisan identification (from Strong Democrat to Strong Republican) and the seven-point scale of political ideology (from Extremely Liberal to Extremely Conservative).
+These results are presented in Figure XXX below.
+
+```{r}
+bind_rows(bind_rows(group_meansc5) %>% mutate(year = 1956), 
+          bind_rows(group_meansc7) %>% mutate(year = 1972),
+          bind_rows(group_meansc9) %>% mutate(year = 1992), 
+          bind_rows(group_meansc2k) %>% mutate(year = 2000),
+          bind_rows(group_meansc06) %>% mutate(year = 2006),
+          bind_rows(group_meansc08) %>% mutate(year = 2008),
+          bind_rows(group_meansc10) %>% mutate(year = 2010)) %>%
+  filter(var %in% c("polviews", "partyid")) %>%
+  mutate(var = recode(var, "polviews"="Ideology",
+                      "partyid"="Identitification")) %>%
+  ggplot(aes(x = year, y = abs_mean_slope, fill = group)) +
+  geom_point(shape = 21, size = 2) + 
+  facet_wrap(~var) + 
+  theme_bw() + 
+  labs(x = "Year", y = "Abs. Mean Slope", 
+       fill = "Age Group")
+
+```
+
+For partisan identification, we see some evidence for the intensification hypothesis, with people in the 18-25 age range showing higher average absolute slopes than people in the other two age ranges.
+However, though the population as a whole also makes more linear change in partisan identification as well during these more recent years, complicating the overall picture.
+There is no evidence that people in the 26-33 age range increase more than the general population in this time frame, suggesting no evidence of the elongation of the formative period.
+
+We see no corresponding pattern for political ideology.
+Respondents in the 18 to 25 age range look remarkably similar to respondents in other age groups, and there is no obvious trajectory across cohorts.
+In one panel (the GSS panel beginning in 2006), people in this age range show higher rates of change, but this pattern does not hold even for the other GSS panels that overlap this panel.
+People in the 26 to 33 age range cannot be distinguished from people over 34 even in the most recent cohorts, suggesting no evidence for the elongation hypothesis.
+
+## Non-systematic change
+
+```{r}
+bind_rows(mean_var5, mean_var8, mean_var9, mean_var7,
+          mean_var2k, mean_var90, mean_var06,
+          mean_var08, mean_var10) %>%
+  mutate(year = recode(df, "anes2k"=2000, "anes5"=1956,
+                       "anes7"=1972, "anes9"=1992,
+                       "anes8"=1980, "anes90"=1990,
+                       "gss06"=2006, "gss08"=2008,
+                       "gss10"=2010)) %>%
+  filter(year != 1980, year != 1990) %>%
+  ggplot(aes(x = year, y = mean, fill = group)) + 
+  geom_point(size = 2, shape = 21, color = "black") + 
+  theme_bw() + 
+  labs(x = "Year", y = "Abs. Slope", fill = "Age Group")
+
+```
+
+In the early panels (the 1956 and 1972 ANES panels), there is little evidence that the youngest respodents exhibited more or less consistency over time in their attitude reports than the rest of the sample.
+In the 1956 panel, they have the lowest residual variance around their latent trajectories, and in the 1972 panel, they fall between the other two groups.
+However, in more recent cohorts, people in this age group consistently have the most residual variance of the three groups.
+Whether this is due to lack of knowledge of the political system (and therefore measurement error), or whether these vacillations around their latent trajectories are due to "exploration" across attitudes is not clear from these results.
+
+There is no consistent evidence for "elongation" of early adulthood in these data, as people in the 26-33 age group show no consistent trajectory over time.
+
+Turning again to the
+
+```{r}
+bind_rows(bind_rows(variancec5) %>% mutate(year = 1956), 
+          bind_rows(variancec7) %>% mutate(year = 1972),
+          bind_rows(variancec9) %>% mutate(year = 1992), 
+          bind_rows(variancec2k) %>% mutate(year = 2000),
+          bind_rows(variancec06) %>% mutate(year = 2006),
+          bind_rows(variancec08) %>% mutate(year = 2008),
+          bind_rows(variancec10) %>% mutate(year = 2010)) %>%
+  filter(var %in% c("polviews", "partyid")) %>%
+  mutate(var = recode(var, "polviews"="Ideology",
+                      "partyid"="Identitification")) %>%
+  ggplot(aes(x = year, y = mse, fill = group)) +
+  geom_point(shape = 21, size = 2) + 
+  facet_wrap(~var) + 
+  theme_bw() + 
+  labs(x = "Year", y = "Abs. Mean Slope", 
+       fill = "Age Group")
+
+```
 

### 2024-06-17 code/kkapproach/time_change.R
edtis617

@@ -23,11 +23,17 @@ anes20 <- read_dta("~/Dropbox/data/anes/anes2022/anes2022")
 
 anes5_long <- anes5 %>%
   mutate(id = 1:nrow(anes5)) %>%
+  zap_labels() %>%
   select(id, V560193,	V580309,	V600578,	V600837,
          V560088, V580360, V600657, V600835,
-         V560295, V580472, V600688) %>%
+         V560295, V580472, V600688,
+         V560035, V580323, V600622) %>%
   mutate(across(c(V560088, V580360, V600657, V600835), ~ifelse(.x %in% c(7,8,9), NA, .x)),
          across(c(V560295, V580472, V600688), ~ifelse(.x > 97, NA, .x)),
+         across(c(V560035, V580323, V600622),
+                ~recode(.x, 
+                        "1"=1, "2"=1, "3"=1.5, "4"=2, "5"=2, "7"=NA_real_,
+                        "0"=NA_real_, "8"=NA_real_, "9"=NA_real_)),
          age_1 = V560295,
          age_2 = V580472,
          age_3 = V600688,
@@ -120,16 +126,20 @@ anes5_long <- anes5 %>%
          partyid_1 = as.character(V560088), 
          partyid_2 = as.character(V580360), 
          partyid_3 = as.character(V600657), 
-         partyid_4 = as.character(V600835)) %>%
-  select(id, age_1:date_4, partyid_1:partyid_4) %>%
-  pivot_longer(age_1:partyid_4) %>%
+         partyid_4 = as.character(V600835),
+         stayhome_1 = as.character(V560035), 
+         stayhome_2 = as.character(V580323), 
+         stayhome_3 = as.character(V600622)) %>%
+  select(id, age_1:date_4, partyid_1:stayhome_3) %>% 
+  pivot_longer(age_1:stayhome_3) %>%
   separate(name, into = c("measure", "wave")) %>%
   spread(measure, value) %>%
   mutate(date = as.Date(date),
-         partyid = as.numeric(partyid),
+         across(c(age, partyid:stayhome),
+                ~as.numeric(.x)),
          partyid = (partyid/6)*100,
-         age = as.numeric(age)) %>%
-  pivot_longer(partyid) %>%
+         stayhome = (stayhome-1)*100) %>%
+  pivot_longer(partyid:stayhome) %>%
   group_by(id, name) %>%
   mutate(d1_value = lead(date) - date,
          d2_value = lead(date,2) - date,

### 2024-06-17 code/kkapproach/time_change.R
quality control, some edits to draft

@@ -1054,7 +1054,8 @@ anes0_long <- anes0 %>%
                 ~ifelse(.x == 4, 3, .x)),
          across(c(M000682, M000678, M000677, M000681, 
                   M000683, M000685, M000676, M000684), 
-                ~recode(.x, "1"=1, "5"=2, "3"=3, "7"=NA_real_)),
+                ~recode(.x, "1"=1, "5"=2, "3"=3, "7"=3,
+                        "8"=NA_real_, "9"=NA_real_, "0"=NA_real_)),
          across(c(M023052, M025113, M045068,
                   M023054, M025116, M045076,
                   M023045, M025106, M045069,
@@ -1063,7 +1064,8 @@ anes0_long <- anes0 %>%
                   M023049, M025109, M045073, 
                   M023046, M025107, M045070, 
                   M023048, M025109, M045072),
-                ~recode(.x, "1"=1, "2"=3, "3"=2)),
+                ~recode(.x, "1"=1, "2"=3, "3"=2,
+                        "4"=3)),
          across(c(M000523, M023038X, M045058x), 
                 ~ifelse(.x %in% c(7,8,9), NA, .x)),
          across(c(M000446, M001368, M023022), 
@@ -1656,11 +1658,11 @@ anes16_long <- anes16 %>%
                   govins, helpblk, spendserv,
                   defscale),
                 ~(.x-1)/6*100),
-         across(c(stayhome, runfew),
-                ~(.x-1)*100),
-         across(c(wastetax,
+         across(c(stayhome, runfew,
                   natenvir, natsoc, natschools, natfare,
                   natcrime, natpoor),
+                ~(.x-1)*100),
+         across(c(wastetax),
                 ~(.x-1)/2*100),
          across(c(abortion),
                 ~(.x-1)/3*100)) %>%
@@ -1771,7 +1773,10 @@ long_data <- bind_rows(anes5_long %>% mutate(df = "1956-60 ANES"),
           anes20_long %>% mutate(df = "2020-22 ANES")) %>%
   ungroup() %>%
   mutate(dec_diff = (as.numeric(difftime(date, min(date), units = "days")))/3652.5,
-         d = as.numeric(d))
+         d = as.numeric(d)) %>%
+  mutate(age_group = ifelse(age < 26, "18-25", "34-65"),
[…]

### 2024-06-17 manuscript/formative_period_kk.Rmd
quality control, some edits to draft

@@
 
-Analyzing these questions requires data on attitude reports over time for members of different cohorts at different time points, presumably coming from the same population.
-While ideally we would be able to observe different cohorts responding to the same broad range of questions as they pass through the same age periods, we do not know of such a data source.
+Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time.
+Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, coming from the same general population.
+To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years.
+Both the ANES and GSS are designed to generalize to the non-institutionalized U.S.
+adult (18+) population, and can be thought of as tapping the same general population.
+We included a total of 12 panels in our search: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20.
+We evaluated an additional panel conducted by the ANES in 2008-09 (with an additional follow-up in 2010), but no questions overlapped this panel and any other panel.
 
-Instead, we focus on a range of questions tapping a similar domain -- national politics in the United States.
-We draw on data from seven different panels administered since the 1950s.
-These include four major panels administered by the American National Election Study in 1956-58-60, 1972-74-76, 1992-94-96, and 2000-02-04.
-These were all administered over a four-year period, with waves two years apart and contain a diverse array of attitudinal questions.
-We also use the General Social Survey's rotating panels from 2006-08-10, 2008-10-12, 2010-12-14.
-We also include the 1980 Major Panel study from the American National Election Study, a four-wave panel spanning about eight months.
-Finally, we incldue the the 1990-91-92 Merged File, a three-wave panel spanning two years.
-All seven surveys reflect nationally representative samples of U.S.
-households.
+For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points.
+Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14), we did not treat these as separate panels for the purposes of generating three distinct time points.
+In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.
 
-In all panels, we identified questions that tap several categories of political attitudes: (1) political efficacy and alienation; (2) attitudes toward social groups (such as labor unions, blacks, whites, the police, Democrats, liberals, etc.); (3) attitudes toward both domestic and policy issues; (4) ideological identification on a liberal-conservative scale; and (5) partisan identification.
-For the 1956-60, 1972-76, and 1980 panels, we broadly follow Krosnick and Alwin's selection of variables [-@krosnick1989].
-In the 1992-96 panel, we follow Alwin's identification of questions [-@alwin2007].
-For the 1990-92 panel, the 2000-04 panel, and the three GSS panels, we identify questions similar to those identified in the earlier panels.
+This produced a total of ### questions tapping a broad range of topics including political and ideological identification, government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.
 
-This produces a set of 265 panel-attitude measures asked on a variety of scales tapping a diverse array of topics related to national government and politics in the United States.
-They include four- and five-point Likert scales, binary agree/disagree response scales, 100-point "feeling thermometers", and a range of other structures.
+## Methods
 
-Most questions are asked in only one panel, but a small number of questions are replicated exactly across at least two panels (excluding questions duplicated across the GSS panels, where all questions are duplicated).
-There are theoretically relevant trade-offs from focusing only on the same questions over time.
-By focusing on the same questions over time, we control for question content and structure.
-If attributes of questions changes over time, that might affect the estimates of stability or reliability.
-However, the questions asked in major surveys change over time as the dominant issues change.
-By focusing only on questions that are repeated across eras, we limit our attention only to issues that have been salient for an extended period of time.
-These might behave different than the "issues of the day" -- they might get settled earlier in life, be more resistant to change, or be reported more reliably than other kinds of questions.
+When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's ability to report stable attitudes over time or to the structural features of the survey administration.
+Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change.
+Across our panels, duration between questions ranges from a few days (for questions asked in both pre- and post-election surveys in the ANES) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
 
-By focusing on all questions tapping the domains outlined above, the average level of stability and reliability reflect change in "important" issues -- including issues that remain important for a long time and issues that only maintain salience for a few years.
+In the one-item, one-panel case (e.g., partisan identification in the 1972-76 panel), we could model the absolute change from wave-to-wave as a function of the duration between observations.
 
-To ensure that our findings are robust to both sets of issues, we conduct a separate analysis of questions that are repeated across multiple panels (excluding questions replicated just across the GSS).
-These include partisan identification (asked in all panels), political ideology on a seven-point liberal-conservative scale (asked in \## panels), questions about federal spending on different priorities (asked in the 1990 and 2000 ANES panels and the GSS panels), and questions about government efficacy and alienation from the political process (asked in the 1970s, 1990s, and 2000 ANES panels), as well as others.
+$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1*d_i + \epsilon_i $$
 
-Weighting schemes are different across panels. When available, I use weights that account for differential selection. The 1972-76, 1980, and
+In this equation, $d_{i}$ captures the duration of time between $t-1$ and $t$ in years.
+In the 1972-76 panel, respondents are asked their partisan identification three times: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
+If a person participates in all three waves, they have three observations of $|y_{i,t}-y_{i,t-1} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).
 
-## Methods of Analysis
+If respondents have a stable underlying partisan identification they report with error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$.
+In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise).
+Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$).
+For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3, and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.
 
-There are a range of approaches to measure the amount of change a group undergoes.
-The starting point for our analysis if the Life Course Adaption Model (LCAM) formalized by Lersch [-@lersch2023].
-The LCAM is a mixed-effects growth curve model that formalizes several key insights from the literature on attitude change over time.
-First, it assumes that survey respondents have an underlying disposition shaped by early-life socialization.
-Second, it assumes that people can, but do not necessarily, change this disposition over time as a result of aging and general life-course experiences.
-Third, it assumes that people fluctuate around their underlying dispositions over time as a result of random shocks and measurement error.
-Fourth, it assumes respondents can draw on their prior experiences when adapting to new events.
+Our theoretical question deals primarily with whether these coefficients ($\beta_0$ and $\beta_1$) change over time.
+To explore that, we can add an additional term for time and interact the duration term with time.
 
-The model takes the form of:
+$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1d + \beta_2t + \beta_3(d*t) + \epsilon_i $$
 
-$$ c_{i,t}^* = \beta_0 + \beta_ia_{i,t} + u_{0,i} + u_{i,t}a_{i,t} + \epsilon_{i,t}$$
+In this formulation, $\beta_2$ captures the change in the baseline level of random change over time.
+A negative value of $\beta_2$ would suggest that people are becoming more consistent reporters of their true underlying opinion over time.
+$\beta_3$ captures the expected change in absolute change over time for a one-year change in duration.
+A positive $\beta_3$ value would suggest that people are more likely to undergo a "real" change in their opinion over time in later panels than earlier panels.
 
-where $c_{i,t}^*$ is the standardized survey response for respondent $i$ at time $t$, $\beta_0$ is the intercept with random term $u_{0,i}$ capturing early-life experiences.
-$\beta_1$ with random term $u_{i,t}a_{i,t}$ captures the individual-specific slope for age $a_{i,t}$.
-$\epsilon_{i,t}$ is random error around this individual-specific slope for person $i$ at time $t$.
+A challenge for our analysis is that while the panels we explore often have relatively large samples, most questions are only asked in a few panels.
+As such, we lack large numbers of any particular question, meaning the time trend for any specific question is likely to be unstable.
+One potential solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels.
+It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable.
 
-By combining the main effect for time and the individual-level random effects for time, we generate a person-specific slope estimate, or the expected change in that attitude for a one-year change.
-Because we are unconcerned with whether respondents change in the same direction or tend to change in different directions (which could indicate either convergence or polarization, depending on where respondents start), the central quantity of interest for our analysis is the absolute change in respondents' latent trajectories for a year, or $|\beta_ia_{i,t} + u_{i,t}a_{i,t}|$.
-Averaging this quantity across all respondents in a group produces an average absolute change in responses for a year for members of that group.
+Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty.
 
-As a second quantity of interest, we calculate the mean square residual for respondents in a particular group.
-Groups might differ both in their linear (systematic) change, as well as whether they are consistent at responding around that trajectory.
-High variance around a latent trajectory could indicate that respondents are either more error-prone in making responses or more subject to temporary considerations.
-To evaluate this, we calculate:
+$$ |y_{i,j,t}-y_{i,j,t-1} | = \beta_{0j} + \beta_{1j}d + \beta_{2j}t + \beta_{3}(d*t) + \epsilon_i $$
 
-$$\frac{1}{n}\sum{(y_{i,t} - \hat{y_{i,t}})^2} $$
+Where:
+
+$$ \beta_{0j} = \gamma_0 + \upsilon_{0j} $$
+$$ \beta_{1j} = \gamma_1 + \upsilon_{1j} $$
+$$ \beta_{2j} = \gamma_2 + \upsilon_{2j} $$
+$$ \beta_{3j} = \gamma_3 + \upsilon_{3j} $$
+
+In this framework, the first-level model models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above.
+These equations are nested at the level of the item.
+Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for time and duration.
+These effects are assumed to be normally distributed around a main effect.
+In practice, we see minimal variation in the question-specific coefficients for the interaction between duration and time, and as such do not allow these to vary at the question level.
+
+For all models, we normalize response scales between 0 and 100.
 
 # Results
+
+Model 1 in Table XXX presents the main effects and variance components of the equation presented above.
+
+The intercept suggests that, on average, respondents varied by about a fifth of the scale (17 points), controlling for duration.
+This is a rather high level of variance, but consistent with previous studies that find that, on any particular question, survey respondents tend to make large wave-to-wave changes.
+Variance around this intercept is quite large, with a standard deviation of about 5.67, meaning many items vary by as much as 20 percent of the scale, while some vary by less than 10 points.
+
+Second, duration is weakly positively associated with wave-to-wave absolute change.
+For every additional year between waves, respondents' absolute deviance increases about .71 points.
+Notably, variance around this main effect is relatively small (.43), meaning that two thirds of questions have duration effects that fall between .23 and 1.14.
+Broadly speaking, we can say that most wave-to-wave change is non-persisting, consistent with other recent work [@kiley2020; @lersch2023].
+The main effect of year of observation in decades is also small (-.37), but negative.
+In other words, on average, respondents exhibit less wave-to-wave change in their responses the longer a question is asked.
+However, there is significantly more variance around this main effect, with a standard deviation of about .88.
+This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.
 
@@
 
+--
+There are theoretically relevant trade-offs from focusing only on the same questions over time.
+By focusing only on the same questions over time, we control for question content and structure and ensure that the trend over time is due to something other than changes in survey administration.
+Because our panels span more than five decades that saw dramatic improvements in survey question design, questions included in more recent surveys might simply be of higher quality, with less wave-to-wave change, than items included in older surveys.
+
 We face several limitations in trying to make broader claims.
@@
 We have attempted to limit this problem by averaging as many items as possible for each panel, but the number of questions in each panel is highly variable.
+
+--
 

### 2024-06-18 code/old/clean_panels.R
weights to some panels, clean repo

@@ -1,0 +1,1364 @@
+#Cleaning for Reliability Models
+library(tidyverse)
+library(haven)
+library(broom)
+
+
+anes5 <- read_dta("~/Dropbox/data/anes/anes5660/anes_mergedfile_1956to1960.dta") 
+
+clean_anes5 <- anes5 %>% 
+  select(#Negroes fair treatment
+    V560044,V580329,V600628, 
+    #deseg
+    V560074,V580333,V600636,
+    #Job guarantee
+    V560032,V580321,V600620,
+    #Help poor countries
+    V560041,V580327,V600626,
+    #Build schools
+    V560053,V580325,V600624,
+    #Fight communism
+    V560056,V580331,V600630,
+    #Leave utilities to prvt mkt
+    V560059,V580319,V600618,
+    #US better off stay home
+    V560035,V580323,V600622,
+    #Partyid
+    V560088,V580360,V600835,
+    #closerjob
+    V560034, V580322, V600621,
+    #closeraid
+    V560043, V580328, V600627,
+    #closertreat
+    V560046, V580330, V600629,
+    #closerbuild
+    V560055, V580326, V600625,
+    #closerfghtcomm
+    V560058, V580332, V600631,
+    #closerprvtpwr
+    V560061, V580320, V600619,
+    #closerdeseg
+    V560076, V580336, V600639,
+    #Age
+    V560295, V580472, V600688, 
+    #Marital status
+    V560177, V580474, V600690,
+    #Education
+    V560181, V580478, V600694,
+    #Children in household
+    V560178, V580475, V600691,
+    #Occupational code
+    V560120,
+    #sex
+    V560171, V580468, V600684,
+    #race
+    V560172, V580469, V600685,
+    #survey weight
+    V600569) %>%
+  mutate(weight = V600569,
+         age_1 = V560295,
+         age_2 = V580472,
+         age_3 = V600688,
+         student_1 = ifelse(V560120 %in% c(92), 1, 0),
+         marital_1 = V560177, marital_2 = V580474,
+         marital_3 = V600690,
+         marital_1 = ifelse(V560177 == 9, NA, V560177),
+         marital_1 = ifelse(V560177 %in% c(2), "single/nm", marital_1),
+         marital_1 = ifelse(V560177 %in% c(1), "married", marital_1),
[…]

### 2024-06-18 code/clean_panels.R
weights to some panels, clean repo

@@ -1,1364 +1,1817 @@
-#Cleaning for Reliability Models
-library(tidyverse)
+
+#########################################
+#### Cleaning for Stability Analysis ####
+#########################################
+
+#Packages
 library(haven)
+library(tidyverse)
 library(broom)
+library(broom.mixed)
+library(lme4)
 
-
+#Load in ANES panels
 anes5 <- read_dta("~/Dropbox/data/anes/anes5660/anes_mergedfile_1956to1960.dta") 
+anes7 <- read_dta("~/Dropbox/data/anes/anes7276/anes_mergedfile_1972to1976.dta")
+anes8 <- read_dta("~/Dropbox/data/anes/anes1980/anes1980.dta")
+anes90 <- read_dta("~/Dropbox/data/anes/anes9092/anes_mergedfile_1990to1992.dta")
+anes9 <- read_dta("~/Dropbox/data/anes/anes9297/anes_mergedfile_1992to1997.dta") 
+anes0 <- read_dta("~/Dropbox/data/anes/anes0004/anes_mergedfile_2000to2004.dta")
+anes16 <- read_sav("~/Dropbox/data/anes/anes1620.sav")
+anes20 <- read_dta("~/Dropbox/data/anes/anes2022/anes2022")
 
-clean_anes5 <- anes5 %>% 
-  select(#Negroes fair treatment
-    V560044,V580329,V600628, 
-    #deseg
-    V560074,V580333,V600636,
-    #Job guarantee
-    V560032,V580321,V600620,
-    #Help poor countries
-    V560041,V580327,V600626,
-    #Build schools
-    V560053,V580325,V600624,
-    #Fight communism
-    V560056,V580331,V600630,
-    #Leave utilities to prvt mkt
-    V560059,V580319,V600618,
-    #US better off stay home
-    V560035,V580323,V600622,
-    #Partyid
-    V560088,V580360,V600835,
-    #closerjob
-    V560034, V580322, V600621,
-    #closeraid
-    V560043, V580328, V600627,
-    #closertreat
-    V560046, V580330, V600629,
-    #closerbuild
-    V560055, V580326, V600625,
-    #closerfghtcomm
-    V560058, V580332, V600631,
-    #closerprvtpwr
-    V560061, V580320, V600619,
-    #closerdeseg
-    V560076, V580336, V600639,
-    #Age
-    V560295, V580472, V600688, 
-    #Marital status
[…]

### 2024-06-18 code/kkapproach/time_change.R

@@ -1,1919 +1,0 @@
-
-library(haven)
-library(tidyverse)
-library(broom)
-library(broom.mixed)
-library(lme4)
-
-anes5 <- read_dta("~/Dropbox/data/anes/anes5660/anes_mergedfile_1956to1960.dta") 
-anes7 <- read_dta("~/Dropbox/data/anes/anes7276/anes_mergedfile_1972to1976.dta")
-anes8 <- read_dta("~/Dropbox/data/anes/anes1980/anes1980.dta")
-anes90 <- read_dta("~/Dropbox/data/anes/anes9092/anes_mergedfile_1990to1992.dta")
-anes9 <- read_dta("~/Dropbox/data/anes/anes9297/anes_mergedfile_1992to1997.dta") 
-anes0 <- read_dta("~/Dropbox/data/anes/anes0004/anes_mergedfile_2000to2004.dta")
-
-gss6 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel06.dta")
-gss8 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel08.dta")
-gss10 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel10.dta")
-gss20 <- read_dta("~/Dropbox/data/gss2020panel/gss2020panel.dta")
-
-anes16 <- read_sav("~/Dropbox/data/anes/anes1620.sav")
-anes20 <- read_dta("~/Dropbox/data/anes/anes2022/anes2022")
-
-
-anes5_long <- anes5 %>%
-  mutate(id = 1:nrow(anes5)) %>%
-  zap_labels() %>%
-  select(id, V560193,	V580309,	V600578,	V600837,
-         V560088, V580360, V600657, V600835,
-         V560295, V580472, V600688,
-         V560035, V580323, V600622) %>%
-  mutate(across(c(V560088, V580360, V600657, V600835), ~ifelse(.x %in% c(7,8,9), NA, .x)),
-         across(c(V560295, V580472, V600688), ~ifelse(.x > 97, NA, .x)),
-         across(c(V560035, V580323, V600622),
-                ~recode(.x, 
-                        "1"=1, "2"=1, "3"=1.5, "4"=2, "5"=2, "7"=NA_real_,
-                        "0"=NA_real_, "8"=NA_real_, "9"=NA_real_)),
-         age_1 = V560295,
-         age_2 = V580472,
-         age_3 = V600688,
[…]

### 2024-06-18 manuscript/formative_period_kk.Rmd

@@
 For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points.
-Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14), we did not treat these as separate panels for the purposes of generating three distinct time points.
+Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same set of questions, we did not treat these as separate panels for the purposes of generating three distinct time points.
 In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.
@@
 Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change.
-Across our panels, duration between questions ranges from a few days (for questions asked in both pre- and post-election surveys in the ANES) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
+Panel duration varied from less than one year (the 1980 ANES Major Panel Study) to five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
 
-In the one-item, one-panel case (e.g., partisan identification in the 1972-76 panel), we could model the absolute change from wave-to-wave as a function of the duration between observations.
+In the one-item, one-panel case (e.g., partisan identification in the 1972-76 panel), we could model the absolute change from wave-to-wave as a function of an intercept and the duration between observations.
 
@@
 
-If respondents have a stable underlying partisan identification they report with error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$.
+If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$.
 In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise).
 Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$).
-For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3, and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.
+For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.
+More than two observations of the same question in a single panel produces variation in duration, which is why we ensure that each included question has at least one panel with three observations.
 
-Our theoretical question deals primarily with whether these coefficients ($\beta_0$ and $\beta_1$) change over time.
+Our theoretical question deals primarily with the amount of change over time ($\beta_0$ and $\beta_1$) changes over time.
 To explore that, we can add an additional term for time and interact the duration term with time.
@@
 $\beta_3$ captures the expected change in absolute change over time for a one-year change in duration.
-A positive $\beta_3$ value would suggest that people are more likely to undergo a "real" change in their opinion over time in later panels than earlier panels.
+A positive $\beta_3$ value would suggest that people are more likely to undergo a durable or persisting change in their opinion over time in later panels than earlier panels.
 
@@
 
+# Weights
+
+We focus on using weights that accounts for sampling design.
+
+For the 2016-20 ANES panel, almost all questions are asked in either the pre- or post-election version of both surveys.
+The panel includes weights for .
+The single item that was asked in the pre- and post-election waves of the 2016 panel (polviews) is weighted using the pre- election weights.
+
+For the 2020-22 ANES panel.
+
 # Results
@@
 
-The intercept suggests that, on average, respondents varied by about a fifth of the scale (17 points), controlling for duration.
-This is a rather high level of variance, but consistent with previous studies that find that, on any particular question, survey respondents tend to make large wave-to-wave changes.
-Variance around this intercept is quite large, with a standard deviation of about 5.67, meaning many items vary by as much as 20 percent of the scale, while some vary by less than 10 points.
+The intercept term in Model 1 suggests that, on average, respondents varied by about a fifth of the scale (17 points), controlling for duration, on any particular question.
+This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965].
+At the same time, random-effect variance around this main intercept term is quite large, with a standard deviation of about 5.67.
+Many items vary by as much as 20 percent of the scale, while some vary by less than 10 points.
 
-Second, duration is weakly positively associated with wave-to-wave absolute change.
+Second, duration between observations is positively but weakly associated with wave-to-wave absolute change.
 For every additional year between waves, respondents' absolute deviance increases about .71 points.
 Notably, variance around this main effect is relatively small (.43), meaning that two thirds of questions have duration effects that fall between .23 and 1.14.
-Broadly speaking, we can say that most wave-to-wave change is non-persisting, consistent with other recent work [@kiley2020; @lersch2023].
-The main effect of year of observation in decades is also small (-.37), but negative.
-In other words, on average, respondents exhibit less wave-to-wave change in their responses the longer a question is asked.
-However, there is significantly more variance around this main effect, with a standard deviation of about .88.
+Given the comparatively high level of wave-to-wave change independent of duration, our results broadly suggest that most wave-to-wave change is non-persisting, consistent with other recent work [@kiley2020; @lersch2023].
+
+The main effect of year of observation in decades is negative and also small (-.37).
+On average, respondents exhibit less wave-to-wave change in their responses in more recent panels.
+However, question-level random effects vary considerably around this main effect, with a standard deviation of about .88.
 This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.

### 2024-06-20 code/clean_panels.R
Adds new variables and cleans them

@@ -1,152 +1,366 @@
-
 #########################################
 #### Cleaning for Stability Analysis ####
 #########################################
 
-#Packages
+# Packages ====
 library(haven)
 library(tidyverse)
 library(broom)
 library(broom.mixed)
 library(lme4)
 
-#Load in ANES panels
-anes5 <- read_dta("~/Dropbox/data/anes/anes5660/anes_mergedfile_1956to1960.dta") 
-anes7 <- read_dta("~/Dropbox/data/anes/anes7276/anes_mergedfile_1972to1976.dta")
-anes8 <- read_dta("~/Dropbox/data/anes/anes1980/anes1980.dta")
-anes90 <- read_dta("~/Dropbox/data/anes/anes9092/anes_mergedfile_1990to1992.dta")
-anes9 <- read_dta("~/Dropbox/data/anes/anes9297/anes_mergedfile_1992to1997.dta") 
-anes0 <- read_dta("~/Dropbox/data/anes/anes0004/anes_mergedfile_2000to2004.dta")
-anes16 <- read_sav("~/Dropbox/data/anes/anes1620.sav")
-anes20 <- read_dta("~/Dropbox/data/anes/anes2022/anes2022")
+# Help Functions ====
+convert_to_date <- function(date_str) {
+  year <- substr(date_str, 1, 4)
+  month <- substr(date_str, 5, 6)
+  day = substr(date_str, 7, 8)
+  formatted <- as.character(paste(year,month,day,sep="-"))
+  return(formatted)
+}
 
-#Load in GSS panels
-gss6 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel06.dta")
-gss8 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel08.dta")
-gss10 <- read_dta("~/Dropbox/data/gss_data/gsspanels/gsspanel10.dta")
-gss20 <- read_dta("~/Dropbox/data/gss2020panel/gss2020panel.dta")
+# Read in Data ====
+anes5 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1956to1960.dta")
+anes7 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1972to1976.dta")
+anes8 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes1980.dta")
[…]

### 2024-06-20 code/clean_panels.R
weights,cleaning

@@ -28,14 +28,19 @@ gss20 <- read_dta("~/Dropbox/data/gss2020panel/gss2020panel.dta")
 
 
 ### 1956-60 ANES Panel
+#Does not include weights that compensate for sampling
+#design or sample non-response
 anes5_long <- anes5 %>%
   mutate(id = 1:nrow(anes5)) %>%
   zap_labels() %>%
-  select(id, V560193,	V580309,	V600578,	V600837,
+  select(id, V600569, V560193,	V580309,	V600578,	V600837,
          V560088, V580360, V600657, V600835,
          V560295, V580472, V600688,
          V560035, V580323, V600622) %>%
-  mutate(across(c(V560088, V580360, V600657, V600835), ~ifelse(.x %in% c(7,8,9), NA, .x)),
+  mutate(weight_4 = as.character(V600569),
+         weight_3 = weight_4,
+         weight_2 = weight_4,
+         across(c(V560088, V580360, V600657, V600835), ~ifelse(.x %in% c(7,8,9), NA, .x)),
          across(c(V560295, V580472, V600688), ~ifelse(.x > 97, NA, .x)),
          across(c(V560035, V580323, V600622),
                 ~recode(.x, 
@@ -137,36 +142,27 @@ anes5_long <- anes5 %>%
          stayhome_1 = as.character(V560035), 
          stayhome_2 = as.character(V580323), 
          stayhome_3 = as.character(V600622)) %>%
-  select(id, age_1:date_4, partyid_1:stayhome_3) %>% 
-  pivot_longer(age_1:stayhome_3) %>%
+  select(id, weight_2, weight_3, weight_4, age_1:date_4, partyid_1:stayhome_3) %>% 
+  pivot_longer(weight_2:stayhome_3) %>%
   separate(name, into = c("measure", "wave")) %>%
   spread(measure, value) %>%
   mutate(date = as.Date(date),
-         across(c(age, partyid:stayhome),
+         across(c(weight, age, partyid:stayhome),
                 ~as.numeric(.x)),
          partyid = (partyid/6)*100,
          stayhome = (stayhome-1)*100) %>%
   pivot_longer(partyid:stayhome) %>%
-  group_by(id, name) %>%
-  mutate(d1_value = lead(date) - date,
-         d2_value = lead(date,2) - date,
-         d3_value = lead(date,3) - date,
-         a1_value = abs(value - lead(value)),
[…]

### 2024-06-20 code/analyses.R
weights,cleaning

@@ -7,7 +7,60 @@
 
 #Load in long data frame from clean_data.R file
 
-load("./clean_data/long_data.Rdata")
+load("./clean_data/long_difference.Rdata")
+
+# Summarize absolute difference at the wave-pair level for each age group
+# for each question. 
+summaries <- long_difference %>%
+  mutate(wave_pair = paste(t1, t2, df, sep = "-")) %>%
+  group_by(wave_pair, name, age_group) %>%
+  summarise(d = weighted.mean(duration), a = weighted.mean(abs_diff),
+            date = weighted.mean(date)) %>%
+  #Filtering out bad questions
+  filter(name != "natpoor", 
+         name != "nathome") %>% ungroup() %>%
+  mutate(dec_diff = (as.numeric(difftime(date, min(date), units = "days")))/3652.5)
+
+# Multilevel model
+m1 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + 
+             (1 + d + dec_diff + age_group + dec_diff*age_group|name),
+           data = summaries)
+
+# Data to predict
+new.data <- expand_grid(age_group = c(unique(summaries$age_group)),
+                        dec_diff = seq(0, 6.4, by = .1),
+                        d = 0, name = unique(summaries$name))
+
+#Predict data
+new.data$yhat <- predict(m1, newdata = new.data)
+
+#Graph predictions
+new.data %>%
+  mutate(group = paste(age_group, name, sep = "-")) %>%
+  mutate(year = dec_diff*3652.5 + as.Date("1956-11-11")) %>%
+  ggplot(aes(x = year, y = yhat, color = age_group)) + 
+  geom_line(alpha = .3, aes(group = group)) + 
+  geom_smooth(linewidth = 2) + 
+  theme_bw() + 
+  facet_wrap(~age_group) + 
+  labs(x = "Year", y = "Predicted wave-to-wave change",
+       color = "Age Group",
+       title = "Predicted wave-to-wave change by age group",
+       subtitle = "Individual question trajectories and overall trajectory") + 
+  scale_color_brewer(type = "qual", palette = 2) +
+  theme(legend.position = "none")
+
+#Look at questions instead
+new.data %>%
+  mutate(group = paste(age_group, name, sep = "-")) %>%
+  ggplot(aes(x = dec_diff, y = yhat, color = age_group)) + 
+  geom_point(data = summaries, aes(y = a)) + 
[…]

### 2024-06-21 code/clean_panels.R
harmonizing

@@ -19,19 +19,21 @@ convert_to_date <- function(date_str) {
 }
 
 # Read in Data ====
-anes5 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1956to1960.dta")
-anes7 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1972to1976.dta")
-anes8 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes1980.dta")
-anes90 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1990to1992.dta")
-anes9 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1992to1997.dta")
-anes0 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_2000to2004.dta")
-anes16 <- read_sav("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_2016to2020.sav")
-anes20 <- read_csv("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_2022.csv")
+#Load in ANES panels
+anes5 <- read_dta("~/Dropbox/data/anes/anes5660/anes_mergedfile_1956to1960.dta") 
+anes7 <- read_dta("~/Dropbox/data/anes/anes7276/anes_mergedfile_1972to1976.dta")
+anes8 <- read_dta("~/Dropbox/data/anes/anes1980/anes1980.dta")
+anes90 <- read_dta("~/Dropbox/data/anes/anes9092/anes_mergedfile_1990to1992.dta")
+anes9 <- read_dta("~/Dropbox/data/anes/anes9297/anes_mergedfile_1992to1997.dta") 
+anes0 <- read_dta("~/Dropbox/data/anes/anes0004/anes_mergedfile_2000to2004.dta")
+anes16 <- read_sav("~/Dropbox/data/anes/anes1620.sav")
+anes20 <- read_dta("~/Dropbox/data/anes/anes2022/anes2022")
 
-gss6 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/gss/gsspanel06.dta")
[…]

### 2024-06-24 code/get_dem_variables.R
Partial cleaning of demographic variables

@@ -1,0 +1,412 @@
+###############################################
+#### Bringing in the demographic variables ####
+###############################################
+
+# Packages ====
+library(tidyverse)
+library(haven)
+
+# Read in the data ====
+anes5 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1956to1960.dta")
+anes7 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1972to1976.dta")
+anes8 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes1980.dta")
+anes90 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1990to1992.dta")
+anes9 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1992to1997.dta")
+anes0 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_2000to2004.dta")
+anes16 <- read_sav("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_2016to2020.sav")
+anes20 <- read_csv("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_2022.csv")
+
+gss6 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/gss/gsspanel06.dta")
+gss8 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/gss/gsspanel08.dta")
+gss10 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/gss/gsspanel10.dta")
+gss20 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/gss/gsspanel20.dta")
+
+# ANES 1950 demographics ====
+anes5_demog <- anes5 %>%
+  mutate(id = 1:nrow(.)) %>%
+  select(
+    id,
+    #Marital status
+    V560177,
+    V580474,
+    V600690,
+    #Education
+    V560181,
+    V580478,
+    V600694,
+    #Children in household
+    V560178,
+    V580475,
+    V600691,
+    #sex
+    V560171,
+    V580468,
+    V600684,
+    #race
+    V560172,
+    V580469,
+    V600685
+  ) %>%
[…]

### 2024-06-24 code/analyses.R
few more questions

@@ -9,6 +9,19 @@
 
 load("./clean_data/long_difference.Rdata")
 
+#Panel summaries
+psummaries <- long_difference %>%
+  mutate(wave_pair = paste(t1, t2, df, sep = "-")) %>%
+  group_by(name, df) %>%
+  summarise(d = weighted.mean(duration), a = weighted.mean(abs_diff),
+            date = weighted.mean(date), n = n(), sd = sd(abs_diff)) %>%
+  #Filtering out bad questions
+  filter(name != "natpoor", 
+         name != "nathome") %>% ungroup() %>%
+  mutate(dec_diff = (as.numeric(difftime(date, max(date), units = "days")))/3652.5) %>%
+  mutate(se = sd/sqrt(n))
+
+
 # Summarize absolute difference at the wave-pair level for each age group
 # for each question. 
 summaries <- long_difference %>%
@@ -20,7 +33,8 @@ summaries <- long_difference %>%
   filter(name != "natpoor", 
          name != "nathome") %>% ungroup() %>%
   mutate(dec_diff = (as.numeric(difftime(date, max(date), units = "days")))/3652.5) %>%
-  mutate(se = sd/sqrt(n))
+  mutate(se = sd/sqrt(n)) %>%
+  mutate(name = ifelse(name == "nateduc", "natschools", name))
 
 # Multilevel model
 m1 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + 

### 2024-06-24 code/clean_panels.R

@@ -79,7 +79,16 @@ anes5_long <- anes5 %>%
     V600664, # Pay attention campaign 3
     
     V560108, # People like me, no say 1
-    V600673 # People like me, no say 3
+    V600673, # People like me, no say 3
+    
+    V560111, #so many vote, 1
+    V600676, #so many vote, 3
+    
+    V560112, #pol too complicated 1
+    V600677, #pol too complicated 3
+    
+    V560115, #officials dont care 1
+    V600680 #officials dont care 3
          ) %>%
   mutate(weight_4 = as.character(V600569),
          weight_3 = weight_4,
@@ -87,11 +96,13 @@ anes5_long <- anes5 %>%
          across(c(V560088, V580360, V600657, V600835), ~ ifelse(.x %in% c(7, 8, 9), NA, .x)),
          across(c(V560295, V580472, V600688), ~ ifelse(.x > 97, NA, .x)),
          across(c(V560097, V580373, V600664), ~ ifelse(.x %in% c(8,9), NA, .x)), 
-         across(c(V560108, V600673), ~ ifelse(.x %in% c(8,9), NA, .x)),
          across(c(V560035, V580323, V600622),
                 ~recode(.x, 
                         "1"=1, "2"=1, "3"=1.5, "4"=2, "5"=2, "7"=NA_real_,
                         "0"=NA_real_, "8"=NA_real_, "9"=NA_real_)),
+         across(c(V560108, V600673, V560111, V600676, V560112, V600677,
+                  V560115, V600680),
+                ~ifelse(.x == 8, 3, ifelse(.x == 9, NA, .x))),
          age_1 = V560295,
          age_2 = V580472,
          age_3 = V600688,
@@ -191,19 +202,27 @@ anes5_long <- anes5 %>%
          attentioncpg_2 = as.character(V580373), 
          attentioncpg_3 = as.character(V600664), 
          ppllikeme_1 = as.character(V560108), 
-         ppllikeme_3 = as.character(V600673)) %>%
-  select(id, weight_2, weight_3, weight_4, age_1:date_4, partyid_1:ppllikeme_3) %>% 
-  pivot_longer(weight_2:ppllikeme_3) %>%
+         ppllikeme_3 = as.character(V600673),
+         manyvote_1 = as.character(V560111), #so many vote, 1
+         manyvote_3 = as.character(V600676), #so many vote, 3
+         complicated_1 = as.character(V560112), #pol too complicated 1
+         complicated_3 = as.character(V600677), #pol too complicated 3
[…]

### 2024-06-25 manuscript/formative_period_kk.Rmd
kkedits624

@@
 Over the past 70 years, changes in the timing of major life course events such as marriage and childbirth have transformed and extended this life stage.
-We leverage these changes to explore whether life course changes, as well as broader social changes, have affected how likely people of different cohorts are to make durable changes of general political dispositions over a comparable time frame.
+We leverage these changes to explore whether changes in the structure of the life course have affected how likely people of different cohorts are to change their stated opinions over time.
 We find strikingly little difference in the rates of opinion change among young people in different eras.
@@
 When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's ability to report stable attitudes over time or to the structural features of the survey administration.
-Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change.
-Panel duration varied from less than one year (the 1980 ANES Major Panel Study) to five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
+Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experiences more shocks that should shift their opinions.
+In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
 
@@
 
-Model 1 in Table XXX presents the main effects and variance components of the equation presented above.
+Model 1 in Table XXX presents the main effects and variance components of the multilevel model presented above.
 
-The intercept term in Model 1 suggests that, on average, respondents varied by about a fifth of the scale (17 points), controlling for duration, on any particular question.
+The intercept term in Model 1 suggests that, on average, survey respondents are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question.
 This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965].
-At the same time, random-effect variance around this main intercept term is quite large, with a standard deviation of about 5.67.
-Many items vary by as much as 20 percent of the scale, while some vary by less than 10 points.
+At the same time, random-effect variance around this main intercept term is quite large, with a standard deviation of about 5.86.
+On some questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.
 
@@
 For every additional year between waves, respondents' absolute deviance increases about .71 points.
-Notably, variance around this main effect is relatively small (.43), meaning that two thirds of questions have duration effects that fall between .23 and 1.14.
-Given the comparatively high level of wave-to-wave change independent of duration, our results broadly suggest that most wave-to-wave change is non-persisting, consistent with other recent work [@kiley2020; @lersch2023].
+Notably, variance around this main effect is relatively small (s.d.
+= 0.43).
+In other words, while some items demonstrate a stronger effect of duration than others, these differences are of comparable magnitude.
+Given the comparatively high level of wave-to-wave change independent of duration, our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].
 
-The main effect of year of observation in decades is negative and also small (-.37).
+Turning now to the coefficients for age groups, members of the youngest age group (18-25) are notably less stable than respondents in other age groups in the most recent panels.
+Respondents in the 26-33 age group are about 1.5 points more stable on average, respondents in the 34-65 age group are about 3 points more stable on average, and respondents in the 65 and over age group are about 4 points more stable on average, than respondents in the 18 to 25 age group.
+This provides broad evidence for the "impressionable years" and "increasing persistence" hypotheses, as the youngest age groups are less stable than others, followed by the 26-33 age group.
+
+The main effect of year of observation in decades is negative and also small.
 On average, respondents exhibit less wave-to-wave change in their responses in more recent panels.
@@
 This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.
+
+Turning now to the central question of whether the amount of change we observe for each group has changed over time, we turn to the interaction effects between time and age group.
+The coefficient for the reference category (18-25) suggests that respondents in this age group have not become any more or less stable opinion responders since the start of the observation window.
 

### 2024-06-25 code/analyses.R

@@ -36,10 +36,16 @@ summaries <- long_difference %>%
   mutate(se = sd/sqrt(n)) %>%
   mutate(name = ifelse(name == "nateduc", "natschools", name))
 
+
 # Multilevel model
 m1 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + 
              (1 + d + dec_diff + age_group + dec_diff*age_group|name),
            data = summaries %>% filter(sd > 0), weights = 1/se)
+
+m2 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + d*age_group + 
+             (1 + d + dec_diff + age_group + dec_diff*age_group + d*age_group|name),
+           data = summaries %>% filter(sd > 0), weights = 1/se)
+
 
 # Data to predict
 new.data <- expand_grid(age_group = c(unique(summaries$age_group)),
@@ -47,7 +53,7 @@ new.data <- expand_grid(age_group = c(unique(summaries$age_group)),
                         d = 0, name = unique(summaries$name))
 
 #Predict data
-new.data$yhat <- predict(m1, newdata = new.data)
+new.data$yhat <- predict(m2, newdata = new.data)
 
 #Graph predictions
 new.data %>%

### 2024-06-26 code/get_dem_variables.R
Finished getting demographic variables

@@ -64,12 +64,12 @@ anes5_demog <- anes5 %>%
     marital_3 = ifelse(V600690 %in% c(1), "married", marital_3),
     marital_3 = ifelse(V600690 %in% c(3, 4, 5), "other", marital_3),
     #IS there a child under 18 in the household.
-    child_1 = as.character(ifelse(V560178 == 99, NA, V560178)),
-    child_1 = as.character(ifelse(V560178 == 0, 0, 1)),
-    child_2 = as.character(ifelse(V580475 == 99, NA, V580475)),
-    child_2 = as.character(ifelse(V580475 == 0, 0, 1)),
-    child_3 = as.character(ifelse(V600691 == 99, NA, V600691)),
-    child_3 = as.character(ifelse(V600691 == 0, 0, 1)),
+    anychild_1 = as.character(ifelse(V560178 == 99, NA, V560178)),
+    anychild_1 = as.character(ifelse(V560178 == 0, 0, 1)),
+    anychild_2 = as.character(ifelse(V580475 == 99, NA, V580475)),
+    anychild_2 = as.character(ifelse(V580475 == 0, 0, 1)),
+    anychild_3 = as.character(ifelse(V600691 == 99, NA, V600691)),
+    anychild_3 = as.character(ifelse(V600691 == 0, 0, 1)),
     #The third education variable is differently coded
     ed_1 = ifelse(V560181 == 9, NA, V560181),
     ed_1 = ifelse(V560181 %in% c(0, 1, 2, 3, 4), "less than", ed_1),
@@ -144,15 +144,15 @@ anes7_demog <- anes7 %>%
          marital_3 = ifelse(V763370 %in% c(1,7), "married", marital_3),
          marital_3 = ifelse(V763370 %in% c(3,4,5), "other", marital_3),
          #children
-         child_2 = ifelse(V742408 %in% c(0, 9), NA, V742408),
-         child_2 = ifelse(V742408 == 1, 1, child_2),
-         child_2 = ifelse(V742408 == 5, 0, child_2),
-         child_3 = ifelse(V763371 %in% c(0, 9), NA, V763371),
-         child_3 = ifelse(V763371 == 1, 1, child_3),
-         child_3 = ifelse(V763371 == 5, 0, child_3),
[…]

### 2024-06-26 manuscript/formative_period_kk.Rmd

@@
 fontsize: 12pt
-abstract: Early adulthood is consistently identified as a “formative period” in the life course, when people have heightened probability of changing their personal culture.
-This probability has often been attributed to the social conditions of adolescence.
-Over the past 70 years, changes in the timing of major life course events such as marriage and childbirth have transformed and extended this life stage.
-We leverage these changes to explore whether changes in the structure of the life course have affected how likely people of different cohorts are to change their stated opinions over time.
-We find strikingly little difference in the rates of opinion change among young people in different eras.
-People of all ages have similar attitude behaviors over time.
-The overall pattern suggests either that social conditions are not strongly related to attitude stability or that different social shifts have had counter-vailing effects that produce aggregate stability in rates of change.
+abstract: "Early adulthood is consistently identified as a “formative period” in the life course when people have heightened probability of changing their opinions, beliefs, and political preferences.
+This heightened rate of change is often attributed to the social conditions of this life stage.
+In this paper, we explore whether substantical changes in the timing and ubiquity of major life course transitions over the the past 70 years have affected the rate at which early adults and the rest of the population change their opinions on political and social issues.
+Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the 1960s.
+Instead, we find that other age groups have become more stable opinion holders."
 ---
@@
 
-Adolescence and young adulthood are typically identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when peoples' core ideas and attitudes are formed [@krosnick1989; @alwin1991].
-Research regularly finds that people are more likely to change their preferences, attitudes, and opinions in these life stages than other life stages, with opinions and dispositions becoming more stable once these life stages have passed [@sears2013; @kiley2020].
-This pattern, in which young people are open to cultural change and become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social change [@ryder1965; @mannheim1952].
+Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991].
+In support of this idea, regularly finds that people in these life stages are more likely to change their preferences, attitudes, and opinions than people in other life stages [@sears2013; @kiley2020].
+This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].
 
@@
 Explanations in the social sciences tend to highlight the unique social features of these early life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information [@alwin1991; @danigelis2007].
-Alternative explanations in the social sciences also focus on the unique features of middle age that tend to stabilize attitudes during this period, including higher levels of social authority [@eaton2009; @visser2004].
-Work in biology, psychology, and cultural evolution, in contrast, shows that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION].
-While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change in this life stage, the relative contributions of each are unclear.
+Other related explanations focus on the unique features of middle age that might stabilize attitudes during this period, including people holding positions with higher levels of social authority during middle age [@eaton2009; @visser2004].
+At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.
 
-Clarifying the relative importance of different mechanisms that drive the frequency of opinion change during this formative period is important for understanding the mutually reinforcing processes of social change and cultural evolution.
+While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change during early adulthood, the relative contributions of each are unclear, and clarifying their relative importance is important for understanding the mutually reinforcing processes of social change and cultural evolution.
 If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change.
 Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016].
-If the mechanisms are principally cognitive and biological, rather than social, then these changes are unlikely to affect rates of social change in the same way.
+If the mechanisms are principally cognitive and biological, rather than social, then these broader changes are unlikely to affect rates of social change in the same way.
 
-In this paper, we explore how rates of opinion change in different age groups have changed in response to broader social structure and changes in the timing of life course events since the 1950s.
-Over the course of the 20th century, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005].
+In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events.
+Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005].
 Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009].
@@
 
-We take advantage of this change over time to explore whether young adults of different cohorts demonstrate more or less change in their political opinions over time relative to other members of the population.
+Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of age categories.
+Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) are more likely to make changes of opinion than older respondents in the most recent panels but not in the earliest panels.
+This difference emerged over the second half of 20th Century not because young adults become more likely to make changes but because middle-aged and older respondents become less likely to make changes.
 
-We decompose over-time responses into reliability (a lack of random, non-persisting fluctuations) and stability (true consistency in latent dispositions net of these random fluctuations) for a range of political attitudes in nine nationally representative panel data sets spanning from the 1950s through the 2010s.
-We separately estimate these parameters for a comparable set of age groups in each panel to evaluate whether response reliability or stability changed as new cohorts moved through the same age ranges under different social conditions.
-
-Consistent with the "impressionable years" hypothesis, we find that the youngest respondents in each panel (ages 18-25) are consistently the least stable opinion reporters, suggesting they are more likely than other respondents to make durable changes in their views.
-They are also among the least reliable reporters, suggesting considerable short-term but non-persisting change in their views.
-But we find no consistent trajectory in either measure over time, among members of this age group or other members of the population.
-Notably, respondents aged 26 to 33 continue to look more like older age groups than the youngest age group in their rates of opinion stability and reliability, providing little evidence of "lengthening" of this formative window.
-
-We believe these findings have important implications for our understanding of why early adulthood is a significant formative period.
-While these findings do not provide conclusive evidence of one mechanism over the other, they confine the range of potential theories about the relationship between life course stages and attitude change.
-They suggest that there is something relatively timeless -- potentially biological and cognitive factors -- that accounts for increased opinion variance in young people, and that, barring significant evidence of counter-currents that stabilize people, changes in the social conditions of early adulthood have not increased the degree to which young people make meaningful changes of opinion.
+We believe these findings have important implications for our understanding of why early adulthood is a significant formative period for cultural and political attitudes.
 
@@
 
-## Data and Methods
+## Data
 
-Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time.
+Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time than members of earlier cohorts.
 Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, coming from the same general population.
@@
 Both the ANES and GSS are designed to generalize to the non-institutionalized U.S.
-adult (18+) population, and can be thought of as tapping the same general population.
+adult (18+) population, and therefore can be thought of as tapping the same general population.
 We included a total of 12 panels in our search: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20.
-We evaluated an additional panel conducted by the ANES in 2008-09 (with an additional follow-up in 2010), but no questions overlapped this panel and any other panel.
+We evaluated an additional panel conducted by the ANES in 2008-10, but no questions overlapped this panel and any other panel.
 
@@
 In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.
+When question wording changed across panels, we harmonize response options to the question structure with the fewest response options.
 
-This produced a total of ### questions tapping a broad range of topics including political and ideological identification, government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.
+This produced a total of 77 questions tapping a broad range of topics including political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.
 
@@
 
-When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's ability to report stable attitudes over time or to the structural features of the survey administration.
+When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time or to the structural features of the survey administration.
 Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experiences more shocks that should shift their opinions.
@@
 
+In all models, respondents in the 18-25 age group serve as the reference category.
+
 # Weights
@@
 # Results
-
-Model 1 in Table XXX presents the main effects and variance components of the multilevel model presented above.
-
-The intercept term in Model 1 suggests that, on average, survey respondents are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question.
-This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965].
-At the same time, random-effect variance around this main intercept term is quite large, with a standard deviation of about 5.86.
-On some questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.
-
-Second, duration between observations is positively but weakly associated with wave-to-wave absolute change.
-For every additional year between waves, respondents' absolute deviance increases about .71 points.
-Notably, variance around this main effect is relatively small (s.d.
-= 0.43).
-In other words, while some items demonstrate a stronger effect of duration than others, these differences are of comparable magnitude.
-Given the comparatively high level of wave-to-wave change independent of duration, our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].
-
-Turning now to the coefficients for age groups, members of the youngest age group (18-25) are notably less stable than respondents in other age groups in the most recent panels.
-Respondents in the 26-33 age group are about 1.5 points more stable on average, respondents in the 34-65 age group are about 3 points more stable on average, and respondents in the 65 and over age group are about 4 points more stable on average, than respondents in the 18 to 25 age group.
-This provides broad evidence for the "impressionable years" and "increasing persistence" hypotheses, as the youngest age groups are less stable than others, followed by the 26-33 age group.
-
-The main effect of year of observation in decades is negative and also small.
-On average, respondents exhibit less wave-to-wave change in their responses in more recent panels.
-However, question-level random effects vary considerably around this main effect, with a standard deviation of about .88.
-This suggests that while some items show decreased wave-to-wave change over time, others show increased wave-to-wave change over time.
-
-Turning now to the central question of whether the amount of change we observe for each group has changed over time, we turn to the interaction effects between time and age group.
-The coefficient for the reference category (18-25) suggests that respondents in this age group have not become any more or less stable opinion responders since the start of the observation window.
 
@@
 
-## Systematic Change
+## Opinion Changes
 
-We first evaluate the average estimates for absolute mean change for each age group.
-Figure 1 plots the average asbolute slope for each of the three age groups in each panel across all questions asked in that panel.
-Again, which questions are asked in each panel varies significantly.
+Model 1 in Table XXX presents the main effects and variance components of the multilevel model presented above.
+In all models, the 18-25 age group serves as the reference category, and the time variable is centered on the most recent panel (2016-2020), so coefficients reflect expectations for change in recent years, rather than at earlier time points.
+
+The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question.
+This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965].
+At the same time, random-effect variance around this main intercept term is quite large ($\sigma_{\beta_0} = 5.9220$).
+On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.
+
+The coefficients for age groups other than the reference category suggest that, on average, other age groups are more stable than people in the 18-25 age range in the most recent panels.
+Respondents in the 26-33 age group are about 1.1 points more stable on average, respondents in the 34-65 age group are about 2.2 points more stable on average, and respondents in the 65 and over age group are about 3.5 points more stable on average, than respondents in the 18 to 25 age group.
+This provides broad evidence for the "impressionable years" and "increasing persistence" hypotheses, as the youngest age groups are less stable than others, followed by the 26-33 age group.
+
+Duration between observations is positively but weakly associated with wave-to-wave absolute change.
+For every additional year between responses, the expected absolute deviance for respondents between the ages of 18 and 25 increases about .9 points.
+Notably, variance around this main effect is relatively small (s.d.
+= 0.52), meaning each one year duration only increases the amount changed by between .5 and 1.5.
+Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].
+While often called "measurement error," this change conflates both random change due to measurement error and true non-persisting change.
+
+The interactions between age and duration are, for the most part, small and not statistically significant, meaning groups make "real" change at roughly the same rates over time.
+People over 25 make slightly less change than people 18-25 for each additional year of observation, suggesting that the youngest age groups might be making more durable changes of opinion than their older counterparts, but these differences are small and substantively negligible, especially compared to the amount of non-persisting change.
+
+The intensification of early adulthood hypothesis (H3) generally suggests that the youngest respondents should make more change over time in more recent cohorts than their counterparts from earlier cohorts did when they were the same ages.
+The effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significant.
+On average, recent cohorts in this age group make approximately the same amount of wave-to-wave change in recent years as earlier cohorts did, accounting for the kinds of questions they were asked and time between waves.
+However, question-level random effects vary considerably around this main effect, with a standard deviation of about 1.4316.
+This suggests that the overall flat effect obscures quite large question-level differences, with some questions becoming more stable and others becoming less stable.
+We return to consider these question-specific trends later.
+
+Next, the extended early adulthood framework (H1) suggests that we should see a closing of the gap between people in the 18-25 and 26-33 age groups over time.
+The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that people in this age group have also not become any more or less likely to make changes of opinion since the earliest panels, and have also not become any more or less similar to the youngest respondents, and that significant changes in the social environment have not made more recent cohorts any more likely to make attitude changes than earlier cohorts were when they were in this age range.
+
+However, it is important to understand the full picture of change over time before drawing substantive conclusions.
+Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range, respondents in older age groups do show a notable time trend -- people in this age range become more stable opinion holders over time.
+To more clearly illustrate overall trends, Figure XXX plots the predicted time level of change for respondents in each age category over time.
 
 ```{r}
-bind_rows(mean_val5, mean_val8, mean_val9, mean_val7,
-          mean_val2k, mean_val90, mean_val06,
-          mean_val08, mean_val10) %>%
-  mutate(year = recode(df, "anes2k"=2000, "anes5"=1956,
-                       "anes7"=1972, "anes9"=1992,
-                       "anes8"=1980, "anes90"=1990,
-                       "gss06"=2006, "gss08"=2008,
-                       "gss10"=2010)) %>%
-  filter(year != 1980, year != 1990) %>%
-  ggplot(aes(x = year, y = mean, fill = group)) + 
-  geom_point(size = 2, shape = 21, color = "black") + 
+new.data %>%
+  mutate(group = paste(age_group, name, sep = "-")) %>%
+  mutate(year = as.Date("2020-11-12") + dec_diff*3652.5) %>%
+  ggplot(aes(x = year, y = yhat, color = age_group)) + 
+  geom_hline(yintercept = 0, color = "black") + 
+  #geom_line(alpha = .2, aes(group = group)) + 
+  geom_smooth(linewidth = 2) + 
   theme_bw() + 
-  labs(x = "Year", y = "Abs. Slope", fill = "Age Group")
+  #facet_wrap(~age_group) + 
+  labs(x = "Year", y = "Predicted wave-to-wave change",
+       color = "Age Group",
+       title = "Predicted wave-to-wave change by age group",
+       subtitle = "Individual question trajectories and overall trajectory") + 
+  scale_color_brewer(type = "qual", palette = 2) +
+  theme(legend.position = "none")
 ```
 
-Overall, we see few general patterns across age groups over time.
-There is little consistency in which age group displays the largest absolute change in each of the panels or the trajectory of rates of change across the panels.
-The 2006 and 2008 General Social Survey have the highest absolute change over time, but the 2010 panel, the most recent, is highly comparable to the other panels.
-While people in the 18 to 25 age range display higher rates of change than people in the 26-33 and 34 and older age group in the most two recent panels (the 2008 and 2010 GSS panels), suggesting some evidence of an intensification of early adulthood, they fall below people ages 34 and over in the panel begun only two years prior and overlapping (the 2006 GSS panel).
-
-In aggregating questions, we see no evidence for the "elongation" hypothesis that people in the 26 to 33 age range increasingly look like people in the 18 to 25 age range.
-People in the 26 to 33 age range are functionally identical to people over 34 in all panels, and they do not show any evidence of becoming more like people in the 18 to 25 age range over time.
-
-One challenge we noted at the outset is that the composition of questions in each panel changes over time, meaning our understanding of the general trajectories of age groups across cohorts might be biased by changes in question content, questions wording, or survey administration.
-To control for these potential sources of bias, we explore only questions that have been repeated almost verbatim in multiple panels.
-These are the seven-point scale of partisan identification (from Strong Democrat to Strong Republican) and the seven-point scale of political ideology (from Extremely Liberal to Extremely Conservative).
-These results are presented in Figure XXX below.
-
-```{r}
-bind_rows(bind_rows(group_meansc5) %>% mutate(year = 1956), 
-          bind_rows(group_meansc7) %>% mutate(year = 1972),
-          bind_rows(group_meansc9) %>% mutate(year = 1992), 
-          bind_rows(group_meansc2k) %>% mutate(year = 2000),
-          bind_rows(group_meansc06) %>% mutate(year = 2006),
-          bind_rows(group_meansc08) %>% mutate(year = 2008),
-          bind_rows(group_meansc10) %>% mutate(year = 2010)) %>%
-  filter(var %in% c("polviews", "partyid")) %>%
-  mutate(var = recode(var, "polviews"="Ideology",
-                      "partyid"="Identitification")) %>%
-  ggplot(aes(x = year, y = abs_mean_slope, fill = group)) +
-  geom_point(shape = 21, size = 2) + 
-  facet_wrap(~var) + 
-  theme_bw() + 
-  labs(x = "Year", y = "Abs. Mean Slope", 
-       fill = "Age Group")
-
-```
-
-For partisan identification, we see some evidence for the intensification hypothesis, with people in the 18-25 age range showing higher average absolute slopes than people in the other two age ranges.
-However, though the population as a whole also makes more linear change in partisan identification as well during these more recent years, complicating the overall picture.
-There is no evidence that people in the 26-33 age range increase more than the general population in this time frame, suggesting no evidence of the elongation of the formative period.
-
-We see no corresponding pattern for political ideology.
-Respondents in the 18 to 25 age range look remarkably similar to respondents in other age groups, and there is no obvious trajectory across cohorts.
-In one panel (the GSS panel beginning in 2006), people in this age range show higher rates of change, but this pattern does not hold even for the other GSS panels that overlap this panel.
-People in the 26 to 33 age range cannot be distinguished from people over 34 even in the most recent cohorts, suggesting no evidence for the elongation hypothesis.
-
-## Non-systematic change
-
-```{r}
-bind_rows(mean_var5, mean_var8, mean_var9, mean_var7,
-          mean_var2k, mean_var90, mean_var06,
-          mean_var08, mean_var10) %>%
-  mutate(year = recode(df, "anes2k"=2000, "anes5"=1956,
-                       "anes7"=1972, "anes9"=1992,
-                       "anes8"=1980, "anes90"=1990,
-                       "gss06"=2006, "gss08"=2008,
-                       "gss10"=2010)) %>%
-  filter(year != 1980, year != 1990) %>%
-  ggplot(aes(x = year, y = mean, fill = group)) + 
-  geom_point(size = 2, shape = 21, color = "black") + 
-  theme_bw() + 
-  labs(x = "Year", y = "Abs. Slope", fill = "Age Group")
-
-```
-
-In the early panels (the 1956 and 1972 ANES panels), there is little evidence that the youngest respodents exhibited more or less consistency over time in their attitude reports than the rest of the sample.
-In the 1956 panel, they have the lowest residual variance around their latent trajectories, and in the 1972 panel, they fall between the other two groups.
-However, in more recent cohorts, people in this age group consistently have the most residual variance of the three groups.
-Whether this is due to lack of knowledge of the political system (and therefore measurement error), or whether these vacillations around their latent trajectories are due to "exploration" across attitudes is not clear from these results.
-
-There is no consistent evidence for "elongation" of early adulthood in these data, as people in the 26-33 age group show no consistent trajectory over time.
-
-Turning again to the
-
-```{r}
-bind_rows(bind_rows(variancec5) %>% mutate(year = 1956), 
-          bind_rows(variancec7) %>% mutate(year = 1972),
-          bind_rows(variancec9) %>% mutate(year = 1992), 
-          bind_rows(variancec2k) %>% mutate(year = 2000),
-          bind_rows(variancec06) %>% mutate(year = 2006),
-          bind_rows(variancec08) %>% mutate(year = 2008),
-          bind_rows(variancec10) %>% mutate(year = 2010)) %>%
-  filter(var %in% c("polviews", "partyid")) %>%
-  mutate(var = recode(var, "polviews"="Ideology",
-                      "partyid"="Identitification")) %>%
-  ggplot(aes(x = year, y = mse, fill = group)) +
-  geom_point(shape = 21, size = 2) + 
-  facet_wrap(~var) + 
-  theme_bw() + 
-  labs(x = "Year", y = "Abs. Mean Slope", 
-       fill = "Age Group")
-
-```
-
-We first evaluate the average estimates for reliability and stability for each panel and for each age group to establish a baseline of comparison.
-Table 1 presents the average reliability and stability for each panel.
-Across panels, the 1956-58-60 panel shows the lowest average reliability of all the panels.[^badq] The 1980 panel, with the shortest time between waves (about 3 months between each wave), has notably higher reliability than other panels, a fact established in previous work [@krosnick1989].
-Having less time between waves increases the reliability of responses.
-The remaining panels all have comparable average reliabilities, with many within .02 of each other.
-
-[^badq]: This is potentially due to improvements in question-asking since the ANES first administered this survey.
-Many questions in this panel have been critiqued as difficult for average Americans to answer.
-
-Average stability estimates are also generally comparable across panels, ranging withing a couple points of each other.
-Of note here is that differences between the GSS panels are quite large, with the 2008-10-12 panel showing the highest label of stability while the 2010-12-14 panel showing the lowest, even though these panels contain the exact same questions and overlap considerably in terms of time.
-
-Table XXX presents the average reliability and stability for age groups.
-Here we see results broadly consistent with the "impressionable years" hypothesis.
-First, the 18-25 year old group has notably lower stability than other groups, suggesting higher rates of change in their "true" attitudes over time.
-Stability tends to increase up through middle age (42-29) before leveling off.
-Second, reliability is also substantially lower for the youngest age group and increases consistently as groups get larger.
-For both measures, the largest jump is between the youngest age group (18-25) and the one following it (26-33), supporting the impressionable years model over an increasing persistence model.
-
-These results reinforce life-course patterns established in other work and extend it to more recent panels, but they do not say anything about whether changes in the life course have extended or intensified the "impressionable years" by decreasing stability and reliability in the youngest group (18-25) over time or extending the values seen in that group up the age ladder (to respondents in the 26-33 group).
-For that we must examine age groups separately by panel.
-
-## Over-Time Trends
-
-Table XXX summarizes the average reliability and stability estimated for each of the age groups for each of the panels analyzed.
-The "formative period" or "impressionable years" hypothesis and previous work [@alwin2007; @alwin1989] suggests that we should observe low stability for the youngest cohorts and large and approximately equal stability in the older groups.
-The "extended early adulthood" hypotheses suggests that, over time, people in the 26-33 age group should come to more closely resemble people in the 18-25 age group, meaning their average stability will decrease over time as people in this age group continue to make durable changes of culture.
-The "intensification" of early adulthood hypothesis suggests that 18-25 year olds (and possible 26-33 year olds) will move farther apart from the rest of the population, both in terms of stability and reliability, as they make more temporary changes (lower reliability) and durable changes (lower stability) over time.
-
-It is important to note that some age-group/panel combinations have very small sample sizes, which can lead to large sampling errors and produce problematic estimates of reliability.
-Because of this, we remove some reliability and stability estimates that are outside the bounds of plausibility or were not able to be estimated.
-If a question produced an impalusible value for one of the age groups, we remove it for all age groups.
-
-Because information in the table is difficult to process, and because some of the panels have much higher average levels of stability and reliability because of shorter duration, Figure ...
-plots the average reliability estimates by panel, with each panel pegged to year it begain (e.g., the 1972-74-76 panel is plotted at 1972).
-The figure and table show the same overall age trends as the aggregate data.
-Across panels, the youngest respondents (ages 18-25) are the least stable respondents.
-In eight of the nine panels, they have the lowest average stability, and in the remaining panel, they have the second-lowest.
-However, there is no clear age pattern beyond that.
-Respondents in the second-youngest age group (26-33) have the second-lowest stability in only two of the panels (1990 and 2008), but this does not seem to indicate a pattern.
-In other panels, both early and late, these respondents seem indistinguishable from other respondents, especially in the 2006-10 and 2010-14 GSS panels (the most recent).
-
-There is no clear time trajectory for any age group when it comes to stability, even accounting for different average duration across panels.
-Groups do not, on average, appear to be getting more or less stable over time, with responses from the most recent GSS panels looking quite similar to responses from the 1972-76 ANES panel.
-
-Turning next to the question of reliability, the youngest aged group (18-25) are typically among the least reliable respondents in each panel, suggesting they are making higher rates of non-durable change in their attitudes than other member of the population.
-However, the pattern is not as consistent as with stability, and there is not always a clear gap between the youngest and the rest of the population.
-Other groups -- especially the oldest (66-83) -- frequently show less reliability in each panel.
-In terms of reliability, the second-youngest group (26-33) are functionally indistinguishable from middle-aged respondents in each panel, and much closer to those respondents than the youngest, providing no evidence for the "extended formative period" hypothesis.
-Again, as with stability, these findings reflect previous work on reliability and show that patterns continue in more recent panels.
-
-In terms of change over time, there appears to be a slight upward trend in reliability across the panels, even accounting for the two panels with different duration -- the 1980 and the 1990-91-92 ANES panels.
-However, aside from the general upward trend, there do not appear to be distinct time trends for any of the age groups.
-
-## Question Fixed Effects
-
-The preceding analysis has assumed that panels are generally comparable, and that average estimates of reliability and stability are comparable across panels as measures of reliability and stability toward "salient" issues for that period.
-We now turn toward exploring the reliability and stability for questions that are asked only in multiple panels.
-Table XXX presents regression coefficient estimates for model regressing stability estimates on time and fixed effects for question and panel, with an interaction between time and age group.
-The table shows that there is no discernible time trend for the stability estimates of 18-25 year olds, nor is their time trend statistically significantly different from the time trends for the other age groups.
-
-Table XXX presents similar coefficient estimates for the model regressing reliability on time and fixed effects for question, with an interaction between time and age group.
-The table also shows no meaningful time trend for respondents 18-25.
-The coefficient for the interaction between time and the 58-65 age group is statistically significant, suggesting people in this group became slightly more stable over time compared to 18-25 year olds, but it is difficult to make a clear judgment about the overall trend, and there are no other statistically significant interaction terms.
+In the earliest panels, the oldest respondents (65 and over) are substantially less stable opinion holders than younger respondents, while the remaining three age groups (18-25, 26-33, and 34-64) appear to change at approximately equal rates.
+As noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change over time.
 

### 2024-06-26 manuscript/formative_period_kk.Rmd

@@
 This heightened rate of change is often attributed to the social conditions of this life stage.
-In this paper, we explore whether substantical changes in the timing and ubiquity of major life course transitions over the the past 70 years have affected the rate at which early adults and the rest of the population change their opinions on political and social issues.
+In this paper, we explore whether substantial changes in the timing and ubiquity of major life-course transitions over the the past 70 years have affected the rate at which early adults and the rest of the population change their opinions on political and social issues.
 Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the 1960s.
@@
 Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991].
-In support of this idea, regularly finds that people in these life stages are more likely to change their preferences, attitudes, and opinions than people in other life stages [@sears2013; @kiley2020].
+In support of this idea, research regularly finds that people in these life stages are more likely to change their preferences, attitudes, and opinions than people in other life stages [@sears2013; @kiley2020].
 This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].
@@
 Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009].
-This suggests both a lengthening of and an intensification of this stage of the life course in ways that potentially affect the rates at which people undergo cultural change.
+These broad social changes potentially lengthen of and intensify this stage of the life course in ways that potentially has consequences for the rate at which people undergo cultural changes.
 
-Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of age categories.
+Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of age categories and question.
 Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) are more likely to make changes of opinion than older respondents in the most recent panels but not in the earliest panels.
-This difference emerged over the second half of 20th Century not because young adults become more likely to make changes but because middle-aged and older respondents become less likely to make changes.
-
-We believe these findings have important implications for our understanding of why early adulthood is a significant formative period for cultural and political attitudes.
+However, this difference emerged over the second half of 20th Century not because young adults become more likely to make changes, as life-course theories might predict, but because middle-aged and older respondents become less likely to make changes.
+Our findings complicate the narrative about social change and cultural change and raise questions about what social forces emerged to stabilize attitudes more in middle age for more recent cohorts.
 
@@
 It could also be the case that both change as a function of age.
-Because these mechanisms have not been disentangled, in this section we outline theoretical processes that have been suggested to contribute to one or both of these processes.
+Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.
 
@@
 
-The first potential mechanism for changes in attitude stability over time is rooted in the biology of the human brain and science of how people process information.
+The first potential mechanism for changes in attitude stability over time is rooted in the biology of the human brain and theories about how people process information.
 This approach tends to suggest that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally, whether they encounter it during adolescence or middle age, and are exposed to information at a (more or less) constant rate [@bartels2014; @gopnik2020; @sherratt2018].
@@
 Supporting this model is evidence that early formative periods have been observed not just in people, but across species.
-In a study of social learning among birds, for instance, Aplin and colleagues [-@aplin2017] show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.
+For example, in a study of social learning among birds, Aplin and colleagues [-@aplin2017] show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.
 
@@
 
-The thrust of the biological arguments is that, irrespective of the social conditions they inhabit, millennia of evolution have led to the development of a cognitive architecture that makes humans relatively open changing dispositions at younger ages and relatively resistant at older ages.
+The thrust of the biological arguments is that, irrespective of the contemporaneous social conditions humans inhabit, millennia of evolution have led to the development of a cognitive architecture that makes them relatively open changing dispositions at younger ages and relatively resistant at older ages.
 However, while biological mechanisms can shape how individuals respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change.
@@
 These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions.
-For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, and racial inequality [@campbell2016; @wodtke2018; @edelmann2023].
+For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic].
 Social networks also differ in this time period.
@@
 People in this "early adulthood" phase are not expected to be stable adults, something should not be taken for granted given social variation in norms around when maturity is reached [@cunningham2007; @elder1994].
+Cultural expectations to engage in new experiences, meet new people, and adopt new personalities could motivate the pursuit of new information during this life stage apart from social structural opportunities.
 
@@
 
-Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the "impressionable years" are principally an effect of how societies choose to structure the life course or the interaction between these life course stages and biological mechanisms.
+Given the distinct social conditions of the early-adult phase of life, there are reasons to believe that the "impressionable years" are principally an effect of how societies choose to structure the life course or the interaction between these life course stages and biological mechanisms.
 Because of this, contemporary historical trends offer a window to disentangle the role of socio-cultural and biological or cognitive factors in explaining why we observe "impressionable years."
@@
 First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts.
-Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood.
-In the United States, both the median age at first marriage and median age of first child birth has consistently risen since the mid-20th Century [@rosenfeld2009].
+Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce.
+In the United States, both the median age at first marriage and median age of first childbirth has consistently risen since the mid-20th Century [@rosenfeld2009].
 We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021].
@@
 
+Similarly, the duration of education has extended, with more people completing high school, college, and, increasingly, graduate degrees before entering the workforce.
+These transitions are also accompanied by heightened rates of geographic mobility, as well as movement between the workforce and education.
+
 If the distinctive nature of "impressionable years" is principally driven by their unique social circumstances, then these shifts have the potential to extend these social circumstances into later ages.
 This, in turn, could have an effect on how long exploration, and therefore openness to durable change, lasts throughout the life-course.
-If this is the case, we should expect to see lower levels of attitude stability (and therefore more change) among age groups that, for earlier cohorts, displayed mainly stability.
+If this is the case, we should expect to see heightened levels of attitude change among age groups that, for earlier cohorts, displayed mainly stability.
 In other words, people who were born in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age.
@@
 An increase in independence might entail more exploration, and greater ability to update attitudes as a result of this exploration.
-This assertion in turn engenders a prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.
+This assertion in turn engenders a second prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.
 
@@
 
-*Hypothesis 2: More recent cohorts will show higher levels of residual variance in the 18 to 25 age range, net of their over-time trajectories, compared to previous generations in the same age range.*
-
-Residual variance reflects not just the degree to which people make errors in their reporting of attitudes over time, but the baseline level of variance around a fixed attitude position that people express.
-In other words, we might expect younger people to bounce around their latent trajectory more if they engage in more exploration of attitudes over time, trying on different identities and ideas.
-
-Finally, the intensification of early adulthood hypothesis expects that the rates of durable attitude change will also increase for more recent cohorts.
-Freedom to explore outside of the watch of family members should result in more respondents making durable changes of opinion over time.
-
-*Hypothesis 3: People aged 18 to 25 will demonstrate lower levels of stability in recent years than they did in earlier years.*
+*Hypothesis 2: More recent cohorts will show higher levels of opinion change in the 18 to 25 age range compared earlier cohorts.*
 
@@
 
+## Broader Social Factors
+
+While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader cultural environment that might affect attitude stability.
+We briefly consider three here.
+
+First, the number of people pursuing and completing higher education increased significantly over the window studied here.
+Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007].
+A population with higher levels of education might be more likely to hold stable opinions regardless of broader social transformations.
+
+Second, we have seen significant expansion of information technology over this window that has dramatically increased access to information about social and political affairs, first through broadcast and cable television and then through the internet and social media.
+People have greater access to information with which to form opinions.
+
+Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina; @baldassarri].
+People now receive significantly clearer signals from elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020].
+This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently.
+
 # Analytical Strategy
@@
 Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time than members of earlier cohorts.
-Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, coming from the same general population.
+Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, with respondents coming from the same general population.
 To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years.
@@
 
-This produced a total of 77 questions tapping a broad range of topics including political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.
+Our search produced a total of 77 questions tapping a broad range of topics including political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.
+
+Table A1 in Appendix A outlines the questions included in our analysis, including their full question wording and response scale.
+For a small number of items where question wording remains consistent across panels but response scales differ, we harmonize by converting more fine-grained response scales to the more general scales, following the ANES time series.
 
@@
 When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time or to the structural features of the survey administration.
-Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experiences more shocks that should shift their opinions.
+Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions.
 In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
@@
 
-$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1*d_i + \epsilon_i $$
+$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \epsilon_i $$
 
-In this equation, $d_{i}$ captures the duration of time between $t-1$ and $t$ in years.
-In the 1972-76 panel, respondents are asked their partisan identification three times: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
+In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations.
+Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question is asked.
+In the 1972-76 ANES panel, respondents were asked their partisan identification three times: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
 If a person participates in all three waves, they have three observations of $|y_{i,t}-y_{i,t-1} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).
 
+Variation in duration between observations within panels lets us separate two kinds of change: change attributable to durable shifts of opinions over time that can be associated with duration between observations and variation that is principally random, or unrelated to the duration between observations.
 If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$.
@@
 For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.
-More than two observations of the same question in a single panel produces variation in duration, which is why we ensure that each included question has at least one panel with three observations.
 
@@
 In the earliest panels, the oldest respondents (65 and over) are substantially less stable opinion holders than younger respondents, while the remaining three age groups (18-25, 26-33, and 34-64) appear to change at approximately equal rates.
-As noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change over time.
+Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable.
+Because, as noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change over time, they become the least stable respondents by the end of the panel.
+
+## Question Differences
+
+There are no questions where members of one group become substantially more stable while another becomes less stable.
 

### 2024-06-27 manuscript/formative_period_kk.Rmd
kkedits627

@@
 Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991].
-In support of this idea, research regularly finds that people in these life stages are more likely to change their preferences, attitudes, and opinions than people in other life stages [@sears2013; @kiley2020].
+In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@sears2013; @kiley2020].
 This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].
 
-However, why people in this life stage appear to show more frequent opinion change relative to older people is unclear.
-Explanations in the social sciences tend to highlight the unique social features of these early life stages -- high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" -- that shape how people are exposed to new sources of information or their ability to update opinions in the face of new information [@alwin1991; @danigelis2007].
+Why people in this life stage appear to show more frequent opinion change relative to older people, however, is unclear.
+Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences,” arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007].
 Other related explanations focus on the unique features of middle age that might stabilize attitudes during this period, including people holding positions with higher levels of social authority during middle age [@eaton2009; @visser2004].
@@
 In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events.
-Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course in more recent cohorts [@buchmann1989; @bruckner2005].
+Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course for more recent cohorts [@buchmann1989; @bruckner2005].
 Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009].
-These broad social changes potentially lengthen of and intensify this stage of the life course in ways that potentially has consequences for the rate at which people undergo cultural changes.
+These broad social changes potentially lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.
 
-Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of age categories and question.
-Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) are more likely to make changes of opinion than older respondents in the most recent panels but not in the earliest panels.
-However, this difference emerged over the second half of 20th Century not because young adults become more likely to make changes, as life-course theories might predict, but because middle-aged and older respondents become less likely to make changes.
-Our findings complicate the narrative about social change and cultural change and raise questions about what social forces emerged to stabilize attitudes more in middle age for more recent cohorts.
+Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of the year they were observed and their age at the time.
+Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) have more variable opinions than older respondents.
+However, these differences are not constant over time.
+They emerged over the second half of 20th Century as middle-aged and older respondents came to report more stable opinions over time, while early adult respondents’ reported opinions continued to change at roughly the same rate over time.
+
+Our findings complicate the narrative about social change and cultural change and raise questions about which social forces emerged to stabilize attitudes in middle age for more recent cohorts.
 
@@
 However, the notion of "impressionable years" is more general than just political preferences.
-The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@elder2016].
+The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but their cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@elder2016].
 Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990].
@@
 
-These models of social imprinting have implications for the kinds of patterns we should observe in a population's attitudes over time and suggest specific forms that within-person opinion change should take over time.
+These aggregate models of social imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time.
 The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course.
 The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a].
-While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality traits [@wagner2019].
+While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [@].
 More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S.
-adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort -- the year of a person's birth - explains much more variance than period -- the year of observation -- for most items in the General Social Survey, suggesting a strong impressionable years pattern.
+adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains much more variance than period, the year of observation, for most items in the General Social Survey, suggesting a strong impressionable years pattern.
 
@@
 
-Why is it research consistently finds the period of time between 14 and 25 to be distinctly "impressionable years?" At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information.
+Why does research consistently finds the period of time between 14 and 25 to be distinctly "impressionable years?" At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information.
 For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change.
@@
 
-While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader cultural environment that might affect attitude stability.
+While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure.
 We briefly consider three here.
 
-First, the number of people pursuing and completing higher education increased significantly over the window studied here.
+First, the number of people pursuing and completing higher education increased significantly over the window studied here [].
 Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007].
@@
 
-Second, we have seen significant expansion of information technology over this window that has dramatically increased access to information about social and political affairs, first through broadcast and cable television and then through the internet and social media.
+Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding.
+Both broadcast and cable television and then through the internet and social media.
 People have greater access to information with which to form opinions.
 
-Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina; @baldassarri].
-People now receive significantly clearer signals from elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020].
+Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina2008; @baldassarri2008].
+People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020].
 This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently.
+Geographic and associational sorting into lifestyle “enclaves” also reinforce these processes, continually providing clearer signals about which opinions people should hold [].
+
+Broadly speaking, these three forces suggest that opinions . Additionally, peoples’ opinions on issues might
 
@@
 
-When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time or to the structural features of the survey administration.
+When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time (our main interest) or to the structural features of the survey administration.
 Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions.
@@
 
-In the one-item, one-panel case (e.g., partisan identification in the 1972-76 panel), we could model the absolute change from wave-to-wave as a function of an intercept and the duration between observations.
+To illustrate our approach, consider a one-item, one-panel case (e.g., partisan identification in the 1972-76 panel).
+We model the absolute change from wave-to-wave as a function of an intercept, the duration between observations in years, age categories, and the interaction between duration and age categories.
 
@@
 Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question is asked.
-In the 1972-76 ANES panel, respondents were asked their partisan identification three times: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
-If a person participates in all three waves, they have three observations of $|y_{i,t}-y_{i,t-1} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).
+In the 1972-76 ANES panel, respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
+If a person participates in all three waves, they have three observations of $|y_{2i}-y_{1i} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).
 
-Variation in duration between observations within panels lets us separate two kinds of change: change attributable to durable shifts of opinions over time that can be associated with duration between observations and variation that is principally random, or unrelated to the duration between observations.
+Variation in duration between observations within panels lets us separate two kinds of change: 1) change attributable to durable shifts of opinions over time that can be associated with duration between observations and 2) change that is principally random, or unrelated to the duration between observations.
 If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$.
 In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise).
-Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$).
+Conversely, $\beta_1$ captures the expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$).
 For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.
 
-Our theoretical question deals primarily with the amount of change over time ($\beta_0$ and $\beta_1$) changes over time.
-To explore that, we can add an additional term for time and interact the duration term with time.
+We can extend equation to incorporate age groups:
 
-$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1d + \beta_2t + \beta_3(d*t) + \epsilon_i $$
+$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(age_{26-33}) + \beta_3(age_{34-64}) + \beta_4(age_{65+}) + \beta_5(d*age_{26-33}) + \beta_6(d*age_{34-64}) + \beta_7(d*age_{65+}) + \epsilon_i $$
 
-In this formulation, $\beta_2$ captures the change in the baseline level of random change over time.
-A negative value of $\beta_2$ would suggest that people are becoming more consistent reporters of their true underlying opinion over time.
-$\beta_3$ captures the expected change in absolute change over time for a one-year change in duration.
-A positive $\beta_3$ value would suggest that people are more likely to undergo a durable or persisting change in their opinion over time in later panels than earlier panels.
+In this framework, $\beta_0$ and $beta_1$ now capture the expected change in responses over time for no time between waves and each additional year, respectively, for respondents ages 18-25 (the reference category).
+$\beta_2$ through $\beta_4$ capture the difference in baseline change between observations for the older groups, and $\beta_5$ through $\beta_7$ capture differences in the effect of an additional year duration for the older age groups.
 
-A challenge for our analysis is that while the panels we explore often have relatively large samples, most questions are only asked in a few panels.
-As such, we lack large numbers of any particular question, meaning the time trend for any specific question is likely to be unstable.
-One potential solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels.
+Our theoretical question deals primarily with whether these short-term rates of change over time themselves change across the broader time span captured across panels.
+To explore that, we can add an additional term for time and interact all previous coefficients outlined above with time.
+
+$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1(d) + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \beta_8(t) + \beta_9(d*t) + \beta_{10}(a_{26-33}*t) + \beta_{11}(a_{34-64}*t) + \beta_{12}(a_{65+}*t) + \beta_{13}(d*a_{26-33}*t) + \beta_{14}(d*a_{34-64}*t) + \beta_{15}(d*a_{65+}*t) + \epsilon_i $$
+
+$\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to time = 0, which we set to be the most recent observation date (2016).
+$\beta_8$ estimates the change in the baseline rate of change for the reference group.
+$\beta_9$ captures change in the effect of duration of observation over time for people in the reference group (ages 18-25).
+Coefficients $\beta_{10}$ through $\beta_{12}$ and $beta_{13}$ capture how differences in the baseline rate of change and duration effects between the reference group and older age categories change over time.
+
+To this point, we have developed our model for the case of a single question.
+However, we lack a large number of observations for most items, meaning separate estimates of the time trends for all these items are likely to be inconclusive.
+One solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels.
 It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable.
 
-Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty.
+Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty. We follow this approach, with the model:
 
-$$ |y_{i,j,t}-y_{i,j,t-1} | = \beta_{0j} + \beta_{1j}d + \beta_{2j}t + \beta_{3}(d*t) + \epsilon_i $$
+$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1(d) + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \beta_8(t) + \beta_9(d*t) + \beta_{10}(a_{26-33}*t) + \beta_{11}(a_{34-64}*t) + \beta_{12}(a_{65+}*t) + \beta_{13}(d*a_{26-33}*t) + \beta_{14}(d*a_{34-64}*t) + \beta_{15}(d*a_{65+}*t) + \epsilon_i $$
 
@@
 
-In this framework, the first-level model models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above.
-These equations are nested at the level of the item.
-Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for time and duration.
+This model allows for each question its own set of coefficient estimates weighted by the amount of information we have about the particular question. Questions asked in many panels
+
+In practice, the model presented in Eq ### is likely too complicated in the face of the data limitations we face.
+
+In this framework, the first-level equation models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above.
+These equations are nested at the level of the survey item.
+Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for duration ($beta_{1j}$), time ($\beta_{2j}$), and their interaction ($\beta_{3j}$).
 These effects are assumed to be normally distributed around a main effect.
-In practice, we see minimal variation in the question-specific coefficients for the interaction between duration and time, and as such do not allow these to vary at the question level.
-
-For all models, we normalize response scales between 0 and 100.
-
-In all models, respondents in the 18-25 age group serve as the reference category.
+In practice, we
 
@@
 For the 2020-22 ANES panel.
+
+## Demographic Variables
+
+To more directly test whether divergences in life course structure drove changes in attitude stability, we include a set of covariates capturing demographic transitions.
+These include 1) whether a respondent is currently enrolled in school; 2) whether a respondent has ever been married; and 3) whether a respondent have ever had a child.
+We also include an indicator for whether a respondent reported completing college.
+
+To distinguish between the influence of life course structure and other factors that might be affecting attitude stability, such as broader social forces, we calculate the expected rate of attitude change in later panels holding constant the proportion of respondents in each group who made certain demographic transitions.
 
@@
 
-Model 1 in Table XXX presents the main effects and variance components of the multilevel model presented above.
-In all models, the 18-25 age group serves as the reference category, and the time variable is centered on the most recent panel (2016-2020), so coefficients reflect expectations for change in recent years, rather than at earlier time points.
+Table XXX presents a comparison of the Bayesian Information Criteria for models allowing for random effects for different coefficients.
+The best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for any higher-order interaction.
+The best-fitting model also omits the third-order interaction between time, age group, and duration of observation.
+In other words, while there is evidence that age groups changed the amount they changed over time, and that the effect of duration changed over time, there is no evidence that groups experienced differnet rates of change in the rate of durable change.
+
+Model 1 in Table XXX presents the main effects and variance components of the best-fitting multilevel model presented above.
+In all models, the 18-25 age group serves as the reference category, and the time variable is centered on the most recent panel (2016-2020), so coefficients reflect expectations for the amount of change observed in the most recent years, rather than at earlier time points.
 
 The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question.
-This is a rather high level of variance, but consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965].
+This is a rather high level of variance, but it is consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965].
 At the same time, random-effect variance around this main intercept term is quite large ($\sigma_{\beta_0} = 5.9220$).
 On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.
+
+Duration between observations is positively associated with wave-to-wave absolute change.
+For every additional year between responses, the expected absolute change for respondents increases 1.1 points.
+As suggested above, there is no evidence to suggest that respondents in different age groups vary systematically in this kind of change.
+Variance around this main effect is relatively small (s.d.
+= 0.36), meaning each one-year duration only increases the amount changed by between .5 and 1.5.
+Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].
 
@@
 
-Duration between observations is positively but weakly associated with wave-to-wave absolute change.
-For every additional year between responses, the expected absolute deviance for respondents between the ages of 18 and 25 increases about .9 points.
-Notably, variance around this main effect is relatively small (s.d.
-= 0.52), meaning each one year duration only increases the amount changed by between .5 and 1.5.
-Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].
-While often called "measurement error," this change conflates both random change due to measurement error and true non-persisting change.
-
-The interactions between age and duration are, for the most part, small and not statistically significant, meaning groups make "real" change at roughly the same rates over time.
 People over 25 make slightly less change than people 18-25 for each additional year of observation, suggesting that the youngest age groups might be making more durable changes of opinion than their older counterparts, but these differences are small and substantively negligible, especially compared to the amount of non-persisting change.
@@
 new.data %>%
+  filter(d == 2) %>%
   mutate(group = paste(age_group, name, sep = "-")) %>%
@@
   geom_hline(yintercept = 0, color = "black") + 
-  #geom_line(alpha = .2, aes(group = group)) + 
+  geom_line(alpha = .2, aes(group = group)) + 
   geom_smooth(linewidth = 2) + 
   theme_bw() + 
-  #facet_wrap(~age_group) + 
+  facet_wrap(~age_group) + 
   labs(x = "Year", y = "Predicted wave-to-wave change",
@@
 Theoretical models that explain the heightened malleability of attitudes during early adulthood by referencing the unique social conditions of that life phase suggest that significant changes in the timing of life course events should produce heightened rates of attitude change during early adulthood (18-25) or should extend rates of attitude updating later into life (ages 26-33).
-As Americans have pushed back marriage, childbirth, and entering the workforce, we should expect these transitions to have some effect on the sources of attitude change, such as meeting new significant others, getting exposed to new kinds of information, or delaying social positions that expect stability.
-We tested that claim by looking at the stability and reliability of political attitude reports among different cohorts as they moved through the same age groups.
+Given that Americans, on average, pushed back marriage, childbirth, and entering the workforce over the course of the 20th Century, we expected these social shifts to have some effect on the sources of attitude change, such as meeting new significant others, getting exposed to new kinds of information, or delaying social positions that expect stability.
+We tested that hypothesis by looking at the stability of political and social attitudes among different cohorts as they moved through the same age groups.
 
-The results from the preceding analyses suggest that the changing life course has had minimal impact on rates of attitude change in the population.
-Despite significant shifts in age of first marriage and age of first child birth, and a much higher proportion of the population attending higher education, attitude behavior as measured by reliability and attitude stability are remarkably similar over time.
-Across panels, we see consistent evidence that ages 18 to 25 have lower attitude stability and potentially lower attitude reliability relative to the rest of the population, a pattern consistent with the "impressionable years" hypothesis that has broad support in the sociological literature.
-However, there is no consistent change over time in either reliability or stability estimates for people in this age group.
-They are different from the rest of the population, but similarly different over time.
-It does not seem to be the case that the emergence of the "Age of Independence" since the middle of the 20th Century has increased rates of attitude change among people in this group.
+The results presented above suggest a more complicated picture than expected.
+Consistent with expectations, the youngest adults look similar to older adults in the earliest panels.
+While members of the youngest age categories (18-25 and 26-33) diverged from older members of the population in the rate at which they changed their opinions, showing more change than older respondents in the most recent panels, this shift was principally driven by an increase in stability among people in middle age and over 65, rather than increased opinion change among the youngest adults.
+We suggested that this aggregate divergence was driven by counter-vailing trends.
+First, consistent with our expectations, delays in life course transitions appear to push the youngest members of the population to make more attitude change over time.
+At the same time, a broad set of social forces worked to stabilize opinions for all members of the population, leading to divergence between the youngest adults and the rest of the population.
+
+The broad pattern of results suggests a broad role for social structures to influence patterns of stable opinion holding.
+At the aggregate level, broad social currents such as increased levels of education, changes in media environments and access to information, and changes in aggregate social sorting appear to affect …
+
+At another level, social forces that structure the life course appear to produce different rates of change for people in different life stages.
 

### 2024-06-27 code/clean_panels.R

@@ -279,7 +279,7 @@ anes7_long <- anes7 %>%
          V720096, V720577, V742236, V763743 #electmatter (1,2,3,5) (1 - good deal, 3 - some, 5 not much) (0,8,9, NA)
          ) %>%
   mutate(weight_2 = "1",
-         weight_3 = as.character(V742003),
+         weight_3 = ifelse(V742003 == 0, "1", as.character(V742003)),
          weight_4 = as.character(V764004),
          weight_5 = weight_4,
          age_1 = V720294, 
@@ -637,7 +637,8 @@ anes8_long <- anes8 %>%
          attentioncpg = (attentioncpg-1)/4*100) %>%
   pivot_longer(c(abortion,attentioncpg, defscale:spendserv)) %>%
   filter(!is.na(age), !is.na(value), !is.na(date)) %>% arrange(id, name, date) %>%
-  mutate(df = "1980 ANES")
+  mutate(df = "1980 ANES") %>%
+  mutate(weight = 1)
 
 # Clean Data 1990-92 ANES ====
 
@@ -2633,7 +2634,6 @@ long_difference <- long_scores %>%
          age_group = ifelse(age > 25 & age < 34, "26-33", age_group),
          age_group = ifelse(age > 65, "65+", age_group))
 
-long_difference %>% group_by(name, df) %>% summarise(meana = mean(abs_diff)) %>% View()
 
 # Save dataset ====
 #save(long_data, file = "~/Dropbox/extended_formative_windows/clean_data/long_data.Rdata")

### 2024-06-27 code/get_dem_variables.R

@@ -6,20 +6,38 @@
 library(tidyverse)
 library(haven)
 
-# Read in the data ====
-anes5 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1956to1960.dta")
-anes7 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1972to1976.dta")
-anes8 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes1980.dta")
-anes90 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1990to1992.dta")
-anes9 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1992to1997.dta")
-anes0 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_2000to2004.dta")
-anes16 <- read_sav("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_2016to2020.sav")
-anes20 <- read_csv("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_2022.csv")
+# Nico Read in the data ====
+# anes5 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1956to1960.dta")
+# anes7 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1972to1976.dta")
+# anes8 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes1980.dta")
+# anes90 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1990to1992.dta")
+# anes9 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_1992to1997.dta")
+# anes0 <- read_dta("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_2000to2004.dta")
+# anes16 <- read_sav("~/Library/CloudStorage/Box-Box/panel_surveys/anes/anes_mergedfile_2016to2020.sav")
[…]

### 2024-06-27 code/analyses.R
kkedits627

@@ -21,7 +21,6 @@ psummaries <- long_difference %>%
   mutate(dec_diff = (as.numeric(difftime(date, max(date), units = "days")))/3652.5) %>%
   mutate(se = sd/sqrt(n))
 
-
 # Summarize absolute difference at the wave-pair level for each age group
 # for each question. 
 summaries <- long_difference %>%
@@ -29,7 +28,8 @@ summaries <- long_difference %>%
   mutate(wave_pair = paste(t1, t2, df, sep = "-")) %>%
   group_by(wave_pair, name, age_group) %>%
   summarise(d = weighted.mean(duration), a = weighted.mean(abs_diff),
-            date = weighted.mean(date), n = n(), sd = sd(abs_diff)) %>%
+            date = weighted.mean(date), n = n(), sd = sd(abs_diff),
+            weighted_n = sum(weight)) %>%
   #Filtering out bad questions
   filter(name != "natpoor", 
          name != "nathome") %>% ungroup() %>%
@@ -54,28 +54,65 @@ m2 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + d*age_group +
            data = summaries %>% filter(sd > 0), weights = 1/se)
 
 m4 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + d*age_group + 
-             d*age_group*dec_diff +
+             d*dec_diff + d*age_group*dec_diff +
              (1 + d + dec_diff + age_group + dec_diff*age_group + d*age_group + 
-                d*age_group*dec_diff|name),
+                d*dec_diff + d*age_group*dec_diff|name),
            data = summaries %>% filter(sd > 0), weights = 1/se)
 
+#Remove third-order interaction and random effects around third-order interaction term
+m4.1 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + d*age_group +
+             d*dec_diff +
+             (1 + d + dec_diff + age_group + dec_diff*age_group + d*age_group + 
+                d*dec_diff|name),
+           data = summaries %>% filter(sd > 0), weights = 1/se)
+#Significant improvement in model fit (>400 decrease in BIC)
 
-m3 <- lmer(a ~ d + dec_diff + 
-             (1 + d + dec_diff |name),
-           data = psummaries %>% filter(sd > 0), weights = 1/se)
+#Remove random effects for d*dec_diff
[…]

### 2024-06-28 manuscript/formative_period_kk.Rmd
drafting, some changes to demo cleaning

@@
 In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@sears2013; @kiley2020].
-This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].
+This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins the dynamics of cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].
 
 Why people in this life stage appear to show more frequent opinion change relative to older people, however, is unclear.
-Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences,” arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007].
-Other related explanations focus on the unique features of middle age that might stabilize attitudes during this period, including people holding positions with higher levels of social authority during middle age [@eaton2009; @visser2004].
+Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences," arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007].
+Other related explanations focus on the social-structural features of middle age that might stabilize attitudes [@eaton2009; @visser2004].
 At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.
 
-While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change during early adulthood, the relative contributions of each are unclear, and clarifying their relative importance is important for understanding the mutually reinforcing processes of social change and cultural evolution.
+While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change during early adulthood, the relative contributions of each are unclear, and clarifying their relative contributions is important for understanding the mutually reinforcing processes of social change and cultural evolution.
 If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change.
@@
 In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events.
-Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course for more recent cohorts [@buchmann1989; @bruckner2005].
-Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009].
-These broad social changes potentially lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.
+Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts [@buchmann1989; @bruckner2005].
+The "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009].
+These broad social changes lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.
 
@@
 However, these differences are not constant over time.
-They emerged over the second half of 20th Century as middle-aged and older respondents came to report more stable opinions over time, while early adult respondents’ reported opinions continued to change at roughly the same rate over time.
+They emerged over the second half of 20th Century as middle-aged and older respondents came to report more stable opinions over time, while early adult respondents' reported opinions continued to change at roughly the same rate over time.
 
 Our findings complicate the narrative about social change and cultural change and raise questions about which social forces emerged to stabilize attitudes in middle age for more recent cohorts.
+Life-course stages can potentially explain the divergence of opinion stability between the youngest age groups and older age groups over time, but they cannot explain the overall trajectory of increased stability.
 
@@
 The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a].
-While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [@].
+While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [\@].
 More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S.
@@
 While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure.
-We briefly consider three here.
+We briefly consider three broader changes here.
 
-First, the number of people pursuing and completing higher education increased significantly over the window studied here [].
+First, the number of people pursuing and completing higher education increased significantly over the window studied here [@schofer2005; @hout2012].
 Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007].
-A population with higher levels of education might be more likely to hold stable opinions regardless of broader social transformations.
+A population with higher levels of education, on average, might be more likely to hold stable opinions regardless of transformations in life course stages.
 
 Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding.
-Both broadcast and cable television and then through the internet and social media.
-People have greater access to information with which to form opinions.
+Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire informaiton that makes opinion-holding easier.
 
@@
 People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020].
-This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently.
-Geographic and associational sorting into lifestyle “enclaves” also reinforce these processes, continually providing clearer signals about which opinions people should hold [].
+This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently [@zaller1992].
+Geographic and social sorting into lifestyle "enclaves" also reinforce these processes, continually providing clearer signals about which opinions people should hold [@dellaposta2015].
 
-Broadly speaking, these three forces suggest that opinions . Additionally, peoples’ opinions on issues might
+Broadly speaking, these three forces suggest that opinions should broadly become more stable over time, regardless of life-course stages.
+Additionally, opinions on certain issues might become more solidified the longer issues remain in the public, as social movements and organizations become solified around them.
 
@@
 
-Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time than members of earlier cohorts.
+Political and Social Attitudes: Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time than members of earlier cohorts.
 Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, with respondents coming from the same general population.
@@
 In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.
-When question wording changed across panels, we harmonize response options to the question structure with the fewest response options.
+When question wording changed across panels, we harmonized response options to the question structure with the fewest response options.
 
 Our search produced a total of 77 questions tapping a broad range of topics including political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.
+For reasons elaborated below, we scaled all items to have a minimum of 0 and a maximum of 100.
 
@@
 
+### Demographic Measures
+
+The central question of our analysis pertains to age, and as such we classify respondents into a set of age brackets depending on their age when they answered survey questions.
+Given our theoretical questions, we classify people into four age categories: 18-25, 26-33, 34-64, and 64-80.
+Because of differences in both question structure and age structure across panels, we drop respondents who are over 80.
+
+We are also interested in the effect of having undergone a set of life-course transitions on attitude stability to more directly test whether divergences in life course structure drove changes in attitude stability.
+To this end, we include a set of indicator variables capturing whether people have undergone certain demographic transitions.
+These include 1) whether a respondent is currently enrolled in school; 2) whether a respondent has ever been married; and 3) whether a respondent has ever had a child.
+We also include an indicator for whether a respondent reported completing college to separately explore whether higher rates of college completion explain changes over time across age groups.
+
 ## Methods
 
-When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time (our main interest) or to the structural features of the survey administration.
-Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions.
+When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time (our main theoretical interest) or to the structural features of the survey administration.
+Specifically, variation in the amount of change people makes in their responses observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions.
 In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
@@
 
-In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations.
-Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question is asked.
+In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations by the same respondent.
+Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question was asked.
 In the 1972-76 ANES panel, respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
@@
 If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$.
-In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise).
+In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change.
 Conversely, $\beta_1$ captures the expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$).
@@
 
-We can extend equation to incorporate age groups:
+Given our focus on age-based differences, we can extend equation to incorporate age groups as indicator variables:
 
-$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(age_{26-33}) + \beta_3(age_{34-64}) + \beta_4(age_{65+}) + \beta_5(d*age_{26-33}) + \beta_6(d*age_{34-64}) + \beta_7(d*age_{65+}) + \epsilon_i $$
+$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \epsilon_i $$
 
@@
 
-$\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to time = 0, which we set to be the most recent observation date (2016).
+$\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to $t = 0$, which we set to be the most recent observation date (2016).
 $\beta_8$ estimates the change in the baseline rate of change for the reference group.
@@
 
-$$ \beta_{0j} = \gamma_0 + \upsilon_{0j} $$
-$$ \beta_{1j} = \gamma_1 + \upsilon_{1j} $$
-$$ \beta_{2j} = \gamma_2 + \upsilon_{2j} $$
-$$ \beta_{3j} = \gamma_3 + \upsilon_{3j} $$
-
-This model allows for each question its own set of coefficient estimates weighted by the amount of information we have about the particular question. Questions asked in many panels
-
-In practice, the model presented in Eq ### is likely too complicated in the face of the data limitations we face.
+$$ \beta_{kj} = \gamma_k + \upsilon_{kj} $$ This model allows for each survey question to have its own set of coefficient estimates weighted by the amount of information we have about the particular question.
+Questions asked in many panels will contribute more to the overall sense of the aggregate trend in opinion stability.
 
@@
 These equations are nested at the level of the survey item.
-Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for duration ($beta_{1j}$), time ($\beta_{2j}$), and their interaction ($\beta_{3j}$).
-These effects are assumed to be normally distributed around a main effect.
-In practice, we
+In practice, the model presented in Eq \### is likely too complicated in the face of the data limitations we face.
+As part of our analysis, we compare the fully elaborated model to more Criterion models using model-selection criteria, specifically the Bayesian Information Criterion.
+For example, it could be the case that little predictive information is gained by allowing questions to have separate slopes for the relationship between age categories and duration over time.
 
-# Weights
+### Demographic Transitions
+
+To distinguish between the influence of life course structure and other factors that might be affecting attitude stability, such as broader social forces, we calculate the expected rate of attitude change in later panels holding constant the proportion of respondents in each group who made certain demographic transitions.
+
+## Weights
 
@@
 For the 2020-22 ANES panel.
-
-## Demographic Variables
-
-To more directly test whether divergences in life course structure drove changes in attitude stability, we include a set of covariates capturing demographic transitions.
-These include 1) whether a respondent is currently enrolled in school; 2) whether a respondent has ever been married; and 3) whether a respondent have ever had a child.
-We also include an indicator for whether a respondent reported completing college.
-
-To distinguish between the influence of life course structure and other factors that might be affecting attitude stability, such as broader social forces, we calculate the expected rate of attitude change in later panels holding constant the proportion of respondents in each group who made certain demographic transitions.
 
@@
 
-Table XXX presents a comparison of the Bayesian Information Criteria for models allowing for random effects for different coefficients.
-The best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for any higher-order interaction.
-The best-fitting model also omits the third-order interaction between time, age group, and duration of observation.
-In other words, while there is evidence that age groups changed the amount they changed over time, and that the effect of duration changed over time, there is no evidence that groups experienced differnet rates of change in the rate of durable change.
+Table XXX presents a comparison of the Akaike Information Criteria (AIC) and Bayesian Information Criteria (BIC) for the full model elaborated above (M1) and a series of more simplified model.
+The best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for any of the higher-order interactions.
+The best-fitting model (M5) omits the third-order interaction between time, age group, and duration of observation and an the interaction between age group and duration.
+Further attempts to simplify the model, including dropping random effects for age categories and duration, resulted in increases in the BIC.
+
+| Variables                      | M1       | M2       | M3       | M4       | M5       |
+|------------|------------|------------|------------|------------|------------|
+| Intercept                      | Y        | Y        | Y        | Y        | Y        |
+| RE                             | Y        | Y        | Y        | Y        | Y        |
+| Time                           | Y        | Y        | Y        | Y        | Y        |
+| RE                             | Y        | Y        | Y        | Y        | Y        |
+| Age Groups                     | Y        | Y        | Y        | Y        | Y        |
+| RE                             | Y        | Y        | Y        | Y        | Y        |
+| Duration                       | Y        | Y        | Y        | Y        | Y        |
+| RE                             | Y        | Y        | Y        | Y        | Y        |
+| Time \* Age Group              | Y        | Y        | Y        | Y        | Y        |
+| RE                             | Y        | Y        | Y        | N        | N        |
+| Time \* Duration               | Y        | Y        | Y        | Y        | Y        |
+| RE                             | Y        | Y        | N        | N        | N        |
+| Age Groups \* Duration         | Y        | Y        | Y        | Y        | N        |
+| RE                             | Y        | Y        | N        | N        | N        |
+| Age Groups \* Duration \* Time | Y        | N        | N        | N        | N        |
+| RE                             | Y        | N        | N        | N        | N        |
+|                                |          |          |          |          |          |
+| BIC                            | 20714.39 | 20250.97 | 19996.61 | 19862.23 | 19845.22 |
+| AIC                            | 19757.88 | 19675.82 | 19627.76 | 19643.42 | 19645.17 |
+
+: Model fit comparison
 
@@
 
+| Col1             | Col2  | Col3 | Col4 |
+|------------------|-------|------|------|
+| Fixed Effects    |       |      |      |
+| Intercept        | 17.45 |      |      |
+| Time (decade)    | -0.26 |      |      |
+| Age(26-33)       | -1.60 |      |      |
+| Age(34-64)       | -3.17 |      |      |
+| Age(65+)         | -4.10 |      |      |
+| Duration (year)  | 1.13  |      |      |
+| Time\*Age(26-33) | -0.09 |      |      |
+| Time\*Age(34-64) | -0.54 |      |      |
+| Time\*Age(65+)   | -1.33 |      |      |
+|                  |       |      |      |
+| Random Effects   | S.D.  |      |      |
+| Intercept        | 5.35  |      |      |
+| Time             | 1.14  |      |      |
+| Duration         | 0.37  |      |      |
+| Age(26-33)       | 0.89  |      |      |
+| Age(34-64)       | 1.81  |      |      |
+| Age(64+)         | 2.92  |      |      |
+
+: Coefficient Estimates
+
 The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question.
@@
 At the same time, random-effect variance around this main intercept term is quite large ($\sigma_{\beta_0} = 5.9220$).
-On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.
+On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time with low levels of measurement error or random change.
 
@@
 As suggested above, there is no evidence to suggest that respondents in different age groups vary systematically in this kind of change.
-Variance around this main effect is relatively small (s.d.
+Random effect variance around this main effect is relatively small (s.d.
 = 0.36), meaning each one-year duration only increases the amount changed by between .5 and 1.5.
@@
 
+Coefficients in Model 1 allow us to test several of the hypotheses outlined above.
 The intensification of early adulthood hypothesis (H3) generally suggests that the youngest respondents should make more change over time in more recent cohorts than their counterparts from earlier cohorts did when they were the same ages.
-The effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significant.
-On average, recent cohorts in this age group make approximately the same amount of wave-to-wave change in recent years as earlier cohorts did, accounting for the kinds of questions they were asked and time between waves.
-However, question-level random effects vary considerably around this main effect, with a standard deviation of about 1.4316.
-This suggests that the overall flat effect obscures quite large question-level differences, with some questions becoming more stable and others becoming less stable.
+The effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significantly different from 0.
+On average, recent cohorts in this age group make approximately the same amount of wave-to-wave change in recent years as earlier cohorts did, accounting for the kinds of questions they were asked and time between responses.
+However, question-level random effects vary considerably around this main effect, with a standard deviation of about 1.43.
+This suggests that the overall flat trend obscures quite large question-level differences, with some questions becoming more stable and others becoming less stable.
 We return to consider these question-specific trends later.
@@
 Next, the extended early adulthood framework (H1) suggests that we should see a closing of the gap between people in the 18-25 and 26-33 age groups over time.
-The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that people in this age group have also not become any more or less likely to make changes of opinion since the earliest panels, and have also not become any more or less similar to the youngest respondents, and that significant changes in the social environment have not made more recent cohorts any more likely to make attitude changes than earlier cohorts were when they were in this age range.
+The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that people in this age group have not become any more or less likely to make changes of opinion since the earliest panels, have not become any more or less similar to the youngest respondents.
 
 However, it is important to understand the full picture of change over time before drawing substantive conclusions.
-Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range, respondents in older age groups do show a notable time trend -- people in this age range become more stable opinion holders over time.
+Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range, respondents in older age groups (34-64 and 65+) do show a notable time trend -- people in these age ranges become more stable opinion holders over time.
 To more clearly illustrate overall trends, Figure XXX plots the predicted time level of change for respondents in each age category over time.
@@
 
-In the earliest panels, the oldest respondents (65 and over) are substantially less stable opinion holders than younger respondents, while the remaining three age groups (18-25, 26-33, and 34-64) appear to change at approximately equal rates.
-Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable.
+In the earliest panels, the oldest respondents (65 and over) are substantially less stable opinion holders than younger respondents, changing by about 5 points more than members of the other groups, while the remaining three age groups (18-25, 26-33, and 34-64) appear to change at approximately equal rates.
+Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable, with the latter becoming almost 10 points more stable on average.
 Because, as noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change over time, they become the least stable respondents by the end of the panel.
+
+## Life Course Shifts
+
+While coefficient estimates for the early adult age groups do not suggest that they have become more variable opinion holders over time, the increase in stability for middle-aged and older respondents suggests that broad social forces might counteract life-course the effect of transitions.
+
+To evaluate the specific effect of life-course changes, we attempt to control for these transitions.
+We re-estimate Model \### above, this time incorporating a series of indicator variables for whether people were students, had children, and were married.
+We also include an indicator variable for whether the respondent completed college to capture the effect of ...
+.
+
+Using these estimates, we calculate the expected value of change had life-course transitions occurred at roughly the same rates.
 
@@
 
-There are no questions where members of one group become substantially more stable while another becomes less stable.
+As noted above, the overall trajectory of increasing attitude stability masks significant heterogeneity across questions.
+To more clearly illustrate these differences in trends, Figure XXX plots the estimated trajectories for each item for respondents in the youngest (18-25) and middle-aged (34-64) age categories.
+Items are ordered by their predicted time trends.
+
+```{r}
+slopes <- as.data.frame(ranef(m4.5)$name) %>%
+  rownames_to_column(var = "name") %>%
+  mutate(slope = dec_diff) %>%
+  select(name, slope)
+
+#Look at questions instead
+new.data %>%
+  filter(d == 1, age_group %in% c("18-25", "34-65")) %>%
+  group_by(name) %>%
+  left_join(slopes) %>% 
+  #left_join(test, by = c("name"="name")) %>%
+  mutate(group = paste(age_group, name, sep = "-")) %>%
+  ggplot(aes(x = dec_diff, y = yhat, color = age_group)) + 
+  #geom_point(data = summaries, aes(y = a)) + 
+  geom_line(alpha = .7, aes(group = group)) + 
+  theme_bw() + 
+  facet_wrap(~reorder(name, slope), nrow = 11, ncol = 7) + 
+  labs(x = "Year", y = "Predicted wave-to-wave change",
+       color = "Age Group") + 
+  theme()
+
+```
+
+First, at a high level, there are very few questions where the directional trend diverges for age groups -- where young people become more stable opinion holders while old people become less stable opinion holders, or vice-versa.
+In general, both groups either become more stable or less stable, just at different rates over time.
+
+It is hard to draw broad conclusions about which items become more stable and which become more variable over time.
+Some questions dealing with social issues have become much more stable, including whether homosexuals should be protected from discrimination, whether men and women should have equal roles in society, and whether the enduring effects of slavery and discrimination continue to hinder Black Americans' social and economic progress.
+A broad set of questions about equality in American have also decreased in within-person change over time.
+Many general political views, such as support for capital punishment, whether the government should help people cover medical costs, and spending on the environment, have also decreased in change over time.
+Some of these shifts reflect the emergence of social consensus, while others reflect clear partisan polarization.
+
+A set of items tapping political efficacy (ppllikeme; manyvote; complicated) show increased change over time.
+Feeling thermometers about social groups, including racial groups, the poor, homosexuals, liberals and conservatives, have either also remained at similar levels of change or seen an increase in variance.
+Policy issues that see increased variance include spending on crime and whether organizations should preferentially hire Black Americans to account for past discrimination.
 
@@
 The broad pattern of results suggests a broad role for social structures to influence patterns of stable opinion holding.
-At the aggregate level, broad social currents such as increased levels of education, changes in media environments and access to information, and changes in aggregate social sorting appear to affect …
+At the aggregate level, broad social currents such as increased levels of education, changes in media environments and access to information, and changes in aggregate social sorting appear to affect ...
 
@@
 
---
-There are theoretically relevant trade-offs from focusing only on the same questions over time.
+|                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 |
+|------------------------------------------------------------------------|
+| There are theoretically relevant trade-offs from focusing only on the same questions over time.
 By focusing only on the same questions over time, we control for question content and structure and ensure that the trend over time is due to something other than changes in survey administration.
 Because our panels span more than five decades that saw dramatic improvements in survey question design, questions included in more recent surveys might simply be of higher quality, with less wave-to-wave change, than items included in older surveys.
-
-We face several limitations in trying to make broader claims.
+|
+| We face several limitations in trying to make broader claims.
 First, many of our age-group estimates of stability and reliability are based on relatively small samples and as a result are not highly reliabile.
@@
 We have attempted to limit this problem by averaging as many items as possible for each panel, but the number of questions in each panel is highly variable.
-
---
+|
 

### 2024-06-28 code/get_dem_variables.R

@@ -118,6 +118,7 @@ anes5_demog <- anes5 %>%
   select(
     id, marital_1:race_3
   ) %>% 
+  mutate_all(as.character) %>% 
   pivot_longer(marital_1:race_3) %>%
   separate(name, into = c("measure", "wave")) %>% 
   pivot_wider(
@@ -584,7 +585,8 @@ anes20_demog <- anes20 %>%
       profile_racethnicity %in% c(3,4) ~ 3
     )
   ) %>% 
-  select(id, marital:race)
+  select(id, marital:race) %>%
+  mutate_all(as.character)
 
 # GSS 2006-2010 demographics ====
 

### 2024-06-28 code/analyses.R
drafting, some changes to demo cleaning

@@ -9,6 +9,25 @@
 
 load("./clean_data/long_difference.Rdata")
 
+source("./code/get_dem_variables.R")
+
+long_demo <- bind_rows(anes5_demog %>% mutate(df = "1956-60 ANES"), 
+                       anes7_demog %>% mutate(df = "1972-76 ANES"), 
+                       anes8_demographic %>% mutate(df = "1980 ANES"),
+                       anes9_demog %>% mutate(df = "1990-92 ANES"), 
+                       anes90_demog %>% mutate(df = "1992-97 ANES"), 
+                       anes0_demog %>% mutate(df = "2000-04 ANES"), 
+                       gss6_demog %>% mutate(df = "2006-10 GSS"), 
+                       gss8_demog %>% mutate(df = "2008-12 GSS"), 
+                       gss10_demog %>% mutate(df = "2010-14 GSS"), 
+                       anes16_demog %>% mutate(df = "2016-20 ANES"), 
+                       anes20_demog %>% mutate(df = "2020-22 ANES"), 
+                       gss20_demog %>% mutate("2016-20 GSS")) %>%
+  mutate(evermarried = ifelse(marital %in% c("married", "other"), 1, 0),
+         ba = ifelse(ed == "ba", 1, 0)) 
+
+
+
 #Panel summaries
 psummaries <- long_difference %>%
   mutate(wave_pair = paste(t1, t2, df, sep = "-")) %>%
@@ -53,6 +72,7 @@ m2 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + d*age_group +
              (1 + d + dec_diff + age_group + dec_diff*age_group + d*age_group|name),
            data = summaries %>% filter(sd > 0), weights = 1/se)
 
+#Fully elaborated model
 m4 <- lmer(a ~ d + dec_diff + age_group + dec_diff*age_group + d*age_group + 
              d*dec_diff + d*age_group*dec_diff +
              (1 + d + dec_diff + age_group + dec_diff*age_group + d*age_group + 
@@ -146,21 +166,27 @@ new.data %>%
        subtitle = "Individual question trajectories and overall trajectory") + 
   scale_color_brewer(type = "qual", palette = 2) 
 
+
+slopes <- as.data.frame(ranef(m4.5)$name) %>%
+  rownames_to_column(var = "name") %>%
+  mutate(slope = dec_diff) %>%
[…]

### 2024-07-01 code/get_dem_variables.R
Added student status to all DSs where available

@@ -63,7 +63,9 @@ anes5_demog <- anes5 %>%
     #race
     V560172,
     V580469,
-    V600685
+    V600685,
+    #student status 
+    V560120
   ) %>%
   mutate(
     marital_1 = V560177,
@@ -113,13 +115,14 @@ anes5_demog <- anes5 %>%
     race_1 = as.character(V560172),
     race_2 = as.character(V580469),
     race_3 = as.character(V600685),
-    
+    # Student status 
+    student_1 = as.character(ifelse(V560120 %in% c(92), 1, 0))
   )  %>% 
   select(
-    id, marital_1:race_3
+    id, marital_1:student_1
   ) %>% 
   mutate_all(as.character) %>% 
-  pivot_longer(marital_1:race_3) %>%
+  pivot_longer(marital_1:student_1) %>%
   separate(name, into = c("measure", "wave")) %>% 
   pivot_wider(
     names_from = measure, 
@@ -191,11 +194,12 @@ anes7_demog <- anes7 %>%
            V763513 == 1 ~ 1, 
            V763513 == 2 ~ 2, 
            V763513 %in% c(3,4,5,6,7) ~ 3, 
-           V763513 == 9 ~ NA_real_)) %>% 
-  select(id, oldestkid_2:race_3) %>% 
+           V763513 == 9 ~ NA_real_), 
+         student_1 = as.character(ifelse(V720306 == 8, 1, 0))) %>% 
+  select(id, oldestkid_2:student_1) %>% 
   select(-c(anychild_1a)) %>% 
   mutate_all(as.character) %>% 
-  pivot_longer(oldestkid_2:race_3) %>%
+  pivot_longer(oldestkid_2:student_1) %>%
   separate(name, into = c("measure", "wave")) %>% 
   pivot_wider(
     names_from = measure, 
@@ -214,10 +218,10 @@ anes8_demographic <- anes8 %>%
            VMP0326 == 9 ~ NA_character_
          ), 
          marital_2 = case_when(
-            VMP2377 %in% c(1,7) ~ "married", 
-            VMP2377 == 2 ~ "single/nm", 
-            VMP2377 %in% c(3,4,5) ~ "other", 
-            VMP2377 == 9 ~ NA_character_
+           VMP2377 %in% c(1,7) ~ "married", 
+           VMP2377 == 2 ~ "single/nm", 
+           VMP2377 %in% c(3,4,5) ~ "other", 
[…]

### 2024-07-02 manuscript/formative_period_kk.Rmd
clean demo 70, 90, 92

@@
 We also include an indicator for whether a respondent reported completing college to separately explore whether higher rates of college completion explain changes over time across age groups.
+
+For the 1990-92 panel, respondents did not re-report their employment status, marital status, educational attainment, and parental status for the 1991 wave.
+Respondents who participated in that wave were coded as according to their response in the 1990 wave for these items.
+
+Children:
+
+Broadly speaking, questions tapping marital status tend to capture the same .
+In contrast, questions tapping the number of children a respondent has vary significantly across panels and even within panels.
+Specifically, questions occasionally tap how many children are currently living with the respondent and occasionally tap how many children the respondent has ever had.
+While these items are highly correlated for respondents with young children, these two approaches to the question produce significant differences in rates among parents whose children have left home.
+They also produce differences among respondents who live in
+
+1990 pre: Number of children in house
+1991: copied from 1990
+1992: any children?
+1992 post: copies from 1992pre
+1993: copied from 1992 pre
+1994: any children ever
+96: Any children under 18 in household
 

### 2024-07-02 code/get_dem_variables.R

@@ -133,10 +133,10 @@ anes5_demog <- anes5 %>%
 anes7_demog <- anes7 %>% 
   mutate(id = 1:nrow(.), 
          across(c(V742410:V742417,V763373:V763378), ~ifelse(.x %in% c(0,98,99), NA, .x)),
-         oldestkid_2 = pmax(V742410,V742411,V742412,
+         oldestkid_3 = pmax(V742410,V742411,V742412,
                             V742413,V742414,V742415,
                             V742416,V742417, na.rm = TRUE),
-         oldestkid_3 = pmax(V763373,V763374,V763375,
+         oldestkid_4 = pmax(V763373,V763374,V763375,
                             V763376,V763377,V763378, na.rm = TRUE), 
          #resp's education
          ed_1 = ifelse(V720300 %in% c(98,99), NA, V720300),
@@ -144,67 +144,91 @@ anes7_demog <- anes7 %>%
                                       41,42,43), "less than", ed_1),
          ed_1 = ifelse(V720300 %in% c(51,61,71,50), "hs", ed_1),
          ed_1 = ifelse(V720300 %in% c(81,82,83,84,85,86), "ba", ed_1),
-         ed_2 = ifelse(V742423 %in% c(98,99), NA, V742423),
-         ed_2 = ifelse(V742423 %in% c(0,1,2,3,4), "less than", ed_2),
-         ed_2 = ifelse(V742423 %in% c(5,6,7,8), "hs", ed_2),
-         ed_2 = ifelse(V742423 %in% c(9,10), "ba", ed_2),
-         ed_3 = ifelse(V763389 %in% c(98, 99), NA, V763389),
-         ed_3 = ifelse(V763389 %in% c(0,1,2,3,4), "less than", ed_3),
-         ed_3 = ifelse(V763389 %in% c(5,6,7,8), "hs", ed_3),
-         ed_3 = ifelse(V763389 %in% c(9,10), "ba", ed_3),
+         ed_2 = ed_1,
+         ed_3 = ifelse(V742423 %in% c(98,99), NA, V742423),
+         ed_3 = ifelse(V742423 %in% c(0,1,2,3,4), "less than", ed_3),
+         ed_3 = ifelse(V742423 %in% c(5,6,7,8), "hs", ed_3),
+         ed_3 = ifelse(V742423 %in% c(9,10), "ba", ed_3),
+         ed_4 = ifelse(V763389 %in% c(98, 99), NA, V763389),
[…]

### 2024-07-02 code/get_dem_variables.R
clean demog 00, 16 anes

@@ -563,12 +563,14 @@ anes0_demog <- anes0 %>%
       M000909 %in% c(2,3,4,6) ~ "other", 
       M000909 %in% c(8,9) ~ NA_character_
     ), 
+    marital_2 = marital_1,
     marital_3 = case_when(
       M023127A == 1 ~ "married", 
       M023127A == 5 ~ "single/nm", 
       M023127A %in% c(2,3,4,6) ~ "other", 
       M023127A %in% c(8,9) ~ NA_character_
     ), 
+    marital_4 = marital_3,
     marital_5 = case_when(
       M045176 == 1 ~ "married", 
       M045176 == 5 ~ "single/nm", 
@@ -585,21 +587,52 @@ anes0_demog <- anes0 %>%
       M001024 > 15, NA_real_, 
       M001024
     ), 
+    anychild_2 = anychild_1,
+    anychild_3 = anychild_2,
+    anychild_4 = anychild_3,
+    anychild_5 = anychild_4,
     # Education 
-    educ_2 = case_when(
+    educ_1 = case_when(
+      M000913 %in% c(1,2) ~ "less than",
+      M000913 %in% c(3,4,5) ~ "hs",
+      M000913 %in% c(6,7) ~ "ba",
+      M000913 %in% c(0,9) ~ NA_character_
+    ),
+    educ_2 = educ_1,
+    educ_3 = case_when(
       M023131 %in% c(1,2) ~ "less than", 
       M023131 %in% c(3,4,5) ~ "hs", 
       M023131 %in% c(6,7) ~ "ba", 
       M023131 %in% c(0,9) ~ NA_character_
     ), 
+    educ_4 = educ_3,
+    educ_5 = educ_4,
     # Sex 
-    sex_2 = case_when(
+    sex_1 = M001029,
+    sex_2 = sex_1,
+    sex_3 = case_when(
       M023153 == 1 ~ 1, 
       M023153 == 2 ~ 2, 
       M023153 %in% c(0,9) ~ NA_real_
     ), 
+    sex_4 = sex_3,
+    sex_5 = sex_4,
     # Race 
+    race_1 = case_when(
+      M001006a == 50 ~ 1,
+      M001006a == 10 ~ 2,
+      M001006a %in% c(20,30,40,60,
+                      75,76,79,80,90) ~ 3
+    ),
+    race_2 = race_1,
     race_3 = case_when(
+      M023150 %in% c(1,12,13,14,15) ~ 2,
[…]

### 2024-07-03 code/get_dem_variables.R
small updates

@@ -113,9 +113,9 @@ anes5_demog <- anes5 %>%
     ed_3 = ifelse(V600694 %in% c(81, 82), "ba", ed_3),
     ed_4 = ed_3,
     #1 = male, 2 = female
-    sex_1 = as.character(V560171),
-    sex_2 = as.character(V580468),
-    sex_3 = as.character(V600684),
+    sex_1 = as.character(ifelse(V560171 == 9, NA, V560171)),
+    sex_2 = as.character(ifelse(V580468 == 9, NA, V580468)),
+    sex_3 = as.character(ifelse(V600684 == 9, NA, V600684)),
     sex_4 = sex_3,
     #1 = white, 2 = negro, 3 = other.
     race_1 = as.character(V560172),
@@ -137,7 +137,8 @@ anes5_demog <- anes5 %>%
   pivot_wider(
     names_from = measure, 
     values_from = value
-  )
+  ) %>%
+  mutate(df = "1956-60 ANES")
 
 # ANES 1970 demographics ====
 anes7_demog <- anes7 %>% 
@@ -246,7 +247,8 @@ anes7_demog <- anes7 %>%
     names_from = measure, 
     values_from = value
   ) %>%
-  arrange(id, wave)
+  arrange(id, wave)%>%
+  mutate(df = "1972-76 ANES")
 
 # ANES 1980 demographics ====
 
@@ -273,15 +275,15 @@ anes8_demog <- anes8 %>%
          ), 
          marital_4 = marital_3,
          # Education Summary 
-         educ_1 = case_when(
+         ed_1 = case_when(
            VMP0334 %in% c(1,2,3,4) ~ "less than",
            VMP0334 %in% c(5,6,7,8) ~ "hs", 
            VMP0334 %in% c(9,10) ~ "ba", 
            VMP0334 %in% c(98,99) ~ NA_character_
          ), 
-         educ_2 = educ_1,
-         educ_3 = educ_2,
-         educ_4 = educ_3,
+         ed_2 = ed_1,
+         ed_3 = ed_2,
+         ed_4 = ed_3,
          # Children -- sum total amount of children 
          across(c(VMP0589,VMP0590,VMP0591,VMP0592,
                   VMP2435,VMP2436,VMP2437,VMP2438,
@@ -337,7 +339,8 @@ anes8_demog <- anes8 %>%
   pivot_wider(
     names_from = measure, 
[…]

### 2024-07-03 code/exploring_demogs.R
demog imputation

@@ -1,0 +1,49 @@
+
+
+
+long_demog <- bind_rows(anes5_demog,
+          anes7_demog,
+          anes8_demog,
+          anes90_demog,
+          anes9_demog,
+          anes0_demog,
+          anes16_demog,
+          anes20_demog,
+          gss6_demog,
+          gss8_demog,
+          gss10_demog,
+          gss20_demog) %>%
+  mutate(evermarried = recode(marital, "married"=1, "other"=1, "single/nm"=0),
+         female = recode(sex, "1"=0, "2"=1),
+         ba = recode(ed, "hs"=0, "less than"=0, "ba"=1),
+         everkid = as.numeric(everkid),
+         kidinhouse = as.numeric(kidinhouse),
+         student = as.numeric(student),
+         id = as.numeric(id)) %>%
+  select(df, id, wave, evermarried, student, everkid, kidinhouse, ba)
+
+
+
+test <- left_join(long_scores, long_demog,
+          by = c("df"="df", "id"="id", "wave"="wave"))
+
+one_obs <- test %>%
+  group_by(df, id, wave) %>%
+  slice(1) %>%
+  mutate(age_group = ifelse(age < 26, "18-25", ifelse(age > 25 & age < 34, "26-33",
+                                                      ifelse(age > 33 & age < 65, "33-65", 
+                                                             "65+"))))
+
+one_obs %>%
+  select(-c(name, value)) %>%
+  pivot_longer(evermarried:ba) %>%
+  group_by(df, wave, name, age_group) %>% 
+  summarise(mean_value = weighted.mean(value,na.rm=TRUE),
+            date = mean(date),
+            n = n(),
+            nmiss = sum(is.na(value)))%>% #View()
+  ggplot(aes(x = date, y = mean_value, color = age_group)) +
+  facet_wrap(~name) + 
+  geom_point() + 
+  geom_line() + 
+  theme_bw()

### 2024-07-03 code/get_dem_variables.R
demog imputation

@@ -87,13 +87,13 @@ anes5_demog <- anes5 %>%
     marital_3 = ifelse(V600690 %in% c(3, 4, 5), "other", marital_3),
     marital_4 = marital_3,
     #IS there a child under 18 in the household.
-    anychild_1 = as.character(ifelse(V560178 == 99, NA, V560178)),
-    anychild_1 = as.character(ifelse(V560178 == 0, 0, 1)),
-    anychild_2 = as.character(ifelse(V580475 == 99, NA, V580475)),
-    anychild_2 = as.character(ifelse(V580475 == 0, 0, 1)),
-    anychild_3 = as.character(ifelse(V600691 == 99, NA, V600691)),
-    anychild_3 = as.character(ifelse(V600691 == 0, 0, 1)),
-    anychild_4 = anychild_3,
+    kidinhouse_1 = as.character(ifelse(V560178 == 99, NA, V560178)),
+    kidinhouse_1 = as.character(ifelse(V560178 == 0, 0, 1)),
+    kidinhouse_2 = as.character(ifelse(V580475 == 99, NA, V580475)),
+    kidinhouse_2 = as.character(ifelse(V580475 == 0, 0, 1)),
+    kidinhouse_3 = as.character(ifelse(V600691 == 99, NA, V600691)),
+    kidinhouse_3 = as.character(ifelse(V600691 == 0, 0, 1)),
+    kidinhouse_4 = kidinhouse_3,
     #The third education variable is differently coded
     ed_1 = ifelse(V560181 == 9, NA, V560181),
     ed_1 = ifelse(V560181 %in% c(0, 1, 2, 3, 4), "less than", ed_1),
@@ -118,9 +118,9 @@ anes5_demog <- anes5 %>%
     sex_3 = as.character(ifelse(V600684 == 9, NA, V600684)),
     sex_4 = sex_3,
     #1 = white, 2 = negro, 3 = other.
-    race_1 = as.character(V560172),
-    race_2 = as.character(V580469),
-    race_3 = as.character(V600685),
+    race_1 = ifelse(V560172 == 7, 3, ifelse(V560172 == 9, NA, V560172)),
+    race_2 = ifelse(V580469 == 7, 3, ifelse(V580469 == 9, NA, V580469)),
+    race_3 = ifelse(V600685 == 7, 3, ifelse(V600685 == 9, NA, V600685)),
     race_4 = race_3,
     # Student status 
     student_1 = as.character(ifelse(V560120 == 92, 1, 0)),
[…]

### 2024-07-03 manuscript/formative_period_kk.Rmd
drafting

--- 2024-06-28 manuscript/formative_period_kk.Rmd
@@
 
-While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure.
+While we have parimarily focused on changes in the timing and structure of life course stages over the past 70 years, that time frame has also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure.
 We briefly consider three broader changes here.
@@
 Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007].
-A population with higher levels of education, on average, might be more likely to hold stable opinions regardless of transformations in life course stages.
+The specific mechanism that links educational attainment and attitude stability is unclear, with some perspectives suggesting educational attainment improves cognitive function and memory while other perspectives suggest that educational attainment makes people pay more attention to current events.
+It could also be that education cultivates in people an expectation that they hold opinions in the first place.
+Regardless of the specific mechanism, a population with higher levels of education, on average, might be more likely to hold stable opinions regardless of transformations in life course stages.
 
 Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding.
-Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire informaiton that makes opinion-holding easier.
+Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire information that makes opinion-holding easier.
 
@@
 People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020].
+While clarity aroud political issue positions is most commonly studied, the same forces can also align other social beliefs and identities.
 This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently [@zaller1992].
@@
 For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points.
-Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same set of questions, we did not treat these as separate panels for the purposes of generating three distinct time points.
+Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same set of questions, we did not treat these as separate panels for the purposes of identifying three distinct time points.
 In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.
-When question wording changed across panels, we harmonized response options to the question structure with the fewest response options.
+We allowed for question wording to vary slightly across panels (e.g., feeling thermometers about "blacks" instead of "negros").
+When the same or similar questions had slightly different response scales, we harmonized response options to the question structure with the fewest response options.
+This was most common when one panel asked whether people agreed or disagreed while another included options for "strong" agreement and disagreement.
+We broadly followed the ANES time series' lead in which questions to harmonize.
 
@@
 Table A1 in Appendix A outlines the questions included in our analysis, including their full question wording and response scale.
-For a small number of items where question wording remains consistent across panels but response scales differ, we harmonize by converting more fine-grained response scales to the more general scales, following the ANES time series.
 
@@
 We are also interested in the effect of having undergone a set of life-course transitions on attitude stability to more directly test whether divergences in life course structure drove changes in attitude stability.
-To this end, we include a set of indicator variables capturing whether people have undergone certain demographic transitions.
+To this end, we include a set of indicator variables capturing whether the respondent had undergone certain demographic transitions.
 These include 1) whether a respondent is currently enrolled in school; 2) whether a respondent has ever been married; and 3) whether a respondent has ever had a child.
 We also include an indicator for whether a respondent reported completing college to separately explore whether higher rates of college completion explain changes over time across age groups.
+
+It was quite common for respondents to have missing data on demographic variables for a particular wave of any panel, even if they participated in the panel.
+This was most common when a question occurred in the post-election wave of the ANES, but demographic variables were captured in the pre-election wave.
+Similarly, some demographic items were only asked in one wave of a panel.
+When respondents were missing on a demographic item, we imputed the missing variable using first earlier responses and then, if there were no earlier responses, later responses.
+
+Broadly speaking, questions tapping marital status, student status, and degree attainment tended to be asked in the same way with the same response options or could be easily harmonized.
+In contrast, generating a variable for whether a respondent had children was more challenging, as what information was being sought varied significantly across panels and even within panels.
+Specifically, questions occasionally tap how many children were currently living with the respondent and at other times tapped whether the respondent had any children or how many children the respondent had ever had.
+While these items are highly correlated for respondents with young children, these two approaches to the question produce significant differences in rates among parents whose children have left home.
+
+To generate a measure of whether respondents had children, we ...
+
+1990 pre: Number of children in house
+1991: copied from 1990
+1992: any children?
+1992 post: copies from 1992pre
+1993: copied from 1992 pre
+1994: any children ever
+96: Any children under 18 in household
 
@@
 These equations are nested at the level of the survey item.
-In practice, the model presented in Eq \### is likely too complicated in the face of the data limitations we face.
-As part of our analysis, we compare the fully elaborated model to more Criterion models using model-selection criteria, specifically the Bayesian Information Criterion.
+In practice, the model presented in Eq \### is likely too complicated given our data limitations, and might not be necessary.
+As part of our analysis, we compare the fully elaborated model to more parsimonious models using model-selection criteria, specifically the Bayesian information criterion (BIC) and Akaike information criterion [@raftery1995].
 For example, it could be the case that little predictive information is gained by allowing questions to have separate slopes for the relationship between age categories and duration over time.
@@
 
+Once we select the most parsimonious model to capture attitude stability over time, we then include variables capturing demographic transitions to evaluate their relationship to attitude stability over time.
 To distinguish between the influence of life course structure and other factors that might be affecting attitude stability, such as broader social forces, we calculate the expected rate of attitude change in later panels holding constant the proportion of respondents in each group who made certain demographic transitions.
 
-## Weights
+For each wave pair, we include measures of demographic status at the first observation.
+For example, if a respondent is unmarried at the first wave and gets married by the second, they are counted as "never married" for the purposes of our analysis.
+If they are observed a third time, they would be counted as "married" for that pair of waves.
 
-We focus on using weights that accounts for sampling design.
+### Weights
 
-For the 2016-20 ANES panel, almost all questions are asked in either the pre- or post-election version of both surveys.
-The panel includes weights for .
-The single item that was asked in the pre- and post-election waves of the 2016 panel (polviews) is weighted using the pre- election weights.
+Finally, a key issue in drawing valid inferences about the U.S.
+population over time is to ensure that samples generalize to the same theoretical population -- the U.S.
+adult non-institutionalized population.
+However, none of the panels used in our analysis were simple random samples of the adult population.
+All employ different complex sampling strategies.
 
-For the 2020-22 ANES panel.
+To account for this, we attempt to use weights that account for these sampling strategies as well as, when available, non-response.
 
@@
 
-The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question.
+The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (about 18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question.
 This is a rather high level of variance, but it is consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965].
@@
 For every additional year between responses, the expected absolute change for respondents increases 1.1 points.
-As suggested above, there is no evidence to suggest that respondents in different age groups vary systematically in this kind of change.
+As suggested above by the model comparison, there is no evidence to suggest that respondents in different age groups vary systematically in this kind of change.
 Random effect variance around this main effect is relatively small (s.d.
-= 0.36), meaning each one-year duration only increases the amount changed by between .5 and 1.5.
+= 0.36), meaning each one-year duration only increases the amount changed by between .5 and 1.5 for most questions.
 Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].
@@
 The coefficients for age groups other than the reference category suggest that, on average, other age groups are more stable than people in the 18-25 age range in the most recent panels.
-Respondents in the 26-33 age group are about 1.1 points more stable on average, respondents in the 34-65 age group are about 2.2 points more stable on average, and respondents in the 65 and over age group are about 3.5 points more stable on average, than respondents in the 18 to 25 age group.
+Respondents in the 26-33 age group are about 1.1 points more stable on average, respondents in the 34-65 age group are about 2.2 points more stable on average, and respondents in the 65 and over age group are about 3.5 points more stable on average, than respondents in the 18-25 age group.
 This provides broad evidence for the "impressionable years" and "increasing persistence" hypotheses, as the youngest age groups are less stable than others, followed by the 26-33 age group.
 
-People over 25 make slightly less change than people 18-25 for each additional year of observation, suggesting that the youngest age groups might be making more durable changes of opinion than their older counterparts, but these differences are small and substantively negligible, especially compared to the amount of non-persisting change.
-
 Coefficients in Model 1 allow us to test several of the hypotheses outlined above.
-The intensification of early adulthood hypothesis (H3) generally suggests that the youngest respondents should make more change over time in more recent cohorts than their counterparts from earlier cohorts did when they were the same ages.
+The intensification of early adulthood hypothesis (H3) generally suggests that the most recent cohorts should make the most change when they are in the youngest age range than cohorts that came before them.
 The effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significantly different from 0.
@@
 Next, the extended early adulthood framework (H1) suggests that we should see a closing of the gap between people in the 18-25 and 26-33 age groups over time.
-The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that people in this age group have not become any more or less likely to make changes of opinion since the earliest panels, have not become any more or less similar to the youngest respondents.
+The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that while people in this age group are more stable, on average, than then youngest respondents, more recent cohorts are not any more or less likely to make changes of opinion in this age range than earlier cohorts, and more recent cohorts are not any more or less similar to younger adults than earlier cohorts.
 
-However, it is important to understand the full picture of change over time before drawing substantive conclusions.
-Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range, respondents in older age groups (34-64 and 65+) do show a notable time trend -- people in these age ranges become more stable opinion holders over time.
+However, before drawing broader conclusions about the lack of importance of life course shifts, it is important to understand the full picture of change over time.
+Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range whose rate of opinion change remains flat across cohorts, respondents in older age groups (34-64 and 65+) do show a notable time trend -- more recent cohorts are more stable opinion reports in these age ranges than earlier cohorts.
 To more clearly illustrate overall trends, Figure XXX plots the predicted time level of change for respondents in each age category over time.
@@
 To evaluate the specific effect of life-course changes, we attempt to control for these transitions.
-We re-estimate Model \### above, this time incorporating a series of indicator variables for whether people were students, had children, and were married.
+We re-estimate Model \### above, this time incorporating a series of indicator variables for whether respondents were students, whether they had children, and whether they were married.
 We also include an indicator variable for whether the respondent completed college to capture the effect of ...
@@
 Using these estimates, we calculate the expected value of change had life-course transitions occurred at roughly the same rates.
+
+```{r}
+new.data %>%
+  filter(d == 2) %>%
+  mutate(time = recode(time, "1"="Counterfactual", "t"="Observed")) %>%
+  mutate(time = ifelse(dec_diff == -6.4, "Start", time)) %>%
+  group_by(age_group, dec_diff, time) %>%
+  summarise(yhat = mean(yhat)) %>%
+  ungroup() %>%
+  select(-dec_diff) %>%
+  spread(time, yhat) %>%
+  pivot_longer(Counterfactual:Observed) %>%
+  mutate(Start.Time = -6.4, Finish.Time = 0) %>%
+  ggplot(aes(x = Start.Time, xend = Finish.Time, y = Start, yend = value,
+             linetype = name, color = age_group)) +
+  geom_segment() + 
+  geom_line(aes(color = age_group)) + 
+  #geom_point(size = 3) +
+  theme_bw() + 
+  scale_shape_manual(values = c(21,22)) + 
+  scale_fill_brewer(type = "qual", palette = 2,
+                     guide = "none") + 
+  scale_color_brewer(type = "qual", palette = 2,
+                    guide = "none") + 
+  labs(x = "Age Group", y = "Expected change for 2-year duration",
+       shape = "Estimate",
+       title = "Counterfactual expectation of change",
+       subtitle = "Holding demographic variables constant at 1956 levels") + 
+  facet_wrap(~age_group)
+
+```
+
+Assuming that rates of marriage, student status, and degree attainment remained constant at their 1956 level over time for each age group, age groups would be expected to display more similar rates of change in the most recent panels than they actually do.
+In particular, the youngest respondents would display lower rates of change over time and other age groups would display higher rates of change over time.
+In fact, consistent with the extended early adulthood framework, the youngest respondents would be slightly less consistent opinion holders in more recent years than they actually were.
+
+This divergence between observed and counterfactual trajectories is principally driven by changes in college-going, rather than delays in marriage.
+The dramatic increase in college-going for all age groups produces more stable attitudes over time.
+However, because young people have completed college at lower rates than older respondents in the most recent panels (likely because many of them are still enrolled), they do not benefit as much from this shift (though being currently enrolled as a student also increases attitude stability).
+Young respondents' dramatically lower rates of marriage in the most recent panels counteracts the increased stability that comes from higher levels of educational attainment.
 
@@
 Theoretical models that explain the heightened malleability of attitudes during early adulthood by referencing the unique social conditions of that life phase suggest that significant changes in the timing of life course events should produce heightened rates of attitude change during early adulthood (18-25) or should extend rates of attitude updating later into life (ages 26-33).
-Given that Americans, on average, pushed back marriage, childbirth, and entering the workforce over the course of the 20th Century, we expected these social shifts to have some effect on the sources of attitude change, such as meeting new significant others, getting exposed to new kinds of information, or delaying social positions that expect stability.
+Given that Americans, on average, pushed back marriage, childbirth, and entering the workforce over the course of the second half of the 20th Century and the early 21st Century, we posited that these shifts should, through various pathways, increase over-time variance in survey responses among the youngest members of the population.
 We tested that hypothesis by looking at the stability of political and social attitudes among different cohorts as they moved through the same age groups.
 
-The results presented above suggest a more complicated picture than expected.
-Consistent with expectations, the youngest adults look similar to older adults in the earliest panels.
-While members of the youngest age categories (18-25 and 26-33) diverged from older members of the population in the rate at which they changed their opinions, showing more change than older respondents in the most recent panels, this shift was principally driven by an increase in stability among people in middle age and over 65, rather than increased opinion change among the youngest adults.
-We suggested that this aggregate divergence was driven by counter-vailing trends.
-First, consistent with our expectations, delays in life course transitions appear to push the youngest members of the population to make more attitude change over time.
-At the same time, a broad set of social forces worked to stabilize opinions for all members of the population, leading to divergence between the youngest adults and the rest of the population.
+The results presented above suggest a more complicated picture than anticipated.
+Consistent with expectations, the youngest adults look similar to older adults in the earliest panels in terms of both demographics and attitude stability.
+While members of the youngest age categories (18-25 and 26-33) diverged over time from older members of the population in the rate at which they changed their opinions, showing more change than older respondents in the most recent panels, this shift was principally driven by an increase in stability among people in middle age and over 65, rather than increased opinion change among the youngest adults.
+We suggested that this aggregate divergence was driven by countervailing trends.
+On one hand, delays in life course transitions push the youngest members of the population to make more attitude change over time.
+Second, broader social forces worked to stabilize opinions for all members of the population, leading to divergence between the youngest adults and the rest of the population.
 
-The broad pattern of results suggests a broad role for social structures to influence patterns of stable opinion holding.
-At the aggregate level, broad social currents such as increased levels of education, changes in media environments and access to information, and changes in aggregate social sorting appear to affect ...
+We tested this proposition using data on life-course transitions and measures of educational attainment.
+We found that, consistent with our revised perspective of countervailing forces, age groups' rates of attitude change over time would have been more similar if these social changes did not happen.
+Specifically, increased rates of educational attainment account for much of the stabilizing shifts among older Americans, while delayed rates of marriage and childbearing decreased attitude stability among the youngest members of the population.
 
-At another level, social forces that structure the life course appear to produce different rates of change for people in different life stages.
+Our results suggest that broad social conditions play a substantial role in shaping the rate at which people change their attitudes.
+Since the 1950s, Americans have become meaningfully more stable opinion holders, in part because they have attained higher levels of education.
+Our approach likely under-estimates the importance of this shift, as we focus principally on rising rates of college attainment.
+However, in the earliest panels, a substantial proportion (> 70 percent) of the oldest respondents and more than 50 percent of middle aged respondents lacked a high school degree.
 
-Similarly, as life course transitions have been pushed back since the middle of the 20th Century, we expected to see decreased reliability and stability extend into the next age group (ages 26-33).
-However we see no clear evidence of changes in either stability or reliability.
-Attitude behavior of people in this group is consistently closer to that of people in the next oldest group (34 to 41) than to people the 18 to 25 age group.
-While social life for people in this group increasingly resembles the younger group, their attitude behavior tends to resemble older groups.
+Changes in the timing of life-course events were less important than changes in education for explaining, but they do play a role.
+Our findings suggest that, in the absence of these delays, young people would make attitude changes at a comparable rate to that of older people.
 
-It does not appear that the lack of change is attributable to changes in questions over time.
-We observe the same patterns in stability and reliability of attitudes whether we focus only on questions that are replicated across panels .
+At the same time, it is important not to overstate the importance of these broader social transitions.
+Despite dramatic shifts in educational attainment, access to information, and political organization, people make changes in many social and political attitudes at the same rates they did 70 years ago.
 
@@
 
-|                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 |
-|------------------------------------------------------------------------|
-| There are theoretically relevant trade-offs from focusing only on the same questions over time.
+There are theoretically relevant trade-offs from focusing only on the same questions over time.
 By focusing only on the same questions over time, we control for question content and structure and ensure that the trend over time is due to something other than changes in survey administration.
 Because our panels span more than five decades that saw dramatic improvements in survey question design, questions included in more recent surveys might simply be of higher quality, with less wave-to-wave change, than items included in older surveys.
-|
-| We face several limitations in trying to make broader claims.
+
+We face several limitations in trying to make broader claims.
 First, many of our age-group estimates of stability and reliability are based on relatively small samples and as a result are not highly reliabile.
@@
 We have attempted to limit this problem by averaging as many items as possible for each panel, but the number of questions in each panel is highly variable.
-|
 

### 2024-07-03 code/analyses.R

@@ -8,24 +8,11 @@
 #Load in long data frame from clean_data.R file
 
 load("./clean_data/long_difference.Rdata")
+load("./clean_data/long_demog.Rdata")
 
 source("./code/get_dem_variables.R")
 
-long_demo <- bind_rows(anes5_demog %>% mutate(df = "1956-60 ANES"), 
-                       anes7_demog %>% mutate(df = "1972-76 ANES"), 
-                       anes8_demographic %>% mutate(df = "1980 ANES"),
-                       anes9_demog %>% mutate(df = "1990-92 ANES"), 
-                       anes90_demog %>% mutate(df = "1992-97 ANES"), 
-                       anes0_demog %>% mutate(df = "2000-04 ANES"), 
-                       gss6_demog %>% mutate(df = "2006-10 GSS"), 
-                       gss8_demog %>% mutate(df = "2008-12 GSS"), 
-                       gss10_demog %>% mutate(df = "2010-14 GSS"), 
-                       anes16_demog %>% mutate(df = "2016-20 ANES"), 
-                       anes20_demog %>% mutate(df = "2020-22 ANES"), 
-                       gss20_demog %>% mutate("2016-20 GSS")) %>%
-  mutate(evermarried = ifelse(marital %in% c("married", "other"), 1, 0),
-         ba = ifelse(ed == "ba", 1, 0)) 
-
+long_diff_demo <- left_join(long_difference, long_demog, by = c("df"="df", "id"="id", "t1"="wave"))
 
 
 #Panel summaries
@@ -42,12 +29,16 @@ psummaries <- long_difference %>%
 
 # Summarize absolute difference at the wave-pair level for each age group
 # for each question. 
-summaries <- long_difference %>%
-  filter(age <= 80) %>%
+summaries <- long_diff_demo %>%
+  filter(age <= 80, !is.na(ba), !is.na(evermarried), !is.na(student)) %>%
   mutate(wave_pair = paste(t1, t2, df, sep = "-")) %>%
   group_by(wave_pair, name, age_group) %>%
-  summarise(d = weighted.mean(duration), a = weighted.mean(abs_diff),
-            date = weighted.mean(date), n = n(), sd = sd(abs_diff),
+  summarise(d = weighted.mean(duration, weight = weight), a = weighted.mean(abs_diff, weight = weight),
+            pct_ba = weighted.mean(ba, weight = weight),
[…]

### 2024-07-03 code/get_dem_variables.R
drafting

@@ -1683,3 +1683,36 @@ gss20_demog <- gss20 %>%
     values_from = value
   ) %>%
   mutate(df = "2016-20 GSS")
+
+long_demog <- bind_rows(anes5_demog,
+                        anes7_demog,
+                        anes8_demog,
+                        anes90_demog,
+                        anes9_demog,
+                        anes0_demog,
+                        anes16_demog,
+                        anes20_demog,
+                        gss6_demog,
+                        gss8_demog,
+                        gss10_demog,
+                        gss20_demog) %>%
+  mutate(evermarried = recode(marital, "married"=1, "other"=1, "single/nm"=0),
+         female = recode(sex, "1"=0, "2"=1),
+         ba = recode(ed, "hs"=0, "less than"=0, "ba"=1),
+         lt = recode(ed, "hs"=0, "less than"=1, "ba"=0),
+         everkid = as.numeric(everkid),
+         kidinhouse = as.numeric(kidinhouse),
+         student = as.numeric(student),
+         id = as.numeric(id)) %>%
+  select(df, id, wave, evermarried, student, everkid, kidinhouse, ba, lt)
+
+save(long_demog, file = "~/Dropbox/extended_formative_windows/clean_data/long_demog.Rdata")
+
+rm(anes5, anes7, anes8, anes90, anes9, anes0, anes16, anes20,
+   gss6, gss8, gss10, gss20)
+rm(anes5_demog, anes7_demog, anes8_demog, anes90_demog, anes9_demog, anes0_demog, anes16_demog, anes20_demog,
+   gss6_demog, gss8_demog, gss10_demog, gss20_demog)
+
+
+
+

### 2024-07-05 code/get_dem_variables.R
long_demog fix

@@ -1611,7 +1611,8 @@ long_demog <- bind_rows(anes5_demog,
          rac_black = ifelse(race == "2", 1, 0),
          rac_other = ifelse(race == "3", 1, 0)) %>%
   select(df, id, wave, evermarried, student, everkid, kidinhouse, ba, lt, rac_black,
-         rac_other, female) 
+         rac_other, female) %>%
+  mutate(dec_diff = (as.numeric(difftime(date, max(date), units = "days")))/3652.5)
 
 save(long_demog, file = "~/Dropbox/extended_formative_windows/clean_data/long_demog.Rdata")
 

### 2024-07-08 manuscript/manuscript.Rmd
drafting

--- 2024-07-03 manuscript/formative_period_kk.Rmd
@@
 fontsize: 12pt
-abstract: "Early adulthood is consistently identified as a “formative period” in the life course when people have heightened probability of changing their opinions, beliefs, and political preferences.
-This heightened rate of change is often attributed to the social conditions of this life stage.
+abstract: "Early adulthood is consistently identified as a “formative period” in the life course when people demonstrate a heightened probability of changing their opinions, beliefs, and political preferences.
+This heightened rate of change is often attributed to the social conditions of this life stage but could also reflect biological or cognitive differences between young adults and older adults.
 In this paper, we explore whether substantial changes in the timing and ubiquity of major life-course transitions over the the past 70 years have affected the rate at which early adults and the rest of the population change their opinions on political and social issues.
 Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the 1960s.
-Instead, we find that other age groups have become more stable opinion holders."
+Instead, we find that other age groups have become more stable opinion holders.
+These aggregate trends obscure two countervailing patterns: decreases in stability due to life-course changes among younger adults and increases in stability across all age groups due to increases in educational attainment, as well as a broader stability trend that remains unexplained.
+Our results suggest a braod role in social contexts in shaping attitude stability."
 ---
@@
 
-Why people in this life stage appear to show more frequent opinion change relative to older people, however, is unclear.
+Why people in these life stages appear to show more frequent opinion change relative to older people, however, is unclear.
 Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences," arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007].
-Other related explanations focus on the social-structural features of middle age that might stabilize attitudes [@eaton2009; @visser2004].
+Related explanations conversely focus on the social-structural features of middle age that might stabilize attitudes [@eaton2009; @visser2004].
 At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.
 
-While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change during early adulthood, the relative contributions of each are unclear, and clarifying their relative contributions is important for understanding the mutually reinforcing processes of social change and cultural evolution.
-If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change.
+It is likely true that both social and cognitive/biological mechanisms play a role in facilitating opinion change during adolescence and early adulthood, but the relative contributions of each are unclear, and clarifying their relative contributions is important for understanding the mutually reinforcing processes of social change and cultural evolution.
+If the mechanisms are principally cognitive and biological, rather than social, then broader changes in the structure of different life stages are unlikely to affect rates of social change in the same way.
+If, conversely, openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change.
 Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016].
-If the mechanisms are principally cognitive and biological, rather than social, then these broader changes are unlikely to affect rates of social change in the same way.
 
-In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events.
 Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts [@buchmann1989; @bruckner2005].
-The "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009].
+Rates of high school and college completion have also increased significantly.
+As a result, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009].
 These broad social changes lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.
 
-Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of the year they were observed and their age at the time.
-Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) have more variable opinions than older respondents.
+In this paper, we test whether these broader shifts in social structure and the timing of life course transitions have affected the rate at which young people change their attitudes about political and social issues.
+Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we estimate the amount of change in opinions as a function of age and cohort.
+Consistent with the "impressionable years" hypothesis, we find that people in age ranges associated with this "early adulthood" life stage (18-25 and 26-33) have more variable opinions than older respondents.
 However, these differences are not constant over time.
-They emerged over the second half of 20th Century as middle-aged and older respondents came to report more stable opinions over time, while early adult respondents' reported opinions continued to change at roughly the same rate over time.
+They emerged over the second half of 20th Century as more recent cohorts exhibited more stable opinions in middle and old age compared to their predecessors, while early adult respondents' reported opinions continued to change at roughly the same rate as in earlier cohorts.
 
-Our findings complicate the narrative about social change and cultural change and raise questions about which social forces emerged to stabilize attitudes in middle age for more recent cohorts.
-Life-course stages can potentially explain the divergence of opinion stability between the youngest age groups and older age groups over time, but they cannot explain the overall trajectory of increased stability.
+Furthermore, we show that the aggregate patterns in attitude change over time are principally due to forces other than changes in life-course structure.
+While decreased rates of marriage explain a small proportion of the aggregate pattern among early adults, the largest driver of change in attitude stability comes from increased rates of education across all age groups.
+Furthermore, much of the shift in aggregate attitude stability remains unexplained, suggesting broader social structural features such as increased access to information with which to form opinions, political and cultural polarization, and more are the key drivers of attitude stability.
 
@@
 The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021].
-The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on their personal culture -- worldviews, attitudes, dispositions, behaviors, and preferences [@gerber1998].
-In the study of political socialization, this perspective enjoys a longstanding history.
-"Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014].
+The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on what Lizardo [-@lizardo2017] calls "personal culture" -- worldviews, attitudes, dispositions, behaviors, and preferences that manifest at the individual level.
+In political socialization, "generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@gerber1998; @mannheim1952; @bartels2014].
 However, the notion of "impressionable years" is more general than just political preferences.
-The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but their cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@elder2016].
+The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but their cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@ryder; @fosse2023a; @elder2016].
 Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990].
@@
 
-These aggregate models of social imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time.
-The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course.
+These aggregate models of cultural imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time.
+At a basic level, the "impressionable years" hypothesis requires that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age.
 The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a].
 While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [\@].
+These tests tend to find higher rates of attitude change among the youngest adults -- people ages 18-25 -- and little difference between adults of other age ranges.
 More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S.
-adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains much more variance than period, the year of observation, for most items in the General Social Survey, suggesting a strong impressionable years pattern.
+adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains more variance than period, the year of observation, for most items in the General Social Survey, suggesting an impressionable-years pattern.
 
-In recent years, researchers have attempted to use new approaches to quantify "openness" to social influence across the life course, especially around politics [@bartels2014; @ghitza2022].
-This work has produced remarkably similar patterns with openness peaking from a person's mid-teens through age 25 and declining quickly after that, lending more credence to the "impressionable years" hypothesis [a pattern that seems to hold for music tastes as well @way2019].
-However, this work tends to assume a constant age-based rate of openness across cohorts -- that 20-year-olds in 1950 are open to influence at the same rate as 20-year-olds in 2010 -- which has not been tested.
+In recent years, researchers have also developed new approaches to quantify "openness" to social influence across the life course, especially around politics [@bartels2014; @ghitza2022].
+This work has produced remarkably similar patterns to previous work with openness peaking from a person's mid-teens through age 25 and declining quickly after that, lending more credence to the "impressionable years" hypothesis [a pattern that seems to hold for music tastes as well @way2019].
+However, this work tends to assume a constant age-based rate of openness across cohorts -- that 20-year-olds in 1955 are open to influence at the same rate as 20-year-olds in 2015 -- which has not been tested.
 
@@
 
-Why does research consistently finds the period of time between 14 and 25 to be distinctly "impressionable years?" At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information.
-For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for age-based patterns in opinion change.
+While the empirical pattern of early openness followed by relative stability appears quite robust, its cause is not clear.
+At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information.
+For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for the observed age-based patterns in opinion change.
 Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern.
-It could also be the case that both change as a function of age.
+It could also be the case that both these processes change as a function of age.
 Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.
 
-## Biological Mechanisms
+## Biological and Cognitive Mechanisms
 
@@
 
-It is also possible that there is something unique about human cognition during these periods making people ascribe more weight to the information acquired within them than to information acquired at other points.
+It is also possible that there is something unique about human cognition during these periods that makes people ascribe more weight to the information acquired within them than to information acquired at other points.
 Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life [@mata2015].
@@
 
-The thrust of the biological arguments is that, irrespective of the contemporaneous social conditions humans inhabit, millennia of evolution have led to the development of a cognitive architecture that makes them relatively open changing dispositions at younger ages and relatively resistant at older ages.
-However, while biological mechanisms can shape how individuals respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change.
+The thrust of the biological and cognitive arguments is that, irrespective of the contemporaneous social conditions humans inhabit, millennia of evolution have led to the development of a cognitive architecture that makes them relatively open to changing dispositions at younger ages and relatively resistant at older ages, regardless of their social environments.
+However, while biological mechanisms can shape how people respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change.
 For that, we must turn attention to the structuring of the social world.
@@
 Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct, characterized by sustained social learning and constant changes of social contexts.
-Here we briefly consider some of the ways that the "impressionable years" differ in ways that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.
+Here we briefly consider some of the ways that the "impressionable years" differ from other life stages that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.
 
@@
 People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons.
-Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups.
+Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups [NEED CITES HERE].
 
@@
 These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions.
-For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic].
-Social networks also differ in this time period.
-Young peoples' networks tend to be larger and include more friends than kin, for example [@marsden1987].
+For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic2021].
+Personal social networks also differ in this time period.
+Young peoples' networks tend to be larger and include more friends than kin [@marsden1987].
 It could be the case that these larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].
@@
 
-While the distinct social and cultural conditions of adolescence and early adulthood might expose young people to new information at higher rates than older people, the impact of this information on peoples' attitudes is contingent on their ability to incorporate it into their cultural repertoire.
+While the distinct social and cultural conditions of adolescence and early adulthood might expose young people to new information at higher rates than older people, the impact of this information on people's attitudes is contingent on their ability to incorporate it into their cultural repertoire.
 The social conditions of different life stages might make it easier or more difficult for people to act on these new sources of information in ways that facilitate opinion change.
 
-Social networks not only shape peoples' exposure to new information, but they also affect the degree to which people adopt new beliefs and attitudes.
-People are much less likely to defect from their traditional beliefs and adopt new beliefs if they are embedded in dense networks [@stark1980].
-Consistent with this, mid-life is associated with denser, more homogenous networks [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004].
+Social networks not only shape peoples' exposure to new information, but they can also affect the degree to which people adopt new beliefs and attitudes.
+People are much less likely to defect from their traditional beliefs and adopt new beliefs if they are embedded in dense networks that reinforce the holding of existing beliefs, even when they are exposed to new ideas [@stark1980].
+Consistent with this, mid-life is associated with denser, more homogenous networks that include larger numbers of family [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004].
 If young people exist in less dense, more heterogeneous social networks, it might be easier for them to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks.
@@
 
-Previous work on opinion stability over the life course suggests that people tend to value stability in figures holding social power.
-As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information [@eaton2009].
-This potentially explains another phenomenon: that people appear to open up to changing their minds again as they enter later life and retirement, when these expectations presumably diminish [@danigelis2007].
+Previous work on opinion stability over the life course suggests that people tend to value stability in figures holding social power, which tend to occur more frequently during middle age.
+As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information or forcing them to develop stable opinions on issues where previously they felt capable of ambivalence [@eaton2009].
+This potentially explains another phenomenon: that people appear to open up to changing their minds on some issues as they enter later life and retirement, when these social expectations of stability presumably diminish [@danigelis2007].
+
+Existing research suggests that the structure of social life during adolesence, early adulthood, and later life stages shape both exposure to new information and a person's ability to update attitudes, opinions, and beliefs in the face of that information.
+This further suggests that changes in these life course stages should matter for the amount of opinion change we observe.
 
 # Changing Life Course
-
-Given the distinct social conditions of the early-adult phase of life, there are reasons to believe that the "impressionable years" are principally an effect of how societies choose to structure the life course or the interaction between these life course stages and biological mechanisms.
-Because of this, contemporary historical trends offer a window to disentangle the role of socio-cultural and biological or cognitive factors in explaining why we observe "impressionable years."
 
@@
 
-First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts.
+First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for successive cohorts as they pass through the same ages.
 Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce.
@@
 
-Similarly, the duration of education has extended, with more people completing high school, college, and, increasingly, graduate degrees before entering the workforce.
+Similarly, the duration of education has extended, with more people completing high school, college, and, increasingly, graduate degrees before entering the workforce or movement back into education after several years in the workforce.
 These transitions are also accompanied by heightened rates of geographic mobility, as well as movement between the workforce and education.
 
-If the distinctive nature of "impressionable years" is principally driven by their unique social circumstances, then these shifts have the potential to extend these social circumstances into later ages.
-This, in turn, could have an effect on how long exploration, and therefore openness to durable change, lasts throughout the life-course.
+If the distinctive nature of "impressionable years" is principally driven by their distinct social circumstances, then these shifts have the potential to extend these social circumstances into later ages.
+This, in turn, could have an effect on how long exploration, and therefore openness to change, lasts throughout the life-course.
 If this is the case, we should expect to see heightened levels of attitude change among age groups that, for earlier cohorts, displayed mainly stability.
 In other words, people who were born in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age.
-Specifically, we expect the attitude behavior of people aged 25-33 to more closely resemble middle-aged respondents in earlier cohorts [@krosnic1991; @alwin1991] but younger respondents in more recent cohorts.
-This produces our first hypothesis:
+Specifically, we expect cohorts who entered adulthood in the late 1950s to be fairly comparable to older adults in their rates attitude change but that subsequent cohorts will diverge from older adults, making more change over time.
 
-*Hypothesis 1: More recent cohorts will show higher rates of attitude change in the age range of 25 to 33 than cohorts born in earlier years.*
+*Hypothesis 1: More recent cohorts will show higher rates of attitude change in the age range of 18 to 25 than cohorts born in earlier years.*
+
+We expect a similar but more muted trajectory for the attitude behavior of people aged 25-33.
+Again, we expect the oldest cohorts to resemble middle-aged respondents when they enter this age range, as has been shown in previous research [@krosnic1991; @alwin1991], but that more recent cohorts will come to more closely resemble the 18-25 age group, diverging from middle aged respondents.
+
+*Hypothesis 2: More recent cohorts will show higher rates of attitude change in the age range of 26 to 33 than cohorts born in earlier years.*
+
+Consistent with life-course structuring theories, we expect these divergences will principally be explained by changes in the timing of life course transitions such as marriage and childbirth.
+Controlling for these life-course transitions should eliminate the gap between the youngest adults and later adults.
+
+*Hypothesis 3: Controlling for rates of marriage and childbirth, more recent cohorts' rates of attitude change in early adulthood (18-33) will not be different from the rates of attitude change among contemporaneous middle aged adults.*
 
@@
 
-Rosenfeld's work suggests that the social conditions that shape where we spend our impressionable years might also shape how much exploration we can engage in.
-An increase in independence might entail more exploration, and greater ability to update attitudes as a result of this exploration.
+Rosenfeld's work suggests that the social conditions that shape where people spend their impressionable years might also shape how much cultural exploration they can engage in.
+An increase in independence might entail more exploration and greater ability to update attitudes as a result of this exploration.
 This assertion in turn engenders a second prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.
@@
 
-*Hypothesis 2: More recent cohorts will show higher levels of opinion change in the 18 to 25 age range compared earlier cohorts.*
+*Hypothesis 4: Controlling for life course transitions, more recent cohorts of early adults (18-33) will make higher rates of attitude change than contemporaneous middle-aged adults.*
 
@@
 
-While we have parimarily focused on changes in the timing and structure of life course stages over the past 70 years, that time frame has also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure.
+Finally, while we have primarily focused on changes in the timing and structure of life course stages over the past 70 years, that time frame has also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure.
 We briefly consider three broader changes here.
 
-First, the number of people pursuing and completing higher education increased significantly over the window studied here [@schofer2005; @hout2012].
+First, the number of people completing high school and pursuing and completing higher levels of education has increased significantly over the window studied here [@schofer2005; @hout2012].
 Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007].
 The specific mechanism that links educational attainment and attitude stability is unclear, with some perspectives suggesting educational attainment improves cognitive function and memory while other perspectives suggest that educational attainment makes people pay more attention to current events.
-It could also be that education cultivates in people an expectation that they hold opinions in the first place.
+It could also be the case that education cultivates in people a normative expectation that they should hold opinions.
 Regardless of the specific mechanism, a population with higher levels of education, on average, might be more likely to hold stable opinions regardless of transformations in life course stages.
@@
 Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire information that makes opinion-holding easier.
+To the extent that stable opinions are in part a reflection of signals from elites [@converse1965; @zaller1992], then we should expect these shifts to stabilize opinions across all age ranges.
 
 Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina2008; @baldassarri2008].
-People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020].
-While clarity aroud political issue positions is most commonly studied, the same forces can also align other social beliefs and identities.
+People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta [-@dellaposta2020] calls an "oil spill" of polarization.
+While clarity around political issue positions is most commonly studied, the same forces can also align other social beliefs and identities.
 This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently [@zaller1992].
@@
 Broadly speaking, these three forces suggest that opinions should broadly become more stable over time, regardless of life-course stages.
-Additionally, opinions on certain issues might become more solidified the longer issues remain in the public, as social movements and organizations become solified around them.
+If this is the case, we might not observe an absolute decrease in stability for early adults, but rather a relative decrease, as extension of the life course works to decrease stability for members of these age groups while broader social forces stabilize opinions for all groups.
 
@@
 
-Political and Social Attitudes: Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time than members of earlier cohorts.
-Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, with respondents coming from the same general population.
+Testing our hypotheses requires data on opinion and attitude reports over time for members of different cohorts in the same age ranges, with respondents coming from the same general population.
+We also require information about life-course transitions standardized across these panels.
 To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years.
@@
 
-For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points.
-Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same set of questions, we did not treat these as separate panels for the purposes of identifying three distinct time points.
+For reasons explained below, we identified social and political attitude questions that were asked in at least three different panels with at least one of those panels asking respondents to answer the question at three or more time points.
+We sought questions that tapped a respondent's report of their own belief, avoiding questions that asked respondents to report how others, including politicians and political parties, would respond.
+Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the exact same set of questions, we did not treat these as separate panels for the purposes of identifying three distinct time points.
 In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.
-We allowed for question wording to vary slightly across panels (e.g., feeling thermometers about "blacks" instead of "negros").
+We allowed for question wording to vary slightly across panels (e.g., feeling thermometers about the "women's liberation movement" were combined with questions about the "women's movement").
 When the same or similar questions had slightly different response scales, we harmonized response options to the question structure with the fewest response options.
@@
 
-The central question of our analysis pertains to age, and as such we classify respondents into a set of age brackets depending on their age when they answered survey questions.
+The central question of our analysis pertains to age and cohort, and as such we classify respondents into a set of age brackets depending on their age when they answered survey questions.
+Age was reported at least once in all panels.
+If respondents' were not asked their age or birth date in a specific wave, we calculated it based on other age reports.
 Given our theoretical questions, we classify people into four age categories: 18-25, 26-33, 34-64, and 64-80.
-Because of differences in both question structure and age structure across panels, we drop respondents who are over 80.
+Because of differences in both question structure and age structure across panels, we dropped respondents who are over 80.
 
-We are also interested in the effect of having undergone a set of life-course transitions on attitude stability to more directly test whether divergences in life course structure drove changes in attitude stability.
+Our theoretical questions also focus on the effect of having undergone a set of life-course transitions on attitude stability.
 To this end, we include a set of indicator variables capturing whether the respondent had undergone certain demographic transitions.
@@
 
-It was quite common for respondents to have missing data on demographic variables for a particular wave of any panel, even if they participated in the panel.
+Broadly speaking, questions tapping marital status, student status, and degree attainment were asked in the same way with the same response options in each panel or could be easily harmonized.
+It was quite common for respondents to have missing data on one or more of these demographic variables for a particular wave of any panel, even if they participated in the panel.
 This was most common when a question occurred in the post-election wave of the ANES, but demographic variables were captured in the pre-election wave.
-Similarly, some demographic items were only asked in one wave of a panel.
-When respondents were missing on a demographic item, we imputed the missing variable using first earlier responses and then, if there were no earlier responses, later responses.
+If this occurred, we carried over the pre-election response, as these waves tended to be separated by only a few weeks.
 
-Broadly speaking, questions tapping marital status, student status, and degree attainment tended to be asked in the same way with the same response options or could be easily harmonized.
-In contrast, generating a variable for whether a respondent had children was more challenging, as what information was being sought varied significantly across panels and even within panels.
-Specifically, questions occasionally tap how many children were currently living with the respondent and at other times tapped whether the respondent had any children or how many children the respondent had ever had.
-While these items are highly correlated for respondents with young children, these two approaches to the question produce significant differences in rates among parents whose children have left home.
-
-To generate a measure of whether respondents had children, we ...
-
-1990 pre: Number of children in house
-1991: copied from 1990
-1992: any children?
-1992 post: copies from 1992pre
-1993: copied from 1992 pre
-1994: any children ever
-96: Any children under 18 in household
+In contrast to the other demographic variables, generating an indicator variable for whether a respondent ever had a child was more challenging, as question wording and structure varied significantly across panels.
+Specifically, survey administrators occasionally asked respondents how many children were currently living with them, while at other times asked whether the respondent had any children and how many.
+While these items are highly correlated for respondents with young children, these different approaches to the question produce significant differences in rates among parents whose children have left home.
+To account for this, we generated two variables.
+First, we created an indicator variable for whether they reported having at least one child in the house under 18 ("kidinhouse").
+Respondents who were not asked the question in a particular wave (or temporally proximate wave) were coded as missing.
+Second, we gave respondents an indicator for whether they had ever had a child.
+We coded this as 1 if respondents were asked if they ever had a child and reported that they did or if they were asked if they had a child living with them and reported that they did.
+We coded this as 0 if respondents were asked whether they ever had a child and said no.
+Respondents were coded as missing if they were asked how many children under 18 they had living with them and said none, or if they were not asked either relevant question in a particular wave or temporally proximate wave.
 
 ## Methods
+
+Our outcome of interest is the absolute change in responses a respondent makes across a pair of observations which, due to scaling, can be between 0 and 100.
 
@@
 In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations by the same respondent.
-Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question was asked.
-In the 1972-76 ANES panel, respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
+Importantly, respondents might have multiple observation pairs for each panel depending on the number of times a question was asked.
+In the 1972-76 ANES panel, for example, respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
 If a person participates in all three waves, they have three observations of $|y_{2i}-y_{1i} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).
@@
 Variation in duration between observations within panels lets us separate two kinds of change: 1) change attributable to durable shifts of opinions over time that can be associated with duration between observations and 2) change that is principally random, or unrelated to the duration between observations.
-If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$.
+If respondents have a stable underlying partisan identification they report with random error, and no real change occurs between waves beyond that, then in expectation, duration should be unrelated to absolute change in responses over time, and $E(\beta_1) = 0$.
 In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change.
@@
 
-Given our focus on age-based differences, we can extend equation to incorporate age groups as indicator variables:
+Given our focus on age-based differences, we can extend equations to incorporate age groups as indicator variables:
 
@@
 
-$\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to $t = 0$, which we set to be the most recent observation date (2016).
+In this formulation, $\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to $t = 0$, which we set to be the most recent observation date (2016).
 $\beta_8$ estimates the change in the baseline rate of change for the reference group.
@@
 To this point, we have developed our model for the case of a single question.
-However, we lack a large number of observations for most items, meaning separate estimates of the time trends for all these items are likely to be inconclusive.
+However, because different panels tend to ask different questions, we lack a large number of observations for any particular item, meaning separate estimates of the time trends for all these items are likely to be inconclusive.
 One solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels.
@@
 
-Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty. We follow this approach, with the model:
+Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty. We follow this approach, using the model:
 
-$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1(d) + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \beta_8(t) + \beta_9(d*t) + \beta_{10}(a_{26-33}*t) + \beta_{11}(a_{34-64}*t) + \beta_{12}(a_{65+}*t) + \beta_{13}(d*a_{26-33}*t) + \beta_{14}(d*a_{34-64}*t) + \beta_{15}(d*a_{65+}*t) + \epsilon_i $$
+$$ |y_{2i}-y_{1i} | = \beta_{0j} + \beta_{1j}(d) + \beta_{2j}(a_{26-33}) + \beta_{3j}(a_{34-64}) + \beta_{4j}(a_{65+}) + \beta_{5j}(d*a_{26-33}) + \beta_{6j}(d*a_{34-64}) + \beta_{7j}(d*a_{65+}) + \beta_{8j}(t) + \beta_{9j}(d*t) + \beta_{10j}(a_{26-33}*t) + \beta_{11j}(a_{34-64}*t) + \beta_{12j}(a_{65+}*t) + \beta_{13j}(d*a_{26-33}*t) + \beta_{14j}(d*a_{34-64}*t) + \beta_{15j}(d*a_{65+}*t) + \epsilon_i $$
 
@@
 
-$$ \beta_{kj} = \gamma_k + \upsilon_{kj} $$ This model allows for each survey question to have its own set of coefficient estimates weighted by the amount of information we have about the particular question.
-Questions asked in many panels will contribute more to the overall sense of the aggregate trend in opinion stability.
+$$ \beta_{kj} = \gamma_k + \upsilon_{kj} $$
 
@@
 These equations are nested at the level of the survey item.
+This model allows for each survey question to have its own set of coefficient estimates weighted by the amount of information we have about the particular question.
+Questions asked in many panels, including partisan identification (seven point scale from "Strong Democrat" to "Strong Republican", asked in a total of 32 waves across all panels) and ideological identification (seven-point scale from "Extremely Liberal" to "Extremely Conservative", asked in a total of 28 waves across all but one panel), will contribute more to the overall calculation of the aggregate trend in opinion stability than questions asked in only three panels.
+Given the lack of confidence in each individual question trend, we focus the majority of our attention on $\gamma_k$ estimates.
+
 In practice, the model presented in Eq \### is likely too complicated given our data limitations, and might not be necessary.
@@
 Once we select the most parsimonious model to capture attitude stability over time, we then include variables capturing demographic transitions to evaluate their relationship to attitude stability over time.
-To distinguish between the influence of life course structure and other factors that might be affecting attitude stability, such as broader social forces, we calculate the expected rate of attitude change in later panels holding constant the proportion of respondents in each group who made certain demographic transitions.
-
 For each wave pair, we include measures of demographic status at the first observation.
@@
 If they are observed a third time, they would be counted as "married" for that pair of waves.
+
+Finally, to distinguish between the influence of life-course structure and other factors that might affect attitude stability, such as broader social forces, we calculate a counter-factual expectation of the rate of attitude change in later panels if the proportion of respondents in each group who made certain demographic transitions was constant over time at 1956 levels.
 
@@
 
-While not the central question of our analysis, we first highlight changes in three life-course measures -- proportion of respondents who are married, proportion of respondents who have at least one child, and proportion of respondents who have a college degree -- across the panels we include in our analysis to illustrate the substantial shifts social conditions of early adulthood.
-The measurement of children varies between panel, sometimes asking respondents how many children they have ever had and sometimes asking respondents how many children currently live in their home.
-These responses are often similar for people in younger age groups, but produces very different responses in respondents in older age groups who have children that no longer live with them.
-For that reason, we hold respondents 50 and older off of the figure.
+While not the central question of our analysis, we first highlight changes in four life-course transitions: proportion of respondents who are married, proportion of respondents who reported having a child, the proportion of respondents with a child at home, and proportion of respondents who are currently students.
+We also plot the proportion of respondents at three different education levels: less than high school, high school degree but no college degree, and college degree.
 
-In the early panel (1956-60), respondents in the lowest age group look very similar to respondent in the older groups on all major measures.
-However, they quickly diverge by the 1970s, displaying the lowest probability of being married and the lowest probability of having children by quite a larger margin.
-Over time, the next youngest agr group (26-33) starts to diverge toward the youngest age group, especially in terms of marriage.
+In the early panel (1956-60), respondents in the youngest age groups look very similar to respondents in the older groups on major measures.
+They are equally likely to be married, to have had children, and to be students.
+However, people in the 18-25 age group quickly diverge from the others by the 1970s, displaying the lowest probability of being married and the lowest probability of having children by quite a larger margin.
+Over time, the next youngest age group (26-33) starts to diverge toward the youngest age group, especially in terms of marriage.
 
-These trends set the stage for what to expect in terms of reliability and stability.
-To the extent that these measures as proxies for life stages that influence attitude stability and reliability, then we should see the youngest age group look similar to the older age groups in the 1950s, and diverge by the 1970s, continuing to diverge over time.
+Conversely, the middle age and older age groups see a significant increase in the proportion who hold high school and then college degrees.
+By the mid-1990s, the proportion of middle aged and older respondents who had college degrees significantly exceeded the proportion of youngest age group who has college degrees, likely because many people aged 18-25 were still completing education.
+
+These trends set the stage for what to expect in terms of attitude change for different age groups.
+To the extent that transitions such as marriage and having children themselves stabilize attitudes or serve as proxies for broader life course factors that influence attitude stability, then we should see the youngest age group look similar to the older age groups in the 1950s, and diverge by the 1970s, continuing to diverge over time.
 Respondents in the 26-33 age group should remain similar to older respondents for most panels, but begin to diverge by the 1990s and especially by the 2000s.
@@
 
-The coefficients for age groups other than the reference category suggest that, on average, other age groups are more stable than people in the 18-25 age range in the most recent panels.
+Coefficients in Model 1 allow us to test several of the hypotheses outlined above.
+First, the coefficients for age groups other than the reference category suggest that, on average, other age groups are more stable than people in the 18-25 age range in the most recent panels.
 Respondents in the 26-33 age group are about 1.1 points more stable on average, respondents in the 34-65 age group are about 2.2 points more stable on average, and respondents in the 65 and over age group are about 3.5 points more stable on average, than respondents in the 18-25 age group.
@@
 
-Coefficients in Model 1 allow us to test several of the hypotheses outlined above.
-The intensification of early adulthood hypothesis (H3) generally suggests that the most recent cohorts should make the most change when they are in the youngest age range than cohorts that came before them.
-The effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significantly different from 0.
+Turning now to how these groups have changed over time, the effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significantly different from 0.
+To clarify, the lack of statistical significance for the time effect not because there is a large standard error on the term.
+It is estimated with a high degree of precision and is very close to 0.
 On average, recent cohorts in this age group make approximately the same amount of wave-to-wave change in recent years as earlier cohorts did, accounting for the kinds of questions they were asked and time between responses.
@@
 
-Next, the extended early adulthood framework (H1) suggests that we should see a closing of the gap between people in the 18-25 and 26-33 age groups over time.
+Next, the extended early adulthood framework (H2) suggests that we should see a closing of the gap between people in the 18-25 and 26-33 age groups over time.
 The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that while people in this age group are more stable, on average, than then youngest respondents, more recent cohorts are not any more or less likely to make changes of opinion in this age range than earlier cohorts, and more recent cohorts are not any more or less similar to younger adults than earlier cohorts.
@@
 Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range whose rate of opinion change remains flat across cohorts, respondents in older age groups (34-64 and 65+) do show a notable time trend -- more recent cohorts are more stable opinion reports in these age ranges than earlier cohorts.
-To more clearly illustrate overall trends, Figure XXX plots the predicted time level of change for respondents in each age category over time.
+Both time trends are negative, substantively large, and statistically significant.
+To more clearly illustrate these trends, Figure XXX plots the predicted time level of change for respondents in each age category over time.
 
@@
 In the earliest panels, the oldest respondents (65 and over) are substantially less stable opinion holders than younger respondents, changing by about 5 points more than members of the other groups, while the remaining three age groups (18-25, 26-33, and 34-64) appear to change at approximately equal rates.
-Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable, with the latter becoming almost 10 points more stable on average.
+Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable, with the latter becoming almost 10 points more stable on average from the 1950s to the 2020s.
 Because, as noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change over time, they become the least stable respondents by the end of the panel.
@@
 
-While coefficient estimates for the early adult age groups do not suggest that they have become more variable opinion holders over time, the increase in stability for middle-aged and older respondents suggests that broad social forces might counteract life-course the effect of transitions.
+While coefficient estimates for the early adult age groups do not suggest that these respondents have become more variable opinion holders over time, the increase in stability for middle-aged and older respondents suggests that broad social forces could be counteracting life-course transitions for members of these age groups.
 
@@
 We re-estimate Model \### above, this time incorporating a series of indicator variables for whether respondents were students, whether they had children, and whether they were married.
-We also include an indicator variable for whether the respondent completed college to capture the effect of ...
-.
+We also include an indicator variable for two levels of education -- whether respondents have less than a high school degree and whether respondents completed college -- to evaluate whether broader shifts in attitude stability in the population might be due to shifts in educational attainment.
+These results are presented at Model 2 in Table XXX above.
 
-Using these estimates, we calculate the expected value of change had life-course transitions occurred at roughly the same rates.
+These results provide mixed evidence of life-course shifts having a meaningful effect on attitude stability.
+Respondents who have been married are slightly more stable than unmarried respondent who were never married, but respondents who have had children are less stable than respondents who never had children.
+Having children at home increases change even more.
+Taken together, these coefficients suggest that changes in the timing of life course transitions are insufficient to explain the divergence between young adults and older adults in terms of stability.
+
+On the other hand, coefficients related to education show a clearer picture.
+Higher levels of education are uniformly associated with increased attitude stability (positive coefficient on having less than a high school degree and a negative coefficient on having a college degree), suggesting that changing levels of education in the population -- and the increase in education level of middle-aged respondents relative to younger respondents -- likely accounts for a larger share of the divergence in stability.
+
+Using these estimates, we calculate the expected value of change had life-course transitions occurred at roughly the same rates that they did in the late 1950s.
+These estimated counter-factual values are plotted in figure XXX.
 
@@
 Given that Americans, on average, pushed back marriage, childbirth, and entering the workforce over the course of the second half of the 20th Century and the early 21st Century, we posited that these shifts should, through various pathways, increase over-time variance in survey responses among the youngest members of the population.
-We tested that hypothesis by looking at the stability of political and social attitudes among different cohorts as they moved through the same age groups.
+We tested a series of hypotheses related to this theory by looking at the stability of political and social attitudes among different cohorts as they moved through the same age ranges.
 
@@
 Consistent with expectations, the youngest adults look similar to older adults in the earliest panels in terms of both demographics and attitude stability.
-While members of the youngest age categories (18-25 and 26-33) diverged over time from older members of the population in the rate at which they changed their opinions, showing more change than older respondents in the most recent panels, this shift was principally driven by an increase in stability among people in middle age and over 65, rather than increased opinion change among the youngest adults.
+While members of the youngest age categories (18-25 and 26-33) diverged over time from older members of the population in the rate at which they changed their opinions, with younger respondents showing more change than older respondents in the most recent panels, this shift was principally driven by an increase in stability among people in middle age and over 65, rather than increased opinion change among the youngest adults.
 We suggested that this aggregate divergence was driven by countervailing trends.
@@
 We found that, consistent with our revised perspective of countervailing forces, age groups' rates of attitude change over time would have been more similar if these social changes did not happen.
-Specifically, increased rates of educational attainment account for much of the stabilizing shifts among older Americans, while delayed rates of marriage and childbearing decreased attitude stability among the youngest members of the population.
+Specifically, delayed rates of marriage decreased attitude stability among the youngest members of the population while increased rates of educational attainment account for much of the stabilizing shifts among older Americans.
 
-Our results suggest that broad social conditions play a substantial role in shaping the rate at which people change their attitudes.
-Since the 1950s, Americans have become meaningfully more stable opinion holders, in part because they have attained higher levels of education.
-Our approach likely under-estimates the importance of this shift, as we focus principally on rising rates of college attainment.
-However, in the earliest panels, a substantial proportion (> 70 percent) of the oldest respondents and more than 50 percent of middle aged respondents lacked a high school degree.
-
-Changes in the timing of life-course events were less important than changes in education for explaining, but they do play a role.
-Our findings suggest that, in the absence of these delays, young people would make attitude changes at a comparable rate to that of older people.
+In contrast to purely biological or cognitive explanations for differing rates of attitude stability among adolescents and early adults, our results suggest that broad social conditions play a substantial role in shaping the rate at which people change their attitudes.
+When early adults' lives resembled those of middle-aged respondents in the 1950s, their rates of attitude change were also similar.
+As social lives began to diverge, younger adults began to change their opinions at higher rates than older adults.
+At a broader level, since 1950s, Americans have become meaningfully more stable opinion holders, in part because they have attained higher levels of education.
+Our approach likely under-estimates the importance of this shift, as we collapse education into only three categories -- ignoring differences among people with less-than a high school degree in early panels and dramatic expansions of post-graduate degree attainment in recent decades.
 
 At the same time, it is important not to overstate the importance of these broader social transitions.
-Despite dramatic shifts in educational attainment, access to information, and political organization, people make changes in many social and political attitudes at the same rates they did 70 years ago.
+The average expected attitude change in 2020 is within a few points of the average expected attitude change.
+Despite dramatic shifts in educational attainment, access to information, and political organization, not to mention changes in survey administration, people make changes in many social and political attitudes at the same rates they did 70 years ago.
 
@@
 
-There are theoretically relevant trade-offs from focusing only on the same questions over time.
-By focusing only on the same questions over time, we control for question content and structure and ensure that the trend over time is due to something other than changes in survey administration.
-Because our panels span more than five decades that saw dramatic improvements in survey question design, questions included in more recent surveys might simply be of higher quality, with less wave-to-wave change, than items included in older surveys.
+One limitation of our design is that we focus on questions that are asked repeatedly.
+By focusing only on the same questions over time, we control for question content and structure and ensure that the trend we identify over time is due to something other than changes in survey administration or question quality.
+However, by only focusing on the same questions, we limit the kinds of issues about which we can make claims to issues that receive significant public attention for a long period of time, including general ideological positions and positions on long-standing issues, such as the government's role in the health care sector.
+We ignore large swaths of issues that emerge and disappear before the next panel but were very important in political debates, such as approaches to fight "stagflation." It could be the case that such issues have a different relationship between age and stability than issues that dominate political debates for decades.
+It could also be the case that stability of attitudes is principally a function of how long they have been in the public sphere ().
 
-We face several limitations in trying to make broader claims.
-First, many of our age-group estimates of stability and reliability are based on relatively small samples and as a result are not highly reliabile.
-This is particularly true for the youngest and oldest age groups, which tend to be the smallest in each panel, meaning we have the least confidence in each individual group-question estimate for these groups.
-We have attempted to limit this problem by averaging as many items as possible for each panel, but the number of questions in each panel is highly variable.
-
-Second, while we look at more than 250 questions in panels spanning 60 years, we do not actually have a large number of distinct time points.
-In total, we only have nine estimates of "average stability" and "average reliability" for each age group.
-Four of our nine panels fall within the same decade, and three of these panels overlap.
-Again, given the large amount of unreliability in our estimates for groups, we cannot rule out the possibility that clearer trends would emerge with more data.
-It is not clear how to improve this limitation, as we can not go back in time and generate new panels.
-
-Relatedly, most the panels span two presidential election years, in which respondents might be exposed to more information about national politics than usual that might produce heightened attitude change.
-Because of that, our results might not speak to other kinds of questions related to religion, and the patterns we observe might not be applicable to all times.
-
-Second, the high diversity of questions comes at the cost of not having many questions asked in multiple panels.
-Outside of the GSS panels (which span eight years total), most questions examined here are only asked in one panel.
-Of those that are asked in multiple panels, most are asked only twice, which gives us limited power for understanding change in reliability and stability net of change in question wording.
-It also forces us to assume that estimates of reliability and stability are comparable across panels as estimates of general reliability and stability of "political attitudes" at that time -- a claim that might not be valid.
+Second, because most questions are asked in a small number of panels, our actual number of data points for key questions is much smaller than the total number of wave-pair observations we have.
+The modal number of panels for individual questions is 3, meaning our ability to estimate temporal trajectories is highly constrained, and our estimates of question-specific trajectories is imprecise.
+This also makes it difficult to get efficient estimates of higher-order interactions or estimate time trajectories as anything other than linear trends.
+Relatedly, because many questions are only asked more than twice in one panel, it is difficult to get efficient estimates of whether we see changes in the "duration" effect for specific questions.
 

### 2024-07-11 manuscript/manuscript.Rmd
updated analysis and table

@@
 
-| Col1             | Col2  | Col3 | Col4 |
-|------------------|-------|------|------|
-| Fixed Effects    |       |      |      |
-| Intercept        | 17.45 |      |      |
-| Time (decade)    | -0.26 |      |      |
-| Age(26-33)       | -1.60 |      |      |
-| Age(34-64)       | -3.17 |      |      |
-| Age(65+)         | -4.10 |      |      |
-| Duration (year)  | 1.13  |      |      |
-| Time\*Age(26-33) | -0.09 |      |      |
-| Time\*Age(34-64) | -0.54 |      |      |
-| Time\*Age(65+)   | -1.33 |      |      |
-|                  |       |      |      |
-| Random Effects   | S.D.  |      |      |
-| Intercept        | 5.35  |      |      |
-| Time             | 1.14  |      |      |
-| Duration         | 0.37  |      |      |
-| Age(26-33)       | 0.89  |      |      |
-| Age(34-64)       | 1.81  |      |      |
-| Age(64+)         | 2.92  |      |      |
+|                          | Model 1       | Model 2      |
+|--------------------------|---------------|--------------|
+| **Fixed Effects**        |               |              |
+|  Intercept               | 17.77 (0.75)  | 17.46 (0.77) |
+|  Time (decade)           | -0.01 (0.17)  | -0.15 (0.17) |
+|  Age(26-33)              | -0.94 (0.35)  | 0.42 (0.35)  |
+|  Age(34-64)              | -2.49 (0.37)  | -0.82 (0.36) |
+|  Age(65+)                | -3.39 (0.48)  | -1.70 (0.44) |
+|  Duration (year)         | 1.45 (0.10)   | 1.56 (0.10)  |
+|  Time\*Age(26-33)        | -0.02 (0.07)  | 0.18 (0.07)  |
+|  Time\*Age(34-64)        | -0.51 (0.06)  | -0.15 (0.06) |
+|  Time\*Age(65+)          | -1.38 (0.07)  | -0.80 (0.07) |
+| Duration\*Age(26-33)     | -0.17 (0.09)  | -0.28 (0.08) |
+| Duration\*Age(34-64)     | -0.38 (0.08)  | -0.49 (0.07) |
+| Duration\*Age(65+)       | -0.51 (0.02)  | -0.58 (0.09) |
+| Duration\*Time           | 0.14 (0.02)   | 0.15 (0.02)  |
+| Ever married             |               | -0.89 (0.21) |
+| Ever had child           |               | 0.26 (0.14)  |
+| Child under 18 in house  |               | 0.30 (0.10)  |
+| Student                  |               | -0.57 (0.24) |
+| Less than HS             |               | 2.80 (0.35)  |
+| College degree           |               | -2.42 (0.24) |
+|                          |               |              |
+| **Random Effects**       | **S.D.**      | **S.D.**     |
+|  Intercept               | 5.99          | 6.24         |
+|  Time                    | 1.32          | 1.31         |
+|  Duration                | 0.50          | 0.50         |
+|  Age(26-33)              | 1.42          | 1.32         |
+|  Age(34-64)              | 2.26          | 1.99         |
+|  Age(64+)                | 3.06          | 2.41         |
+|  Ever married            |               | 1.60         |
+|  Ever had child          |               | 0.88         |
+|  Child under 18 in house |               | 0.67         |
+|  Student                 |               | 1.72         |
+|  Less than HS            |               | 1.04         |
+|  College degree          |               | 2.98         |
 
-: Coefficient Estimates
+: Table \@ref(tab:2): Coefficient estimates from multilevel linear regressions of absolute change between responses on age group, duration, and question.
 

### 2024-07-23 manuscript/manuscript.Rmd
small updates

--- 2024-07-08 manuscript/manuscript.Rmd
@@
 | Variables                      | M1       | M2       | M3       | M4       | M5       |
-|------------|------------|------------|------------|------------|------------|
+|--------------------------------|----------|----------|----------|----------|----------|
 | Intercept                      | Y        | Y        | Y        | Y        | Y        |
@@
 
-| Col1             | Col2  | Col3 | Col4 |
-|------------------|-------|------|------|
-| Fixed Effects    |       |      |      |
-| Intercept        | 17.45 |      |      |
-| Time (decade)    | -0.26 |      |      |
-| Age(26-33)       | -1.60 |      |      |
-| Age(34-64)       | -3.17 |      |      |
-| Age(65+)         | -4.10 |      |      |
-| Duration (year)  | 1.13  |      |      |
-| Time\*Age(26-33) | -0.09 |      |      |
-| Time\*Age(34-64) | -0.54 |      |      |
-| Time\*Age(65+)   | -1.33 |      |      |
-|                  |       |      |      |
-| Random Effects   | S.D.  |      |      |
-| Intercept        | 5.35  |      |      |
-| Time             | 1.14  |      |      |
-| Duration         | 0.37  |      |      |
-| Age(26-33)       | 0.89  |      |      |
-| Age(34-64)       | 1.81  |      |      |
-| Age(64+)         | 2.92  |      |      |
+|                         | Model 1      | Model 2      |
+|-------------------------|--------------|--------------|
+| **Fixed Effects**       |              |              |
+| Intercept               | 17.77 (0.75) | 17.46 (0.77) |
+| Time (decade)           | -0.01 (0.17) | -0.15 (0.17) |
+| Age(26-33)              | -0.94 (0.35) | 0.42 (0.35)  |
+| Age(34-64)              | -2.49 (0.37) | -0.82 (0.36) |
+| Age(65+)                | -3.39 (0.48) | -1.70 (0.44) |
+| Duration (year)         | 1.45 (0.10)  | 1.56 (0.10)  |
+| Time\*Age(26-33)        | -0.02 (0.07) | 0.18 (0.07)  |
+| Time\*Age(34-64)        | -0.51 (0.06) | -0.15 (0.06) |
+| Time\*Age(65+)          | -1.38 (0.07) | -0.80 (0.07) |
+| Duration\*Age(26-33)    | -0.17 (0.09) | -0.28 (0.08) |
+| Duration\*Age(34-64)    | -0.38 (0.08) | -0.49 (0.07) |
+| Duration\*Age(65+)      | -0.51 (0.02) | -0.58 (0.09) |
+| Duration\*Time          | 0.14 (0.02)  | 0.15 (0.02)  |
+| Ever married            |              | -0.89 (0.21) |
+| Ever had child          |              | 0.26 (0.14)  |
+| Child under 18 in house |              | 0.30 (0.10)  |
+| Student                 |              | -0.57 (0.24) |
+| Less than HS            |              | 2.80 (0.35)  |
+| College degree          |              | -2.42 (0.24) |
+|                         |              |              |
+| **Random Effects**      | **S.D.**     | **S.D.**     |
+| Intercept               | 5.99         | 6.24         |
+| Time                    | 1.32         | 1.31         |
+| Duration                | 0.50         | 0.50         |
+| Age(26-33)              | 1.42         | 1.32         |
+| Age(34-64)              | 2.26         | 1.99         |
+| Age(64+)                | 3.06         | 2.41         |
+| Ever married            |              | 1.60         |
+| Ever had child          |              | 0.88         |
+| Child under 18 in house |              | 0.67         |
+| Student                 |              | 1.72         |
+| Less than HS            |              | 1.04         |
+| College degree          |              | 2.98         |
 
-: Coefficient Estimates
+: Table \@ref(tab:2): Coefficient estimates from multilevel linear regressions of absolute change between responses on age group, duration, and question.
 

### 2024-07-29 manuscript/manuscript.tex
updated draft, clean folders

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Minion Pro}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\setlength{\parindent}{4em}
\setlength{\parskip}{0em}
\usepackage{graphicx}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Extended Early Adulthood and Attitude Change},
  pdfauthor={Kevin Kiley, North Carolina State University; Nicolas Restrepo, University of California, Davis},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Extended Early Adulthood and Attitude Change}
\author{Kevin Kiley, North Carolina State University \and Nicolas Restrepo, University of California, Davis}
\date{July 29, 2024}

\begin{document}
\maketitle
\begin{abstract}
Early adulthood is consistently identified as a ``formative period'' in the life course when people demonstrate a heightened probability of changing their opinions, beliefs, and political preferences. This heightened rate of change is often attributed to the social conditions of this life stage but could also reflect ontogenetic differences between young adults and older adults. In this paper, we draw on data for more than 70 questions that were asked repeatedly across multiple panel surveys stretching from the 1950s to the 2020s to explore whether changes in the social conditions of early adulthood have affected the rate at which early adults and the rest of the population change their opinions on political and social issues. Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the late 1950s. Instead, we find that other age groups have become more stable opinion holders over this time. This aggregate trend reflects three countervailing patterns: small decreases in stability due to life-course changes among younger adults, larger increases in stability across all age groups due to increased educational attainment, and a general increase in stability that remains unexplained. Our results suggest broad roles for both individual and larger social contexts in shaping when in the life course people experience attitude change.
\end{abstract}

\doublespace

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Adolescence and early adulthood are frequently identified in social science literature as ``impressionable years'' -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making at later life stages (Krosnick and Alwin 1989; Alwin and Krosnick 1991a). In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages (Kiley and Vaisey 2020; Krosnick and Alwin 1989), and that opinions formed in these age ranges tend to persist well into adulthood (Sears and Brown 2013; Sears and Funk 1999). This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins the dynamics of cohortization that are frequently invoked in larger models of social and cultural change (Ryder 1965; Mannheim 1952; Voas and Chaves 2016).

Why people in these life stages appear to show more frequent opinion change relative to older people, however, is unclear. Social-science explanations tend to argue that the unique social features of different life stages shape whether people are open to or closed off from change during these ages. These features include high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing ``new experiences'' in early adulthood (Alwin and Krosnick 1991a; Danigelis, Hardy, and Cutler 2007) and normative expectations around stability for middle-aged people in high-status roles (Eaton et al. 2009; Visser and Mirabile 2004). At the same time, interdisciplinary work on learning across the life-course finds that an early period of active exploration followed by relative stability is a broader phenomenon across organisms, which might be an evolutionary response to the tension between learning and using information in high-information environments (Gopnik 2020; Frankenhuis and Walasek 2020; Spreng and Turner 2019). These findings suggest that the differences between early adults and the rest of the population might be more rooted in ontogenetic mechanisms than social mechanisms.

It is likely true that both social and ontogenetic mechanisms play a role in facilitating opinion change during adolescence and early adulthood, but clarifying their relative contributions is important for understanding the broader relationship between social change and cultural evolution. If the mechanisms driving heightened rates of opinion change early in adulthood are principally ontogenetic, then changes in the structure of different life stages would be unlikely to affect rates of cultural evolution. If openness to belief and opinion change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing (Cherlin 2004, 2021; Robbins, Dechter, and Kornrich 2022), could shape the relative importance of within-person change and cohort effects for explaining broader cultural change (Kiley and Vaisey 2020; Vaisey and Lizardo 2016).

To better understand the mechanisms that drive openness to cultural change, we take advantage of changes in the social conditions of life stages over time. Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, establishing an independent household, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts than they did for earlier cohorts (Buchmann 1989; Brückner and Mayer 2005). As a result, ``early adulthood'' has emerged as a distinct life stage, in which people are relatively detached from family and other role obligations and have more freedom to explore ideas and lifestyles (Rosenfeld 2009). This extension of the social conditions typically thought to open people up to change should be reflected in how much people change over time.

In this paper, we test whether these shifts in social structure have affected the rates at which people change their opinions about diverse social and political issues during early adulthood. Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we quantify broad trends in attitude change and stability over time. Consistent with the ``impressionable years'\,' hypothesis, we find that early adults (aged 18-25 and 26-33) have more variable opinions than older respondents. However, these differences are not constant across cohorts. They emerged over the second half of the 20th Century because more recent cohorts became more stable opinion holders in middle-age than their predecessors, not because they demonstrated more opinion change in early adulthood than their predecessors.

We show that changes in the timing of life-course transitions do not appear to explain these aggregate patterns of attitude change over time. While decreased rates of marriage in early adulthood explain a small proportion of the aggregate pattern among early adults, the largest driver of change in attitude stability comes from increased levels of educational attainment across all age groups. At the same time, much of the increase in attitude stability over time remains unexplained, suggesting broader social structural features, such as increased access to information with which to form opinions or political and cultural polarization, might also be the key drivers of attitude stability.

\hypertarget{impressionable-years}{%
\section{Impressionable Years}\label{impressionable-years}}

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization (Dinas 2014; Elder 1974; Guhin, Calarco, and Miller-Idriss 2021). These frameworks argue that the circumstances in which people are brought up leave an enduring mark on what Lizardo (2017) calls ``personal culture'' -- worldviews, attitudes, dispositions, behaviors, and preferences that manifest at the individual level. In theories of political socialization, for example, ``generational effects'' are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups (Gerber and Green 1998; Mannheim 1952; Bartels and Jackman 2014). However, the notion of ``impressionable years'' is more general than political preferences. The often-invoked concept of cohort effects rests on the assumption that peoples' early years play a key role in shaping a wide variety of tastes, preferences, habits, and dispositions, which then impacts their choices and life outcomes well beyond the time when this period has passed (Ryder 1965; Fosse 2023; Elder and George 2016). Similarly, practice theories associated with Bourdieu suggest that the ``past conditions of production'' durably shape peoples' cognition, reactions, and behavior well into adulthood (Bourdieu 1990). Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding social theories of behavior.

Evidence for the impressionable years hypothesis exists at the individual and aggregate levels. At the individual level, the ``impressionable years'' hypothesis requires that people demonstrate opinion change early in life followed by a longer period of relative stability, especially in middle age. While different attitudes occasionally follow different patterns, this general trend has been shown to be true for political attitudes including ideology, partisan identification, interest in politics, and specific policy preferences (Krosnick and Alwin 1989; Alwin and Krosnick 1991a; Sears and Funk 1999; Prior 2010); social attitudes around race, gender, and sexuality (Danigelis, Hardy, and Cutler 2007); and other dispositions such as personality traits (Wagner, Lüdtke, and Robitzsch 2019) and cultural tastes (Mulder et al. 2010; Ma 2021). More generally, recent research has found that durable change of social opinion is quite rare among U.S. adults (Kiley and Vaisey 2020). At the aggregate level, cohort effects are consistently strong predictors of a wide variety of social and political attitudes, and Vaisey and Lizardo (2016) find that the year of a person's birth explains more variance in opinion differences than the year of observation for most items in the General Social Survey, suggesting an impressionable-years pattern.

In recent years, researchers have developed new approaches to quantify openness to social influence across the life course, especially around politics (Bartels and Jackman 2014; Ghitza, Gelman, and Auerbach 2022). This work has produced remarkably similar patterns to previous work, finding that openness peaks from a person's mid-teens through age 25 and declines quickly after that, lending more credence to the ``impressionable years'' hypothesis (a pattern that Way (2019) suggests also holds for music tastes). However, this work tends to assume a constant age-based rate of openness across cohorts -- i.e.~that 20-year-olds in 1955 are open to influence at the same rate as 20-year-olds in 2015. This assumption remains untested. Regardless, there is broad evidence for a consistent pattern of malleability early in life followed by relative stability.

\hypertarget{theoretical-models}{%
\section{Theoretical Models}\label{theoretical-models}}

While the empirical pattern of early openness followed by relative stability appears robust, its causes are not clear. Changing political, social, and cultural attitudes involve at least two distinct processes that can be difficult to disentangle in observational data: exposure to new information and change conditional on exposure to new information. For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for the observed age-based patterns in opinion change. Conversely, it could be the case that rates of exposure to new ideas changes over the life course, while openness to changing one's mind as a function of exposure does not, producing the same pattern. It could also be the case that both these processes change as a function of age. Because these mechanisms have not been disentangled and because there are reasons to assume that both processes are operative, in this section we outline theoretical mechanisms that have been suggested to contribute to one or both of these processes.

\hypertarget{ontogenetic-mechanisms}{%
\subsection{Ontogenetic Mechanisms}\label{ontogenetic-mechanisms}}

One class of explanations for heightened rates of attitude change in early adulthood posits that attitudinal stability and change across the life course is primarily a function of humans' cognitive architecture, which remains more or less stable across socio-historical contexts. One class theories posit that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally and are exposed to information at a (more or less) constant rate as they age (Bartels and Jackman 2014; Gopnik 2020; Sherratt and Morand-Ferron 2018). Because people have already encountered a lot of information by the time they reach adulthood, each new piece of information encountered later has less of an impact on their overall dispositions (Bartels and Jackman 2014). From this perspective, the distinctive thing about our early years is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.

A second theoretical framework suggests that development of the human mind changes across the life course and that there is something unique about human cognition during early periods that makes humans ascribe more weight to the information acquired within them than to information acquired at other points in the life course. Maturational arguments suggest that animal brains exhibit more malleability during childhood and adolescence than later in life (Mata and Helversen 2015, Spreng and Turner), and researchers observe evidence of this increased malleability in a wide variety of species. Younger birds, for example, are able to retool their habits in light of new information while older birds were not (Aplin, Sheldon, and McElreath 2017). Evolutionary theories posit that this pattern is a response to conditions in which environments are unstable and information is plentiful yet unreliable (Fawcett and Frankenhuis 2015; Frankenhuis and Walasek 2020; Walasek, Frankenhuis, and Panchanathan 2021), three circumstances common in human social environments. Simulations show that in these circumstances, it makes sense for agents to explore early on, to find something that works, and then to stick to strategies that have worked well in the past.

The thrust of these ontogenetic arguments is that humans have evolved a cognitive architecture that makes them relatively open to changing dispositions at younger ages and relatively resistant at older ages regardless of their contemporaneous social conditions. However, while these mechanisms can shape how humans respond to new information when they encounter it, they cannot dictate the rate at which people are exposed to new information, which is a fundamentally social process.

\hypertarget{life-course-and-social-roles}{%
\subsection{Life Course and Social Roles}\label{life-course-and-social-roles}}

Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct. There are at least four dimensions on which early adulthood differs from other life stages in ways that could produce heightened rates of opinion change, regardless of ontogenetic factors.

First, people in these life stages tend to move across social contexts at a higher rate than people in other age groups. In the 11-year window from ages 14 to 25 in which existing work suggests people tend to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year potentially exposing people to new classmates, teachers and other role models, and sources of information and restructuring core discussion networks (Small, Deeds Pamphile, and McMahan 2015). People in this age range also show the highest rates of geographic mobility (Geist and McManus 2008). Assuming that new social contexts shape exposure to new information, this heightened contextual mobility might explain part of the difference in attitude stability between age groups (Billari and Liefbroer 2010). Social, contextual, and geographic mobility might also free people to adopt new opinions as they move to new contexts and are unburdened by previous friends' or family members' expectations.

Second, in addition to inhabiting a greater diversity of social contexts during these years, people in adolescence and early adulthood spend their time in social contexts explicitly designed to affect attitudes and opinions. Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55. These contexts expose young adults to new information in ways that can lead to new opinions. For example, college attendance appears to change some opinions about civil liberties, politics, gender egalitarianism, racial inequality, and morality (Campbell and Horowitz 2016; Wodtke 2018; Edelmann and Vaisey 2023; Broćić and Miles 2021).

Third, the structure of personal social networks is distinct in this age range, with young peoples' networks tending to be larger and including more friends than kin compared to networks in middle age (Marsden 1987). Having a higher number of ``weak ties'' in their networks can expose younger adults to new information at higher rates (Granovetter 1973; Burt 2005). And because people are less likely to adopt new beliefs if they are embedded in dense networks that reinforce the holding of existing beliefs (Stark and Bainbridge 1980), the denser, more homogenous networks that characterize middle-age (Marsden 1987; DiPrete et al. 2011) might make it more difficult for people in this life stage to update opinions even if they receive new information (Visser and Mirabile 2004). Conversely, young peoples' less-dense networks might allow them to close off parts of their lives to people who might not be supportive of new attitudes.

Fourth, there is significant social variation in norms around when maturity is reached (Cunningham 2007; Elder 1994), and contemporary early adults have little cultural pressure to resemble middle-aged adults. Cultural narratives encourage exploration and self-discovery during early adulthood, pushing young adults to pursue new experiences, meet new people, and adopt new personalities during this ``age of independence'' (Rosenfeld 2009). These cultural expectations might also shape whether people feel capable of updating opinions in the face of new information. People appear to value stability in people holding social roles with authority (Eaton et al. 2009). As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information. This potentially explains why some work finds that people appear to open up to changing their minds as they enter later life, when these social expectations of stability potentially diminish (Danigelis, Hardy, and Cutler 2007).

As the above suggests, life stages vary in myriad ways that could affect how people are exposed to new information and whether they can act on that information in ways that produce changes of opinion. As a result, changes in the duration of these life stages could have significant ramifications for how much opinion change we see in a population.

\hypertarget{changing-life-course}{%
\section{Changing Life Course}\label{changing-life-course}}

The social and cultural conditions of early adulthood have changed considerably in the past seven decades in the United States (Buchmann 1989; Rosenfeld 2009; Rosenfeld and Kim 2005), as expansion of higher education and increased possibility for geographic mobility have reshaped expectations around life-course trajectories (Rosenfeld 2009). While cultural norms previously encouraged marriage, stable employment, and childbearing shortly after the completion of secondary education, relaxation of these norms means that early adulthood now might involve prolonged higher education, moving to a different city, or dating without a stable romantic partner Billari and Liefbroer (2010). There are two related trends that are particularly important: the elongation of young adulthood and the intensification of this life stage.

\hypertarget{timing}{%
\subsection{Timing}\label{timing}}

The period between adolescence and ``settling down'' has gotten longer over time (Billari and Liefbroer 2010). Buchmann (1989) finds that young adults increasingly postpone the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce. In the United States, the median age at first marriage and median age of first childbirth both occur about eight years later than they did in the mid-20th Century (Rosenfeld 2009), and rates of non-marital childbearing have also increased across all demographic groups (Cherlin 2021). Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts than they did for previous cohorts.

Similarly, more recent cohorts spend more years pursuing education, with a larger proportion of the U.S. population completing high school, college, and, increasingly, graduate degrees before entering the workforce (Schofer and Meyer 2005). This shift has been particularly dramatic for women, whose educational attainment and labor-force participation are much higher than 70 years ago (England 2010). People now also move between the workforce and education at higher rates than they did in the past (Rosenfeld 2009).

If the heightened rates of attitude change during peoples' ``impressionable years'' is principally driven by the distinct social circumstances of adolescence and early adulthood, then the delay of full adulthood has the potential to extend these social circumstances into later ages, extending heightened rates of attitude change later in the life course. While people born in the mid-20th century might have displayed mainly opinion stability by the time they reached their mid-twenties, when they married and had children, people from more contemporary cohorts, unmarried and childless at the same ages, might show higher rates of opinion change. Specifically, we expect cohorts who entered adulthood in the late 1950s to be fairly comparable to contemporaneous middle-aged adults in their rates of attitude change, but that these age groups will diverge in their rates of change in more recent years. This leads us to our first hypothesis:

\emph{Hypothesis 1: More recent cohorts will show higher rates of opinion change in the age range of 18 to 25 than cohorts born in earlier years.}

We focus on people 18 to 25 for two reasons. First, existing work has explored attitude stability and change among people in this age range in earlier cohorts (Krosnick 1991; Alwin and Krosnick 1991a), providing a baseline for what to expect. Second, this group has seen the most significant change in life course transitions of all age groups. Whereas most people in earlier cohorts were married with children by the midpoint of this age range, the median age of marriage and first childbirth is now outside this group for both men and women. We expect a similar but more muted trajectory for attitude behavior of people aged 25-33. Again, we expect the oldest cohorts to resemble middle-aged respondents when they enter this age range, as has been shown in previous research, but that more recent cohorts will come to more closely resemble the 18-25 age group. This produces our second hypothesis:

\emph{Hypothesis 2: More recent cohorts will show higher rates of attitude change in the age range of 26 to 33 than cohorts born in earlier years.}

If these differences are principally driven by the structure of the life course at these stages, then we would expect that, controlling for these life-course transitions, the gap between early adults and older adults will disappear. This leads to our third hypothesis:

\emph{Hypothesis 3: Controlling for rates of marriage, childbirth, and having a child at home, more recent cohorts' rates of attitude change in early adulthood (18-33) will not be different from the rates of attitude change among contemporaneous middle aged adults.}

\hypertarget{intensifying-impressionable-years}{%
\subsection{Intensifying Impressionable Years}\label{intensifying-impressionable-years}}

Rosenfeld (2009) argues that one of the processes underlying the rapid rise in same-sex and interracial marriages is that young adults have increasingly escaped their parents' authority. When a young person lives with, or is financially dependent on, his or her parents, the latter have more say about their child's choice of partner. Social and cultural changes that have increased geographic mobility and independence from parental authority mean people in this life stage have been able to engage in courtship outside the gaze of their parents, making choices that would not have been sanctioned by them.

This work suggests that increased independence might enable greater exploration and freedom to update attitudes beyond that facilitated life-course transitions. This framework suggests that freedom of exploration is not just dependent on life-course roles (marriage, parenthood, etc.) but also influenced by broader cultural expectations. In other words, married 25-year-old men and women in 1956 might have experienced an additional set of social constraints that married 25-year-old men and women in 2016 do not experience, leaving the latter much more open to developing new opinions. The ``intensification'' of impressionable years suggests a fourth hypothesis that challenges the third:

\emph{Hypothesis 4: Controlling for life course transitions, more recent cohorts of early adults (18-33) will still make higher rates of attitude change than contemporaneous middle-aged adults.}

\hypertarget{broader-social-factors}{%
\subsection{Broader Social Factors}\label{broader-social-factors}}

We have focused primarily on changes in the timing and structure of early adulthood over the past 70 years, but that time frame has also seen significant shifts in the broader social and cultural environment that might affect attitude stability independent of changes in life course structure.

First, the number of people completing high school and pursuing higher levels of education increased significantly during the 20th and early 21st Centuries (Schofer and Meyer 2005; Hout 2012). Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts (Alwin and Krosnick 1991b; Alwin 2007). The specific mechanism that links educational attainment and attitude stability is unclear, with some perspectives suggesting educational attainment improves cognitive function and memory while other perspectives suggest that educational attainment makes people pay more attention to current events. It could also be the case that education cultivates in people a normative expectation that they should hold stable opinions. Regardless of the specific mechanism, people with higher levels of education might be more likely to hold stable opinions regardless of where they are in the life course, and differences in educational attainment across age groups could produce differences in attitude stability. This leads to our fifth hypothesis:

\emph{Hypothesis 5: Cohorts with higher levels of education will demonstrate higher levels of attitude stability at all age ranges.}

Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding. Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire information that makes stable opinion holding easier. To the extent that stable opinions are in part a reflection of signals from elites (Converse 1964; Zaller 1992), then we should expect these shifts to stabilize opinions across all age ranges.

Third, the past 70 years have seen elite polarization around political and social affairs, as well broader public partisan realignment (Fiorina and Abrams 2008; Baldassarri and Gelman 2008). People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities ``go together'' in the population, resulting in what DellaPosta (2020) calls an ``oil spill'' of polarization. While clarity around political issue positions is most commonly studied, the same forces can also align other social beliefs and identities. Geographic and social sorting into lifestyle ``enclaves'' also reinforce these processes, continually providing clearer signals about which opinions people should hold (DellaPosta, Shi, and Macy 2015). This issue and identity alignment might make it easier for the public to identify what beliefs people like them ``should'' hold and hold them consistently (Zaller 1992).

Broadly speaking, these forces suggest that all age groups' opinions will be more stable in successive cohorts, regardless of life-course stages. If this is the case, we might not observe an absolute decrease in stability for early adults, but rather a relative decrease, as extension of the life course works to decrease stability for members of these age groups while broader social forces stabilize opinions for all groups. This produces our final hypothesis:

\emph{Hypothesis 6: At all ages, more recent cohorts will demonstrate higher levels of attitude stability, controlling for educational attainment and life-course transitions, than earlier cohorts.}

We have outlined hypotheses derived from the theoretical model that suggests that social conditions are the key driver of higher rates of attitude change among early adults and adolescents. To the extent that we observe significant changes in the social context of early adulthood but no corresponding changes in rates of attitude change or response stability across cohorts, we would take this as evidence that social circumstances play a relatively small role in explaining rates of attitude change compared to ontogenetic factors.

\hypertarget{analytical-strategy}{%
\section{Analytical Strategy}\label{analytical-strategy}}

Our hypotheses can be adjudicated with a set of theoretical estimands (Lundberg, Johnson, and Stewart 2021). Our comparisons suggest a set of theoretical estimands regarding the difference in opinion change for Americans in the same age range born in different cohorts. We seek to compare the average amount of opinion change we observe for a particular cohort during a particular age range to the average amount of opinion change for a cohort born 10 years later, conditional on age range. To be clear, we do not seek to estimate this quantity assuming only the year someone was born is different and holding all other factors constant. It is impossible to separate out the age at which people of a certain cohort answer questions from the year in which they answer them (Fosse and Winship 2019). Cohorts also differ in other ways, including size and demographic composition, that might be relevant for explaining differences in the rates of change they demonstrate. This set of theoretical estimands is principally descriptive, with variation in the social experiences of people in successive cohorts and even the demographic composition of the cohort itself being potential explanations for whatever gaps we observe.

\begin{equation*}
 \begin{aligned}
  \tau_1 = \frac{1}{n}\sum_{c=1}(Y_{c}-Y_{c+10})
 \end{aligned}
 \end{equation*}

In this theoretical estimand, \(Y_{c,a}\) captures the expected amount of change in responses for a cohort at age range \(a\), and \(Y_{c+10,a}\) captures the expected amount of change in responses for a cohort born 10 years later at the same age range, with a fixed duration between observations. We seek to average the quantity across cohorts captured in panel surveys from the 1950s onward.

Our second set of theoretical estimands captures how differences in the rate of change between age groups itself changes over time, specifically whether the gap between how much early adults (either 18-25 or 26-33) change over time and how much middle-aged adults (34-64) change over time itself is different in different years. Again, we do not assume that the composition of any age group is fixed, nor that later cohorts experience the same social environment as earlier cohorts.

\begin{equation*}
 \begin{aligned}
   \tau_2 = \frac{1}{n}\sum_{t=1}([Y_{t,\text{ma}} - Y_{{t},\text{ea}}] - [Y_{{t-10},\text{ma}} - Y_{{t-10},\text{ea}}])
 \end{aligned}
 \end{equation*}

In this estimand, \(Y_{t,\text{ea}}\) reflect the average change in responses for early adults at a particular time \(t\) and \(Y_{t,\text{ma}}\) reflects the average change in responses for middle aged respondents at the same time. \(Y_{t,\text{ea}}\) and \(Y_{{t-10},\text{ea}}])\) reflect these same rates observed in the population 10 years earlier. In contrast to the previous estimand, this estimand is indexed by t, or time in years, not cohort, and we average over all times captured since the late 1950s.
Finally, hypotheses 3, 4, 5, and 6, imply theoretical estimands reflecting the difference between cohorts as they pass through the same age range, conditional on cohorts having the same levels of certain demographic characteristics. Specifically, they focus on whether differences in life course transitions and educational attainment mediate the observed differences between cohorts at different time points.

\begin{equation*}
 \begin{aligned}
   \tau_3(e,l) = \frac{1}{n}\sum_{c=1}(Y_{c,a}(E=e,L=l)-Y_{c+10,a}(E=e,L=l))
 \end{aligned}
 \end{equation*}

This framework repeats that of \(\tau_1\), comparing a focal cohort to the cohort that came 10 years later, while holding both cohorts' levels of educational attainment, \(E=e\), and life course transition rates, \(L=l\) to the level of the earlier cohort. We again average this difference across all cohorts observed since the late-1950s. In comparison to \(\tau_1\) and \(\tau_2\), however, this estimand requires a counterfactual component that is not directly observed.

\hypertarget{data}{%
\subsection{Data}\label{data}}

Generating empirical estimands for these theoretical estimands requires data on opinion and attitude reports over time for members of different cohorts in the same age ranges, with respondents coming from the same general population. We also require information about life-course transitions standardized across panels. To construct such a data source, we combine panel surveys conducted by the American National Election Studies (ANES) and the General Social Survey (GSS) spanning about 60 years. Both the ANES and GSS are designed to generalize to the non-institutionalized U.S. adult (18+) population, and therefore can be thought of as tapping the same general population. We included 12 panels in our final dataset: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20.

For reasons related to our modeling strategy explained below, we identified social and political attitude questions asked in at least three different panels with at least one of those panels asking respondents to answer the question at three or more time points. Questions had to ask respondents to report their own opinions, not the positions or opinions of others. Because the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same questions, we did not treat these as separate panels for the purposes of identifying three distinct time points. Questions needed to appear in these panels and an ANES panel to be included in our analysis. We allowed for question wording to vary slightly across panels (e.g., feeling thermometers about the ``women's liberation movement'' were combined with feeling thermometers about the ``women's movement''). When similar questions had different response scales, we harmonized response options to the question structure with the fewest response options. This was most common when one panel asked whether people agreed or disagreed while another included options for ``strong'' agreement and disagreement. We followed the ANES time series Cumulative Data File codebook in deciding which questions to harmonize.

Our search produced a total of 77 questions tapping political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views of government efficacy and trust in government, views on social issues and social change, and sentiment toward various political and social groups. For reasons elaborated below, we scaled items to have a minimum of 0 and a maximum of 100. Table A1 in Appendix A outlines the questions included in our analyses, including their full question wording, number of waves and panels, and response scales.

\hypertarget{opinion-change}{%
\subsubsection{Opinion Change}\label{opinion-change}}

Our outcome of interest is the amount of change in a person's responses across two instances of being asked the same question, meaning our unit of analysis is a pair of waves by the same person, rather than a single observation or a single person. If a person responds to a question K times in a panel, he or she will have \((K*(K-1))/2\) observations for that item in our data set. Because we are unconcerned with the direction of change between responses, we take the absolute value of the difference between each two observations of a person's response to the same question. Because all questions are scaled between 0 and 100, this takes on a value between 0 and 100.

\hypertarget{demographic-measures}{%
\subsubsection{Demographic Measures}\label{demographic-measures}}

The central questions in our analysis pertain to age, cohort, and the time of the interview. All surveys included the date when the survey interview took place.\footnote{In the rare cases that a survey interview for a single wave of the survey occurred on two distinct days, we classified the date according to the day the interview started.} Age was reported at least once in all panels. If respondents' were not asked their age or birth date in a specific wave, we calculated it based on other age reports and the date of the interview. We classify respondents into a set of age brackets depending on their age when they answered the survey questions: 18-25, 26-33, 34-64, and 64-80. Because of differences in the wording of age questions across panels, we dropped respondents who were older than 80 in the first observation of a wave pair.

Our second set of theoretical questions focuses on the relationship between life-course transitions and attitude change. Questions tapping marital status, student status, and degree attainment were asked in similar ways and could be easily harmonized. We generated indicator variables capturing whether the respondent is currently enrolled in school and has ever been married. We also include indicators for whether a respondent reported completing college or did not complete high school. It was common for respondents to have missing data on one or more of these demographic variables for a particular wave of any panel, even if they participated in the wave. This was most common when a question was asked in the pre-election wave of the ANES but not in the post-election wave. When this occurred, we carried over the pre-election response, as these waves tended to be separated by only a few weeks. If these questions were missing from a wave and someone had previously reported ever being married or ever having a child, we carried over that report to subsequent waves. If their status at the time of an interview was unclear, we coded the response as missing.

Generating an indicator for whether a respondent ever had a child was more challenging, as question wording and structure varied across panels. Some waves asked respondents how many children were currently living with them while others asked whether the respondent had any children and how many. To account for this, we generated two variables. We created an indicator variable for whether a respondent reported having at least one child in the house under 18. Respondents who were not asked the question in a wave (or temporally proximate wave) were coded as missing. Second, we gave respondents an indicator for whether they had ever had a child. We coded this as 1 if respondents were asked if they ever had a child and reported that they did or if they were asked if they had a child living with them and reported that they did. We coded this as 0 if respondents were asked whether they ever had a child and said no. Respondents were coded as missing if they were asked how many children under 18 they had living with them and said none, or if they were not asked either relevant question in a particular wave or temporally proximate wave.

We imputed missing demographic data for respondents using multiple imputation with chained equations. Multiple imputation approaches the question of missing data as one of prediction. It uses existing relationships in the data to create a predicted value for missing data, which is consistent with the patterns in the data. But it recognizes that there is uncertainty in these predictions, so it iterates over this process several times. Different imputed values reflect uncertainty in the predictions that can be drawn from relationships in the data. In the analysis stage, all predicted values are used to incorporate the uncertainty in the imputed values. We used other demographic variables as well as attitude difference reports to impute missing values, though no outcomes were imputed this way. We also included sex (male or female) and race (white, black, or other), both recorded in all waves, as predictors in these imputation equations. We created a set of 5 different imputed datasets to acknowledge uncertainty in this process, and our final analyses are conducted across all imputed datasets.

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in overall attitude stability (our main theoretical interest) or to the structural features of the survey. Centrally, the amount of change people make in their responses over time is likely affected by the duration of time between responses, with longer duration between responses leading to more wave-to-wave change as people experience more shocks that could shift their opinions. In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years. Accounting for these differences is a key part of our analytical approach.

Consider a one-item, one-panel case (e.g., partisan identification in the 1972-76 panel). We model the absolute change between two responses to the same question by the same person as a function of an intercept and the duration between observations in years. In this equation, \(i\) indexes a pair of observations, not a person.

\begin{equation*}
 \begin{aligned}
   |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \epsilon_i
 \end{aligned}
 \end{equation*}

In this equation, \(d_{i}\) captures the duration of time between \(t_1\) and \(t_2\) in years for each pair of observations by the same respondent. Importantly, respondents might have multiple observation pairs for each panel depending on the number of times they answered a question. In the 1972-76 ANES panel respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart. If a person participates in all three waves, they would have three observations of \(|y_{2i}-y_{1i} |\) and \(d_i\): 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).

Variation in duration between observations within panels lets us separate two kinds of change: 1) change that is principally random, or unrelated to the duration between observations, captured in \(\beta_0\), and 2) change associated with duration between observations, captured in \(\beta_1\). If respondents have a stable underlying partisan identification they report with random error and no real change occurs between waves beyond that, duration should be unrelated to absolute change in responses over time in expectation, and \(E(\beta_1) = 0\). If a respondent shifts from ``strong Democrat'' (point 0 on the scale) to ``Independent'' (point 3 on the scale) between 1972 and 1974, and stays at ``Independent'' for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for \(\beta_1\).

Given our focus on age-based differences, we extend this equation to incorporate age groups as predictors of absolute difference:

\begin{equation*}
 \begin{aligned}
   |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \\
   \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \epsilon_i
 \end{aligned}
 \end{equation*}

In this framework, \(\beta_0\) and \(beta_1\) now capture the expected change in responses that is unrelated to time and each additional year, respectively, for respondents ages 18-25 (the reference category). \(\beta_2\) through \(\beta_4\) capture the difference in baseline change between observations for the older groups and \(\beta_5\) through \(\beta_7\) capture differences in the effect of an additional year duration for the older age groups.

Our theoretical questions deal primarily with whether these relationships change over time and whether differences between groups change across the broader time span captured across panels. To explore whether the rate of change changes over time, we add an additional term for time (in decades) and interact all previous coefficients outlined above with time.

\begin{equation*}
 \begin{aligned}
   |y_{2i}-y_{1i} | = \beta_0 + \beta_1(d) + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \\ 
   \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \\
   \beta_8(t) + \beta_9(d*t) + \\
   \beta_{10}(a_{26-33}*t) + \beta_{11}(a_{34-64}*t) + \beta_{12}(a_{65+}*t) + \\
   \beta_{13}(d*a_{26-33}*t) + \beta_{14}(d*a_{34-64}*t) + \beta_{15}(d*a_{65+}*t) + \\
   \epsilon_i
 \end{aligned}
 \end{equation*}

In this formulation, \(\beta_0\) through \(\beta_7\) retain their general meaning as outlined above, but apply to \(t = 0\), which we set to be the most recent observation date (2020). \(\beta_8\) estimates the change in the baseline rate of change for the reference group. \(\beta_9\) captures change in the effect of duration of observation over time for people in the reference group (ages 18-25). Coefficients \(\beta_{10}\) through \(\beta_{12}\) and \(beta_{13}\) capture how differences in the baseline rate of change and duration effects between the reference group and older age categories change over time.\footnote{We index time by the year of observation in decades, but it would be functionally similar to index time by cohort.}

To this point, we have developed our model for the case of a single question. While a small number of questions are asked in almost every panel, most are only asked in a few panels. As such, separate estimates of time trends for all items are likely to have a low degree of reliability. One solution is to pool questions and estimate a single time trend for stability in attitudes over time. That approach is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels. It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable. Multilevel models (Gelman and Hill 2007; Raudenbush and Bryk 2002) provide an alternative solution to estimate variation in time trends in the presence of uncertainty. We follow this approach, using the model:

\begin{equation*}
 \begin{aligned}
   |y_{2ij}-y_{1ij} | = \beta_{0j} + \beta_{1j}(d) + \beta_{2j}(a_{26-33}) + \beta_{3j}(a_{34-64}) + \beta_{4j}(a_{65+}) + \\ 
   \beta_{5j}(d*a_{26-33}) + \beta_{6j}(d*a_{34-64}) + \beta_{7j}(d*a_{65+}) + \\
   \beta_{8j}(t) + \beta_{9j}(d*t) + \\
   \beta_{10j}(a_{26-33}*t) + \beta_{11j}(a_{34-64}*t) + \beta_{12j}(a_{65+}*t) + \\
   \beta_{13j}(d*a_{26-33}*t) + \beta_{14j}(d*a_{34-64}*t) + \beta_{15j}(d*a_{65+}*t) + \\
   \epsilon_i
 \end{aligned}
 \end{equation*}

Where:

\[ \beta_{kj} = \gamma_k + \upsilon_{kj} \]

\[ \gamma_k \sim \mathcal{N}(0,\sigma_{\gamma_k}^2) \]

\[ \epsilon_i \sim \mathcal{N}(0,\sigma_{\epsilon}^2) \]

In this framework, the first-level equation models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above. These equations are nested at the level of the survey item. This approach produces an aggregate estimate for each coefficient across all questions (\(\gamma_k\)) while accounting for variation in which questions are asked at different time points. It also allows for each survey question to have its own set of coefficient estimates weighted by the amount of information we have about the particular question. Questions asked in many panels, including partisan identification (asked in a total of 32 waves across all panels) and ideological identification (asked in a total of 28 waves across all but one panel), will contribute more to the overall calculation of the aggregate trend in opinion stability and age differences than a question asked in only three panels. Given the lack of confidence in each individual question's estimates, we focus the majority of our attention on \(\gamma_k\) estimates.

The full model outlined above is likely too complicated given our data limitations, and some interaction terms or random effects might not be necessary if certain effects or differences are constant over time, do not vary by age group, or are similar across questions. For example, it could be the case that little predictive information is gained by allowing questions to have separate slopes for the relationship between age categories and duration over time. We compare the fully elaborated model to more parsimonious models using model-selection criteria, specifically the Bayesian information criterion (BIC) and Akaike information criterion (AIC) (Raftery 1995).

Coefficients generated in this model can be used to provide empirical estimates of our theoretical estimands. In this framework, \(\gamma_{8}\) estimates the expected difference in absolute change for someone ages 18-25 born in a particular year and someone born 10 years earlier at the same age bracket, providing a quantification of \(\tau_1\). Similarly, \(gamma_{11}\) captures the change in the difference between the youngest age group (18-25) and the middle age group (34-64) for a 10-year difference in time, providing a quantification of \(\tau_2\).

Once we select the best-fitting model to capture trends in attitude change over time, we add variables capturing demographic transitions and educational attainment. For each wave pair, we include measures of demographic status at the first observation for a pair of observations.\footnote{For example, if a respondent got married between waves, they are counted as ``never married'' for that pair of observations, since they are unmarried at time 1. If they are observed a third time, they would be counted as having ever been married.}

\begin{equation*}
 \begin{aligned}
   |y_{2ij}-y_{1ij} | = \beta_{0j} + \beta_{1j}(d) + \beta_{2j}(a_{26-33}) + … + \beta_{15j}(d*a_{65+}*t) + \\ 
   \beta_{15j}(d*a_{65+}*t) + \beta_{16j}(evermarried) + \beta_{17j}(everkid) + \\
   \beta_{18j}(kidinhome) + \beta_{19j}(student) + \beta_{20j}(<HS) + \\
   \beta_{21j}(BA) + \epsilon_i
 \end{aligned}
 \end{equation*}

These variables are all centered at their grand means, and like coefficients for age group, time, and duration, are allowed to vary by question. Coefficients produced in this model allow us to generate an estimate of \(\tau_3\), or what a cohort would look like if we set their rates of marriage, parental status, and educational attainment to that of another cohort.

Most panels include weights that account for differential sampling and panel attrition. When available, we use panel weights that ensure that the second wave of the pair of observations is reweighted to make the sample representative of the general population. The 1980 Major Panel Study does not include weights, and as such all respondents in that panel are weighted equally.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{life-course-changes}{%
\subsection{Life Course Changes}\label{life-course-changes}}

Figure 1 plots the proportion of respondents who are married, who reported ever having a child, who reported having a child at home, and who are currently students for each panel wave included in our aggregate data set. Figure 1 also includes the proportion of respondents at two education levels: those with less than a high school education and those with a college degree (with the remaining population being those with a high school degree but nothing more).

\begin{figure}[]
\begin{center}
\caption{Proportion of respondents in different age categories reporting life course milestones or levels of educational attainment, by panel wave.}

\includegraphics[width=1\linewidth]{../figures/demo_variables}

\end{center}
\footnotesize{\textit{Notes: Proportions calculated using imputed data.} }
\end{figure}

In the 1956-60 ANES panel, respondents in the youngest age groups are equally likely to be married, to have had children, and to be students as middle-aged respondents. At the same time, despite being younger, respondents in the 18-25 and 26-33 age groups had higher levels of educational attainment, on average, than their older peers, indicated by a lower proportion having less than a higher school degree (meaning more had a high school degree). People in the 18-25 age group diverged from the others by the 1970s, displaying a lower probability of being married and lower probability of having children, a trend that continues through the most recent panels. The next youngest age group (26-33) also diverged from the middle age group, especially in their rates of marriage and having children, but did so more slowly.

Figure 1 also shows an increase in the proportion of middle age and older age groups who complete high school and college. By the mid-1990s, the proportion of middle aged and older respondents who had college degrees significantly exceeded the proportion of the youngest age group who had college degrees, likely because many people aged 18-25 were still completing education.

These trends set the stage for what to expect in terms of opinion change for different age groups. If marriage and having children serve as proxies for broader life course factors that influence opinion stability, we should see the youngest age group look similar to the older age groups in the 1950s. We then would expect this younger group to diverge from older groups in terms of attitudinal stability as time passes. Respondents in the 26-33 age group should also start similar to middle-aged respondents and diverge, albeit more slowly. To the extent that educational attainment is related to attitude stability, we should observe increased attitude stability among all groups, but especially among middle aged and older respondents who have seen the most dramatic increases in educational attainment.

\hypertarget{opinion-changes}{%
\subsection{Opinion Changes}\label{opinion-changes}}

Comparing the fully elaborated model outlined above to more parsimonious models, the best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for higher-order interactions. This model also omits the three-way interactions between time, duration of observation, and age groups. Further attempts to simplify the model, including dropping two-way interactions or random effects for age categories and duration, resulted in increases in the BIC.

Model 1 in Table 1 presents the main effects and variance components of the best-fitting multilevel model presented above. In all models, the 18-25 age group serves as the reference category, and the time variable is centered on the most recent wave pair (starting in 2020), so coefficients reflect expectations for the amount of change observed in the most recent years, rather than at earlier time points.

\pagebreak

\begin{longtable}[]{@{}lll@{}}
\caption{Coefficient estimates from multilevel linear regressions of absolute change between responses on age group, duration, and question.}\tabularnewline
\toprule\noalign{}
& Model 1 & Model 2 \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& Model 1 & Model 2 \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Fixed Effects} & & \\
Intercept & 17.77 (0.75) & 17.46 (0.77) \\
Time (decade) & -0.01 (0.17) & -0.15 (0.17) \\
Age(26-33) & -0.94 (0.35) & 0.42 (0.35) \\
Age(34-64) & -2.49 (0.37) & -0.82 (0.36) \\
Age(65+) & -3.39 (0.48) & -1.70 (0.44) \\
Duration (year) & 1.45 (0.10) & 1.56 (0.10) \\
Time*Age(26-33) & -0.02 (0.07) & 0.18 (0.07) \\
Time*Age(34-64) & -0.51 (0.06) & -0.15 (0.06) \\
Time*Age(65+) & -1.38 (0.07) & -0.80 (0.07) \\
Duration*Age(26-33) & -0.17 (0.09) & -0.28 (0.08) \\
Duration*Age(34-64) & -0.38 (0.08) & -0.49 (0.07) \\
Duration*Age(65+) & -0.51 (0.02) & -0.58 (0.09) \\
Duration*Time & 0.14 (0.02) & 0.15 (0.02) \\
Ever married & & -0.89 (0.21) \\
Ever had child & & 0.26 (0.14) \\
Child under 18 in house & & 0.30 (0.10) \\
Student & & -0.57 (0.24) \\
Less than HS & & 2.80 (0.35) \\
College degree & & -2.42 (0.24) \\
& & \\
\textbf{Random Effects} & \textbf{S.D.} & \textbf{S.D.} \\
Intercept & 5.99 & 6.24 \\
Time & 1.32 & 1.31 \\
Duration & 0.50 & 0.50 \\
Age(26-33) & 1.42 & 1.32 \\
Age(34-64) & 2.26 & 1.99 \\
Age(64+) & 3.06 & 2.41 \\
Ever married & & 1.60 \\
Ever had child & & 0.88 \\
Child under 18 in house & & 0.67 \\
Student & & 1.72 \\
Less than HS & & 1.04 \\
College degree & & 2.98 \\
\end{longtable}

The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to change by almost a fifth of the response scale on average (about 18 points) in the most recent years of the survey on any particular question, regardless of the duration between observations. To be clear, very few wave pairs have a duration close to 0, and this coefficient primarily reflects the fact that this amount of absolute change will be present regardless of whether responses are 2 or 4 years apart. This high level of change is consistent with previous studies that find that survey respondents tend to make large, seemingly random, wave-to-wave changes (Ansolabehere, Rodden, and Snyder 2008; Converse 1964). At the same time, random-effect variance around this intercept term is large (\(\sigma_{\beta_0} = 5.99\)). On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others people tend to respond with the same answer over time with low levels of measurement error or random change.

Duration between observations is positively associated with wave-to-wave absolute change for the youngest age group (duration coefficient = 1.45), and while the interaction between duration and age group is negative for older age groups, duration is still positively related to the amount of change for all. Random effect variance around this main effect of duration is relatively small (s.d. = 0.50), meaning each one-year duration only increases the amount changed by between .5 and 1.5 for most questions, and it is never negatively associated with change. Given the high level of wave-to-wave change independent of duration, our results reinforce long-standing findings that most wave-to-wave change is non-persisting or measurement error (Converse 1964; Zaller 1992; Kiley and Vaisey 2020; Lersch 2023).

The coefficients for age groups other than the reference category suggest that, on average, older age groups are more stable than people in the 18-25 range in the most recent panels. In these panels, respondents in the 26-33 age group are about 1 point more stable on average, respondents in the 34-65 age group are about 2.5 points more stable on average, and respondents in the 65 and over age group are about 3.4 points more stable on average, than respondents in the 18-25 age group. The negative coefficients attached to the interactions between duration and age categories further compound this difference, as longer durations produce more change among the younger respondents than older respondents. This provides broad evidence for the ``impressionable years'' hypothesis, with the youngest age group being more likely to change than others, followed by the 26-33 age group.

Turning now to how these groups have changed over time, the coefficient for time in decades for people in the 18-25 age group (\(\gamma = -0.01\), \(t = 0.75\)) is substantively small and not statistically significantly different from 0. Question-level random effects vary considerably around this main effect, with a standard deviation of about 1.43, meaning the overall flat trend obscures large question-level differences, with responses to some questions becoming more stable and responses to others becoming less stable. We consider these question-specific trends later. It is important to note that because the effect of duration also increased over time (duration*time coefficient = 0.14), the most recent cohorts to pass into the 18-25 age range are expected to make slightly more change over time in recent years, on average, than previous cohorts did when they were in this age range, but the trend is not pronounced.
The extended early adulthood framework hypothesized that the 26-33 age group would come to resemble the youngest age group over time. The interaction between year of observation and the 26-33 age category is small and not statistically significant (\(\gamma = -0.02\)), suggesting that while people in this age group are more stable, on average, than the youngest respondents, more recent cohorts are not any more or less likely to make changes of opinion in this age range than earlier cohorts and not any more or less similar to the youngest adults (18 to 25) than earlier cohorts.

Before concluding that this stability of rates of attitude change across cohorts of early adults reflects a lack of importance of life course shifts, it is important to understand the full picture of change over time. Model 1 shows that, in contrast to people in the 18-25 and 26-33 age ranges, whose rate of opinion change was relatively stable across cohorts, older age groups (34-64 and 65+) show a notable negative time trend across cohorts. Both time trends are negative, substantively large (-0.5/decade for middle age and -1.3 for old age), and statistically significant, meaning more recent cohorts are more stable opinion reporters while in these age ranges than earlier cohorts were. To more clearly illustrate these trends, Figure 2 plots the overall predicted level of change for respondents in each age category over time, assuming a two-year duration between observations (the modal duration between waves). It also includes separate trajectories for each individual question to highlight variation in the trajectories across these items for the different age groups.

\begin{figure}[]
\begin{center}
\caption{Predicted absolute wave-to-wave change over time for respondents in different age categories.}

\includegraphics[width=1\linewidth]{../figures/trajectory_plot}

\end{center}
\footnotesize{\textit{Notes: Bold line plots average trajectory across all questions. Thin lines plot individual question trajectories. Expected values predicted with duration = 2, the modal duration between repeated observations across all panels.}}
\end{figure}

Figure 2 shows that because of the interaction between duration and time, more recent cohorts show more change over time in the 18-25 and 26-33 age ranges than previous cohorts. In the earliest panels, the oldest respondents (65 and over) were substantially less stable opinion holders than younger respondents, changing by about 5 points more than members of the other groups, while the remaining three age groups (18-25, 26-33, and 34-64) appeared to change at approximately equal rates (about 20 points over a two-year duration). Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable, with the latter becoming almost 10 points more stable on average from 1956 to 2020. Because, as noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change across cohorts, they become the least stable respondents by the most recent panels.

Figure 2 also highlights the variance of trajectories for individual questions, especially for young adults. While older respondents' slopes for each question are almost uniformly negative, there is much more variation for respondents in trajectories for younger age ranges. We return to these differences below.

While Model 1 suggests that there has not been a substantial increase in attitude change over time for the youngest age groups, as expected by the theoretical framework, the increase in stability for middle-aged and older respondents suggests that broad social forces could be counteracting the influence of delayed life-course transitions for members of these age groups. Most centrally increased educational attainment over time might obscure effects of delayed life-course transitions.

Model 2 re-estimates Model 1 incorporating a series of indicator variables for whether respondents were students, whether they had children, and whether they were married, and for two levels of education (less than a high school degree and completed college). These results are presented as Model 2 in Table 1 above.

These results provide mixed evidence that life-course shifts have a meaningful effect on attitude stability. First, the gaps between age groups become smaller and the time trends become weaker when these demographic factors are included in the model, suggesting they account for part of the overall picture. Respondents who have been married are about one point (-0.89) more stable on average than respondents who were never married, but respondents who have had children (0.26) or currently have children at home (0.30) are less stable than respondents who never had children. Taken together, these coefficients suggest that changes in the timing of life course transitions are likely insufficient to explain the divergence between young adults and older adults in terms of stability.

On the other hand, coefficients related to education show a clearer picture. Higher levels of educational attainment are uniformly associated with increased attitude stability, with having less than a high school degree increasing absolute change by about 2.8 points on average and having a college degree increasing absolute change by about 2.4 points on average, relative to having only a high school education, suggesting that changing levels of education in the population -- and the increased education level of middle-aged respondents relative to younger respondents -- likely accounts for a larger share of the divergence in stability.

Using these estimates, we calculate the expected value of change under the assumption that life-course transitions and educational attainment stayed at the same rates they did in the late 1950s. These estimated counter-factual trends are plotted in Figure 3.

\begin{figure}[]
\begin{center}
\caption{Observed trajectories of wave-to-wave response change over time (solid lines) compared to counterfactual trajectories assuming no change in rates of marriage, childbearing, student status, and educational attainment from 1956-60 panel (dashed lines), for each age group.}

\includegraphics[width=1\linewidth]{../figures/counterfactual_plot}

\end{center}
\footnotesize{\textit{Notes: Predictions generated using Model 2 coefficients. Covariate values set to average proportions from 1956-60 ANES panel and 2016-20 GSS panel. Expected values predicted with duration = 2, the modal duration between repeated observations across all panels.}}
\end{figure}

Assuming that rates of marriage, student status, and educational attainment remained constant at their average 1956-60 level over time for each age group, all groups would have been expected to display more change over time than we observe, a reflection of the overall increase in stability due to increased educational attainment. Both middle aged and older respondents would still be more stable in recent years than they were in the earliest years, but, consistent with the theory of extended early adulthood, the youngest respondents and those in the 26-33 age range would have been less consistent opinion holders in more recent years than they actually were.

Figure 3 suggests that most of the change in opinion stability over time is due to shifts in educational attainment. This is clearly illustrated by older and middle aged adults -- who saw the largest shifts in educational attainment but minimal shifts in other demographics -- also having the largest shifts from their late-1950s baselines. However, there do appear to be additional stabilizing forces that principally affected middle-aged and older adults beyond educational attainment. It could be the case that the three-category measure of education is insufficient to capture the full range of educational attainment shifts among people in this group. However, it could also be that other forces also drove attitude stability in this period.

\hypertarget{question-differences}{%
\subsection{Question Differences}\label{question-differences}}

As noted above, Model 1 suggests that the overall trajectory of increasing attitude stability masks heterogeneity across questions and age groups, though the trajectories for individual questions are estimated with a low degree of confidence. To illustrate these different trends for different questions, Figure 4 plots the estimated trajectories for each item for respondents in the youngest (18-25) and middle-aged (34-64) age categories, holding constant duration at 2 years. Items are ordered by their predicted time trends. It is important to note that most of these trajectories are estimated with a low degree of confidence, and that the linear trends for each question are likely not accurate outside of the windows in which we observe the questions.

\begin{figure}[]
\begin{center}
\caption{Expected rate of change for two-year duration, by year, for ages 18-25 (pink lines) and 34-64 (blue lines).  }

\includegraphics[width=1\linewidth]{../figures/question_plot}

\end{center}
\footnotesize{\textit{Notes: Predicted values generated using coefficients in Model 1. Expected values predicted with duration = 2, the modal duration between repeated observations across all panels.}}
\end{figure}

It is hard to draw broad conclusions about which items become more stable and which become more variable over time. Several questions showing the largest increases in stability deal with social issues, including whether homosexuals should be protected from discrimination, whether men and women should have equal roles in society, and whether the enduring effects of slavery and discrimination continue to hinder Black Americans' social and economic progress. A related set of questions about equality in America have also decreased in within-person change over time. Many general political views, such as support for capital punishment, whether the government should help people cover medical costs, and spending on the environment, have also decreased in change over time. Some of these shifts reflect the emergence of social consensus, as observed on several ``culture war'' issues like gay marriage (Baldassarri and Park 2020), while others reflect clearer partisan polarization with elites of each party providing clearer signals about which positions should be held by partisans (Kozlowski and Murphy 2021).

A set of items tapping political efficacy (``ppllikeme''; ``manyvote''; and ``complicated'') show increased change over time. Feeling thermometers about social groups, including racial groups, the poor, homosexuals, liberals and conservatives, have either also remained at similar levels of change or seen an increase in variance. Policy issues that see increased variance include spending on crime and whether organizations should preferentially hire Black Americans to account for past discrimination. It is unclear what is driving increased inconsistency among these items over time.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The dramatic changes in the timing of life course transitions that have occurred in the United States in the last few decades provide a valuable opportunity for exploring the mechanisms that underpin attitudinal change across different life stages. Given that Americans, on average, pushed back marriage, childbirth, and entering the workforce over the course of the second half of the 20th Century and the early 21st Century, we posited that these shifts should, through various pathways, increase over-time change in survey responses among more recent cohorts of American early adults compared to their predecessors when they were the same age. We tested a series of hypotheses related to this framework by looking at the stability of political and social opinions among different cohorts as they moved through the same age ranges.

The results presented above suggest a more complicated picture than anticipated. Consistent with expectations, early adults look similar to middle-aged adults in the earliest panels in terms of both demographic characteristics and attitude stability. Members of the youngest age categories (18-25 and 26-33) diverged over time from older members of the population, showing higher rates of opinion change than them in more recent panels. This shift, however, was principally driven by an increase in stability among people in older age groups, rather than increased opinion change among the early adults. We suggested that this aggregate divergence was driven by countervailing trends. On the one hand, delays in life course transitions push the youngest adults to make more attitude changes over time. On the other hand, broader social forces -- like the increase in educational attainment -- stabilized opinions for all members of the population.

We examined the relative importance of these countervailing forces using data on life-course transitions and measures of educational attainment. We found that, consistent with our revised perspective, increased educational attainment stabilized opinions for all age groups, with contemporary older (65-80) adults exhibiting much higher levels of attitude stability than their counterparts from previous cohorts. Life course shifts such as delayed rates of marriage and childbirth and delayed entry into the workforce had contradictory effects on the overall rate of attitude change. While marriage made change less likely, ever having children or having children at home increased the rate of change. The effects of these transitions were small and, thus, did not significantly affect rates of stability in the aggregate. In general, then, increased educational attainment has stabilized attitudes across all age groups.

Our analyses recovered trends that are compatible with the idea of an extended formative period for opinions and beliefs but suggest that the mechanisms behind it are counterintuitive. It is true that, in the contemporary United States, early adults change their opinions more than older adults. It is also true that this gap is wider now than it had been in previous decades. However, this is not because contemporary early adults are changing at a higher rate than their counterparts in previous time periods. It is rather that older age groups have become considerably more stable, in great part due to an increase in educational attainment. Our approach likely under-estimates the importance of this shift, as we collapse education into only three categories -- ignoring differences among people with less-than a high school degree in early panels and dramatic expansions of post-graduate degree attainment in recent decades. So the perceived differences in opinion change between age groups can be partly explained by the fact that early adults have not stabilized as much relative to older age groups.

Our analyses are particularly relevant when considered against the backdrop of how robust findings about formative periods are across disciplines and domains. The pattern of a period of exploration followed by stability is evident in different species, and its evolution has been widely theorized. Our results suggest that a person's openness to changing their opinions and beliefs is not only a result of maturational processes like brain plasticity or even of how information is processed but rather is partly shaped by the social structuring of the life-course and the border social environment. When early adults' lives resembled those of middle-aged respondents in the 1950s, their rates of opinion change were also similar. Their rates of attitude change diverged as their social conditions and experiences diverged. Many theories of socialization place great weight on the importance of the ``early years'' as a moment of exploration and attitudinal formation. Our work shows that social factors can affect the extent to which this exploration can take place, how long these ``early years'' might last, and how other countervailing processes might stabilize attitudes.

At the same time, it is important not to overstate the importance of these broader social factors in shaping opinion stability. The average expected opinion change over time in 2020 is within a few points of the average expected attitude change on a 100-point scale for any particular survey question, and we do see wide variation in the trajectories of opinion stability across different questions. Despite dramatic shifts in educational attainment, access to information, and political organization, not to mention changes in survey administration, people make changes in many social and political attitudes at about the same rates they did 70 years ago. And most of this change reflects temporary shifts of opinion or measurement error. Even early adults, who appear to make the most durable change according to our models, durably change only about an average of 3 points on a 100-point scale every two years.

\hypertarget{limitations}{%
\subsection{Limitations}\label{limitations}}

As we noted when defining our theoretical estimands, it is difficult to understand the questions we posed in a purely causal framework. Respondents born in different cohorts face not only diverse social circumstances upon birth -- different opportunities for education, different media environments, different access to technologies -- but also have different compositions at birth. More recent cohorts are more racially and ethnically diverse than the earliest cohorts, are less likely to be born into two-parent households, are less likely to be born to religious parents, and vary in other ways that are likely related to opinion formation. All of these factors make it difficult to conceptualize what it would mean to hold everything but birth year constant. Our estimates here should not be interpreted as causal effects of only being born in later years, holding these kinds of features constant, and it is plausible that coefficients estimated above model are biased by these compositional shifts.

Similarly, it is important to be cautious in our interpretation of the relationship between increased educational attainment and widespread stabilization. What we observe as an increase in educational attainment likely also captures other time trends such as increased levels of material wealth, curricular changes in public schools, or more diverse and integrated classrooms. Thus, when interpreting the effect of educational attainment on stabilization, we should bear in mind that there are many potential correlates with the former and that, for that reason, establishing a clear causal connection is difficult.

Beyond this central limitation, one additional limitation of our design is that we focus on questions that are asked repeatedly across multiple panels. By focusing only on the same questions over time, we control for question content and structure and ensure the trend we identify over time is due to something other than changes in survey administration or question quality. However, this approach limits the kinds of issues about which we can make claims to issues that receive significant public attention for a long period of time, including general ideological positions and positions on long-standing issues, such as the government's role in the health care sector. We ignore large swaths of issues that emerge and disappear before the next panel but were very important in political debates, such as approaches to fight inflation in the 1970s, differences over the Iraq and Afghanistan wars in the 2000s, and many questions about social justice in recent years. It could be the case that such issues have a different relationship between age and stability than issues that dominate political debates for decades.

Similarly, because most questions are asked in a small number of panels, our actual number of data points for key questions is much smaller than the total number of wave-pair observations we have. The modal number of panels for individual questions is 3, meaning our ability to estimate time trajectories is constrained and our estimates of question-specific trajectories is imprecise. This also makes it difficult to estimate time trajectories as anything other than linear trends. It is almost certainly true that changes in attitude stability are not a linear function of time. Our estimated effect of time should be thought of as an average of what are likely heterogeneous effects of different eras that should be explored more in future research.

Despite these limitations, we believe our results speak to central questions across multiple subfields in sociology and suggest further avenues of investigation.

\hypertarget{implications-for-cultural-sociology}{%
\subsection{Implications for Cultural Sociology}\label{implications-for-cultural-sociology}}

First, our results reinforce recent findings that suggest that attitudes, beliefs, and opinions -- what Lizardo (2017) calls ``declarative personal culture'' -- are mostly stable in adulthood, with most change observed between responses by the same person being transit, non-durable change or measurement error (Kiley and Vaisey 2020; Lersch 2023). Adults make large non-persistent changes over time but small persistent changes, on the order of 1-2 percent of a response scale a year. These estimates support work that suggests that most aggregate cultural change must be driven by cohort replacement processes rather than changes of opinion driven by shifts in the contemporaneous environment, and that understanding why and how cultural change occurs requires greater attention to the ``past conditions of production'' of childhood, adolescence, and early adulthood (Vaisey and Lizardo 2016), and reinforce calls for greater attention to early life socialization (Guhin, Calarco, and Miller-Idriss 2021).

We did find that the rate of durable change in adulthood did increase slightly over the course of the window studied here, especially for younger adults, suggesting more room for durable attitude change at later life-course stages.

Relatedly, it is worth highlighting that the main trend we observe is not about attitudinal change, but rather about increased stability. The most consistent finding we have is about how social factors -- in this case educational attainment -- might act to stabilize opinions, rather than how they might lead to more openness to change. Like previous work on attitude stability (Kiley and Vaisey 2020; Lersch 2023; Vaisey and Kiley 2021), these findings highlight the need to focus on the social processes that make holding stable opinions more likely (or even possible). Sociologists interested in personal declarative culture are often interested in how it changes. It seems, however, that our efforts should also be directed at explaining what makes attitudinal stability possible to maintain, or the ``social scaffolding'' that constrains cultural beliefs (Lizardo 2010; Martin and Desmond 2010), and the aggregate consequences for social contexts that allow for more or less change.

\hypertarget{implications-for-sociology-of-aging-and-the-life-course}{%
\subsection{Implications for Sociology of Aging and the Life Course}\label{implications-for-sociology-of-aging-and-the-life-course}}

We focused on the period of transition into adulthood because it has been theorized as a key moment for attitudinal formation. However, our results show that patterns of attitudinal updating have changed perhaps more for individuals in older age-groups, where opinions have become more stable. Sociologists of aging have long criticized visions of old age as a period of passivity, arguing that it has been reconceptualized as a period for renewed exploration. Our results challenge previous work that suggests that attitudes follow a ``life stages'' model in which people are open to change both when they are young and when they are old. While we find that older Americans had high rates of change in early panels, this appears principally to be because cohorts that were old in the middle of the 20th Century had very low levels of formal education, which made them inconsistent opinion holders. As successive cohorts of older Amerians have attained higher levels of education, their opinions have become much more stable than younger Americans.

Our analyses also have important implications for how we understand the development of attitudinal change and formation across the life course during the past 70 years. The fact that the amount of opinion change -- both durable and non-durable -- has changed across cohorts challenges models of social learning and political imprinting that assume a cohort-invariant propensity to adjust opinions in the face of social influence at different ages (Ghitza, Gelman, and Auerbach 2022; Bartels and Jackman 2014). Our results suggest that social conditions such as access to education have the capacity to change how susceptible people are to broader social environments across the life course, and have shifted the windows in which Americans learn over time.

\hypertarget{implications-for-political-sociology}{%
\subsection{Implications for Political Sociology}\label{implications-for-political-sociology}}

The pattern of increased opinion stability in middle age reinforces the argument that political dispositions form primarily through an imprinting process, in which preferences are formed early in life and are relatively resistant to change during adulthood (Ghitza, Gelman, and Auerbach 2022; Sears and Funk 1999). While adults do occasionally make changes of opinion (Lersch 2023), experiences during adulthood do not appear to be sufficient to produce widespread changes of political opinion. At the same time, it is not the case that people have clear, stable attitudes on all issues as adults. While adults' expression of their partisan and ideological identification are quite stable over time, Americans on average demonstrate a high degree of inconsistency on a number of issues, such as whether they view government to be effective, trustworthy, and responsive to peoples' interests.

Our results further show that over-time trends in the stability of opinions on political issues are quite varied. Opinions on so-called ``social issues,'' such as support for the rights of homosexuals and views on the role of women in society, appear to have become more stable over time, potentially reflecting the transition from political dissensus to consensus (Baldassarri and Park 2020). On the other hand, Americans' opinions on long-standing questions about economic redistribution, such as views on the government's role in health care and government spending on social insurance programs, are as variable as they were decades ago. This suggests that duration as an issue of public disagreement is not sufficient to produce either consensus or stable opinions.

Finally, our results suggest a key role of education in shaping public opinion on political issues. Consistent with previous work that finds that educated Americans hold more ideologically aligned issue positions (Boutyline and Vaisey 2017), our strongest pattern of results suggests that education also stabilizes opinions over time. It seems likely that these trends reinforce each other, with ideologically aligned opinions being easier to hold.

\hypertarget{implications-for-sociology-of-education}{%
\subsection{Implications for Sociology of Education}\label{implications-for-sociology-of-education}}

Our results suggest that educational attainment, of factors studied here, plays a very strong role in stabilizing opinions, though the specific mechanisms are unclear. In general, question-specific effects for attaining a college degree (and having less than a high school degree) were almost uniformly negative (positive) and statistically significantly different from 0 for most questions, and there were almost no questions where higher levels of educational attainment were positively associated with change. However, it is unclear why educational attainment has this effect. We reiterate calls for greater exploration into the specific mechanisms that produce this greater level of stability among people with high levels of education, be it changes to intelligence, cultural expectations, changes in information consumption habits, or something else.

Some work suggests that highly educated Americans are more culturally homogeneous and inhabit more culturally homogeneous environments (Lynn and Ellerbach 2017; Lynn, Shi, and Kiley 2024; Harding 2009). It could be that cultural homogeneity and stability reinforce each other, as these people receive consistent signals about what kinds of opinions they should hold, while people in more heterogeneous environments receive conflicting signals.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This paper leveraged delays in the timing of life course transitions and the emergence of an ``age of independence'' over the past 70 years to adjudicate whether heightened rates of attitude change in early adulthood are principally ontogenetic in origin, and therefore relatively stable over time, or whether they emerge due to the distinct social circumstances of adolescence and early adulthood. We expected to see more attitude changes in early adulthood and an extension of the ``impressionable years'' pattern up the age ladder in more recent cohorts as early adulthood became more socially distinct from middle age.

Instead, we found that other social processes, particularly the dramatic expansion of educational attainment during the second half of the 20th Century, affected the opinions of middle-aged and older respondents much more substantially than any life-course changes. This expansion of education has stabilized opinions across the board, reducing random wave-to-wave change (what is often called measurement error) while also slightly increasing the rates at which people made changes that appear to persist over time.

In this way, our findings provide broad support for arguments that the unique rates of attitude change across the life course are shaped by social structures, rather than just ontogenetic factors, and we call for much more exploration by social scientists into the social factors that enable stable opinion holding.

\hypertarget{references}{%
\section{References}\label{references}}

\singlespace

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-alwin2007}{}}%
Alwin, Duane F. 2007. \emph{Margins of {Error}: {A} {Study} of {Reliability} in {Survey} {Measurement}}. Hoboken, N.J.: John Wiley \& Sons.

\leavevmode\vadjust pre{\hypertarget{ref-alwin1991}{}}%
Alwin, Duane F., and Jon A. Krosnick. 1991a. {``Aging, {Cohorts}, and the {Stability} of {Sociopolitical} {Orientations} {Over} the {Life} {Span}.''} \emph{American Journal of Sociology} 97 (1): 169--95. \url{https://www.jstor.org/stable/2781642}.

\leavevmode\vadjust pre{\hypertarget{ref-alwin1991a}{}}%
---------. 1991b. {``The {Reliability} of {Survey} {Attitude} {Measurement}: {The} {Influence} of {Question} and {Respondent} {Attributes}.''} \emph{Sociological Methods \& Research} 20 (1): 139--81. \url{https://doi.org/10.1177/0049124191020001005}.

\leavevmode\vadjust pre{\hypertarget{ref-ansolabehere2008}{}}%
Ansolabehere, Stephen, Jonathan Rodden, and James M. Snyder. 2008. {``The {Strength} of {Issues}: {Using} {Multiple} {Measures} to {Gauge} {Preference} {Stability}, {Ideological} {Constraint}, and {Issue} {Voting}.''} \emph{American Political Science Review} 102 (2): 215--32. \url{https://doi.org/10.1017/S0003055408080210}.

\leavevmode\vadjust pre{\hypertarget{ref-aplin2017}{}}%
Aplin, Lucy M., Ben C. Sheldon, and Richard McElreath. 2017. {``Conformity Does Not Perpetuate Suboptimal Traditions in a Wild Population of Songbirds.''} \emph{Proceedings of the National Academy of Sciences} 114 (30): 7830--37. \url{https://doi.org/10.1073/pnas.1621067114}.

\leavevmode\vadjust pre{\hypertarget{ref-baldassarri2008}{}}%
Baldassarri, Delia, and Andrew Gelman. 2008. {``Partisans Without {Constraint}: {Political} {Polarization} and {Trends} in {American} {Public} {Opinion}.''} \emph{American Journal of Sociology} 114 (2): 408--46.

\leavevmode\vadjust pre{\hypertarget{ref-baldassarri2020}{}}%
Baldassarri, Delia, and Barum Park. 2020. {``Was {There} a {Culture} {War}? {Partisan} {Polarization} and {Secular} {Trends} in {US} {Public} {Opinion}.''} \emph{The Journal of Politics} 82 (3): 809--27. \url{https://doi.org/10.1086/707306}.

\leavevmode\vadjust pre{\hypertarget{ref-bartels2014}{}}%
Bartels, Larry M., and Simon Jackman. 2014. {``A Generational Model of Political Learning.''} \emph{Electoral Studies} 33 (March): 7--18. \url{https://doi.org/10.1016/j.electstud.2013.06.004}.

\leavevmode\vadjust pre{\hypertarget{ref-billari2010}{}}%
Billari, Francesco C., and Aart C. Liefbroer. 2010. {``Towards a New Pattern of Transition to Adulthood?''} \emph{Advances in Life Course Research}, Demographic {Perspectives} on the {Transition} to {Adulthood}, 15 (2): 59--75. \url{https://doi.org/10.1016/j.alcr.2010.10.003}.

\leavevmode\vadjust pre{\hypertarget{ref-bourdieu1990}{}}%
Bourdieu, Pierre. 1990. \emph{The Logic of Practice}. Stanford, Calif. : Stanford University Press, 1990.

\leavevmode\vadjust pre{\hypertarget{ref-boutyline2017}{}}%
Boutyline, Andrei, and Stephen Vaisey. 2017. {``Belief {Network} {Analysis}: {A} {Relational} {Approach} to {Understanding} the {Structure} of {Attitudes}.''} \emph{American Journal of Sociology} 122 (5): 1371--1447. \url{https://doi.org/10.1086/691274}.

\leavevmode\vadjust pre{\hypertarget{ref-brocic2021}{}}%
Broćić, Miloš, and Andrew Miles. 2021. {``College and the {`{Culture} {War}'}: {Assessing} {Higher} {Education}'s {Influence} on {Moral} {Attitudes}.''} \emph{American Sociological Review}, September, 00031224211041094. \url{https://doi.org/10.1177/00031224211041094}.

\leavevmode\vadjust pre{\hypertarget{ref-bruckner2005}{}}%
Brückner, Hannah, and Karl Ulrich Mayer. 2005. {``De-{Standardization} of the {Life} {Course}: {What} It {Might} {Mean}? {And} If It {Means} {Anything}, {Whether} It {Actually} {Took} {Place}?''} \emph{Advances in Life Course Research} 9 (January): 27--53. \url{https://doi.org/10.1016/S1040-2608(04)09002-1}.

\leavevmode\vadjust pre{\hypertarget{ref-buchmann1989}{}}%
Buchmann, Marlis. 1989. \emph{The Script of Life in Modern Society: Entry into Adulthood in a Changing World}. Chicago: University of Chicago Press.

\leavevmode\vadjust pre{\hypertarget{ref-burt2005}{}}%
Burt, Ronald S. 2005. \emph{Brokerage and {Closure}: {An} {Introduction} to {Social} {Capital}}. OUP Oxford.

\leavevmode\vadjust pre{\hypertarget{ref-campbell2016}{}}%
Campbell, Colin, and Jonathan Horowitz. 2016. {``Does {College} {Influence} {Sociopolitical} {Attitudes}?''} \emph{Sociology of Education} 89 (1): 40--58. \url{https://doi.org/10.1177/0038040715617224}.

\leavevmode\vadjust pre{\hypertarget{ref-cherlin2004}{}}%
Cherlin, Andrew J. 2004. {``The Deinstitutionalization of {American} Marriage.''} \emph{Journal of Marriage and Family} 66 (4): 848--61. \url{https://doi.org/10.1111/j.0022-2445.2004.00058.x}.

\leavevmode\vadjust pre{\hypertarget{ref-cherlin2021}{}}%
---------. 2021. {``Rising Nonmarital First Childbearing Among College-Educated Women: {Evidence} from Three National Studies.''} \emph{Proceedings of the National Academy of Sciences} 118 (37): e2109016118. \url{https://doi.org/10.1073/pnas.2109016118}.

\leavevmode\vadjust pre{\hypertarget{ref-converse1964}{}}%
Converse, Philip E. 1964. {``The Nature of Belief Systems in Mass Publics (1964).''} In \emph{Ideology and Discontent}, edited by D. E. Apter, 18:206--61. New York: Free Press.

\leavevmode\vadjust pre{\hypertarget{ref-cunningham2007}{}}%
Cunningham, Hugh. 2007. \emph{The {Invention} of Childhood}. London: BBC Books.

\leavevmode\vadjust pre{\hypertarget{ref-danigelis2007}{}}%
Danigelis, Nicholas L., Melissa Hardy, and Stephen J. Cutler. 2007. {``Population {Aging}, {Intracohort} {Aging}, and {Sociopolitical} {Attitudes}.''} \emph{American Sociological Review} 72 (5): 812--30. \url{https://doi.org/10.1177/000312240707200508}.

\leavevmode\vadjust pre{\hypertarget{ref-dellaposta2020}{}}%
DellaPosta, Daniel. 2020. {``Pluralistic {Collapse}: {The} {`{Oil} {Spill}'} {Model} of {Mass} {Opinion} {Polarization}.''} \emph{American Sociological Review} 85 (3): 507--36. \url{https://doi.org/10.1177/0003122420922989}.

\leavevmode\vadjust pre{\hypertarget{ref-dellaposta2015}{}}%
DellaPosta, Daniel, Yongren Shi, and Michael Macy. 2015. {``Why {Do} {Liberals} {Drink} {Lattes}?''} \emph{American Journal of Sociology} 120 (5): 1473--1511. \url{https://doi.org/10.1086/681254}.

\leavevmode\vadjust pre{\hypertarget{ref-dinas2014}{}}%
Dinas, Elias. 2014. {``Why {Does} the {Apple} {Fall} {Far} from the {Tree}? {How} {Early} {Political} {Socialization} {Prompts} {Parent}-{Child} {Dissimilarity}.''} \emph{British Journal of Political Science} 44 (4): 827--52. \url{https://doi.org/10.1017/S0007123413000033}.

\leavevmode\vadjust pre{\hypertarget{ref-diprete2011}{}}%
DiPrete, Thomas A., Andrew Gelman, Tyler McCormick, Julien Teitler, and Tian Zheng. 2011. {``Segregation in {Social} {Networks} {Based} on {Acquaintanceship} and {Trust}.''} \emph{American Journal of Sociology} 116 (4): 1234--83. \url{https://doi.org/10.1086/659100}.

\leavevmode\vadjust pre{\hypertarget{ref-eaton2009}{}}%
Eaton, Asia A., Penny S. Visser, Jon A. Krosnick, and Sowmya Anand. 2009. {``Social {Power} and {Attitude} {Strength} {Over} the {Life} {Course}.''} \emph{Personality and Social Psychology Bulletin} 35 (12): 1646--60. \url{https://doi.org/10.1177/0146167209349114}.

\leavevmode\vadjust pre{\hypertarget{ref-edelmann2023}{}}%
Edelmann, Achim, and Stephen Vaisey. 2023. {``Gender Differential Effect of College on Political Orientation over the Last 40 Years in the {U}.{S}.---{A} Propensity Score Weighting Approach.''} Edited by Supat Chupradit. \emph{PLOS ONE} 18 (1): e0279273. \url{https://doi.org/10.1371/journal.pone.0279273}.

\leavevmode\vadjust pre{\hypertarget{ref-elder1974}{}}%
Elder, Glen H. 1974. \emph{Children of the {Great} {Depression} : Social Change in Life Experience}. Chicago: University of Chicago Press. \url{https://find.library.duke.edu/catalog/DUKE000108847}.

\leavevmode\vadjust pre{\hypertarget{ref-elder1994}{}}%
---------. 1994. {``Time, {Human} {Agency}, and {Social} {Change}: {Perspectives} on the {Life} {Course}.''} \emph{Social Psychology Quarterly} 57 (1): 4--15. \url{https://doi.org/10.2307/2786971}.

\leavevmode\vadjust pre{\hypertarget{ref-elder2016}{}}%
Elder, Glen H., and Linda K. George. 2016. {``Age, {Cohorts}, and the {Life} {Course}.''} In \emph{Handbook of the {Life} {Course}: {Volume} {II}}, edited by Michael J. Shanahan, Jeylan T. Mortimer, and Monica Kirkpatrick Johnson, 59--85. Handbooks of {Sociology} and {Social} {Research}. Cham: Springer International Publishing. \url{https://doi.org/10.1007/978-3-319-20880-0_3}.

\leavevmode\vadjust pre{\hypertarget{ref-england2010}{}}%
England, Paula. 2010. {``{THE} {GENDER} {REVOLUTION}: {Uneven} and {Stalled}.''} \emph{Gender and Society} 24 (2): 149--66. \url{https://www.jstor.org/stable/27809263}.

\leavevmode\vadjust pre{\hypertarget{ref-fiorina2008}{}}%
Fiorina, Morris P, and Samuel J Abrams. 2008. {``Political {Polarization} in the {American} {Public},''} 29.

\leavevmode\vadjust pre{\hypertarget{ref-fosse2023a}{}}%
Fosse, Ethan. 2023. {``Theorizing in the {Cohort} {Mode}: {On} {Ryder}'s {Processual} {Account} of {Social} {Change}.''} \emph{Canadian Studies in Population} 50 (3): 5. \url{https://doi.org/10.1007/s42650-023-00075-9}.

\leavevmode\vadjust pre{\hypertarget{ref-fosse2019}{}}%
Fosse, Ethan, and Christopher Winship. 2019. {``Analyzing {Age}-{Period}-{Cohort} {Data}: {A} {Review} and {Critique}.''} \emph{Annual Review of Sociology} 45 (1): 467--92. \url{https://doi.org/10.1146/annurev-soc-073018-022616}.

\leavevmode\vadjust pre{\hypertarget{ref-frankenhuis2020}{}}%
Frankenhuis, Willem E., and Nicole Walasek. 2020. {``Modeling the Evolution of Sensitive Periods.''} \emph{Developmental Cognitive Neuroscience} 41 (February): 100715. \url{https://doi.org/10.1016/j.dcn.2019.100715}.

\leavevmode\vadjust pre{\hypertarget{ref-geist2008}{}}%
Geist, Claudia, and Patricia A. McManus. 2008. {``Geographical Mobility over the Life Course: Motivations and Implications.''} \emph{Population, Space and Place} 14 (4): 283--303. \url{https://doi.org/10.1002/psp.508}.

\leavevmode\vadjust pre{\hypertarget{ref-gelman2007}{}}%
Gelman, Andrew, and Jennifer Hill. 2007. \emph{Data Analysis Using Regression and Multilevel/Hierarchical Models}. 23rd printing. Analytical Methods for Social Research. Cambridge: Cambridge Univ. Press.

\leavevmode\vadjust pre{\hypertarget{ref-gerber1998}{}}%
Gerber, Alan, and Donald P. Green. 1998. {``Rational {Learning} and {Partisan} {Attitudes}.''} \emph{American Journal of Political Science} 42 (3): 794--818. \url{https://doi.org/10.2307/2991730}.

\leavevmode\vadjust pre{\hypertarget{ref-ghitza2022}{}}%
Ghitza, Yair, Andrew Gelman, and Jonathan Auerbach. 2022. {``The {Great} {Society}, {Reagan}'s {Revolution}, and {Generations} of {Presidential} {Voting}.''} \emph{American Journal of Political Science}, September, ajps.12713. \url{https://doi.org/10.1111/ajps.12713}.

\leavevmode\vadjust pre{\hypertarget{ref-gopnik2020}{}}%
Gopnik, Alison. 2020. {``Childhood as a Solution to Explore--Exploit Tensions.''} \emph{Philosophical Transactions of the Royal Society B: Biological Sciences} 375 (1803): 20190502. \url{https://doi.org/10.1098/rstb.2019.0502}.

\leavevmode\vadjust pre{\hypertarget{ref-granovetter1973}{}}%
Granovetter, Mark S. 1973. {``The {Strength} of {Weak} {Ties}.''} \emph{American Journal of Sociology} 78 (6): 1360--80. \url{http://www.jstor.org/stable/2776392}.

\leavevmode\vadjust pre{\hypertarget{ref-guhin2021}{}}%
Guhin, Jeffrey, Jessica McCrory Calarco, and Cynthia Miller-Idriss. 2021. {``Whatever {Happened} to {Socialization}?''} \emph{Annual Review of Sociology} 47 (1): 109--29. \url{https://doi.org/10.1146/annurev-soc-090320-103012}.

\leavevmode\vadjust pre{\hypertarget{ref-harding2009}{}}%
Harding, David J. 2009. {``Violence, {Older} {Peers}, and the {Socialization} of {Adolescent} {Boys} in {Disadvantaged} {Neighborhoods}.''} \emph{American Sociological Review} 74 (3): 445--64. \url{https://doi.org/10.1177/000312240907400306}.

\leavevmode\vadjust pre{\hypertarget{ref-hout2012}{}}%
Hout, Michael. 2012. {``Social and {Economic} {Returns} to {College} {Education} in the {United} {States}.''} \emph{Annual Review of Sociology} 38 (1): 379--400. \url{https://doi.org/10.1146/annurev.soc.012809.102503}.

\leavevmode\vadjust pre{\hypertarget{ref-kiley2020}{}}%
Kiley, Kevin, and Stephen Vaisey. 2020. {``Measuring {Stability} and {Change} in {Personal} {Culture} {Using} {Panel} {Data}.''} \emph{American Sociological Review} 85 (3): 477--506. \url{https://doi.org/10.1177/0003122420921538}.

\leavevmode\vadjust pre{\hypertarget{ref-kozlowski2021}{}}%
Kozlowski, Austin C., and James P. Murphy. 2021. {``Issue Alignment and Partisanship in the {American} Public: {Revisiting} the {`Partisans Without Constraint'} Thesis.''} \emph{Social Science Research} 94 (February): 102498. \url{https://doi.org/10.1016/j.ssresearch.2020.102498}.

\leavevmode\vadjust pre{\hypertarget{ref-krosnick1991}{}}%
Krosnick, Jon A. 1991. {``Response Strategies for Coping with the Cognitive Demands of Attitude Measures in Surveys.''} \emph{Applied Cognitive Psychology} 5 (3): 213--36. \url{https://doi.org/10.1002/acp.2350050305}.

\leavevmode\vadjust pre{\hypertarget{ref-krosnick1989}{}}%
Krosnick, Jon A., and Duane F. Alwin. 1989. {``Aging and {Susceptibility} to {Attitude} {Change}.''} \emph{Journal of Personality and Social Psychology} 57 (October): 416--25. \url{https://doi.org/10.1037//0022-3514.57.3.416}.

\leavevmode\vadjust pre{\hypertarget{ref-lersch2023}{}}%
Lersch, Philipp M. 2023. {``Change in {Personal} {Culture} over the {Life} {Course}.''} \emph{American Sociological Review} 88 (2): 220--51. \url{https://doi.org/10.1177/00031224231156456}.

\leavevmode\vadjust pre{\hypertarget{ref-lizardo2010}{}}%
Lizardo, Omar. 2010. {``Beyond the Antinomies of Structure: {Levi}-{Strauss}, {Giddens}, {Bourdieu}, and {Sewell}.''} \emph{Theory and Society} 39 (6): 651--88. \url{https://doi.org/10.1007/s11186-010-9125-1}.

\leavevmode\vadjust pre{\hypertarget{ref-lizardo2017}{}}%
---------. 2017. {``Improving {Cultural} {Analysis}: {Considering} {Personal} {Culture} in Its {Declarative} and {Nondeclarative} {Modes}.''} \emph{American Sociological Review} 82 (1): 88--115. \url{https://doi.org/10.1177/0003122416675175}.

\leavevmode\vadjust pre{\hypertarget{ref-lundberg2021}{}}%
Lundberg, Ian, Rebecca Johnson, and Brandon M Stewart. 2021. {``What {Is} {Your} {Estimand}? {Defining} the {Target} {Quantity} {Connects} {Statistical} {Evidence} to {Theory}.''} \emph{American Sociological Review} 86 (3): 532--65.

\leavevmode\vadjust pre{\hypertarget{ref-lynn2017}{}}%
Lynn, Freda B., and George Ellerbach. 2017. {``A {Position} with a {View}: {Educational} {Status} and the {Construction} of the {Occupational} {Hierarchy}.''} \emph{American Sociological Review} 82 (1): 32--58. \url{https://doi.org/10.1177/0003122416671743}.

\leavevmode\vadjust pre{\hypertarget{ref-lynn2024}{}}%
Lynn, Freda B., Yongren Shi, and Kevin Kiley. 2024. {``Intersectional {Group} {Agreement} on the {Occupational} {Order}.''} \emph{Social Psychology Quarterly}, June, 01902725241256378. \url{https://doi.org/10.1177/01902725241256378}.

\leavevmode\vadjust pre{\hypertarget{ref-ma2021}{}}%
Ma, Xiangyu. 2021. {``What Are the Temporal Dynamics of Taste?''} \emph{Poetics} 84 (February): 101514. \url{https://doi.org/10.1016/j.poetic.2020.101514}.

\leavevmode\vadjust pre{\hypertarget{ref-mannheim1952}{}}%
Mannheim, Karl. 1952. {``The {Problem} of {Generations}.''} In \emph{Essays on the {Sociology} of {Knowledge}}, 276--322. New York: Oxford University Press.

\leavevmode\vadjust pre{\hypertarget{ref-marsden1987}{}}%
Marsden, Peter V. 1987. {``Core {Discussion} {Networks} of {Americans}.''} \emph{American Sociological Review} 52 (1): 122--31. \url{https://doi.org/10.2307/2095397}.

\leavevmode\vadjust pre{\hypertarget{ref-martin2010}{}}%
Martin, John Levi, and Matthew Desmond. 2010. {``Political {Position} and {Social} {Knowledge}.''} \emph{Sociological Forum} 25 (1): 1--26. \url{https://doi.org/10.1111/j.1573-7861.2009.01154.x}.

\leavevmode\vadjust pre{\hypertarget{ref-mata2015}{}}%
Mata, Rui, and Bettina von Helversen. 2015. {``Search and the {Aging} {Mind}: {The} {Promise} and {Limits} of the {Cognitive} {Control} {Hypothesis} of {Age} {Differences} in {Search}.''} \emph{Topics in Cognitive Science} 7 (3): 416--27. \url{https://doi.org/10.1111/tops.12139}.

\leavevmode\vadjust pre{\hypertarget{ref-mulder2010}{}}%
Mulder, Juul, Tom F. M. Ter Bogt, Quinten A. W. Raaijmakers, Saoirse Nic Gabhainn, and Paul Sikkema. 2010. {``From Death Metal to {R}\&{B}? {Consistency} of Music Preferences Among {Dutch} Adolescents and Young Adults.''} \emph{Psychology of Music} 38 (1): 67--83. \url{https://doi.org/10.1177/0305735609104349}.

\leavevmode\vadjust pre{\hypertarget{ref-prior2010}{}}%
Prior, Markus. 2010. {``You've {Either} {Got} {It} or {You} {Don}'t? {The} {Stability} of {Political} {Interest} over the {Life} {Cycle}.''} \emph{The Journal of Politics} 72 (3): 747--66. \url{https://doi.org/10.1017/s0022381610000149}.

\leavevmode\vadjust pre{\hypertarget{ref-raftery1995}{}}%
Raftery, Adrian E. 1995. {``Bayesian {Model} {Selection} in {Social} {Research}.''} \emph{Sociological Methodology} 25: 111--63. \url{https://doi.org/10.2307/271063}.

\leavevmode\vadjust pre{\hypertarget{ref-raudenbush2002}{}}%
Raudenbush, Stephen W., and Anthony S. Bryk. 2002. \emph{Hierarchical Linear Models: Applications and Data Analysis Methods}. 2nd ed. Advanced Quantitative Techniques in the Social Sciences 1. Thousand Oaks: Sage Publications.

\leavevmode\vadjust pre{\hypertarget{ref-robbins2022}{}}%
Robbins, Blaine G., Aimée Dechter, and Sabino Kornrich. 2022. {``Assessing the {Deinstitutionalization} of {Marriage} {Thesis}: {An} {Experimental} {Test}.''} \emph{American Sociological Review} 87 (2): 237--74. \url{https://doi.org/10.1177/00031224221080960}.

\leavevmode\vadjust pre{\hypertarget{ref-rosenfeld2009}{}}%
Rosenfeld, Michael J. 2009. \emph{The Age of Independence: Interracial Unions, Same-Sex Unions, and the Changing {American} Family}. Cambridge, Mass.: Harvard University Press.

\leavevmode\vadjust pre{\hypertarget{ref-rosenfeld2005}{}}%
Rosenfeld, Michael J., and Byung-Soo Kim. 2005. {``The {Independence} of {Young} {Adults} and the {Rise} of {Interracial} and {Same}-{Sex} {Unions}.''} \emph{American Sociological Review} 70 (4): 541--62. \url{https://doi.org/10.1177/000312240507000401}.

\leavevmode\vadjust pre{\hypertarget{ref-ryder1965}{}}%
Ryder, Norman B. 1965. {``The {Cohort} as a {Concept} in the {Study} of {Social} {Change}.''} \emph{American Sociological Review} 30 (6): 843--61. \url{https://doi.org/10.2307/2090964}.

\leavevmode\vadjust pre{\hypertarget{ref-schofer2005}{}}%
Schofer, Evan, and John W. Meyer. 2005. {``The {Worldwide} {Expansion} of {Higher} {Education} in the {Twentieth} {Century}.''} \emph{American Sociological Review} 70 (6): 898--920. \url{http://www.jstor.org/stable/4145399}.

\leavevmode\vadjust pre{\hypertarget{ref-sears2013}{}}%
Sears, David O., and Christia Brown. 2013. \emph{Childhood and {Adult} {Political} {Development}}. Oxford University Press. \url{https://doi.org/10.1093/oxfordhb/9780199760107.013.0003}.

\leavevmode\vadjust pre{\hypertarget{ref-sears1999}{}}%
Sears, David O., and Carolyn L. Funk. 1999. {``Evidence of the {Long}-{Term} {Persistence} of {Adults}' {Political} {Predispositions}.''} \emph{The Journal of Politics} 61 (1): 1--28. \url{https://doi.org/10.2307/2647773}.

\leavevmode\vadjust pre{\hypertarget{ref-sherratt2018}{}}%
Sherratt, Thomas N., and Julie Morand-Ferron. 2018. {``The Adaptive Significance of Age-Dependent Changes in the Tendency of Individuals to Explore.''} \emph{Animal Behaviour} 138: 59--67. \url{https://doi.org/10.1016/j.anbehav.2018.01.025}.

\leavevmode\vadjust pre{\hypertarget{ref-small2015}{}}%
Small, Mario Luis, Vontrese Deeds Pamphile, and Peter McMahan. 2015. {``How Stable Is the Core Discussion Network?''} \emph{Social Networks} 40 (January): 90--102. \url{https://doi.org/10.1016/j.socnet.2014.09.001}.

\leavevmode\vadjust pre{\hypertarget{ref-spreng2019}{}}%
Spreng, R. Nathan, and Gary R. Turner. 2019. {``The {Shifting} {Architecture} of {Cognition} and {Brain} {Function} in {Older} {Adulthood}.''} \emph{Perspectives on Psychological Science} 14 (4): 523--42. \url{https://doi.org/10.1177/1745691619827511}.

\leavevmode\vadjust pre{\hypertarget{ref-stark1980}{}}%
Stark, Rodney, and William Sims Bainbridge. 1980. {``Networks of {Faith}: {Interpersonal} {Bonds} and {Recruitment} to {Cults} and {Sects}.''} \emph{American Journal of Sociology} 85 (6): 1376--95. \url{http://www.jstor.org/stable/2778383}.

\leavevmode\vadjust pre{\hypertarget{ref-vaisey2021}{}}%
Vaisey, Stephen, and Kevin Kiley. 2021. {``A {Model}-{Based} {Method} for {Detecting} {Persistent} {Cultural} {Change} {Using} {Panel} {Data}.''} \emph{Sociological Science} 8 (March): 83--95. \url{https://doi.org/10.15195/v8.a5}.

\leavevmode\vadjust pre{\hypertarget{ref-vaisey2016}{}}%
Vaisey, Stephen, and Omar Lizardo. 2016. {``Cultural {Fragmentation} or {Acquired} {Dispositions}? {A} {New} {Approach} to {Accounting} for {Patterns} of {Cultural} {Change}.''} \emph{Socius} 2 (January): 2378023116669726. \url{https://doi.org/10.1177/2378023116669726}.

\leavevmode\vadjust pre{\hypertarget{ref-visser2004}{}}%
Visser, Penny S., and Robert R. Mirabile. 2004. {``Attitudes in the Social Context: {The} Impact of Social Network Composition on Individual-Level Attitude Strength.''} \emph{Journal of Personality and Social Psychology}, 779--95.

\leavevmode\vadjust pre{\hypertarget{ref-voas2016}{}}%
Voas, David, and Mark Chaves. 2016. {``Is the {United} {States} a {Counterexample} to the {Secularization} {Thesis}?''} \emph{American Journal of Sociology} 121 (5): 1517--56.

\leavevmode\vadjust pre{\hypertarget{ref-wagner2019}{}}%
Wagner, Jenny, Oliver Lüdtke, and Alexander Robitzsch. 2019. {``Does Personality Become More Stable with Age? {Disentangling} State and Trait Effects for the Big Five Across the Life Span Using Local Structural Equation Modeling.''} \emph{Journal of Personality and Social Psychology} 116 (4): 666--80. \url{https://doi.org/10.1037/pspp0000203}.

\leavevmode\vadjust pre{\hypertarget{ref-way2019}{}}%
Way, Samuel F., Santiago Gil, Ian Anderson, and Aaron Clauset. 2019. {``Environmental {Changes} and the {Dynamics} of {Musical} {Identity}.''} \emph{Proceedings of the International AAAI Conference on Web and Social Media} 13 (July): 527--36. \url{https://doi.org/10.1609/icwsm.v13i01.3250}.

\leavevmode\vadjust pre{\hypertarget{ref-wodtke2018}{}}%
Wodtke, Geoffrey T. 2018. {``The {Effects} of {Education} on {Beliefs} about {Racial} {Inequality}.''} \emph{Social Psychology Quarterly} 81 (4): 273--94. \url{https://doi.org/10.1177/0190272518804145}.

\leavevmode\vadjust pre{\hypertarget{ref-zaller1992}{}}%
Zaller, John. 1992. \emph{The {Nature} and {Origins} of {Mass} {Opinion}}. Cambridge {Studies} in {Public} {Opinion} and {Political} {Psychology}. Cambridge: Cambridge University Press. \url{https://doi.org/10.1017/CBO9780511818691}.

\end{CSLReferences}

\end{document}

### 2024-07-29 manuscript/manuscript.Rmd

@@
 output: 
-  bookdown::word_document2: 
-    fig_caption: yes
-    reference_docx: word-styles-reference-02.docx
   bookdown::pdf_document2:
@@
     number_sections: true
+  bookdown::word_document2: 
+    fig_caption: yes
+    reference_docx: word-styles-reference-02.docx
 bibliography: ["kileybib.bib"]
@@
    - \setlength{\parskip}{0em}
-   - \usepackage[markers,nolists]{endfloat}
    - \usepackage{graphicx}
@@
 fontsize: 12pt
-abstract: "Early adulthood is consistently identified as a “formative period” in the life course when people demonstrate a heightened probability of changing their opinions, beliefs, and political preferences.
-This heightened rate of change is often attributed to the social conditions of this life stage but could also reflect biological or cognitive differences between young adults and older adults.
-In this paper, we explore whether substantial changes in the timing and ubiquity of major life-course transitions over the the past 70 years have affected the rate at which early adults and the rest of the population change their opinions on political and social issues.
-Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the 1960s.
-Instead, we find that other age groups have become more stable opinion holders.
-These aggregate trends obscure two countervailing patterns: decreases in stability due to life-course changes among younger adults and increases in stability across all age groups due to increases in educational attainment, as well as a broader stability trend that remains unexplained.
-Our results suggest a braod role in social contexts in shaping attitude stability."
+abstract: 'Early adulthood is consistently identified as a "formative period" in the life course when people demonstrate a heightened probability of changing their opinions, beliefs, and political preferences.
+This heightened rate of change is often attributed to the social conditions of this life stage but could also reflect ontogenetic differences between young adults and older adults.
+In this paper, we draw on data for more than 70 questions that were asked repeatedly across multiple panel surveys stretching from the 1950s to the 2020s to explore whether changes in the social conditions of early adulthood have affected the rate at which early adults and the rest of the population change their opinions on political and social issues.
+Contrary to expectations, we find that early adults have not become more likely to change attitudes and opinions since the late 1950s.
+Instead, we find that other age groups have become more stable opinion holders over this time.
+This aggregate trend reflects three countervailing patterns: small decreases in stability due to life-course changes among younger adults, larger increases in stability across all age groups due to increased educational attainment, and a general increase in stability that remains unexplained.
+Our results suggest broad roles for both individual and larger social contexts in shaping when in the life course people experience attitude change.'
 ---
@@
 
+\doublespace
+
 # Introduction
 
-Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991].
-In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@sears2013; @kiley2020].
-This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins the dynamics of cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].
+Adolescence and early adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making at later life stages [@krosnick1989; @alwin1991].
+In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@kiley2020; @krosnick1989], and that opinions formed in these age ranges tend to persist well into adulthood [@sears2013; @sears1999].
+This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins the dynamics of cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952; @voas2016].
 
 Why people in these life stages appear to show more frequent opinion change relative to older people, however, is unclear.
-Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences," arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007].
-Related explanations conversely focus on the social-structural features of middle age that might stabilize attitudes [@eaton2009; @visser2004].
-At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.
+Social-science explanations tend to argue that the unique social features of different life stages shape whether people are open to or closed off from change during these ages.
+These features include high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences" in early adulthood [@alwin1991; @danigelis2007] and normative expectations around stability for middle-aged people in high-status roles [@eaton2009; @visser2004].
+At the same time, interdisciplinary work on learning across the life-course finds that an early period of active exploration followed by relative stability is a broader phenomenon across organisms, which might be an evolutionary response to the tension between learning and using information in high-information environments [@gopnik2020; @frankenhuis2020; @spreng2019].
+These findings suggest that the differences between early adults and the rest of the population might be more rooted in ontogenetic mechanisms than social mechanisms.
 
-It is likely true that both social and cognitive/biological mechanisms play a role in facilitating opinion change during adolescence and early adulthood, but the relative contributions of each are unclear, and clarifying their relative contributions is important for understanding the mutually reinforcing processes of social change and cultural evolution.
-If the mechanisms are principally cognitive and biological, rather than social, then broader changes in the structure of different life stages are unlikely to affect rates of social change in the same way.
-If, conversely, openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change.
-Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016].
+It is likely true that both social and ontogenetic mechanisms play a role in facilitating opinion change during adolescence and early adulthood, but clarifying their relative contributions is important for understanding the broader relationship between social change and cultural evolution.
+If the mechanisms driving heightened rates of opinion change early in adulthood are principally ontogenetic, then changes in the structure of different life stages would be unlikely to affect rates of cultural evolution.
+If openness to belief and opinion change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing [@cherlin2004; @cherlin2021; @robbins2022], could shape the relative importance of within-person change and cohort effects for explaining broader cultural change [@kiley2020; @vaisey2016].
 
-Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts [@buchmann1989; @bruckner2005].
-Rates of high school and college completion have also increased significantly.
-As a result, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009].
-These broad social changes lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.
+To better understand the mechanisms that drive openness to cultural change, we take advantage of changes in the social conditions of life stages over time.
+Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, establishing an independent household, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts than they did for earlier cohorts [@buchmann1989; @bruckner2005].
+As a result, "early adulthood" has emerged as a distinct life stage, in which people are relatively detached from family and other role obligations and have more freedom to explore ideas and lifestyles [@rosenfeld2009].
+This extension of the social conditions typically thought to open people up to change should be reflected in how much people change over time.
 
-In this paper, we test whether these broader shifts in social structure and the timing of life course transitions have affected the rate at which young people change their attitudes about political and social issues.
-Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we estimate the amount of change in opinions as a function of age and cohort.
-Consistent with the "impressionable years" hypothesis, we find that people in age ranges associated with this "early adulthood" life stage (18-25 and 26-33) have more variable opinions than older respondents.
-However, these differences are not constant over time.
-They emerged over the second half of 20th Century as more recent cohorts exhibited more stable opinions in middle and old age compared to their predecessors, while early adult respondents' reported opinions continued to change at roughly the same rate as in earlier cohorts.
+In this paper, we test whether these shifts in social structure have affected the rates at which people change their opinions about diverse social and political issues during early adulthood.
+Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we quantify broad trends in attitude change and stability over time.
+Consistent with the "impressionable years'' hypothesis, we find that early adults (aged 18-25 and 26-33) have more variable opinions than older respondents.
+However, these differences are not constant across cohorts.
+They emerged over the second half of the 20th Century because more recent cohorts became more stable opinion holders in middle-age than their predecessors, not because they demonstrated more opinion change in early adulthood than their predecessors.
 
-Furthermore, we show that the aggregate patterns in attitude change over time are principally due to forces other than changes in life-course structure.
-While decreased rates of marriage explain a small proportion of the aggregate pattern among early adults, the largest driver of change in attitude stability comes from increased rates of education across all age groups.
-Furthermore, much of the shift in aggregate attitude stability remains unexplained, suggesting broader social structural features such as increased access to information with which to form opinions, political and cultural polarization, and more are the key drivers of attitude stability.
+We show that changes in the timing of life-course transitions do not appear to explain these aggregate patterns of attitude change over time.
+While decreased rates of marriage in early adulthood explain a small proportion of the aggregate pattern among early adults, the largest driver of change in attitude stability comes from increased levels of educational attainment across all age groups.
+At the same time, much of the increase in attitude stability over time remains unexplained, suggesting broader social structural features, such as increased access to information with which to form opinions or political and cultural polarization, might also be the key drivers of attitude stability.
 
@@
 
-The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021].
-The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on what Lizardo [-@lizardo2017] calls "personal culture" -- worldviews, attitudes, dispositions, behaviors, and preferences that manifest at the individual level.
-In political socialization, "generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@gerber1998; @mannheim1952; @bartels2014].
-However, the notion of "impressionable years" is more general than just political preferences.
-The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but their cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@ryder; @fosse2023a; @elder2016].
+The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @elder1974; @guhin2021].
+These frameworks argue that the circumstances in which people are brought up leave an enduring mark on what Lizardo [-@lizardo2017] calls "personal culture" -- worldviews, attitudes, dispositions, behaviors, and preferences that manifest at the individual level.
+In theories of political socialization, for example, "generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@gerber1998; @mannheim1952; @bartels2014].
+However, the notion of "impressionable years" is more general than political preferences.
+The often-invoked concept of cohort effects rests on the assumption that peoples' early years play a key role in shaping a wide variety of tastes, preferences, habits, and dispositions, which then impacts their choices and life outcomes well beyond the time when this period has passed [@ryder1965; @fosse2023a; @elder2016].
 Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990].
-Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.
+Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding social theories of behavior.
 
-These aggregate models of cultural imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time.
-At a basic level, the "impressionable years" hypothesis requires that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age.
-The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a].
-While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [\@].
-These tests tend to find higher rates of attitude change among the youngest adults -- people ages 18-25 -- and little difference between adults of other age ranges.
-More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S.
-adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains more variance than period, the year of observation, for most items in the General Social Survey, suggesting an impressionable-years pattern.
+Evidence for the impressionable years hypothesis exists at the individual and aggregate levels.
+At the individual level, the "impressionable years" hypothesis requires that people demonstrate opinion change early in life followed by a longer period of relative stability, especially in middle age.
+While different attitudes occasionally follow different patterns, this general trend has been shown to be true for political attitudes including ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; and other dispositions such as personality traits [@wagner2019] and cultural tastes [@mulder2010; @ma2021].
+More generally, recent research has found that durable change of social opinion is quite rare among U.S.
+adults [@kiley2020].
+At the aggregate level, cohort effects are consistently strong predictors of a wide variety of social and political attitudes, and Vaisey and Lizardo [-@vaisey2016] find that the year of a person's birth explains more variance in opinion differences than the year of observation for most items in the General Social Survey, suggesting an impressionable-years pattern.
 
-In recent years, researchers have also developed new approaches to quantify "openness" to social influence across the life course, especially around politics [@bartels2014; @ghitza2022].
-This work has produced remarkably similar patterns to previous work with openness peaking from a person's mid-teens through age 25 and declining quickly after that, lending more credence to the "impressionable years" hypothesis [a pattern that seems to hold for music tastes as well @way2019].
-However, this work tends to assume a constant age-based rate of openness across cohorts -- that 20-year-olds in 1955 are open to influence at the same rate as 20-year-olds in 2015 -- which has not been tested.
+In recent years, researchers have developed new approaches to quantify openness to social influence across the life course, especially around politics [@bartels2014; @ghitza2022].
+This work has produced remarkably similar patterns to previous work, finding that openness peaks from a person's mid-teens through age 25 and declines quickly after that, lending more credence to the "impressionable years" hypothesis (a pattern that Way [-@way2019] suggests also holds for music tastes).
+However, this work tends to assume a constant age-based rate of openness across cohorts – i.e.
+that 20-year-olds in 1955 are open to influence at the same rate as 20-year-olds in 2015.
+This assumption remains untested.
+Regardless, there is broad evidence for a consistent pattern of malleability early in life followed by relative stability.
 
@@
 
-While the empirical pattern of early openness followed by relative stability appears quite robust, its cause is not clear.
-At the outset, it is important to note that changing cultural dispositions like attitudes involves at least two distinct processes that can be difficult to disentangle: exposure to new information and openness to change conditional on exposure to new information.
+While the empirical pattern of early openness followed by relative stability appears robust, its causes are not clear.
+Changing political, social, and cultural attitudes involve at least two distinct processes that can be difficult to disentangle in observational data: exposure to new information and change conditional on exposure to new information.
 For example, it could be the case that people of all ages are exposed to new ideas at similar rates, with acceptance of a new idea being a function of age, which would account for the observed age-based patterns in opinion change.
@@
 
-## Biological and Cognitive Mechanisms
+## Ontogenetic Mechanisms
 
-The first potential mechanism for changes in attitude stability over time is rooted in the biology of the human brain and theories about how people process information.
-This approach tends to suggest that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally, whether they encounter it during adolescence or middle age, and are exposed to information at a (more or less) constant rate [@bartels2014; @gopnik2020; @sherratt2018].
-However, because people have already encountered a lot of information by the time they reach adulthood, each new piece of information encountered later has less of an impact on their overall dispositions [@bartels2014].
+One class of explanations for heightened rates of attitude change in early adulthood posits that attitudinal stability and change across the life course is primarily a function of humans’ cognitive architecture, which remains more or less stable across socio-historical contexts.
+One class theories posit that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally and are exposed to information at a (more or less) constant rate as they age [@bartels2014; @gopnik2020; @sherratt2018].
+Because people have already encountered a lot of information by the time they reach adulthood, each new piece of information encountered later has less of an impact on their overall dispositions [@bartels2014].
 From this perspective, the distinctive thing about our early years is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.
 
-It is also possible that there is something unique about human cognition during these periods that makes people ascribe more weight to the information acquired within them than to information acquired at other points.
-Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life [@mata2015].
-Supporting this model is evidence that early formative periods have been observed not just in people, but across species.
-For example, in a study of social learning among birds, Aplin and colleagues [-@aplin2017] show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.
+A second theoretical framework suggests that development of the human mind changes across the life course and that there is something unique about human cognition during early periods that makes humans ascribe more weight to the information acquired within them than to information acquired at other points in the life course.
+Maturational arguments suggest that animal brains exhibit more malleability during childhood and adolescence than later in life [@mata2015, Spreng and Turner], and researchers observe evidence of this increased malleability in a wide variety of species.
+Younger birds, for example, are able to retool their habits in light of new information while older birds were not [@aplin2017].
+Evolutionary theories posit that this pattern is a response to conditions in which environments are unstable and information is plentiful yet unreliable (Fawcett and Frankenhuis 2015; Frankenhuis and Walasek 2020; Walasek, Frankenhuis, and Panchanathan 2021), three circumstances common in human social environments.
+Simulations show that in these circumstances, it makes sense for agents to explore early on, to find something that works, and then to stick to strategies that have worked well in the past.
 
-In light of empirical evidence about these sensitive windows across species, there has been modeling work about why periods of heightened sensitivity would emerge and what contexts would facilitate their emergence (Fawcett and Frankenhuis 2015).
-The central argument from this work is that that the emergence and timing of these sensitive windows depends on the interaction between the access to information, the reliability of information, and the stability of the environment, and that early formative periods are likely to evolve when environments are unstable and information is plentiful yet unreliable (Frankenhuis and Walasek 2020).
-In these circumstances, it makes sense for agents to explore early on, to find something that works, and then to stick to strategies that have worked well in the past.
-Access to different types of information might vary across time and this might explain the emergence of sensitive periods at later stages of the life-course (Walasek, Frankenhuis, and Panchanathan 2021).
-For example, the importance of musical tastes and fashion choices becomes important around adolescence, and, at this stage, we are awash with at best contradictory information about what one should or should not like, or should or should not wear.
-Sensitive periods around the development of these tastes might peak around this period, when there are more signals and when the payoffs are more salient.
-
-The thrust of the biological and cognitive arguments is that, irrespective of the contemporaneous social conditions humans inhabit, millennia of evolution have led to the development of a cognitive architecture that makes them relatively open to changing dispositions at younger ages and relatively resistant at older ages, regardless of their social environments.
-However, while biological mechanisms can shape how people respond to new information when they encounter it, these processes cannot dictate the rate at which individuals are exposed to new information, which might also matter for rates of opinion change.
-For that, we must turn attention to the structuring of the social world.
+The thrust of these ontogenetic arguments is that humans have evolved a cognitive architecture that makes them relatively open to changing dispositions at younger ages and relatively resistant at older ages regardless of their contemporaneous social conditions.
+However, while these mechanisms can shape how humans respond to new information when they encounter it, they cannot dictate the rate at which people are exposed to new information, which is a fundamentally social process.
 
@@
 
-Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct, characterized by sustained social learning and constant changes of social contexts.
-Here we briefly consider some of the ways that the "impressionable years" differ from other life stages that might expose people to new information at higher rates and also affect their willingness to update beliefs, beyond what cognition might allow.
+Humans' early years might be distinct in terms of cognitive plasticity, but they are also socially distinct.
+There are at least four dimensions on which early adulthood differs from other life stages in ways that could produce heightened rates of opinion change, regardless of ontogenetic factors.
 
-First, people in this age group tend to move across social contexts at a much higher rates than people in other age groups.
-In the 11-year window from ages 14 to 25, in which people tend to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year potentially exposing people to new classmates, teachers and other role models, and sources of information.
-People in this age window also show the highest rates of residential mobility, moving for educational and occupational reasons.
-Assuming that new social contexts shape exposure to new information, this heightened contextual mobility during this stage might explain part of the difference between age groups [NEED CITES HERE].
+First, people in these life stages tend to move across social contexts at a higher rate than people in other age groups.
+In the 11-year window from ages 14 to 25 in which existing work suggests people tend to show the highest rates of opinion updating, young people move from middle school to high school to either higher education or the workforce or both, with each year potentially exposing people to new classmates, teachers and other role models, and sources of information and restructuring core discussion networks [@small2015].
+People in this age range also show the highest rates of geographic mobility [@geist2008].
+Assuming that new social contexts shape exposure to new information, this heightened contextual mobility might explain part of the difference in attitude stability between age groups [@billari2010].
+Social, contextual, and geographic mobility might also free people to adopt new opinions as they move to new contexts and are unburdened by previous friends’ or family members’ expectations.
 
-Second, people in the "impressionable years" window spend their time in different social contexts than people in middle age.
+Second, in addition to inhabiting a greater diversity of social contexts during these years, people in adolescence and early adulthood spend their time in social contexts explicitly designed to affect attitudes and opinions.
 Data from the American Time Use Study shows that people under age 25 spend a larger portion of the day on leisure time and educational activities and a smaller part of the day doing work and work-related activities than people between the ages of 25 and 55.
-These contexts, especially education, are designed to expose people to new information in ways that can lead to new opinions.
-For example, there is some evidence that college attendance leads to changes in opinions about civil liberties, political views, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic2021].
-Personal social networks also differ in this time period.
-Young peoples' networks tend to be larger and include more friends than kin [@marsden1987].
-It could be the case that these larger "weak tie" networks expose younger people to more new information that drives opinion change [@granovetter1973; @burt2005].
+These contexts expose young adults to new information in ways that can lead to new opinions.
+For example, college attendance appears to change some opinions about civil liberties, politics, gender egalitarianism, racial inequality, and morality [@campbell2016; @wodtke2018; @edelmann2023; @brocic2021].
 
-Third, these are periods are also culturally distinct, with exploration and self-discovery expected and encouraged.
-People in this "early adulthood" phase are not expected to be stable adults, something should not be taken for granted given social variation in norms around when maturity is reached [@cunningham2007; @elder1994].
-Cultural expectations to engage in new experiences, meet new people, and adopt new personalities could motivate the pursuit of new information during this life stage apart from social structural opportunities.
+Third, the structure of personal social networks is distinct in this age range, with young peoples' networks tending to be larger and including more friends than kin compared to networks in middle age [@marsden1987].
+Having a higher number of "weak ties" in their networks can expose younger adults to new information at higher rates [@granovetter1973; @burt2005].
+And because people are less likely to adopt new beliefs if they are embedded in dense networks that reinforce the holding of existing beliefs [@stark1980], the denser, more homogenous networks that characterize middle-age [@marsden1987; @diprete2011] might make it more difficult for people in this life stage to update opinions even if they receive new information [@visser2004].
+Conversely, young peoples’ less-dense networks might allow them to close off parts of their lives to people who might not be supportive of new attitudes.
 
-## Updating Conditional on Exposure
+Fourth, there is significant social variation in norms around when maturity is reached [@cunningham2007; @elder1994], and contemporary early adults have little cultural pressure to resemble middle-aged adults.
+Cultural narratives encourage exploration and self-discovery during early adulthood, pushing young adults to pursue new experiences, meet new people, and adopt new personalities during this “age of independence” [@rosenfeld2009].
+These cultural expectations might also shape whether people feel capable of updating opinions in the face of new information.
+People appear to value stability in people holding social roles with authority [@eaton2009].
+As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information.
+This potentially explains why some work finds that people appear to open up to changing their minds as they enter later life, when these social expectations of stability potentially diminish [@danigelis2007].
 
-While the distinct social and cultural conditions of adolescence and early adulthood might expose young people to new information at higher rates than older people, the impact of this information on people's attitudes is contingent on their ability to incorporate it into their cultural repertoire.
-The social conditions of different life stages might make it easier or more difficult for people to act on these new sources of information in ways that facilitate opinion change.
-
-Social networks not only shape peoples' exposure to new information, but they can also affect the degree to which people adopt new beliefs and attitudes.
-People are much less likely to defect from their traditional beliefs and adopt new beliefs if they are embedded in dense networks that reinforce the holding of existing beliefs, even when they are exposed to new ideas [@stark1980].
-Consistent with this, mid-life is associated with denser, more homogenous networks that include larger numbers of family [@marsden1987; @diprete2011], which experimental research suggests reduces the propensity to adopt new beliefs [@visser2004].
-If young people exist in less dense, more heterogeneous social networks, it might be easier for them to update their networks to reflect new beliefs rather than constrain their beliefs to their social networks.
-Similarly, if information flows less freely through weak-tie networks, these structures might allow respondents to limit their exposure to or close off parts of their lives to people who might not be supportive of new attitudes.
-
-Previous work on opinion stability over the life course suggests that people tend to value stability in figures holding social power, which tend to occur more frequently during middle age.
-As people move into roles with more social power in mid-life, both in family and occupational contexts, they potentially feel this expectation from others and act accordingly, closing themselves off to updating beliefs even as they are exposed to new information or forcing them to develop stable opinions on issues where previously they felt capable of ambivalence [@eaton2009].
-This potentially explains another phenomenon: that people appear to open up to changing their minds on some issues as they enter later life and retirement, when these social expectations of stability presumably diminish [@danigelis2007].
-
-Existing research suggests that the structure of social life during adolesence, early adulthood, and later life stages shape both exposure to new information and a person's ability to update attitudes, opinions, and beliefs in the face of that information.
-This further suggests that changes in these life course stages should matter for the amount of opinion change we observe.
+As the above suggests, life stages vary in myriad ways that could affect how people are exposed to new information and whether they can act on that information in ways that produce changes of opinion.
+As a result, changes in the duration of these life stages could have significant ramifications for how much opinion change we see in a population.
 
@@
 
-In the United States, the last seven decades have seen considerable social and cultural changes around adolescence and young adulthood [@buchmann1989; @rosenfeld2007; @rosenfeld2005].
-The expansion of higher education and increased possibilities for mobility are among the many social processes that have led to changing expectations for the trajectories that young adults should follow [@rosenfeld2007].
-These changes have made life course trajectories less predictable, and the "script" of young adulthood has become less rigid: it might involve prolonged higher education, moving to a different city, or dating without a stable romantic partner [@buchmann1989].
-For the purposes of this paper, there are two related trends that are particularly important: the elongation of young adulthood and the intensification of this life stage.
+The social and cultural conditions of early adulthood have changed considerably in the past seven decades in the United States [@buchmann1989; @rosenfeld2009; @rosenfeld2005], as expansion of higher education and increased possibility for geographic mobility have reshaped expectations around life-course trajectories [@rosenfeld2009].
+While cultural norms previously encouraged marriage, stable employment, and childbearing shortly after the completion of secondary education, relaxation of these norms means that early adulthood now might involve prolonged higher education, moving to a different city, or dating without a stable romantic partner [@buchmann1989, @billari2010].
+There are two related trends that are particularly important: the elongation of young adulthood and the intensification of this life stage.
 
@@
 
-First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for successive cohorts as they pass through the same ages.
-Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce.
-In the United States, both the median age at first marriage and median age of first childbirth has consistently risen since the mid-20th Century [@rosenfeld2009].
-We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021].
-Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts.
+The period between adolescence and "settling down" has gotten longer over time [@billari2010].
+Buchmann [-@buchmann1989] finds that young adults increasingly postpone the rituals associated with adulthood, including marriage, childbirth, and entry into the workforce.
+In the United States, the median age at first marriage and median age of first childbirth both occur about eight years later than they did in the mid-20th Century [@rosenfeld2009], and rates of non-marital childbearing have also increased across all demographic groups [@cherlin2021].
+Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts than they did for previous cohorts.
 
-Similarly, the duration of education has extended, with more people completing high school, college, and, increasingly, graduate degrees before entering the workforce or movement back into education after several years in the workforce.
-These transitions are also accompanied by heightened rates of geographic mobility, as well as movement between the workforce and education.
+Similarly, more recent cohorts spend more years pursuing education, with a larger proportion of the U.S.
+population completing high school, college, and, increasingly, graduate degrees before entering the workforce [@schofer2005].
+This shift has been particularly dramatic for women, whose educational attainment and labor-force participation are much higher than 70 years ago [@england2010].
+People now also move between the workforce and education at higher rates than they did in the past [@rosenfeld2009].
 
-If the distinctive nature of "impressionable years" is principally driven by their distinct social circumstances, then these shifts have the potential to extend these social circumstances into later ages.
-This, in turn, could have an effect on how long exploration, and therefore openness to change, lasts throughout the life-course.
-If this is the case, we should expect to see heightened levels of attitude change among age groups that, for earlier cohorts, displayed mainly stability.
-In other words, people who were born in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age.
-Specifically, we expect cohorts who entered adulthood in the late 1950s to be fairly comparable to older adults in their rates attitude change but that subsequent cohorts will diverge from older adults, making more change over time.
+If the heightened rates of attitude change during peoples’ "impressionable years" is principally driven by the distinct social circumstances of adolescence and early adulthood, then the delay of full adulthood has the potential to extend these social circumstances into later ages, extending heightened rates of attitude change later in the life course.
+While people born in the mid-20th century might have displayed mainly opinion stability by the time they reached their mid-twenties, when they married and had children, people from more contemporary cohorts, unmarried and childless at the same ages, might show higher rates of opinion change.
+Specifically, we expect cohorts who entered adulthood in the late 1950s to be fairly comparable to contemporaneous middle-aged adults in their rates of attitude change, but that these age groups will diverge in their rates of change in more recent years.
+This leads us to our first hypothesis:
 
-*Hypothesis 1: More recent cohorts will show higher rates of attitude change in the age range of 18 to 25 than cohorts born in earlier years.*
+*Hypothesis 1: More recent cohorts will show higher rates of opinion change in the age range of 18 to 25 than cohorts born in earlier years.*
 
-We expect a similar but more muted trajectory for the attitude behavior of people aged 25-33.
-Again, we expect the oldest cohorts to resemble middle-aged respondents when they enter this age range, as has been shown in previous research [@krosnic1991; @alwin1991], but that more recent cohorts will come to more closely resemble the 18-25 age group, diverging from middle aged respondents.
+We focus on people 18 to 25 for two reasons.
+First, existing work has explored attitude stability and change among people in this age range in earlier cohorts [@krosnick1991; @alwin1991], providing a baseline for what to expect.
+Second, this group has seen the most significant change in life course transitions of all age groups.
+Whereas most people in earlier cohorts were married with children by the midpoint of this age range, the median age of marriage and first childbirth is now outside this group for both men and women.
+We expect a similar but more muted trajectory for attitude behavior of people aged 25-33.
+Again, we expect the oldest cohorts to resemble middle-aged respondents when they enter this age range, as has been shown in previous research, but that more recent cohorts will come to more closely resemble the 18-25 age group.
+This produces our second hypothesis:
 
@@
 
-Consistent with life-course structuring theories, we expect these divergences will principally be explained by changes in the timing of life course transitions such as marriage and childbirth.
-Controlling for these life-course transitions should eliminate the gap between the youngest adults and later adults.
+If these differences are principally driven by the structure of the life course at these stages, then we would expect that, controlling for these life-course transitions, the gap between early adults and older adults will disappear.
+This leads to our third hypothesis:
 
-*Hypothesis 3: Controlling for rates of marriage and childbirth, more recent cohorts' rates of attitude change in early adulthood (18-33) will not be different from the rates of attitude change among contemporaneous middle aged adults.*
+*Hypothesis 3: Controlling for rates of marriage, childbirth, and having a child at home, more recent cohorts' rates of attitude change in early adulthood (18-33) will not be different from the rates of attitude change among contemporaneous middle aged adults.*
 
@@
 
-A second trend in early adulthood relates to just how much exploration young adults can engage in during these "impressionable years." In his work, Rosenfeld [-@rosenfeld2007] argues that one of the processes underlying the rapid rise in same-sex and interracial marriages in the United States is that young adults are increasingly escaping the controlling gaze of their parents.
+Rosenfeld [-@rosenfeld2009] argues that one of the processes underlying the rapid rise in same-sex and interracial marriages is that young adults have increasingly escaped their parents’ authority.
 When a young person lives with, or is financially dependent on, his or her parents, the latter have more say about their child's choice of partner.
@@
 
-Rosenfeld's work suggests that the social conditions that shape where people spend their impressionable years might also shape how much cultural exploration they can engage in.
-An increase in independence might entail more exploration and greater ability to update attitudes as a result of this exploration.
-This assertion in turn engenders a second prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.
+This work suggests that increased independence might enable greater exploration and freedom to update attitudes beyond that facilitated life-course transitions.
+This framework suggests that freedom of exploration is not just dependent on life-course roles (marriage, parenthood, etc.) but also influenced by broader cultural expectations.
+In other words, married 25-year-old men and women in 1956 might have experienced an additional set of social constraints that married 25-year-old men and women in 2016 do not experience, leaving the latter much more open to developing new opinions.
+The "intensification" of impressionable years suggests a fourth hypothesis that challenges the third:
 
-This approach suggests that it is not the timing of life events that is important for rates of change, but the broader social conditions of early adulthood that might make people more malleable in recent cohorts.
-In other words, unmarried, childless 25-year-old men and women in 1956 might have experienced a set of social constraints that unmarried, childless 25-year-old men and women in 2008 do not experience, leaving the latter much more open to developing new opinions than their predecessors.
-The "intensification" of impressionable years suggests two hypotheses.
-
-*Hypothesis 4: Controlling for life course transitions, more recent cohorts of early adults (18-33) will make higher rates of attitude change than contemporaneous middle-aged adults.*
-
-## Biological and Cognitive Factors
-
-We have framed our hypotheses as expectations derived from the framework suggesting that social conditions play a large role in explaining the uniquely high rates of attitude change among early adults and adolescents.
-However, to the extent that we observe significant changes in the social context of early adulthood but no corresponding changes in rates of attitude change or response stability across cohorts, we take this as evidence that social circumstances play a relatively small role in explaining change in favor of biological and cognitive factors.
+*Hypothesis 4: Controlling for life course transitions, more recent cohorts of early adults (18-33) will still make higher rates of attitude change than contemporaneous middle-aged adults.*
 
@@
 
-Finally, while we have primarily focused on changes in the timing and structure of life course stages over the past 70 years, that time frame has also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure.
-We briefly consider three broader changes here.
+We have focused primarily on changes in the timing and structure of early adulthood over the past 70 years, but that time frame has also seen significant shifts in the broader social and cultural environment that might affect attitude stability independent of changes in life course structure.
 
-First, the number of people completing high school and pursuing and completing higher levels of education has increased significantly over the window studied here [@schofer2005; @hout2012].
+First, the number of people completing high school and pursuing higher levels of education increased significantly during the 20th and early 21st Centuries [@schofer2005; @hout2012].
 Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007].
 The specific mechanism that links educational attainment and attitude stability is unclear, with some perspectives suggesting educational attainment improves cognitive function and memory while other perspectives suggest that educational attainment makes people pay more attention to current events.
-It could also be the case that education cultivates in people a normative expectation that they should hold opinions.
-Regardless of the specific mechanism, a population with higher levels of education, on average, might be more likely to hold stable opinions regardless of transformations in life course stages.
+It could also be the case that education cultivates in people a normative expectation that they should hold stable opinions.
+Regardless of the specific mechanism, people with higher levels of education might be more likely to hold stable opinions regardless of where they are in the life course, and differences in educational attainment across age groups could produce differences in attitude stability.
+This leads to our fifth hypothesis:
+
+*Hypothesis 5: Cohorts with higher levels of education will demonstrate higher levels of attitude stability at all age ranges.*
 
 Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding.
-Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire information that makes opinion-holding easier.
-To the extent that stable opinions are in part a reflection of signals from elites [@converse1965; @zaller1992], then we should expect these shifts to stabilize opinions across all age ranges.
+Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire information that makes stable opinion holding easier.
+To the extent that stable opinions are in part a reflection of signals from elites [@converse1964; @zaller1992], then we should expect these shifts to stabilize opinions across all age ranges.
 
-Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina2008; @baldassarri2008].
+Third, the past 70 years have seen elite polarization around political and social affairs, as well broader public partisan realignment [@fiorina2008; @baldassarri2008].
 People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta [-@dellaposta2020] calls an "oil spill" of polarization.
 While clarity around political issue positions is most commonly studied, the same forces can also align other social beliefs and identities.
+Geographic and social sorting into lifestyle "enclaves" also reinforce these processes, continually providing clearer signals about which opinions people should hold [@dellaposta2015].
 This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently [@zaller1992].
-Geographic and social sorting into lifestyle "enclaves" also reinforce these processes, continually providing clearer signals about which opinions people should hold [@dellaposta2015].
 
-Broadly speaking, these three forces suggest that opinions should broadly become more stable over time, regardless of life-course stages.
+Broadly speaking, these forces suggest that all age groups’ opinions will be more stable in successive cohorts, regardless of life-course stages.
 If this is the case, we might not observe an absolute decrease in stability for early adults, but rather a relative decrease, as extension of the life course works to decrease stability for members of these age groups while broader social forces stabilize opinions for all groups.
+This produces our final hypothesis:
+
+*Hypothesis 6: At all ages, more recent cohorts will demonstrate higher levels of attitude stability, controlling for educational attainment and life-course transitions, than earlier cohorts.*
+
+We have outlined hypotheses derived from the theoretical model that suggests that social conditions are the key driver of higher rates of attitude change among early adults and adolescents.
+To the extent that we observe significant changes in the social context of early adulthood but no corresponding changes in rates of attitude change or response stability across cohorts, we would take this as evidence that social circumstances play a relatively small role in explaining rates of attitude change compared to ontogenetic factors.
 
@@
 
+Our hypotheses can be adjudicated with a set of theoretical estimands [@lundberg2021].
+Our comparisons suggest a set of theoretical estimands regarding the difference in opinion change for Americans in the same age range born in different cohorts.
+We seek to compare the average amount of opinion change we observe for a particular cohort during a particular age range to the average amount of opinion change for a cohort born 10 years later, conditional on age range.
+To be clear, we do not seek to estimate this quantity assuming only the year someone was born is different and holding all other factors constant.
+It is impossible to separate out the age at which people of a certain cohort answer questions from the year in which they answer them [@fosse2019].
+Cohorts also differ in other ways, including size and demographic composition, that might be relevant for explaining differences in the rates of change they demonstrate.
+This set of theoretical estimands is principally descriptive, with variation in the social experiences of people in successive cohorts and even the demographic composition of the cohort itself being potential explanations for whatever gaps we observe.
+
+\begin{equation*}
+ \begin{aligned}
+  \tau_1 = \frac{1}{n}\sum_{c=1}(Y_{c}-Y_{c+10})
+ \end{aligned}
+ \end{equation*}
+
+In this theoretical estimand, $Y_{c,a}$ captures the expected amount of change in responses for a cohort at age range $a$, and $Y_{c+10,a}$ captures the expected amount of change in responses for a cohort born 10 years later at the same age range, with a fixed duration between observations.
+We seek to average the quantity across cohorts captured in panel surveys from the 1950s onward.
+
+Our second set of theoretical estimands captures how differences in the rate of change between age groups itself changes over time, specifically whether the gap between how much early adults (either 18-25 or 26-33) change over time and how much middle-aged adults (34-64) change over time itself is different in different years.
+Again, we do not assume that the composition of any age group is fixed, nor that later cohorts experience the same social environment as earlier cohorts.
+
+\begin{equation*}
+ \begin{aligned}
+   \tau_2 = \frac{1}{n}\sum_{t=1}([Y_{t,\text{ma}} - Y_{{t},\text{ea}}] - [Y_{{t-10},\text{ma}} - Y_{{t-10},\text{ea}}])
+ \end{aligned}
+ \end{equation*}
+
+In this estimand, $Y_{t,\text{ea}}$ reflect the average change in responses for early adults at a particular time $t$ and $Y_{t,\text{ma}}$ reflects the average change in responses for middle aged respondents at the same time.
+$Y_{t,\text{ea}}$ and $Y_{{t-10},\text{ea}}])$ reflect these same rates observed in the population 10 years earlier.
+In contrast to the previous estimand, this estimand is indexed by t, or time in years, not cohort, and we average over all times captured since the late 1950s.
+Finally, hypotheses 3, 4, 5, and 6, imply theoretical estimands reflecting the difference between cohorts as they pass through the same age range, conditional on cohorts having the same levels of certain demographic characteristics.
+Specifically, they focus on whether differences in life course transitions and educational attainment mediate the observed differences between cohorts at different time points.
+
+\begin{equation*}
+ \begin{aligned}
+   \tau_3(e,l) = \frac{1}{n}\sum_{c=1}(Y_{c,a}(E=e,L=l)-Y_{c+10,a}(E=e,L=l))
+ \end{aligned}
+ \end{equation*}
+
+This framework repeats that of $\tau_1$, comparing a focal cohort to the cohort that came 10 years later, while holding both cohorts’ levels of educational attainment, $E=e$, and life course transition rates, $L=l$ to the level of the earlier cohort.
+We again average this difference across all cohorts observed since the late-1950s.
+In comparison to $\tau_1$ and $\tau_2$, however, this estimand requires a counterfactual component that is not directly observed.
+
 ## Data
 
-Testing our hypotheses requires data on opinion and attitude reports over time for members of different cohorts in the same age ranges, with respondents coming from the same general population.
-We also require information about life-course transitions standardized across these panels.
-To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years.
+Generating empirical estimands for these theoretical estimands requires data on opinion and attitude reports over time for members of different cohorts in the same age ranges, with respondents coming from the same general population.
+We also require information about life-course transitions standardized across panels.
+To construct such a data source, we combine panel surveys conducted by the American National Election Studies (ANES) and the General Social Survey (GSS) spanning about 60 years.
 Both the ANES and GSS are designed to generalize to the non-institutionalized U.S.
 adult (18+) population, and therefore can be thought of as tapping the same general population.
-We included a total of 12 panels in our search: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20.
-We evaluated an additional panel conducted by the ANES in 2008-10, but no questions overlapped this panel and any other panel.
+We included 12 panels in our final dataset: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20.
 
-For reasons explained below, we identified social and political attitude questions that were asked in at least three different panels with at least one of those panels asking respondents to answer the question at three or more time points.
-We sought questions that tapped a respondent's report of their own belief, avoiding questions that asked respondents to report how others, including politicians and political parties, would respond.
-Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the exact same set of questions, we did not treat these as separate panels for the purposes of identifying three distinct time points.
-In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis.
-We allowed for question wording to vary slightly across panels (e.g., feeling thermometers about the "women's liberation movement" were combined with questions about the "women's movement").
-When the same or similar questions had slightly different response scales, we harmonized response options to the question structure with the fewest response options.
+For reasons related to our modeling strategy explained below, we identified social and political attitude questions asked in at least three different panels with at least one of those panels asking respondents to answer the question at three or more time points.
+Questions had to ask respondents to report their own opinions, not the positions or opinions of others.
+Because the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same questions, we did not treat these as separate panels for the purposes of identifying three distinct time points.
+Questions needed to appear in these panels and an ANES panel to be included in our analysis.
+We allowed for question wording to vary slightly across panels (e.g., feeling thermometers about the "women's liberation movement" were combined with feeling thermometers about the "women's movement").
+When similar questions had different response scales, we harmonized response options to the question structure with the fewest response options.
 This was most common when one panel asked whether people agreed or disagreed while another included options for "strong" agreement and disagreement.
-We broadly followed the ANES time series' lead in which questions to harmonize.
+We followed the ANES time series Cumulative Data File codebook in deciding which questions to harmonize.
 
-Our search produced a total of 77 questions tapping a broad range of topics including political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.
-For reasons elaborated below, we scaled all items to have a minimum of 0 and a maximum of 100.
+Our search produced a total of 77 questions tapping political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views of government efficacy and trust in government, views on social issues and social change, and sentiment toward various political and social groups.
+For reasons elaborated below, we scaled items to have a minimum of 0 and a maximum of 100.
+Table A1 in Appendix A outlines the questions included in our analyses, including their full question wording, number of waves and panels, and response scales.
 
-Table A1 in Appendix A outlines the questions included in our analysis, including their full question wording and response scale.
+### Opinion Change
+
+Our outcome of interest is the amount of change in a person’s responses across two instances of being asked the same question, meaning our unit of analysis is a pair of waves by the same person, rather than a single observation or a single person.
+If a person responds to a question K times in a panel, he or she will have $(K*(K-1))/2$ observations for that item in our data set.
+Because we are unconcerned with the direction of change between responses, we take the absolute value of the difference between each two observations of a person’s response to the same question.
+Because all questions are scaled between 0 and 100, this takes on a value between 0 and 100.
 
@@
 
-The central question of our analysis pertains to age and cohort, and as such we classify respondents into a set of age brackets depending on their age when they answered survey questions.
-Age was reported at least once in all panels.
-If respondents' were not asked their age or birth date in a specific wave, we calculated it based on other age reports.
-Given our theoretical questions, we classify people into four age categories: 18-25, 26-33, 34-64, and 64-80.
-Because of differences in both question structure and age structure across panels, we dropped respondents who are over 80.
+The central questions in our analysis pertain to age, cohort, and the time of the interview.
+All surveys included the date when the survey interview took place.[^1] Age was reported at least once in all panels.
+If respondents' were not asked their age or birth date in a specific wave, we calculated it based on other age reports and the date of the interview.
+We classify respondents into a set of age brackets depending on their age when they answered the survey questions: 18-25, 26-33, 34-64, and 64-80.
+Because of differences in the wording of age questions across panels, we dropped respondents who were older than 80 in the first observation of a wave pair.
 
-Our theoretical questions also focus on the effect of having undergone a set of life-course transitions on attitude stability.
-To this end, we include a set of indicator variables capturing whether the respondent had undergone certain demographic transitions.
-These include 1) whether a respondent is currently enrolled in school; 2) whether a respondent has ever been married; and 3) whether a respondent has ever had a child.
-We also include an indicator for whether a respondent reported completing college to separately explore whether higher rates of college completion explain changes over time across age groups.
+[^1]: In the rare cases that a survey interview for a single wave of the survey occurred on two distinct days, we classified the date according to the day the interview started.
 
-Broadly speaking, questions tapping marital status, student status, and degree attainment were asked in the same way with the same response options in each panel or could be easily harmonized.
-It was quite common for respondents to have missing data on one or more of these demographic variables for a particular wave of any panel, even if they participated in the panel.
-This was most common when a question occurred in the post-election wave of the ANES, but demographic variables were captured in the pre-election wave.
-If this occurred, we carried over the pre-election response, as these waves tended to be separated by only a few weeks.
+Our second set of theoretical questions focuses on the relationship between life-course transitions and attitude change.
+Questions tapping marital status, student status, and degree attainment were asked in similar ways and could be easily harmonized.
+We generated indicator variables capturing whether the respondent is currently enrolled in school and has ever been married.
+We also include indicators for whether a respondent reported completing college or did not complete high school.
+It was common for respondents to have missing data on one or more of these demographic variables for a particular wave of any panel, even if they participated in the wave.
+This was most common when a question was asked in the pre-election wave of the ANES but not in the post-election wave.
+When this occurred, we carried over the pre-election response, as these waves tended to be separated by only a few weeks.
+If these questions were missing from a wave and someone had previously reported ever being married or ever having a child, we carried over that report to subsequent waves.
+If their status at the time of an interview was unclear, we coded the response as missing.
 
-In contrast to the other demographic variables, generating an indicator variable for whether a respondent ever had a child was more challenging, as question wording and structure varied significantly across panels.
-Specifically, survey administrators occasionally asked respondents how many children were currently living with them, while at other times asked whether the respondent had any children and how many.
-While these items are highly correlated for respondents with young children, these different approaches to the question produce significant differences in rates among parents whose children have left home.
+Generating an indicator for whether a respondent ever had a child was more challenging, as question wording and structure varied across panels.
+Some waves asked respondents how many children were currently living with them while others asked whether the respondent had any children and how many.
 To account for this, we generated two variables.
-First, we created an indicator variable for whether they reported having at least one child in the house under 18 ("kidinhouse").
-Respondents who were not asked the question in a particular wave (or temporally proximate wave) were coded as missing.
+We created an indicator variable for whether a respondent reported having at least one child in the house under 18.
+Respondents who were not asked the question in a wave (or temporally proximate wave) were coded as missing.
 Second, we gave respondents an indicator for whether they had ever had a child.
@@
 
+We imputed missing demographic data for respondents using multiple imputation with chained equations.
+Multiple imputation approaches the question of missing data as one of prediction.
+It uses existing relationships in the data to create a predicted value for missing data, which is consistent with the patterns in the data.
+But it recognizes that there is uncertainty in these predictions, so it iterates over this process several times.
+Different imputed values reflect uncertainty in the predictions that can be drawn from relationships in the data.
+In the analysis stage, all predicted values are used to incorporate the uncertainty in the imputed values.
+We used other demographic variables as well as attitude difference reports to impute missing values, though no outcomes were imputed this way.
+We also included sex (male or female) and race (white, black, or other), both recorded in all waves, as predictors in these imputation equations.
+We created a set of 5 different imputed datasets to acknowledge uncertainty in this process, and our final analyses are conducted across all imputed datasets.
+
 ## Methods
 
-Our outcome of interest is the absolute change in responses a respondent makes across a pair of observations which, due to scaling, can be between 0 and 100.
+When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in overall attitude stability (our main theoretical interest) or to the structural features of the survey.
+Centrally, the amount of change people make in their responses over time is likely affected by the duration of time between responses, with longer duration between responses leading to more wave-to-wave change as people experience more shocks that could shift their opinions.
+In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
+Accounting for these differences is a key part of our analytical approach.
 
-When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time (our main theoretical interest) or to the structural features of the survey administration.
-Specifically, variation in the amount of change people makes in their responses observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions.
-In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.
+Consider a one-item, one-panel case (e.g., partisan identification in the 1972-76 panel).
+We model the absolute change between two responses to the same question by the same person as a function of an intercept and the duration between observations in years.
+In this equation, $i$ indexes a pair of observations, not a person.
 
-To illustrate our approach, consider a one-item, one-panel case (e.g., partisan identification in the 1972-76 panel).
-We model the absolute change from wave-to-wave as a function of an intercept, the duration between observations in years, age categories, and the interaction between duration and age categories.
-
-$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \epsilon_i $$
+\begin{equation*}
+ \begin{aligned}
+   |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \epsilon_i
+ \end{aligned}
+ \end{equation*}
 
 In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations by the same respondent.
-Importantly, respondents might have multiple observation pairs for each panel depending on the number of times a question was asked.
-In the 1972-76 ANES panel, for example, respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
-If a person participates in all three waves, they have three observations of $|y_{2i}-y_{1i} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).
+Importantly, respondents might have multiple observation pairs for each panel depending on the number of times they answered a question.
+In the 1972-76 ANES panel respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
+If a person participates in all three waves, they would have three observations of $|y_{2i}-y_{1i} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).
 
-Variation in duration between observations within panels lets us separate two kinds of change: 1) change attributable to durable shifts of opinions over time that can be associated with duration between observations and 2) change that is principally random, or unrelated to the duration between observations.
-If respondents have a stable underlying partisan identification they report with random error, and no real change occurs between waves beyond that, then in expectation, duration should be unrelated to absolute change in responses over time, and $E(\beta_1) = 0$.
-In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change.
-Conversely, $\beta_1$ captures the expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$).
-For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.
+Variation in duration between observations within panels lets us separate two kinds of change: 1) change that is principally random, or unrelated to the duration between observations, captured in $\beta_0$, and 2) change associated with duration between observations, captured in $\beta_1$.
+If respondents have a stable underlying partisan identification they report with random error and no real change occurs between waves beyond that, duration should be unrelated to absolute change in responses over time in expectation, and $E(\beta_1) = 0$.
+If a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$.
 
-Given our focus on age-based differences, we can extend equations to incorporate age groups as indicator variables:
+Given our focus on age-based differences, we extend this equation to incorporate age groups as predictors of absolute difference:
 
-$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \epsilon_i $$
+\begin{equation*}
+ \begin{aligned}
+   |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \\
+   \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \epsilon_i
+ \end{aligned}
+ \end{equation*}
 
-In this framework, $\beta_0$ and $beta_1$ now capture the expected change in responses over time for no time between waves and each additional year, respectively, for respondents ages 18-25 (the reference category).
-$\beta_2$ through $\beta_4$ capture the difference in baseline change between observations for the older groups, and $\beta_5$ through $\beta_7$ capture differences in the effect of an additional year duration for the older age groups.
+In this framework, $\beta_0$ and $beta_1$ now capture the expected change in responses that is unrelated to time and each additional year, respectively, for respondents ages 18-25 (the reference category).
+$\beta_2$ through $\beta_4$ capture the difference in baseline change between observations for the older groups and $\beta_5$ through $\beta_7$ capture differences in the effect of an additional year duration for the older age groups.
 
-Our theoretical question deals primarily with whether these short-term rates of change over time themselves change across the broader time span captured across panels.
-To explore that, we can add an additional term for time and interact all previous coefficients outlined above with time.
+Our theoretical questions deal primarily with whether these relationships change over time and whether differences between groups change across the broader time span captured across panels.
+To explore whether the rate of change changes over time, we add an additional term for time (in decades) and interact all previous coefficients outlined above with time.
 
-$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1(d) + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \beta_8(t) + \beta_9(d*t) + \beta_{10}(a_{26-33}*t) + \beta_{11}(a_{34-64}*t) + \beta_{12}(a_{65+}*t) + \beta_{13}(d*a_{26-33}*t) + \beta_{14}(d*a_{34-64}*t) + \beta_{15}(d*a_{65+}*t) + \epsilon_i $$
+\begin{equation*}
+ \begin{aligned}
+   |y_{2i}-y_{1i} | = \beta_0 + \beta_1(d) + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \\ 
+   \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \\
+   \beta_8(t) + \beta_9(d*t) + \\
+   \beta_{10}(a_{26-33}*t) + \beta_{11}(a_{34-64}*t) + \beta_{12}(a_{65+}*t) + \\
+   \beta_{13}(d*a_{26-33}*t) + \beta_{14}(d*a_{34-64}*t) + \beta_{15}(d*a_{65+}*t) + \\
+   \epsilon_i
+ \end{aligned}
+ \end{equation*}
 
-In this formulation, $\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to $t = 0$, which we set to be the most recent observation date (2016).
+In this formulation, $\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to $t = 0$, which we set to be the most recent observation date (2020).
 $\beta_8$ estimates the change in the baseline rate of change for the reference group.
 $\beta_9$ captures change in the effect of duration of observation over time for people in the reference group (ages 18-25).
-Coefficients $\beta_{10}$ through $\beta_{12}$ and $beta_{13}$ capture how differences in the baseline rate of change and duration effects between the reference group and older age categories change over time.
+Coefficients $\beta_{10}$ through $\beta_{12}$ and $beta_{13}$ capture how differences in the baseline rate of change and duration effects between the reference group and older age categories change over time.[^2]
+
+[^2]: We index time by the year of observation in decades, but it would be functionally similar to index time by cohort.
 
 To this point, we have developed our model for the case of a single question.
-However, because different panels tend to ask different questions, we lack a large number of observations for any particular item, meaning separate estimates of the time trends for all these items are likely to be inconclusive.
-One solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels.
+While a small number of questions are asked in almost every panel, most are only asked in a few panels.
+As such, separate estimates of time trends for all items are likely to have a low degree of reliability.
+One solution is to pool questions and estimate a single time trend for stability in attitudes over time.
+That approach is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels.
 It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable.
+Multilevel models [@gelman2007; @raudenbush2002] provide an alternative solution to estimate variation in time trends in the presence of uncertainty.
+We follow this approach, using the model:
 
-Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty. We follow this approach, using the model:
-
-$$ |y_{2i}-y_{1i} | = \beta_{0j} + \beta_{1j}(d) + \beta_{2j}(a_{26-33}) + \beta_{3j}(a_{34-64}) + \beta_{4j}(a_{65+}) + \beta_{5j}(d*a_{26-33}) + \beta_{6j}(d*a_{34-64}) + \beta_{7j}(d*a_{65+}) + \beta_{8j}(t) + \beta_{9j}(d*t) + \beta_{10j}(a_{26-33}*t) + \beta_{11j}(a_{34-64}*t) + \beta_{12j}(a_{65+}*t) + \beta_{13j}(d*a_{26-33}*t) + \beta_{14j}(d*a_{34-64}*t) + \beta_{15j}(d*a_{65+}*t) + \epsilon_i $$
+\begin{equation*}
+ \begin{aligned}
+   |y_{2ij}-y_{1ij} | = \beta_{0j} + \beta_{1j}(d) + \beta_{2j}(a_{26-33}) + \beta_{3j}(a_{34-64}) + \beta_{4j}(a_{65+}) + \\ 
+   \beta_{5j}(d*a_{26-33}) + \beta_{6j}(d*a_{34-64}) + \beta_{7j}(d*a_{65+}) + \\
+   \beta_{8j}(t) + \beta_{9j}(d*t) + \\
+   \beta_{10j}(a_{26-33}*t) + \beta_{11j}(a_{34-64}*t) + \beta_{12j}(a_{65+}*t) + \\
+   \beta_{13j}(d*a_{26-33}*t) + \beta_{14j}(d*a_{34-64}*t) + \beta_{15j}(d*a_{65+}*t) + \\
+   \epsilon_i
+ \end{aligned}
+ \end{equation*}
 
@@
 
+$$ \gamma_k \sim \mathcal{N}(0,\sigma_{\gamma_k}^2) $$
+
+$$ \epsilon_i \sim \mathcal{N}(0,\sigma_{\epsilon}^2) $$
+
 In this framework, the first-level equation models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above.
 These equations are nested at the level of the survey item.
-This model allows for each survey question to have its own set of coefficient estimates weighted by the amount of information we have about the particular question.
-Questions asked in many panels, including partisan identification (seven point scale from "Strong Democrat" to "Strong Republican", asked in a total of 32 waves across all panels) and ideological identification (seven-point scale from "Extremely Liberal" to "Extremely Conservative", asked in a total of 28 waves across all but one panel), will contribute more to the overall calculation of the aggregate trend in opinion stability than questions asked in only three panels.
-Given the lack of confidence in each individual question trend, we focus the majority of our attention on $\gamma_k$ estimates.
+This approach produces an aggregate estimate for each coefficient across all questions ($\gamma_k$) while accounting for variation in which questions are asked at different time points.
+It also allows for each survey question to have its own set of coefficient estimates weighted by the amount of information we have about the particular question.
+Questions asked in many panels, including partisan identification (asked in a total of 32 waves across all panels) and ideological identification (asked in a total of 28 waves across all but one panel), will contribute more to the overall calculation of the aggregate trend in opinion stability and age differences than a question asked in only three panels.
+Given the lack of confidence in each individual question’s estimates, we focus the majority of our attention on $\gamma_k$ estimates.
 
-In practice, the model presented in Eq \### is likely too complicated given our data limitations, and might not be necessary.
-As part of our analysis, we compare the fully elaborated model to more parsimonious models using model-selection criteria, specifically the Bayesian information criterion (BIC) and Akaike information criterion [@raftery1995].
+The full model outlined above is likely too complicated given our data limitations, and some interaction terms or random effects might not be necessary if certain effects or differences are constant over time, do not vary by age group, or are similar across questions.
 For example, it could be the case that little predictive information is gained by allowing questions to have separate slopes for the relationship between age categories and duration over time.
+We compare the fully elaborated model to more parsimonious models using model-selection criteria, specifically the Bayesian information criterion (BIC) and Akaike information criterion (AIC) [@raftery1995].
 
-### Demographic Transitions
+Coefficients generated in this model can be used to provide empirical estimates of our theoretical estimands.
+In this framework, $\gamma_{8}$ estimates the expected difference in absolute change for someone ages 18-25 born in a particular year and someone born 10 years earlier at the same age bracket, providing a quantification of $\tau_1$.
+Similarly, $gamma_{11}$ captures the change in the difference between the youngest age group (18-25) and the middle age group (34-64) for a 10-year difference in time, providing a quantification of $\tau_2$.
 
-Once we select the most parsimonious model to capture attitude stability over time, we then include variables capturing demographic transitions to evaluate their relationship to attitude stability over time.
-For each wave pair, we include measures of demographic status at the first observation.
-For example, if a respondent is unmarried at the first wave and gets married by the second, they are counted as "never married" for the purposes of our analysis.
-If they are observed a third time, they would be counted as "married" for that pair of waves.
+Once we select the best-fitting model to capture trends in attitude change over time, we add variables capturing demographic transitions and educational attainment.
+For each wave pair, we include measures of demographic status at the first observation for a pair of observations.[^3]
 
-Finally, to distinguish between the influence of life-course structure and other factors that might affect attitude stability, such as broader social forces, we calculate a counter-factual expectation of the rate of attitude change in later panels if the proportion of respondents in each group who made certain demographic transitions was constant over time at 1956 levels.
+[^3]: For example, if a respondent got married between waves, they are counted as “never married” for that pair of observations, since they are unmarried at time 1.
+If they are observed a third time, they would be counted as having ever been married.
 
-### Weights
+\begin{equation*}
+ \begin{aligned}
+   |y_{2ij}-y_{1ij} | = \beta_{0j} + \beta_{1j}(d) + \beta_{2j}(a_{26-33}) + … + \beta_{15j}(d*a_{65+}*t) + \\ 
+   \beta_{15j}(d*a_{65+}*t) + \beta_{16j}(evermarried) + \beta_{17j}(everkid) + \\
+   \beta_{18j}(kidinhome) + \beta_{19j}(student) + \beta_{20j}(<HS) + \\
+   \beta_{21j}(BA) + \epsilon_i
+ \end{aligned}
+ \end{equation*}
 
-Finally, a key issue in drawing valid inferences about the U.S.
-population over time is to ensure that samples generalize to the same theoretical population -- the U.S.
-adult non-institutionalized population.
-However, none of the panels used in our analysis were simple random samples of the adult population.
-All employ different complex sampling strategies.
+These variables are all centered at their grand means, and like coefficients for age group, time, and duration, are allowed to vary by question.
+Coefficients produced in this model allow us to generate an estimate of $\tau_3$, or what a cohort would look like if we set their rates of marriage, parental status, and educational attainment to that of another cohort.
 
-To account for this, we attempt to use weights that account for these sampling strategies as well as, when available, non-response.
+Most panels include weights that account for differential sampling and panel attrition.
+When available, we use panel weights that ensure that the second wave of the pair of observations is reweighted to make the sample representative of the general population.
+The 1980 Major Panel Study does not include weights, and as such all respondents in that panel are weighted equally.
 
@@
 
-While not the central question of our analysis, we first highlight changes in four life-course transitions: proportion of respondents who are married, proportion of respondents who reported having a child, the proportion of respondents with a child at home, and proportion of respondents who are currently students.
-We also plot the proportion of respondents at three different education levels: less than high school, high school degree but no college degree, and college degree.
+Figure 1 plots the proportion of respondents who are married, who reported ever having a child, who reported having a child at home, and who are currently students for each panel wave included in our aggregate data set.
+Figure 1 also includes the proportion of respondents at two education levels: those with less than a high school education and those with a college degree (with the remaining population being those with a high school degree but nothing more).
 
-In the early panel (1956-60), respondents in the youngest age groups look very similar to respondents in the older groups on major measures.
-They are equally likely to be married, to have had children, and to be students.
-However, people in the 18-25 age group quickly diverge from the others by the 1970s, displaying the lowest probability of being married and the lowest probability of having children by quite a larger margin.
-Over time, the next youngest age group (26-33) starts to diverge toward the youngest age group, especially in terms of marriage.
+```{=latex}
+\begin{figure}[]
+\begin{center}
+\caption{Proportion of respondents in different age categories reporting life course milestones or levels of educational attainment, by panel wave.}
+```
+```{r demogs, out.width = "100%", fig.pos = "H", echo = FALSE}
+knitr::include_graphics("../figures/demo_variables.jpeg")
+```
+```{=latex}
+\end{center}
+\footnotesize{\textit{Notes: Proportions calculated using imputed data.} }
+\end{figure}
+```
 
-Conversely, the middle age and older age groups see a significant increase in the proportion who hold high school and then college degrees.
-By the mid-1990s, the proportion of middle aged and older respondents who had college degrees significantly exceeded the proportion of youngest age group who has college degrees, likely because many people aged 18-25 were still completing education.
+In the 1956-60 ANES panel, respondents in the youngest age groups are equally likely to be married, to have had children, and to be students as middle-aged respondents.
+At the same time, despite being younger, respondents in the 18-25 and 26-33 age groups had higher levels of educational attainment, on average, than their older peers, indicated by a lower proportion having less than a higher school degree (meaning more had a high school degree).
+People in the 18-25 age group diverged from the others by the 1970s, displaying a lower probability of being married and lower probability of having children, a trend that continues through the most recent panels.
+The next youngest age group (26-33) also diverged from the middle age group, especially in their rates of marriage and having children, but did so more slowly.
 
-These trends set the stage for what to expect in terms of attitude change for different age groups.
-To the extent that transitions such as marriage and having children themselves stabilize attitudes or serve as proxies for broader life course factors that influence attitude stability, then we should see the youngest age group look similar to the older age groups in the 1950s, and diverge by the 1970s, continuing to diverge over time.
-Respondents in the 26-33 age group should remain similar to older respondents for most panels, but begin to diverge by the 1990s and especially by the 2000s.
+Figure 1 also shows an increase in the proportion of middle age and older age groups who complete high school and college.
+By the mid-1990s, the proportion of middle aged and older respondents who had college degrees significantly exceeded the proportion of the youngest age group who had college degrees, likely because many people aged 18-25 were still completing education.
+
+These trends set the stage for what to expect in terms of opinion change for different age groups.
+If marriage and having children serve as proxies for broader life course factors that influence opinion stability, we should see the youngest age group look similar to the older age groups in the 1950s.
+We then would expect this younger group to diverge from older groups in terms of attitudinal stability as time passes.
+Respondents in the 26-33 age group should also start similar to middle-aged respondents and diverge, albeit more slowly.
+To the extent that educational attainment is related to attitude stability, we should observe increased attitude stability among all groups, but especially among middle aged and older respondents who have seen the most dramatic increases in educational attainment.
 
@@
 
-Table XXX presents a comparison of the Akaike Information Criteria (AIC) and Bayesian Information Criteria (BIC) for the full model elaborated above (M1) and a series of more simplified model.
-The best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for any of the higher-order interactions.
-The best-fitting model (M5) omits the third-order interaction between time, age group, and duration of observation and an the interaction between age group and duration.
-Further attempts to simplify the model, including dropping random effects for age categories and duration, resulted in increases in the BIC.
+Comparing the fully elaborated model outlined above to more parsimonious models, the best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for higher-order interactions.
+This model also omits the three-way interactions between time, duration of observation, and age groups.
+Further attempts to simplify the model, including dropping two-way interactions or random effects for age categories and duration, resulted in increases in the BIC.
 
-| Variables                      | M1       | M2       | M3       | M4       | M5       |
-|--------------------------------|----------|----------|----------|----------|----------|
-| Intercept                      | Y        | Y        | Y        | Y        | Y        |
-| RE                             | Y        | Y        | Y        | Y        | Y        |
-| Time                           | Y        | Y        | Y        | Y        | Y        |
-| RE                             | Y        | Y        | Y        | Y        | Y        |
-| Age Groups                     | Y        | Y        | Y        | Y        | Y        |
-| RE                             | Y        | Y        | Y        | Y        | Y        |
-| Duration                       | Y        | Y        | Y        | Y        | Y        |
-| RE                             | Y        | Y        | Y        | Y        | Y        |
-| Time \* Age Group              | Y        | Y        | Y        | Y        | Y        |
-| RE                             | Y        | Y        | Y        | N        | N        |
-| Time \* Duration               | Y        | Y        | Y        | Y        | Y        |
-| RE                             | Y        | Y        | N        | N        | N        |
-| Age Groups \* Duration         | Y        | Y        | Y        | Y        | N        |
-| RE                             | Y        | Y        | N        | N        | N        |
-| Age Groups \* Duration \* Time | Y        | N        | N        | N        | N        |
-| RE                             | Y        | N        | N        | N        | N        |
-|                                |          |          |          |          |          |
-| BIC                            | 20714.39 | 20250.97 | 19996.61 | 19862.23 | 19845.22 |
-| AIC                            | 19757.88 | 19675.82 | 19627.76 | 19643.42 | 19645.17 |
+Model 1 in Table 1 presents the main effects and variance components of the best-fitting multilevel model presented above.
+In all models, the 18-25 age group serves as the reference category, and the time variable is centered on the most recent wave pair (starting in 2020), so coefficients reflect expectations for the amount of change observed in the most recent years, rather than at earlier time points.
 
-: Model fit comparison
-
-Model 1 in Table XXX presents the main effects and variance components of the best-fitting multilevel model presented above.
-In all models, the 18-25 age group serves as the reference category, and the time variable is centered on the most recent panel (2016-2020), so coefficients reflect expectations for the amount of change observed in the most recent years, rather than at earlier time points.
+\pagebreak
 
@@
 
-: Table \@ref(tab:2): Coefficient estimates from multilevel linear regressions of absolute change between responses on age group, duration, and question.
+: Coefficient estimates from multilevel linear regressions of absolute change between responses on age group, duration, and question.
 
-The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (about 18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question.
-This is a rather high level of variance, but it is consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965].
-At the same time, random-effect variance around this main intercept term is quite large ($\sigma_{\beta_0} = 5.9220$).
-On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time with low levels of measurement error or random change.
+The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to change by almost a fifth of the response scale on average (about 18 points) in the most recent years of the survey on any particular question, regardless of the duration between observations.
+To be clear, very few wave pairs have a duration close to 0, and this coefficient primarily reflects the fact that this amount of absolute change will be present regardless of whether responses are 2 or 4 years apart.
+This high level of change is consistent with previous studies that find that survey respondents tend to make large, seemingly random, wave-to-wave changes [@ansolabehere2008; @converse1964].
+At the same time, random-effect variance around this intercept term is large ($\sigma_{\beta_0} = 5.99$).
+On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others people tend to respond with the same answer over time with low levels of measurement error or random change.
 
-Duration between observations is positively associated with wave-to-wave absolute change.
-For every additional year between responses, the expected absolute change for respondents increases 1.1 points.
-As suggested above by the model comparison, there is no evidence to suggest that respondents in different age groups vary systematically in this kind of change.
-Random effect variance around this main effect is relatively small (s.d.
-= 0.36), meaning each one-year duration only increases the amount changed by between .5 and 1.5 for most questions.
-Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].
+Duration between observations is positively associated with wave-to-wave absolute change for the youngest age group (duration coefficient = 1.45), and while the interaction between duration and age group is negative for older age groups, duration is still positively related to the amount of change for all.
+Random effect variance around this main effect of duration is relatively small (s.d.
+= 0.50), meaning each one-year duration only increases the amount changed by between .5 and 1.5 for most questions, and it is never negatively associated with change.
+Given the high level of wave-to-wave change independent of duration, our results reinforce long-standing findings that most wave-to-wave change is non-persisting or measurement error [@converse1964; @zaller1992; @kiley2020; @lersch2023].
 
-Coefficients in Model 1 allow us to test several of the hypotheses outlined above.
-First, the coefficients for age groups other than the reference category suggest that, on average, other age groups are more stable than people in the 18-25 age range in the most recent panels.
-Respondents in the 26-33 age group are about 1.1 points more stable on average, respondents in the 34-65 age group are about 2.2 points more stable on average, and respondents in the 65 and over age group are about 3.5 points more stable on average, than respondents in the 18-25 age group.
-This provides broad evidence for the "impressionable years" and "increasing persistence" hypotheses, as the youngest age groups are less stable than others, followed by the 26-33 age group.
+The coefficients for age groups other than the reference category suggest that, on average, older age groups are more stable than people in the 18-25 range in the most recent panels.
+In these panels, respondents in the 26-33 age group are about 1 point more stable on average, respondents in the 34-65 age group are about 2.5 points more stable on average, and respondents in the 65 and over age group are about 3.4 points more stable on average, than respondents in the 18-25 age group.
+The negative coefficients attached to the interactions between duration and age categories further compound this difference, as longer durations produce more change among the younger respondents than older respondents.
+This provides broad evidence for the "impressionable years" hypothesis, with the youngest age group being more likely to change than others, followed by the 26-33 age group.
 
-Turning now to how these groups have changed over time, the effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significantly different from 0.
-To clarify, the lack of statistical significance for the time effect not because there is a large standard error on the term.
-It is estimated with a high degree of precision and is very close to 0.
-On average, recent cohorts in this age group make approximately the same amount of wave-to-wave change in recent years as earlier cohorts did, accounting for the kinds of questions they were asked and time between responses.
-However, question-level random effects vary considerably around this main effect, with a standard deviation of about 1.43.
-This suggests that the overall flat trend obscures quite large question-level differences, with some questions becoming more stable and others becoming less stable.
-We return to consider these question-specific trends later.
+Turning now to how these groups have changed over time, the coefficient for time in decades for people in the 18-25 age group ($\gamma = -0.01$, $t = 0.75$) is substantively small and not statistically significantly different from 0.
+Question-level random effects vary considerably around this main effect, with a standard deviation of about 1.43, meaning the overall flat trend obscures large question-level differences, with responses to some questions becoming more stable and responses to others becoming less stable.
+We consider these question-specific trends later.
+It is important to note that because the effect of duration also increased over time (duration*time coefficient = 0.14), the most recent cohorts to pass into the 18-25 age range are expected to make slightly more change over time in recent years, on average, than previous cohorts did when they were in this age range, but the trend is not pronounced.
+The extended early adulthood framework hypothesized that the 26-33 age group would come to resemble the youngest age group over time.
+The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\gamma = -0.02$), suggesting that while people in this age group are more stable, on average, than the youngest respondents, more recent cohorts are not any more or less likely to make changes of opinion in this age range than earlier cohorts and not any more or less similar to the youngest adults (18 to 25) than earlier cohorts.
 
-Next, the extended early adulthood framework (H2) suggests that we should see a closing of the gap between people in the 18-25 and 26-33 age groups over time.
-The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that while people in this age group are more stable, on average, than then youngest respondents, more recent cohorts are not any more or less likely to make changes of opinion in this age range than earlier cohorts, and more recent cohorts are not any more or less similar to younger adults than earlier cohorts.
+Before concluding that this stability of rates of attitude change across cohorts of early adults reflects a lack of importance of life course shifts, it is important to understand the full picture of change over time.
+Model 1 shows that, in contrast to people in the 18-25 and 26-33 age ranges, whose rate of opinion change was relatively stable across cohorts, older age groups (34-64 and 65+) show a notable negative time trend across cohorts.
+Both time trends are negative, substantively large (-0.5/decade for middle age and -1.3 for old age), and statistically significant, meaning more recent cohorts are more stable opinion reporters while in these age ranges than earlier cohorts were.
+To more clearly illustrate these trends, Figure 2 plots the overall predicted level of change for respondents in each age category over time, assuming a two-year duration between observations (the modal duration between waves).
+It also includes separate trajectories for each individual question to highlight variation in the trajectories across these items for the different age groups.
 
-However, before drawing broader conclusions about the lack of importance of life course shifts, it is important to understand the full picture of change over time.
-Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range whose rate of opinion change remains flat across cohorts, respondents in older age groups (34-64 and 65+) do show a notable time trend -- more recent cohorts are more stable opinion reports in these age ranges than earlier cohorts.
-Both time trends are negative, substantively large, and statistically significant.
-To more clearly illustrate these trends, Figure XXX plots the predicted time level of change for respondents in each age category over time.
-
-```{r}
-new.data %>%
-  filter(d == 2) %>%
-  mutate(group = paste(age_group, name, sep = "-")) %>%
-  mutate(year = as.Date("2020-11-12") + dec_diff*3652.5) %>%
-  ggplot(aes(x = year, y = yhat, color = age_group)) + 
-  geom_hline(yintercept = 0, color = "black") + 
-  geom_line(alpha = .2, aes(group = group)) + 
-  geom_smooth(linewidth = 2) + 
-  theme_bw() + 
-  facet_wrap(~age_group) + 
-  labs(x = "Year", y = "Predicted wave-to-wave change",
-       color = "Age Group",
-       title = "Predicted wave-to-wave change by age group",
-       subtitle = "Individual question trajectories and overall trajectory") + 
-  scale_color_brewer(type = "qual", palette = 2) +
-  theme(legend.position = "none")
+```{=latex}
+\begin{figure}[]
+\begin{center}
+\caption{Predicted absolute wave-to-wave change over time for respondents in different age categories.}
+```
+```{r grouptrend, out.width = "100%", fig.pos = "H", echo = FALSE}
+knitr::include_graphics("../figures/trajectory_plot.jpeg")
+```
+```{=latex}
+\end{center}
+\footnotesize{\textit{Notes: Bold line plots average trajectory across all questions.
+Thin lines plot individual question trajectories.
+Expected values predicted with duration = 2, the modal duration between repeated observations across all panels.}}
+\end{figure}
 ```
 
-In the earliest panels, the oldest respondents (65 and over) are substantially less stable opinion holders than younger respondents, changing by about 5 points more than members of the other groups, while the remaining three age groups (18-25, 26-33, and 34-64) appear to change at approximately equal rates.
-Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable, with the latter becoming almost 10 points more stable on average from the 1950s to the 2020s.
-Because, as noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change over time, they become the least stable respondents by the end of the panel.
+Figure 2 shows that because of the interaction between duration and time, more recent cohorts show more change over time in the 18-25 and 26-33 age ranges than previous cohorts.
+In the earliest panels, the oldest respondents (65 and over) were substantially less stable opinion holders than younger respondents, changing by about 5 points more than members of the other groups, while the remaining three age groups (18-25, 26-33, and 34-64) appeared to change at approximately equal rates (about 20 points over a two-year duration).
+Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable, with the latter becoming almost 10 points more stable on average from 1956 to 2020.
+Because, as noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change across cohorts, they become the least stable respondents by the most recent panels.
 
-## Life Course Shifts
+Figure 2 also highlights the variance of trajectories for individual questions, especially for young adults.
+While older respondents’ slopes for each question are almost uniformly negative, there is much more variation for respondents in trajectories for younger age ranges.
+We return to these differences below.
 
-While coefficient estimates for the early adult age groups do not suggest that these respondents have become more variable opinion holders over time, the increase in stability for middle-aged and older respondents suggests that broad social forces could be counteracting life-course transitions for members of these age groups.
+While Model 1 suggests that there has not been a substantial increase in attitude change over time for the youngest age groups, as expected by the theoretical framework, the increase in stability for middle-aged and older respondents suggests that broad social forces could be counteracting the influence of delayed life-course transitions for members of these age groups.
+Most centrally increased educational attainment over time might obscure effects of delayed life-course transitions.
 
-To evaluate the specific effect of life-course changes, we attempt to control for these transitions.
-We re-estimate Model \### above, this time incorporating a series of indicator variables for whether respondents were students, whether they had children, and whether they were married.
-We also include an indicator variable for two levels of education -- whether respondents have less than a high school degree and whether respondents completed college -- to evaluate whether broader shifts in attitude stability in the population might be due to shifts in educational attainment.
-These results are presented at Model 2 in Table XXX above.
+Model 2 re-estimates Model 1 incorporating a series of indicator variables for whether respondents were students, whether they had children, and whether they were married, and for two levels of education (less than a high school degree and completed college).
+These results are presented as Model 2 in Table 1 above.
 
-These results provide mixed evidence of life-course shifts having a meaningful effect on attitude stability.
-Respondents who have been married are slightly more stable than unmarried respondent who were never married, but respondents who have had children are less stable than respondents who never had children.
-Having children at home increases change even more.
-Taken together, these coefficients suggest that changes in the timing of life course transitions are insufficient to explain the divergence between young adults and older adults in terms of stability.
+These results provide mixed evidence that life-course shifts have a meaningful effect on attitude stability.
+First, the gaps between age groups become smaller and the time trends become weaker when these demographic factors are included in the model, suggesting they account for part of the overall picture.
+Respondents who have been married are about one point (-0.89) more stable on average than respondents who were never married, but respondents who have had children (0.26) or currently have children at home (0.30) are less stable than respondents who never had children.
+Taken together, these coefficients suggest that changes in the timing of life course transitions are likely insufficient to explain the divergence between young adults and older adults in terms of stability.
 
 On the other hand, coefficients related to education show a clearer picture.
-Higher levels of education are uniformly associated with increased attitude stability (positive coefficient on having less than a high school degree and a negative coefficient on having a college degree), suggesting that changing levels of education in the population -- and the increase in education level of middle-aged respondents relative to younger respondents -- likely accounts for a larger share of the divergence in stability.
+Higher levels of educational attainment are uniformly associated with increased attitude stability, with having less than a high school degree increasing absolute change by about 2.8 points on average and having a college degree increasing absolute change by about 2.4 points on average, relative to having only a high school education, suggesting that changing levels of education in the population -- and the increased education level of middle-aged respondents relative to younger respondents -- likely accounts for a larger share of the divergence in stability.
 
-Using these estimates, we calculate the expected value of change had life-course transitions occurred at roughly the same rates that they did in the late 1950s.
-These estimated counter-factual values are plotted in figure XXX.
+Using these estimates, we calculate the expected value of change under the assumption that life-course transitions and educational attainment stayed at the same rates they did in the late 1950s.
+These estimated counter-factual trends are plotted in Figure 3.
 
-```{r}
-new.data %>%
-  filter(d == 2) %>%
-  mutate(time = recode(time, "1"="Counterfactual", "t"="Observed")) %>%
-  mutate(time = ifelse(dec_diff == -6.4, "Start", time)) %>%
-  group_by(age_group, dec_diff, time) %>%
-  summarise(yhat = mean(yhat)) %>%
-  ungroup() %>%
-  select(-dec_diff) %>%
-  spread(time, yhat) %>%
-  pivot_longer(Counterfactual:Observed) %>%
-  mutate(Start.Time = -6.4, Finish.Time = 0) %>%
-  ggplot(aes(x = Start.Time, xend = Finish.Time, y = Start, yend = value,
-             linetype = name, color = age_group)) +
-  geom_segment() + 
-  geom_line(aes(color = age_group)) + 
-  #geom_point(size = 3) +
-  theme_bw() + 
-  scale_shape_manual(values = c(21,22)) + 
-  scale_fill_brewer(type = "qual", palette = 2,
-                     guide = "none") + 
-  scale_color_brewer(type = "qual", palette = 2,
-                    guide = "none") + 
-  labs(x = "Age Group", y = "Expected change for 2-year duration",
-       shape = "Estimate",
-       title = "Counterfactual expectation of change",
-       subtitle = "Holding demographic variables constant at 1956 levels") + 
-  facet_wrap(~age_group)
-
+```{=latex}
+\begin{figure}[]
+\begin{center}
+\caption{Observed trajectories of wave-to-wave response change over time (solid lines) compared to counterfactual trajectories assuming no change in rates of marriage, childbearing, student status, and educational attainment from 1956-60 panel (dashed lines), for each age group.}
+```
+```{r counterfactual, out.width = "100%", fig.pos = "H", echo = FALSE}
+knitr::include_graphics("../figures/counterfactual_plot.jpeg")
+```
+```{=latex}
+\end{center}
+\footnotesize{\textit{Notes: Predictions generated using Model 2 coefficients.
+Covariate values set to average proportions from 1956-60 ANES panel and 2016-20 GSS panel.
+Expected values predicted with duration = 2, the modal duration between repeated observations across all panels.}}
+\end{figure}
 ```
 
-Assuming that rates of marriage, student status, and degree attainment remained constant at their 1956 level over time for each age group, age groups would be expected to display more similar rates of change in the most recent panels than they actually do.
-In particular, the youngest respondents would display lower rates of change over time and other age groups would display higher rates of change over time.
-In fact, consistent with the extended early adulthood framework, the youngest respondents would be slightly less consistent opinion holders in more recent years than they actually were.
+Assuming that rates of marriage, student status, and educational attainment remained constant at their average 1956-60 level over time for each age group, all groups would have been expected to display more change over time than we observe, a reflection of the overall increase in stability due to increased educational attainment.
+Both middle aged and older respondents would still be more stable in recent years than they were in the earliest years, but, consistent with the theory of extended early adulthood, the youngest respondents and those in the 26-33 age range would have been less consistent opinion holders in more recent years than they actually were.
 
-This divergence between observed and counterfactual trajectories is principally driven by changes in college-going, rather than delays in marriage.
-The dramatic increase in college-going for all age groups produces more stable attitudes over time.
-However, because young people have completed college at lower rates than older respondents in the most recent panels (likely because many of them are still enrolled), they do not benefit as much from this shift (though being currently enrolled as a student also increases attitude stability).
-Young respondents' dramatically lower rates of marriage in the most recent panels counteracts the increased stability that comes from higher levels of educational attainment.
+Figure 3 suggests that most of the change in opinion stability over time is due to shifts in educational attainment.
+This is clearly illustrated by older and middle aged adults – who saw the largest shifts in educational attainment but minimal shifts in other demographics – also having the largest shifts from their late-1950s baselines.
+However, there do appear to be additional stabilizing forces that principally affected middle-aged and older adults beyond educational attainment.
+It could be the case that the three-category measure of education is insufficient to capture the full range of educational attainment shifts among people in this group.
+However, it could also be that other forces also drove attitude stability in this period.
 
@@
 
-As noted above, the overall trajectory of increasing attitude stability masks significant heterogeneity across questions.
-To more clearly illustrate these differences in trends, Figure XXX plots the estimated trajectories for each item for respondents in the youngest (18-25) and middle-aged (34-64) age categories.
+As noted above, Model 1 suggests that the overall trajectory of increasing attitude stability masks heterogeneity across questions and age groups, though the trajectories for individual questions are estimated with a low degree of confidence.
+To illustrate these different trends for different questions, Figure 4 plots the estimated trajectories for each item for respondents in the youngest (18-25) and middle-aged (34-64) age categories, holding constant duration at 2 years.
 Items are ordered by their predicted time trends.
+It is important to note that most of these trajectories are estimated with a low degree of confidence, and that the linear trends for each question are likely not accurate outside of the windows in which we observe the questions.
 
-```{r}
-slopes <- as.data.frame(ranef(m4.5)$name) %>%
-  rownames_to_column(var = "name") %>%
-  mutate(slope = dec_diff) %>%
-  select(name, slope)
-
-#Look at questions instead
-new.data %>%
-  filter(d == 1, age_group %in% c("18-25", "34-65")) %>%
-  group_by(name) %>%
-  left_join(slopes) %>% 
-  #left_join(test, by = c("name"="name")) %>%
-  mutate(group = paste(age_group, name, sep = "-")) %>%
-  ggplot(aes(x = dec_diff, y = yhat, color = age_group)) + 
-  #geom_point(data = summaries, aes(y = a)) + 
-  geom_line(alpha = .7, aes(group = group)) + 
-  theme_bw() + 
-  facet_wrap(~reorder(name, slope), nrow = 11, ncol = 7) + 
-  labs(x = "Year", y = "Predicted wave-to-wave change",
-       color = "Age Group") + 
-  theme()
-
+```{=latex}
+\begin{figure}[]
+\begin{center}
+\caption{Expected rate of change for two-year duration, by year, for ages 18-25 (pink lines) and 34-64 (blue lines).  }
+```
+```{r questions, out.width = "100%", fig.pos = "H", echo = FALSE}
+knitr::include_graphics("../figures/question_plot.jpeg")
+```
+```{=latex}
+\end{center}
+\footnotesize{\textit{Notes: Predicted values generated using coefficients in Model 1.
+Expected values predicted with duration = 2, the modal duration between repeated observations across all panels.}}
+\end{figure}
 ```
 
-First, at a high level, there are very few questions where the directional trend diverges for age groups -- where young people become more stable opinion holders while old people become less stable opinion holders, or vice-versa.
-In general, both groups either become more stable or less stable, just at different rates over time.
+It is hard to draw broad conclusions about which items become more stable and which become more variable over time.
+Several questions showing the largest increases in stability deal with social issues, including whether homosexuals should be protected from discrimination, whether men and women should have equal roles in society, and whether the enduring effects of slavery and discrimination continue to hinder Black Americans' social and economic progress.
+A related set of questions about equality in America have also decreased in within-person change over time.
+Many general political views, such as support for capital punishment, whether the government should help people cover medical costs, and spending on the environment, have also decreased in change over time.
+Some of these shifts reflect the emergence of social consensus, as observed on several “culture war” issues like gay marriage [@baldassarri2020], while others reflect clearer partisan polarization with elites of each party providing clearer signals about which positions should be held by partisans [@kozlowski2021].
 
-It is hard to draw broad conclusions about which items become more stable and which become more variable over time.
-Some questions dealing with social issues have become much more stable, including whether homosexuals should be protected from discrimination, whether men and women should have equal roles in society, and whether the enduring effects of slavery and discrimination continue to hinder Black Americans' social and economic progress.
-A broad set of questions about equality in American have also decreased in within-person change over time.
-Many general political views, such as support for capital punishment, whether the government should help people cover medical costs, and spending on the environment, have also decreased in change over time.
-Some of these shifts reflect the emergence of social consensus, while others reflect clear partisan polarization.
-
-A set of items tapping political efficacy (ppllikeme; manyvote; complicated) show increased change over time.
+A set of items tapping political efficacy (“ppllikeme”; “manyvote”; and “complicated”) show increased change over time.
 Feeling thermometers about social groups, including racial groups, the poor, homosexuals, liberals and conservatives, have either also remained at similar levels of change or seen an increase in variance.
 Policy issues that see increased variance include spending on crime and whether organizations should preferentially hire Black Americans to account for past discrimination.
+It is unclear what is driving increased inconsistency among these items over time.
 
@@
 
-Theoretical models that explain the heightened malleability of attitudes during early adulthood by referencing the unique social conditions of that life phase suggest that significant changes in the timing of life course events should produce heightened rates of attitude change during early adulthood (18-25) or should extend rates of attitude updating later into life (ages 26-33).
-Given that Americans, on average, pushed back marriage, childbirth, and entering the workforce over the course of the second half of the 20th Century and the early 21st Century, we posited that these shifts should, through various pathways, increase over-time variance in survey responses among the youngest members of the population.
-We tested a series of hypotheses related to this theory by looking at the stability of political and social attitudes among different cohorts as they moved through the same age ranges.
+The dramatic changes in the timing of life course transitions that have occurred in the United States in the last few decades provide a valuable opportunity for exploring the mechanisms that underpin attitudinal change across different life stages.
+Given that Americans, on average, pushed back marriage, childbirth, and entering the workforce over the course of the second half of the 20th Century and the early 21st Century, we posited that these shifts should, through various pathways, increase over-time change in survey responses among more recent cohorts of American early adults compared to their predecessors when they were the same age.
+We tested a series of hypotheses related to this framework by looking at the stability of political and social opinions among different cohorts as they moved through the same age ranges.
 
 The results presented above suggest a more complicated picture than anticipated.
-Consistent with expectations, the youngest adults look similar to older adults in the earliest panels in terms of both demographics and attitude stability.
-While members of the youngest age categories (18-25 and 26-33) diverged over time from older members of the population in the rate at which they changed their opinions, with younger respondents showing more change than older respondents in the most recent panels, this shift was principally driven by an increase in stability among people in middle age and over 65, rather than increased opinion change among the youngest adults.
+Consistent with expectations, early adults look similar to middle-aged adults in the earliest panels in terms of both demographic characteristics and attitude stability.
+Members of the youngest age categories (18-25 and 26-33) diverged over time from older members of the population, showing higher rates of opinion change than them in more recent panels.
+This shift, however, was principally driven by an increase in stability among people in older age groups, rather than increased opinion change among the early adults.
 We suggested that this aggregate divergence was driven by countervailing trends.
-On one hand, delays in life course transitions push the youngest members of the population to make more attitude change over time.
-Second, broader social forces worked to stabilize opinions for all members of the population, leading to divergence between the youngest adults and the rest of the population.
+On the one hand, delays in life course transitions push the youngest adults to make more attitude changes over time.
+On the other hand, broader social forces -- like the increase in educational attainment -- stabilized opinions for all members of the population.
 
-We tested this proposition using data on life-course transitions and measures of educational attainment.
-We found that, consistent with our revised perspective of countervailing forces, age groups' rates of attitude change over time would have been more similar if these social changes did not happen.
-Specifically, delayed rates of marriage decreased attitude stability among the youngest members of the population while increased rates of educational attainment account for much of the stabilizing shifts among older Americans.
+We examined the relative importance of these countervailing forces using data on life-course transitions and measures of educational attainment.
+We found that, consistent with our revised perspective, increased educational attainment stabilized opinions for all age groups, with contemporary older (65-80) adults exhibiting much higher levels of attitude stability than their counterparts from previous cohorts.
+Life course shifts such as delayed rates of marriage and childbirth and delayed entry into the workforce had contradictory effects on the overall rate of attitude change.
+While marriage made change less likely, ever having children or having children at home increased the rate of change.
+The effects of these transitions were small and, thus, did not significantly affect rates of stability in the aggregate.
+In general, then, increased educational attainment has stabilized attitudes across all age groups.
 
-In contrast to purely biological or cognitive explanations for differing rates of attitude stability among adolescents and early adults, our results suggest that broad social conditions play a substantial role in shaping the rate at which people change their attitudes.
-When early adults' lives resembled those of middle-aged respondents in the 1950s, their rates of attitude change were also similar.
-As social lives began to diverge, younger adults began to change their opinions at higher rates than older adults.
-At a broader level, since 1950s, Americans have become meaningfully more stable opinion holders, in part because they have attained higher levels of education.
+Our analyses recovered trends that are compatible with the idea of an extended formative period for opinions and beliefs but suggest that the mechanisms behind it are counterintuitive.
+It is true that, in the contemporary United States, early adults change their opinions more than older adults.
+It is also true that this gap is wider now than it had been in previous decades.
+However, this is not because contemporary early adults are changing at a higher rate than their counterparts in previous time periods.
+It is rather that older age groups have become considerably more stable, in great part due to an increase in educational attainment.
 Our approach likely under-estimates the importance of this shift, as we collapse education into only three categories -- ignoring differences among people with less-than a high school degree in early panels and dramatic expansions of post-graduate degree attainment in recent decades.
+So the perceived differences in opinion change between age groups can be partly explained by the fact that early adults have not stabilized as much relative to older age groups.
 
-At the same time, it is important not to overstate the importance of these broader social transitions.
-The average expected attitude change in 2020 is within a few points of the average expected attitude change.
-Despite dramatic shifts in educational attainment, access to information, and political organization, not to mention changes in survey administration, people make changes in many social and political attitudes at the same rates they did 70 years ago.
+Our analyses are particularly relevant when considered against the backdrop of how robust findings about formative periods are across disciplines and domains.
+The pattern of a period of exploration followed by stability is evident in different species, and its evolution has been widely theorized.
+Our results suggest that a person’s openness to changing their opinions and beliefs is not only a result of maturational processes like brain plasticity or even of how information is processed but rather is partly shaped by the social structuring of the life-course and the border social environment.
+When early adults' lives resembled those of middle-aged respondents in the 1950s, their rates of opinion change were also similar.
+Their rates of attitude change diverged as their social conditions and experiences diverged.
+Many theories of socialization place great weight on the importance of the "early years" as a moment of exploration and attitudinal formation.
+Our work shows that social factors can affect the extent to which this exploration can take place, how long these "early years" might last, and how other countervailing processes might stabilize attitudes.
+
+At the same time, it is important not to overstate the importance of these broader social factors in shaping opinion stability.
+The average expected opinion change over time in 2020 is within a few points of the average expected attitude change on a 100-point scale for any particular survey question, and we do see wide variation in the trajectories of opinion stability across different questions.
+Despite dramatic shifts in educational attainment, access to information, and political organization, not to mention changes in survey administration, people make changes in many social and political attitudes at about the same rates they did 70 years ago.
+And most of this change reflects temporary shifts of opinion or measurement error.
+Even early adults, who appear to make the most durable change according to our models, durably change only about an average of 3 points on a 100-point scale every two years.
 
@@
 
-One limitation of our design is that we focus on questions that are asked repeatedly.
-By focusing only on the same questions over time, we control for question content and structure and ensure that the trend we identify over time is due to something other than changes in survey administration or question quality.
-However, by only focusing on the same questions, we limit the kinds of issues about which we can make claims to issues that receive significant public attention for a long period of time, including general ideological positions and positions on long-standing issues, such as the government's role in the health care sector.
-We ignore large swaths of issues that emerge and disappear before the next panel but were very important in political debates, such as approaches to fight "stagflation." It could be the case that such issues have a different relationship between age and stability than issues that dominate political debates for decades.
-It could also be the case that stability of attitudes is principally a function of how long they have been in the public sphere ().
+As we noted when defining our theoretical estimands, it is difficult to understand the questions we posed in a purely causal framework.
+Respondents born in different cohorts face not only diverse social circumstances upon birth – different opportunities for education, different media environments, different access to technologies – but also have different compositions at birth.
+More recent cohorts are more racially and ethnically diverse than the earliest cohorts, are less likely to be born into two-parent households, are less likely to be born to religious parents, and vary in other ways that are likely related to opinion formation.
+All of these factors make it difficult to conceptualize what it would mean to hold everything but birth year constant.
+Our estimates here should not be interpreted as causal effects of only being born in later years, holding these kinds of features constant, and it is plausible that coefficients estimated above model are biased by these compositional shifts.
 
-Second, because most questions are asked in a small number of panels, our actual number of data points for key questions is much smaller than the total number of wave-pair observations we have.
-The modal number of panels for individual questions is 3, meaning our ability to estimate temporal trajectories is highly constrained, and our estimates of question-specific trajectories is imprecise.
-This also makes it difficult to get efficient estimates of higher-order interactions or estimate time trajectories as anything other than linear trends.
-Relatedly, because many questions are only asked more than twice in one panel, it is difficult to get efficient estimates of whether we see changes in the "duration" effect for specific questions.
+Similarly, it is important to be cautious in our interpretation of the relationship between increased educational attainment and widespread stabilization.
+What we observe as an increase in educational attainment likely also captures other time trends such as increased levels of material wealth, curricular changes in public schools, or more diverse and integrated classrooms.
+Thus, when interpreting the effect of educational attainment on stabilization, we should bear in mind that there are many potential correlates with the former and that, for that reason, establishing a clear causal connection is difficult.
+
+Beyond this central limitation, one additional limitation of our design is that we focus on questions that are asked repeatedly across multiple panels.
+By focusing only on the same questions over time, we control for question content and structure and ensure the trend we identify over time is due to something other than changes in survey administration or question quality.
+However, this approach limits the kinds of issues about which we can make claims to issues that receive significant public attention for a long period of time, including general ideological positions and positions on long-standing issues, such as the government's role in the health care sector.
+We ignore large swaths of issues that emerge and disappear before the next panel but were very important in political debates, such as approaches to fight inflation in the 1970s, differences over the Iraq and Afghanistan wars in the 2000s, and many questions about social justice in recent years.
+It could be the case that such issues have a different relationship between age and stability than issues that dominate political debates for decades.
+
+Similarly, because most questions are asked in a small number of panels, our actual number of data points for key questions is much smaller than the total number of wave-pair observations we have.
+The modal number of panels for individual questions is 3, meaning our ability to estimate time trajectories is constrained and our estimates of question-specific trajectories is imprecise.
+This also makes it difficult to estimate time trajectories as anything other than linear trends.
+It is almost certainly true that changes in attitude stability are not a linear function of time.
+Our estimated effect of time should be thought of as an average of what are likely heterogeneous effects of different eras that should be explored more in future research.
+
+Despite these limitations, we believe our results speak to central questions across multiple subfields in sociology and suggest further avenues of investigation.
+
+## Implications for Cultural Sociology
+
+First, our results reinforce recent findings that suggest that attitudes, beliefs, and opinions – what Lizardo (2017) calls “declarative personal culture” – are mostly stable in adulthood, with most change observed between responses by the same person being transit, non-durable change or measurement error [@kiley2020; @lersch2023].
+Adults make large non-persistent changes over time but small persistent changes, on the order of 1-2 percent of a response scale a year.
+These estimates support work that suggests that most aggregate cultural change must be driven by cohort replacement processes rather than changes of opinion driven by shifts in the contemporaneous environment, and that understanding why and how cultural change occurs requires greater attention to the “past conditions of production” of childhood, adolescence, and early adulthood [@vaisey2016], and reinforce calls for greater attention to early life socialization [@guhin2021].
+
+We did find that the rate of durable change in adulthood did increase slightly over the course of the window studied here, especially for younger adults, suggesting more room for durable attitude change at later life-course stages.
+
+Relatedly, it is worth highlighting that the main trend we observe is not about attitudinal change, but rather about increased stability.
+The most consistent finding we have is about how social factors -- in this case educational attainment -- might act to stabilize opinions, rather than how they might lead to more openness to change.
+Like previous work on attitude stability [@kiley2020; @lersch2023; @vaisey2021], these findings highlight the need to focus on the social processes that make holding stable opinions more likely (or even possible).
+Sociologists interested in personal declarative culture are often interested in how it changes.
+It seems, however, that our efforts should also be directed at explaining what makes attitudinal stability possible to maintain, or the “social scaffolding” that constrains cultural beliefs [@lizardo2010; @martin2010], and the aggregate consequences for social contexts that allow for more or less change.
+
+## Implications for Sociology of Aging and the Life Course
+
+We focused on the period of transition into adulthood because it has been theorized as a key moment for attitudinal formation.
+However, our results show that patterns of attitudinal updating have changed perhaps more for individuals in older age-groups, where opinions have become more stable.
+Sociologists of aging have long criticized visions of old age as a period of passivity, arguing that it has been reconceptualized as a period for renewed exploration.
+Our results challenge previous work that suggests that attitudes follow a “life stages” model in which people are open to change both when they are young and when they are old.
+While we find that older Americans had high rates of change in early panels, this appears principally to be because cohorts that were old in the middle of the 20th Century had very low levels of formal education, which made them inconsistent opinion holders.
+As successive cohorts of older Amerians have attained higher levels of education, their opinions have become much more stable than younger Americans.
+
+Our analyses also have important implications for how we understand the development of attitudinal change and formation across the life course during the past 70 years.
+The fact that the amount of opinion change – both durable and non-durable – has changed across cohorts challenges models of social learning and political imprinting that assume a cohort-invariant propensity to adjust opinions in the face of social influence at different ages [@ghitza2022; @bartels2014].
+Our results suggest that social conditions such as access to education have the capacity to change how susceptible people are to broader social environments across the life course, and have shifted the windows in which Americans learn over time.
+
+## Implications for Political Sociology
+
+The pattern of increased opinion stability in middle age reinforces the argument that political dispositions form primarily through an imprinting process, in which preferences are formed early in life and are relatively resistant to change during adulthood [@ghitza2022; @sears1999].
+While adults do occasionally make changes of opinion [@lersch2023], experiences during adulthood do not appear to be sufficient to produce widespread changes of political opinion.
+At the same time, it is not the case that people have clear, stable attitudes on all issues as adults.
+While adults’ expression of their partisan and ideological identification are quite stable over time, Americans on average demonstrate a high degree of inconsistency on a number of issues, such as whether they view government to be effective, trustworthy, and responsive to peoples’ interests.
+
+Our results further show that over-time trends in the stability of opinions on political issues are quite varied.
+Opinions on so-called “social issues,” such as support for the rights of homosexuals and views on the role of women in society, appear to have become more stable over time, potentially reflecting the transition from political dissensus to consensus [@baldassarri2020].
+On the other hand, Americans’ opinions on long-standing questions about economic redistribution, such as views on the government's role in health care and government spending on social insurance programs, are as variable as they were decades ago.
+This suggests that duration as an issue of public disagreement is not sufficient to produce either consensus or stable opinions.
+
+Finally, our results suggest a key role of education in shaping public opinion on political issues.
+Consistent with previous work that finds that educated Americans hold more ideologically aligned issue positions [@boutyline2017], our strongest pattern of results suggests that education also stabilizes opinions over time.
+It seems likely that these trends reinforce each other, with ideologically aligned opinions being easier to hold.
+
+## Implications for Sociology of Education
+
+Our results suggest that educational attainment, of factors studied here, plays a very strong role in stabilizing opinions, though the specific mechanisms are unclear.
+In general, question-specific effects for attaining a college degree (and having less than a high school degree) were almost uniformly negative (positive) and statistically significantly different from 0 for most questions, and there were almost no questions where higher levels of educational attainment were positively associated with change.
+However, it is unclear why educational attainment has this effect.
+We reiterate calls for greater exploration into the specific mechanisms that produce this greater level of stability among people with high levels of education, be it changes to intelligence, cultural expectations, changes in information consumption habits, or something else.
+
+Some work suggests that highly educated Americans are more culturally homogeneous and inhabit more culturally homogeneous environments [@lynn2017; @lynn2024; @harding2009].
+It could be that cultural homogeneity and stability reinforce each other, as these people receive consistent signals about what kinds of opinions they should hold, while people in more heterogeneous environments receive conflicting signals.
 
@@
 
-This paper set out to test whether the unique rates of attitude change found among young adults are due to the social features of this life stage or due to features of how human brains develop and process information over time.
-We leverage delays in the timing of life course transitions and the emergence of an "age of independence" over the past 70 years to explore whether attitude behavior shifted in response to these changes.
-We expected that we should see more durable attitude change in early adulthood and an extension of the "impressionable years" pattern up the age ladder.
+This paper leveraged delays in the timing of life course transitions and the emergence of an "age of independence" over the past 70 years to adjudicate whether heightened rates of attitude change in early adulthood are principally ontogenetic in origin, and therefore relatively stable over time, or whether they emerge due to the distinct social circumstances of adolescence and early adulthood.
+We expected to see more attitude changes in early adulthood and an extension of the "impressionable years" pattern up the age ladder in more recent cohorts as early adulthood became more socially distinct from middle age.
 
-We found no evidence of either process, broadly suggesting that changes in the timing of life course transitions, as well as broader social and cultural changes around early adulthood have not broadly affected attitude behavior among young adults.
-The pattern of results generally suggests that what makes early adulthood unique when it comes to attitude development is not its social conditions, but how humans process information as they age.
-These years are not formative because they are unfettered, they are formative because they are early.
+Instead, we found that other social processes, particularly the dramatic expansion of educational attainment during the second half of the 20th Century, affected the opinions of middle-aged and older respondents much more substantially than any life-course changes.
+This expansion of education has stabilized opinions across the board, reducing random wave-to-wave change (what is often called measurement error) while also slightly increasing the rates at which people made changes that appear to persist over time.
+
+In this way, our findings provide broad support for arguments that the unique rates of attitude change across the life course are shaped by social structures, rather than just ontogenetic factors, and we call for much more exploration by social scientists into the social factors that enable stable opinion holding.
 
 # References
+
+\singlespace

### 2024-07-29 other_docs/model_kk.Rmd

--- 2024-06-05 code/kkapproach/model_kk.Rmd
@@
 
-Our central questions are 1) whether respondents in the youngest age group (18-25) make more change in more recent panels than in earlier panels and 2) whether respondents in the next-youngest age group (26-34) become more similar to respondents in the youngest age group over time.
-To evaluate these hypotheses, we need a measure of the rate of change for respondents of different age groups over time in different era.
+Panels differ in both the number of times respondets are asked a particular question as well as the duration between observations.
 
-A central challenge for this analysis is that panels differ in the questions they ask, the number of times they ask participants to respond to different questions, and the duration between observations.
+In the one-item, one-panel case (partisan identification in the 1972-76 panel, for example), we could model the absolute change from wave-to-wave as a function of the duration between observations (more time between waves should lead to more change).
 
-In the one-item, one-panel case (partisan identification in the 1972-76 panel, for example), we could model the absolute change from wave-to-wave as a function of the duration between observations.
+$$ y_{i} = \beta_0 + \beta_1*d_i + \epsilon_i $$
 
-$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1*d_i + \epsilon_i $$
+In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise).
+Conversely, $\beta_1$ captures the expected expected change one a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$).
 
-In this equation, $d_{i}$ captures the duration of time between $t-1$ and $t$ in years.
-In the 1972-76 panel, respondents are asked their partisan identification three times: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart.
-If a person participates in all three waves, they have three observations of $|y_{i,t}-y_{i,t-1} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).
+Our theoretical question deals primarily with whether these coefficients change over time.
+In particular, we want to know whether people demonstrate more or less real change over time.
+To explore that, we can interact these measures with time.
 
-If respondents have a stable underlying partisan identification they report with error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$.
-In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise).
-Conversely, $\beta_1$ captures the expected expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$).
-For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3, and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common.
-
-Our theoretical question deals primarily with whether these coefficients ($\beta_0$ and $\beta_1$) change over time.
-To explore that, we can add an additional term for time and interact the duration term with time.
-
-$$ |y_{i,t}-y_{i,t-1} | = \beta_0 + \beta_1d + \beta_2t + \beta_3(d*t) + \epsilon_i $$
+$$ y_{i} = \beta_0 + \beta_1d + \beta_2t + \beta_3(d*t) + \epsilon_i $$
 
@@
 
-A challenge for our analysis is that while the panels we explore often have relatively large samples, most questions are only asked in a few panels.
-As such, we lack large numbers of any particular question, meaning our time trends for any individual question is likely to be unstable.
-One potential solution is to pool questions and estimate a single time trend for stability in attitudes over time, but that is potentially problematic if the kinds of questions asked in later panels are systematically different from the kinds of questions asked in earlier panels.
-It could also obscure the fact that some items might be becoming more stable over time while others are becoming less stable.
+Now how do we model this with multiple questions?
 
-Multilevel models provide an alternative solution to estimate variation in trends in the presence of uncertainty.
-
-$$ |y_{i,j,t}-y_{i,j,t-1} | = \beta_{0j} + \beta_{1j}d + \beta_{2j}t + \beta_{3j}(d*t) + \epsilon_i $$
+$$ y_{ij} = \beta_{0j} + \beta_{1j}d + \beta_{2j}t + \beta_{3j}(d*t) + \epsilon_i $$
 
@@
 
-In this framework, the first-level model models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above.
-The second-level equations represent the items.
-Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for time, duration, and their interaction.
-These effects are assumed to be normally distributed around a main effect.
-
-## Results
-
-```{r}
-long_data %>%
-  group_by(name) %>%
-  nest() %>%
-  mutate(t = map(data, lm_func)) %>%
-  unnest(t) %>%
-  filter(term == "date") %>%
-  unnest(data) %>%
-  mutate(direction = ifelse(estimate < 0, "negative", "positive")) %>%
-  ggplot(aes(x = date, y = a, color = direction)) + 
-  facet_wrap(~reorder(name, estimate)) +
-  geom_smooth(method = "lm") + 
-  theme_bw() + 
-  labs(x = "Year", y = "Absolute difference")
-```
-
-The figure above presents the expected absolute difference over time generated from separate regressions for each item only on date of observation.
-This model does not account for varying duration observed in different panels.
-Generally speaking, most items show a negative slope over time, suggesting that respondents are becoming more stable in their responses over time.
-
-Notably, most of the "feeling thermometer" items show positive slopes over time, suggesting people are becoming more variable in how they feel toward other groups.
-In contrast, general policy positions show negative slopes over time.
+In this framework, each item is assumed to have its own intercept and and effects for time, duration, and their interaction, but these...

### 2025-06-06 code/aggregate_analyses.R
cleaned folder; code for cleaning

@@ -1,254 +1,0 @@
-
-
-
-#########################################
-####      Analyses at Aggregate      ####
-#########################################
-
-#Panel summaries
-psummaries <- long_difference %>%
-  mutate(wave_pair = paste(t1, t2, df, sep = "-")) %>%
-  group_by(name, wave_pair) %>%
-  summarise(d = weighted.mean(duration), a = weighted.mean(abs_diff),
-            date = weighted.mean(date), n = n(), sd = sd(abs_diff)) %>%
-  #Filtering out bad questions
-  filter(name != "natpoor", 
-         name != "nathome") %>% ungroup() %>%
-  mutate(dec_diff = (as.numeric(difftime(date, max(date), units = "days")))/3652.5) %>%
-  mutate(se = sd/sqrt(n))
-
-# Summarize absolute difference at the wave-pair level for each age group
-# for each question. 
-summaries <- long_diff_demo %>%
-  left_join(ik, by = c("df"="df", "id"="id", "t1"="wave")) %>%
-  filter(age <= 80, !is.na(ba), !is.na(evermarried), !is.na(student)) %>%
-  mutate(wave_pair = paste(t1, t2, df, sep = "-")) %>%
-  group_by(wave_pair, name, age_group) %>%
-  summarise(d = weighted.mean(duration, weight = weight), a = weighted.mean(abs_diff, weight = weight),
-            pct_ba = weighted.mean(ba, weight = weight),
-            pct_lt = weighted.mean(lt, weight = weight),
-            pct_married = weighted.mean(evermarried, weight = weight),
-            pct_student = weighted.mean(student, weight = weight),
-            pct_kid = weighted.mean(impute_kid, weight = weight),
-            pct_kih = weighted.mean(impute_kih, weight = weight),
-            date = weighted.mean(date, weight = weight), n = n(), sd = sd(abs_diff),
-            weighted_n = sum(weight)) %>%
-  #Filtering out bad questions
-  filter(name != "natpoor", 
-         name != "nathome") %>% ungroup() %>%
-  mutate(dec_diff = (as.numeric(difftime(date, max(date), units = "days")))/3652.5) %>%
-  mutate(se = sd/sqrt(n)) %>%
-  mutate(name = ifelse(name == "nateduc", "natschools", name))
-
-
-
-
-# Multilevel model
[…]

### 2025-06-06 code/analyses.R

@@ -1,445 +1,0 @@
-
-
-#########################################
-####      Analysis of Stability      ####
-#########################################
-
-
-#Load in long data frame from clean_data.R file
-
-load("./clean_data/long_difference.Rdata")
-load("./clean_data/long_demog.Rdata")
-
-source("./code/get_dem_variables.R")
-
-long_diff_demo <- left_join(long_difference, long_demog, by = c("df"="df", "id"="id", "t1"="wave")) %>%
-  mutate(t = (as.numeric(difftime(date, max(date), units = "days")))/3652.5,
-         a = abs_diff,
-         d = duration) %>%
-  mutate(weight = ifelse(weight == 0, 1, weight)) %>%
-  filter(name != "natfood") %>%
-  mutate(name = ifelse(name == "nateduc", "natschools", name)) %>%
-  select(-c(abs_diff, duration)) %>%
-  filter(age <= 80)
-
-#########################################
-####      Analyses at Individual    ####
-#########################################
-
-library(optimx)
-
-#Don't need to impute missing data for theoretical model
-# Multilevel model
-m0 <- lmer(a ~ d + t + age_group + t*age_group + 
-             (1|name),
-           data = long_diff_demo, weights = weight,
-           REML = FALSE,
-           control = lmerControl(optimizer ="nloptwrap"))
-#BIC: 8800268
-
-m1 <- lmer(a ~ d + t + age_group + t*age_group + 
-             (1 + d|name),
-           data = long_diff_demo, weights = weight,
-           REML = FALSE,
-           control = lmerControl(optimizer ="nloptwrap"))
-# BIC: 8799775
-
-m2 <- lmer(a ~ d + t + age_group + t*age_group + 
-             (1 + d + t|name),
-           data = long_diff_demo, weights = weight,
-           REML = FALSE,
-           control = lmerControl(optimizer ="nloptwrap"))
-#BIC: 8798411
-
-m3a <- lmer(a ~ d + t + age_group + t*age_group + 
-              (1 + d + t + age_group|name),
-            data = long_diff_demo, weights = weight,
-            REML = FALSE,
-            control = lmerControl(optimizer ="nloptwrap"))
-#BIC: 8797630
-
-m3b <- lmer(a ~ d + t + age_group + t*age_group + d*age_group + 
-              (1 + d + t + age_group + t*age_group + d*age_group|name),
[…]

### 2025-06-06 code/cleaning/clean_anes_00_04.R

@@ -1,0 +1,366 @@
+
+# cleaning/clean_anes_00_04.R
+
+source("code/cleaning/utils.R")
+
+
+clean_anes_00_04 <- function(path) {
+  
+  anes00 <- read_and_zap(path)
+  
+  cleaned <- anes00 %>%
+    mutate(id = row_number()) %>%
+    select(id, 
+           WT00PRE, WT00PO, WT02PRE, WT02PO, WT04,
+           M000908,M023126X,M045193,
+           M000008, M000130, M022012C, M024012C, M044013b,
+           M000523, M023038X, M045058x,
+           M000446, M001368, M023022,
+           M001475, M025101, M045158,
+           M001477, M025103, M045160,
+           M001476, M025102, M045159, 
+           M000513a, M023033, M045143,
+           M000694, M001403, M045110,
+           M000545, M001385, M045127,
+           M000682, M023052, M025113, M045068,
+           M000678, M023054, M025116, M045076,
+           M000677, M023045, M025106, M045069, #aids
+           M000681, M023055, M025117, M045077, #socsec
+           M000683, M023047A, M025108A, M045071a, #schools
+           M000685, M023049, M025109, M045073,
+           M000676, M023046, M025107, M045070,
+           M000684, M023048, M025109, M045072,
+           M001308, M025055, M045023,
+           M001309, M025056, M045024,
+           M001310, M025057, M045025,
+           M001311, M025058, M045026,
+           M001313, M025060, M045028,
+           M001312, M025059, M045027,
+           M001306, M025053, M045021,
+           M001316, M025063, M045031,
+           M001321, M025067, M045035,
+           M001314, M025061, M045029,
+           M001323, M025068, M045036,
+           M001324, M025069, M045037,
+           M001537, M025177, M045152, #crooked
+           M001535, M025175, M045150, #wastetax
+           M001534, M025174, M045149, #trustgov
+           M001536, M025176, M045151,#runfew
+           # Attention political
+           M000301,	M001201, M023001,	M025001, 
+           # People like me 
+           M001528,	M025173, M045148, 
+           # Spend poor ppl
+           M000680, M023053A,	M025115A,	M045075a, 
+           # Spend military
+           M025105,	M045081, 
+           # Spend aid to blacks
+           M000687, M023057, M025119,	M045079,
+           # Immigrant number
+           M000510, M045109, 
+           # Favor or oppose laws around homosexuals
+           M001478, M045111, 
[…]

### 2025-06-06 code/cleaning/clean_anes_16_20.R

@@ -1,0 +1,327 @@
+
+# cleaning/clean_anes_16_20.R
+
+source("code/cleaning/utils.R")
+
+
+clean_anes_16_20 <- function(path) {
+  
+  anes16 <- read_sav(path) %>%
+    zap_labels()
+  
+  cleaned <- anes16 %>%
+    mutate(id = row_number()) %>%
+    select(id, V161267, V201507x,
+           V160102,#weight for pre- and post 2016 
+           V200011a, #weight for pre-16 and pre-20 combined
+           V200011b, #weight for post-16 and post-20 combined
+           V164002, V165002, V203053, V203078,
+           V161158x, V201231x,
+           V161126, V162171, V201200,
+           V161153, V201347,
+           V161217, V201235,
+           V161216, V201234,
+           V161189, V201255,
+           V161184, V201252,
+           V161198, V201258,
+           V161232, V201336,
+           V161178, V201246,
+           V161181, V201249,
+           V161212, V201321,
+           V161205, V201300,
+           V161206, V201303,
+           V161209, V201312,
+           V161208, V201309,
+           V161211, V201318,
+           V161095, V201156,
+           V161096, V201157,
+           V162312, V202480,
+           V162314, V202482,
+           V162101, V202164,
+           V162097, V202161,
+           V162100, V202163,
+           V162098, V202162,
+           V162311, V202479,
+           V162103, V202166,
+           V162108, V202170,
+           V162110, V202171, 
+           # Attention Campaigns
+           V161004, V201006, 
+           # People like me Likert
+           V162216, V202213, 
+           # Spend Poor 
+           V161211, V201318,
+           # Less gvt 
+           V162185, V202253, 
+           # Preferential hiring
+           V162238, V202249,
+           # Death penalty
+           V161233, V201343,
+           # Number of migrants
+           V162157, V202232,
+           # Favor Oppose Homo
+           V161229, V201412,
+           # T: asian-americans
+           V162310, V202477,
+           # T: feminists
+           V162096, V202160,
+           # T: illegal immigrants
+           V162313, V202481,
+           # T: Equality of opportunity for success
+           V162243, V202260,
+           # Worry less equality
+           V162244, V202261,
+           # Unequal no problem
+           V162245, V202262,
+           # Equality less problem
+           V162246, V202263,
+           # Moral Relativity
+           V162207, V202264,
[…]

### 2025-06-06 code/cleaning/clean_anes_56_60.R

@@ -1,0 +1,183 @@
+
+# cleaning/clean_anes_1956_60.R
+
+source("code/cleaning/utils.R")
+
+clean_anes_1956_60 <- function(path) {
+  raw <- read_and_zap(path)
+  
+  cleaned <- raw %>%
+    mutate(id = row_number()) %>%
+    select(id, V600569, #weight
+           V560193, # Interview date 1
+           V580309, # Interview date 2
+           V600578, # Interview date 3
+           V600837, # Interview date 4
+           
+           V560088, # Party ID 1 
+           V580360, # Party ID 2
+           V600657, # Party ID 3
+           V600835, # Party ID 4 
+           
+           V560295, # Age 1 
+           V580472, # Age 2
+           V600688, # Age 3
+           
+           V560035, V580323, V600622, # stayhome 
+           
+           V560097, # Pay attention campaign 1
+           V580373, # Pay attention campaign 2
+           V600664, # Pay attention campaign 3
+           
+           V560108, # People like me, no say 1
+           V600673, # People like me, no say 3
+           
+           V560111, #so many vote, 1
+           V600676, #so many vote, 3
+           
+           V560112, #pol too complicated 1
+           V600677, #pol too complicated 3
+           
+           V560115, #officials dont care 1
+           V600680 #officials dont care 3
+    ) %>%
+    mutate(weight_4 = as.character(V600569),
+           weight_3 = weight_4,
+           weight_2 = weight_4,
+           across(c(V560088, V580360, V600657, V600835), ~ ifelse(.x %in% c(7, 8, 9), NA, .x)),
+           across(c(V560295, V580472, V600688), ~ ifelse(.x > 97, NA, .x)),
+           across(c(V560097, V580373, V600664), ~ ifelse(.x %in% c(8,9), NA, .x)), 
+           across(c(V560035, V580323, V600622),
+                  ~recode(.x, 
+                          "1"=1, "2"=1, "3"=1.5, "4"=2, "5"=2, "7"=NA_real_,
+                          "0"=NA_real_, "8"=NA_real_, "9"=NA_real_)),
+           across(c(V560108, V600673, V560111, V600676, V560112, V600677,
+                    V560115, V600680),
+                  ~ifelse(.x == 8, 3, ifelse(.x == 9, NA, .x))),
+           age_1 = V560295,
+           age_2 = V580472,
+           age_3 = V600688,
[…]

### 2025-06-06 code/cleaning/clean_anes_80.R

@@ -1,0 +1,178 @@
+
+# cleaning/clean_anes_80.R
+
+source("code/cleaning/utils.R")
+
+clean_anes_1980 <- function(path) {
+  
+  anes8 <- read_and_zap(path)
+  
+  cleaned <- anes8 %>%
+    mutate(id = row_number()) %>%
+    select(
+      id,
+      VMP0325,
+      VMP2404,
+      VMP3386,
+      VMP0578,
+      VMP2425,
+      VMP3571,
+      VMP4105,
+      VMP0191,
+      VMP2212,
+      VMP3212,
+      VMP4022,
+      VMP0121,
+      VMP2125,
+      VMP3213,
+      VMP2192,
+      VMP3284,
+      VMP0144,
+      VMP2159,
+      VMP3254,
+      VMP0133,
+      VMP2137,
+      VMP3234,
+      VMP0052,
+      VMP2052,
+      VMP3109,
+      VMP0053,
+      VMP2053,
+      VMP3110,
+      # attention political campaigns
+      VMP0005,
+      VMP2001,
+      VMP3001,
+      VMP4001
+    ) %>%
+    mutate(
+      across(c(VMP0325, VMP2404, VMP3386), ~ ifelse(.x == 0, NA, .x)),
+      age_1 = VMP0325,
+      age_2 = VMP2404,
+      age_3 = VMP3386,
+      age_1 = ifelse(is.na(age_1) & !is.na(age_2), age_2, age_1),
+      age_1 = ifelse(is.na(age_1) & !is.na(age_3), age_3, age_1),
+      age_2 = ifelse(is.na(age_2), age_1, age_2),
+      age_3 = ifelse(is.na(age_3), age_1, age_3),
+      age_4 = age_3,
+      across(c(age_1:age_4), ~ as.character(.x))
+    ) %>%
+    mutate(
+      date_str = sprintf("%06d", as.integer(VMP0578)),
+      date_1 = paste(
+        paste0("19", substr(date_str, 5, 6)),
+        substr(date_str, 1, 2),
+        substr(date_str, 3, 4),
+        sep = "-"
+      )
+    ) %>%
+    select(-c(date_str)) %>%
+    mutate(
+      date_str = sprintf("%06d", as.integer(VMP2425)),
+      date_2 = paste(
+        paste0("19", substr(date_str, 5, 6)),
+        substr(date_str, 1, 2),
+        substr(date_str, 3, 4),
[…]

### 2025-06-06 code/cleaning/clean_anes_90_92.R

@@ -1,0 +1,344 @@
+
+# cleaning/clean_anes_90_92.R
+
+source("code/cleaning/utils.R")
+
+
+clean_anes_90_92 <- function(path) {
+  
+  anes90 <- read_and_zap(path)
+  
+  cleaned <- anes90 %>%
+    mutate(id = row_number()) %>%
+    select(id,
+           V923009,
+           V900552, V923903,
+           V900052, V900053, V912032, V923026, V923027, V925005, V925006,
+           V900320, V912333, V923634,
+           V900406, V912450, V923509,
+           V912831, V926139,
+           V900498, V912485, V923604,
+           V900446, V923718,
+           V900447, V923724,
+           V900479, V923732,
+           V900452, V912600, V923701,
+           V900377, V923815,
+           V900379, V923727,
+           V900380, V923811,
+           V900382, V923725,
+           V900383, V923818,
+           V900384, V923730,
+           V900385, V923813,
+           V900439, V912475, V923707,
+           V900151, V912222, V923317,
+           V900152, V912228, V923318,
+           V900155, V912232, V925323,
+           V912231, V925333,
+           V900156, V912220, V925319,
+           V900161, V912226, V925326,
+           V900157, V925316,
+           V912233, V925327,
+           V900158, V912239, V925324,
+           V900438, V923801,
+           V900507, V926123,
+           V900505, V926121,
+           V900504, V912487, V926120,
+           V900506, V912488, V926122, 
+           # Attention political campaigns(1,3)
+           V900062,	V923101, 
+           # People like me Likert (1-4)
+           V900509,	V912489, V926102,
+           # Less gvtment 
+           V900333,	V925729, 
+           # Preferential hiring
+           V900463,	V912558, V925935, 
+           # Death Penalty 
+           V900477,	V925933, 
+           # Immigrants decrease 
+           V912619,	V926235, 
+           # T: asian-americans
+           V912235,	V925339, 
+           # T: Feminists
+           V912240,	V925317, 
+           # T: welfare
+           V900159, V925318, 
+           # T: environmentalists 
+           V912230,	V925329, 
+           # T: illegal immigrants 
+           V912234, V925331, 
+           # Equal Op Success 
+           V900426, V926024, 
+           # Equal rights too far 
+           V900427,	V912705,	V926025, 
+           # Not everyone chance 
+           V900428, V926029, 
+           # Worry less equality
+           V900429,	V926026, 
[…]

### 2025-06-06 code/cleaning/clean_anes_92_97.R

@@ -1,0 +1,525 @@
+
+# cleaning/clean_anes_92_97.R
+
+source("code/cleaning/utils.R")
+
+clean_anes_92_97 <- function(path) {
+  
+  anes92 <- read_and_zap(path)
+  
+  cleaned <- anes92 %>%
+    mutate(id = row_number()) %>%
+    select(id, 
+           V923008a, V923008b, V940006a, 
+           V960005a, V960005b,#weights
+           V923903, V941203, V960605, 
+           V923026, V923027, V925005, V925006, V937014, V937015, 
+           V940033, V940034, V952360, V960012, V960011, V960903, V960902,
+           V970367,
+           V923634, V937370, V940655, V952263a, V960420, V970106, #partyid
+           V923509, V937204, V940839, V952253, V960365, V961269, V970302, #polviews
+           V926139, V960567, V961258, V970011, #trust
+           V923604, V941019, V960410, #stayhome
+           V923718, V940930, V960483, #jobguar
+           V923716, V940950, V960479, #govins
+           V923724, V940936, V960487, V961210, V970193, #helpblk
+           V923732, V941014, V960503, #abortion
+           V923701, V940940, V960450, V970096, #spendserv
+           V923815, V940817, V960561, #environment
+           V923727, V940821, V960498, #aids
+           V923811, V940819, V960560, #socsec
+           V923725, V940822, V960496, #schools
+           V923818, V940823, V960562, #food stamps
+           V923730, V960501, #homeless
+           V923813, V940824, V960564, #child care
+           V923726, V940820, V960497, #welfare
+           V923814, V940825, V960563,
+           V923707, V940929, V960463,
+           V923317, V937141, V940301, V960292,
+           V923318, V937142, V940302, V960293,
+           V925323, V940305, V961029, V970040,
+           V925333, V940313, V961030, V970041, 
+           V925319, V940306, V961031,
+           V925326, V940311, V961032,
+           V925322, V940314, V952115, V961034, V970056,
+           V925316, V940307, V961033, V970055,
+           V925328, V937146, V961027,
+           V925327, V940304, V961037, V970042,
+           V925335, V937145, V940318, V961042, V970054,
+           V940312, V961035, 
+           V925324, V940308, V961039,
+           V960569, V961259, V970012,
+           V923801, V940928, V960543, V970230,
+           V926123, V941036, V961254,
[…]

### 2025-06-06 code/cleaning/clean_gss_06_10.R

@@ -1,0 +1,81 @@
+
+# cleaning/clean_gss_06_10.R
+
+source("code/cleaning/utils.R")
+
+clean_gss_06_10 <- function(path) {
+  
+  gss06 <- read_and_zap(path)
+  
+  cleaned <- gss06 %>%
+    mutate(id = row_number()) %>%
+    select(id, 
+           wtpannr12, wtpannr123,
+           age_1, age_2, age_3,
+           dateintv_1, dateintv_2, dateintv_3,
+           partyid_1, partyid_2, partyid_3,
+           polviews_1, polviews_2, polviews_3,
+           trust_1, trust_2, trust_3,
+           fair_1, fair_2, fair_3,
+           helpful_1, helpful_2, helpful_3,
+           natenvir_1, natenvir_2, natenvir_3,
+           nataid_1, nataid_2, nataid_3,
+           natsoc_1, natsoc_2, natsoc_3,
+           natchld_1, natchld_2, natchld_3,
+           natfare_1, natfare_2, natfare_3,
+           natcrime_1, natcrime_2, natcrime_3,
+           nateduc_1, nateduc_2, nateduc_3, 
+           natarms_1, natarms_2, natarms_3, 
+           natrace_1, natrace_2, natrace_3, 
+           cappun_1, cappun_2, cappun_3, 
+           letin1a_1, letin1a_2, letin1a_3,
+           tax_1, tax_2, tax_3,
+           wrkwayup_1, wrkwayup_2, wrkwayup_3) %>%
+    mutate(weight_2 = as.character(wtpannr12),
+           weight_3 = as.character(wtpannr123),
+           age_1 = ifelse(is.na(age_1) & !is.na(age_2), age_2 - 2, age_1),
+           age_2 = ifelse(is.na(age_2), age_1 + 2, age_2),
+           age_3 = ifelse(is.na(age_3), age_1 + 4, age_3),
+           across(age_1:age_3, ~as.character(.x))) %>%
+    mutate(date_1 = paste(2006,substr(dateintv_1, 1, 1),substr(dateintv_1, 2, 3),
+                          sep = "-"),
+           date_2 = paste(2008,substr(dateintv_2, 1, 1),substr(dateintv_2, 2, 3),
+                          sep = "-"),
+           date_2 = ifelse(date_2 == "2008-NA-NA", NA, date_2),
+           date_3 = paste(2010,substr(dateintv_3, 1, 1),substr(dateintv_3, 2, 3),
+                          sep = "-"),
[…]

### 2025-06-06 code/cleaning/clean_gss_08_12.R

@@ -1,0 +1,82 @@
+
+# cleaning/clean_gss_08_12.R
+
+source("code/cleaning/utils.R")
+
+clean_gss_08_12 <- function(path) {
+  
+  gss08 <- read_and_zap(path)
+  
+  cleaned <- gss08 %>%
+    mutate(id = row_number()) %>%
+    select(id, 
+           wtpannr12, wtpannr123,
+           age_1, age_2, age_3, 
+           dateintv_1, dateintv_2, dateintv_3,
+           partyid_1, partyid_2, partyid_3,
+           polviews_1, polviews_2, polviews_3,
+           trust_1, trust_2, trust_3,
+           fair_1, fair_2, fair_3,
+           helpful_1, helpful_2, helpful_3,
+           natenvir_1, natenvir_2, natenvir_3,
+           nataid_1, nataid_2, nataid_3,
+           natsoc_1, natsoc_2, natsoc_3,
+           natchld_1, natchld_2, natchld_3,
+           natfare_1, natfare_2, natfare_3,
+           natcrime_1, natcrime_2, natcrime_3, 
+           nateduc_1, nateduc_2, nateduc_3, 
+           natarms_1, natarms_2, natarms_3, 
+           natrace_1, natrace_2, natrace_3, 
+           cappun_1, cappun_2, cappun_3, 
+           letin1a_1, letin1a_2, letin1a_3,
+           tax_1, tax_2, tax_3,
+           wrkwayup_1, wrkwayup_2, wrkwayup_3) %>%
+    mutate(weight_2 = as.character(wtpannr12),
+           weight_3 = as.character(wtpannr123),
+           age_1 = ifelse(is.na(age_1) & !is.na(age_2), age_2 - 2, age_1),
+           age_1 = ifelse(is.na(age_1) & !is.na(age_3), age_3 - 4, age_1),
+           age_2 = ifelse(is.na(age_2), age_1 + 2, age_2),
+           age_3 = ifelse(is.na(age_3), age_1 + 4, age_3),
+           across(age_1:age_3, ~as.character(.x))) %>%
+    mutate(date_1 = paste(2008,substr(dateintv_1, 1, 1),substr(dateintv_1, 2, 3),
+                          sep = "-"),
+           date_2 = paste(2010,substr(dateintv_2, 1, 1),substr(dateintv_2, 2, 3),
+                          sep = "-"),
+           date_2 = ifelse(date_2 == "2010-NA-NA", NA, date_2),
[…]

### 2025-06-06 code/cleaning/clean_gss_10_14.R

@@ -1,0 +1,83 @@
+
+
+
+# cleaning/clean_gss_10_14.R
+
+source("code/cleaning/utils.R")
+
+clean_gss_10_14 <- function(path) {
+  
+  gss10 <- read_and_zap(path)
+  
+  cleaned <- gss10 %>%
+    mutate(id = row_number()) %>%
+    select(id, 
+           wtpannr12, WTPANNR123,
+           age_1, age_2, age_3,
+           dateintv_1, dateintv_2, dateintv_3,
+           partyid_1, partyid_2, partyid_3,
+           polviews_1, polviews_2, polviews_3,
+           trust_1, trust_2, trust_3,
+           fair_1, fair_2, fair_3,
+           helpful_1, helpful_2, helpful_3,
+           natenvir_1, natenvir_2, natenvir_3,
+           nataid_1, nataid_2, nataid_3,
+           natsoc_1, natsoc_2, natsoc_3,
+           natchld_1, natchld_2, natchld_3,
+           natfare_1, natfare_2, natfare_3,
+           natcrime_1, natcrime_2, natcrime_3, 
+           nateduc_1, nateduc_2, nateduc_3, 
+           natarms_1, natarms_2, natarms_3, 
+           natrace_1, natrace_2, natrace_3, 
+           cappun_1, cappun_2, cappun_3, 
+           letin1a_1, letin1a_2, letin1a_3,
+           tax_1, tax_2, tax_3,
+           wrkwayup_1, wrkwayup_2, wrkwayup_3) %>%
+    mutate(weight_2 = as.character(wtpannr12),
+           weight_3 = as.character(WTPANNR123),
+           age_1 = ifelse(is.na(age_1) & !is.na(age_2), age_2 - 2, age_1),
+           age_1 = ifelse(is.na(age_1) & !is.na(age_3), age_3 - 4, age_1),
+           age_2 = ifelse(is.na(age_2), age_1 + 2, age_2),
+           age_3 = ifelse(is.na(age_3), age_1 + 4, age_3),
+           across(age_1:age_3, ~as.character(.x))) %>%
+    mutate(date_1 = paste(2010,substr(dateintv_1, 1, 1),substr(dateintv_1, 2, 3),
+                          sep = "-"),
+           date_2 = paste(2012,substr(dateintv_2, 1, 1),substr(dateintv_2, 2, 3),
+                          sep = "-"),
+           date_2 = ifelse(date_2 == "2012-NA-NA", NA, date_2),
[…]

### 2025-06-06 code/cleaning/clean_gss_16_20.R

@@ -1,0 +1,88 @@
+
+# cleaning/clean_gss_16_20.R
+
+source("code/cleaning/utils.R")
+
+clean_gss_16_20 <- function(path) {
+  
+  gss16 <- read_and_zap(path)
+  
+  cleaned <- gss16 %>%
+    mutate(id = row_number()) %>%
+    select(id, 
+           wtssnr_2,
+           age_1a, age_1b, age_2,
+           dateintv_1a, dateintv_1b, dateintv_2,
+           partyid_1a, partyid_1b, partyid_2,
+           polviews_1a, polviews_1b, polviews_2,
+           trust_1a, trust_1b, trust_2,
+           fair_1a, fair_1b, fair_2,
+           helpful_1a, helpful_1b, helpful_2,
+           natenvir_1a, natenvir_1b, natenvir_2,
+           nataid_1a, nataid_1b, nataid_2,
+           natsoc_1a, natsoc_1b, natsoc_2,
+           natchld_1a, natchld_1b, natchld_2,
+           natfare_1a, natfare_1b, natfare_2,
+           natcrime_1a, natcrime_1b, natcrime_2, 
+           nateduc_1a, nateduc_1b, nateduc_2, 
+           natarms_1a, natarms_1b, natarms_2, 
+           natrace_1a, natrace_1b, natrace_2, 
+           cappun_1a, cappun_1b, cappun_2, 
+           letin1a_1a, letin1a_1b, letin1a_2,
+           tax_1a, tax_1b, tax_2,
+           wrkwayup_1a, wrkwayup_1b, wrkwayup_2) %>%
+    mutate(weight_2 = as.character(wtssnr_2)) %>%
+    mutate(age_1a = ifelse(is.na(age_1a) & !is.na(age_2), age_2 - 4, age_1a),
+           age_1b = ifelse(is.na(age_1b) & !is.na(age_2), age_2 - 2, age_1b),
+           age_2 = ifelse(is.na(age_2) & !is.na(age_1a), age_1a + 4, age_2),
+           age_2 = ifelse(is.na(age_2) & !is.na(age_1b), age_1b + 2, age_2),
+           across(age_1a:age_2, ~as.character(.x))) %>%
+    mutate(md = sprintf("%04d", dateintv_1a),
+           date_1a = paste(2016, substr(md, 1, 2),substr(md, 3, 4),
+                           sep = "-"),
+           date_1a = ifelse(date_1a == "2016-00-NA", NA, date_1a),
+           md = sprintf("%04d", dateintv_1b),
[…]

### 2025-06-06 code/old/build_long_dataframe.R

@@ -1,2260 +1,0 @@
-# Packages ----
-library(tidyverse)
-library(patchwork)
-library(brms)
-parallel::detectCores()
-
-# Load Data ----
-load("~/Documents/extended_formative_windows/Data/clean_anes5.Rdata")
-load("~/Documents/extended_formative_windows/Data/clean_anes7.Rdata")
-load("~/Documents/extended_formative_windows/Data/clean_anes9.Rdata")
-load("~/Documents/extended_formative_windows/Data/clean_g6.Rdata")
-
-# PARTYID - ANES56 ----
-# Strict
-a5_pid_s <- clean_anes5 %>%
-  mutate(id = 1:nrow(.)) %>%
-  select(id, age_1, contains("party")) %>%
-  filter(
-    partyid_1 != 7,
-    partyid_1 != 8,
-    partyid_1 != 9,
-    partyid_2 != 7,
-    partyid_2 != 8,
-    partyid_2 != 9,
-    partyid_3 != 7,
-    partyid_3 != 8,
-    partyid_3 != 9
-  ) %>%
-  mutate(
-    change_w2 = abs(partyid_2 - partyid_1),
-    change_w3 = abs(partyid_3 - partyid_2),
-    total_change_s = change_w2 + change_w3, 
-    any_change_s = if_else(total_change_s >= 1, 1, 0), 
-    crossed_midpoint_s = case_when(
-      partyid_1 < 3 & partyid_2 > 3 ~ 1,
-      partyid_1 > 3 & partyid_2 < 3 ~ 1,
-      partyid_2 < 3 & partyid_3 > 3 ~ 1,
-      partyid_2 > 3 & partyid_3 < 3 ~ 1,
-      partyid_1 < 3 & partyid_3 > 3 ~ 1,
-      partyid_1 > 3 & partyid_3 < 3 ~ 1,
-      TRUE ~ 0
-    ), 
-    two_points_s = if_else(
-      abs(partyid_3-partyid_1) >= 2, 1,0
-    )
-  ) %>% 
-  select(id, age_1, total_change_s, any_change_s, crossed_midpoint_s, two_points_s) %>% 
-  mutate(variable = "partyid")
-
-# Recodes 7s and 8s into 3s to be less strict 
-a5_pid_l <- clean_anes5 %>%
-  mutate(id = 1:nrow(.)) %>%
-  select(id, age_1, contains("party")) %>%
-  filter(
[…]

### 2025-06-06 code/old/clean_panels.R

@@ -1,1364 +1,0 @@
-#Cleaning for Reliability Models
-library(tidyverse)
-library(haven)
-library(broom)
-
-
-anes5 <- read_dta("~/Dropbox/data/anes/anes5660/anes_mergedfile_1956to1960.dta") 
-
-clean_anes5 <- anes5 %>% 
-  select(#Negroes fair treatment
-    V560044,V580329,V600628, 
-    #deseg
-    V560074,V580333,V600636,
-    #Job guarantee
-    V560032,V580321,V600620,
-    #Help poor countries
-    V560041,V580327,V600626,
-    #Build schools
-    V560053,V580325,V600624,
-    #Fight communism
-    V560056,V580331,V600630,
-    #Leave utilities to prvt mkt
-    V560059,V580319,V600618,
-    #US better off stay home
-    V560035,V580323,V600622,
-    #Partyid
-    V560088,V580360,V600835,
-    #closerjob
-    V560034, V580322, V600621,
-    #closeraid
-    V560043, V580328, V600627,
-    #closertreat
-    V560046, V580330, V600629,
-    #closerbuild
-    V560055, V580326, V600625,
-    #closerfghtcomm
-    V560058, V580332, V600631,
-    #closerprvtpwr
-    V560061, V580320, V600619,
-    #closerdeseg
-    V560076, V580336, V600639,
-    #Age
-    V560295, V580472, V600688, 
-    #Marital status
-    V560177, V580474, V600690,
-    #Education
-    V560181, V580478, V600694,
-    #Children in household
-    V560178, V580475, V600691,
-    #Occupational code
-    V560120,
-    #sex
-    V560171, V580468, V600684,
-    #race
-    V560172, V580469, V600685,
-    #survey weight
-    V600569) %>%
-  mutate(weight = V600569,
-         age_1 = V560295,
-         age_2 = V580472,
-         age_3 = V600688,
-         student_1 = ifelse(V560120 %in% c(92), 1, 0),
-         marital_1 = V560177, marital_2 = V580474,
-         marital_3 = V600690,
-         marital_1 = ifelse(V560177 == 9, NA, V560177),
-         marital_1 = ifelse(V560177 %in% c(2), "single/nm", marital_1),
-         marital_1 = ifelse(V560177 %in% c(1), "married", marital_1),
[…]

### 2025-06-06 code/old/slope_analysis.R

@@ -1,693 +1,0 @@
-
-library(lme4)
-
-source("./code/clean_panels.R")
-
-canes5 <- canes5 %>% filter(var != "clsrprvtpwr")
-vars <- unique(canes5$var)
-group_meansc5 <- vector(mode = "list", length = length(vars))
-variancec5 <- vector(mode = "list", length = length(vars))
-for (i in 1:length(vars)) {
-  v <- vars[i]
-  df <- canes5 %>%
-    filter(var == v) %>%
-    pivot_longer(y1:y3) %>%
-    filter(!is.na(value), !is.na(age_1)) %>%
-    group_by(id) %>% filter(n() > 1) %>%
-    ungroup() %>%
-    mutate(year = as.numeric(gsub("y", "", name))) %>%
-    select(-name) %>%
-    mutate(value = (value - mean(value,na.rm=TRUE))/(sd(value,na.rm=TRUE))) %>%
-    mutate(year = (year - 1)*2) %>%
-    mutate(group = ifelse(age_1 <= 25, "18-25",
-                          ifelse(age_1 >= 26 & age_1 <= 33, "26-33",
-                                 "34+"))) 
-  #Estimate the effect of one year on standardized response.
-  m1 <- lmer(value ~ year + (1 + year|id), data = df)
-  #Extract individual year effect
-  cm1 <- coef(m1)
-  df$yhat <- predict(m1)
-  #Put them in data frame
-  effects <- data.frame(cm1$id) %>% 
-    select(-X.Intercept.) %>%
-    rownames_to_column(var = "id") %>%
-    mutate(id = as.numeric(id)) %>%
-    mutate(slope = year) %>% select(-year)
-  #Link to ids/groups
-  df <- df %>%
-    left_join(effects, by = c("id"="id"))
-  group_meansc5[[i]] <- df %>% 
-    group_by(id) %>% 
-    slice(1) %>%
-    group_by(group) %>%
-    summarise(n = n(), abs_mean_slope = mean(abs(slope),na.rm=TRUE)) %>%
-    mutate(var = v)
-  
-  variancec5[[i]] <- df %>%
-    mutate(resid = (value - yhat)^2) %>%
-    group_by(group) %>%
-    summarise(ssr = sum(resid), 
-              n_obs = n(),
-              n_id = length(unique(id))) %>%
-    mutate(mse = ssr/n_obs) %>%
-    mutate(var = v)
-  
-  
-}
-
[…]

### 2025-06-06 code/cleaning/clean_anes_16_20.R
workwhts

@@ -270,10 +270,10 @@ clean_anes_16_20 <- function(path) {
            moralrel_4 = as.character(V202264),
            tradfamily_2 = as.character(V162210),
            tradfamily_4 = as.character(V202265),
-           hrdwrklazywhite_2 = as.character(V162345),
-           hrdwrklazywhite_4 = as.character(V202515),
-           hrdwrklazyblack_2 = as.character(V162346),
-           hrdwrklazyblack_4 = as.character(V202516),
+           workwhts_2 = as.character(V162345),
+           workwhts_4 = as.character(V202515),
+           workblks_2 = as.character(V162346),
+           workblks_4 = as.character(V202516),
            slavediff_2 = as.character(V162212),
            slavediff_4 = as.character(V202301),
            wrkwayup_2 = as.character(V162211),
@@ -299,7 +299,7 @@ clean_anes_16_20 <- function(path) {
                   ~as.numeric(.x)),
            across(c(partyid, polviews, jobguar,
                     govins, helpblk, spendserv,
-                    defscale, hrdwrklazywhite, hrdwrklazyblack),
+                    defscale, workwhts, workblks),
                   ~(.x-1)/6*100),
            across(c(stayhome, runfew,
                     natenvir, natsoc, natschools, natfare,
@@ -319,7 +319,7 @@ clean_anes_16_20 <- function(path) {
                     matterhrdwrk, 
                     dontcare), 
                   ~(.x-1)/4*100)) %>%
-    pivot_longer(c(abortion, attentioncpg:cappun, defscale:wastetax, worrylesseq:wrkwayup)) %>%
+    pivot_longer(c(abortion, attentioncpg:cappun, defscale:wastetax, workblks:wrkwayup)) %>%
     filter(!is.na(age), !is.na(value), !is.na(date)) %>% arrange(id, name, date) %>%
     mutate(df = "2016-20 ANES")
   

### 2025-06-06 code/cleaning/clean_anes_92_97.R

@@ -458,10 +458,10 @@ clean_anes_92_97 <- function(path) {
            tradfamily_2 = as.character(V926117), 
            tradfamily_4 = as.character(V941031), 
            tradfamily_7 = as.character(V961249), 
-           hrdwrklazywhite_2 = as.character(V926221), 
-           hrdwrklazywhite_7 = as.character(V961311), 
-           hrdwrklazyblack_2 = as.character(V926222), 
-           hrdwrklazyblack_7 = as.character(V961312), 
+           workwhts_2 = as.character(V926221), 
+           workwhts_7 = as.character(V961311), 
+           workblks_2 = as.character(V926222), 
+           workblks_7 = as.character(V961312), 
            slavediff_2 = as.character(V926129), 
            slavediff_4 = as.character(V941051),
            wrkwayup_2 = as.character(V926126), 
@@ -499,8 +499,8 @@ clean_anes_92_97 <- function(path) {
                   ~as.numeric(.x)),
            partyid = (partyid/6)*100,
            across(c(polviews, jobguar, govins, helpblk,
-                    spendserv, defscale, eqrole, hrdwrklazywhite, 
-                    hrdwrklazyblack),
+                    spendserv, defscale, eqrole, workwhts, 
+                    workblks),
                   ~(.x - 1)/6*100),
            across(c(trust, fair),
                   ~(.x-1)*100),
@@ -517,7 +517,7 @@ clean_anes_92_97 <- function(path) {
                     nathome, natchld, natfare, natcrime, natpoor, 
                     lessgvt),
                   ~(.x - 1)/2*100)) %>% 
-    pivot_longer(c(abortion, attentioncpg:crooked, defscale:wastetax, worrylesseq:wrkwayup)) %>%
+    pivot_longer(c(abortion, attentioncpg:crooked, defscale:wastetax, workblks:wrkwayup)) %>%
     filter(!is.na(age), !is.na(value), !is.na(date)) %>% arrange(id, name, date) %>%
     mutate(df = "1992-97 ANES")
   

### 2025-06-06 code/cleaning/clean_gss_06_10.R

@@ -30,6 +30,8 @@ clean_gss_06_10 <- function(path) {
            cappun_1, cappun_2, cappun_3, 
            letin1a_1, letin1a_2, letin1a_3,
            tax_1, tax_2, tax_3,
+           workblks_1, workblks_2, workblks_3,
+           workwhts_1, workwhts_2, workwhts_3,
            wrkwayup_1, wrkwayup_2, wrkwayup_3) %>%
     mutate(weight_2 = as.character(wtpannr12),
            weight_3 = as.character(wtpannr123),
@@ -51,6 +53,9 @@ clean_gss_06_10 <- function(path) {
                     fair_1, fair_2, fair_3,
                     helpful_1, helpful_2, helpful_3), 
                   ~ifelse(.x == 3, 1.5, .x)),
+           across(c(workblks_1, workblks_2, workblks_3,
+                    workwhts_1, workwhts_2, workwhts_3),
+                  ~ifelse(.x %in% c(8,9), NA, .x)),
            across(partyid_1:wrkwayup_3,
                   ~as.character(.x))) %>%
     select(-c(wtpannr12, wtpannr123, dateintv_1, dateintv_2, dateintv_3)) %>%
@@ -70,9 +75,10 @@ clean_gss_06_10 <- function(path) {
                   ~(.x-1)/2*100), 
            across(c(cappun), 
                   ~(.x)/2*100),
+           across(c(workblks, workwhts), ~(.x - 1)/6*100),
            across(c(letin1a, wrkwayup), 
                   ~(.x - 1)/4*100)) %>%
-    pivot_longer(c(cappun, fair:trust, wrkwayup)) %>%
+    pivot_longer(c(cappun, fair:trust, workblks:wrkwayup)) %>%
     filter(!is.na(age), !is.na(value), !is.na(date)) %>% arrange(id, name, date) %>%
     mutate(df = "2006-10 GSS")
   

### 2025-06-06 code/cleaning/clean_gss_08_12.R

@@ -30,6 +30,8 @@ clean_gss_08_12 <- function(path) {
            cappun_1, cappun_2, cappun_3, 
            letin1a_1, letin1a_2, letin1a_3,
            tax_1, tax_2, tax_3,
+           workblks_1, workblks_2, workblks_3,
+           workwhts_1, workwhts_2, workwhts_3,
            wrkwayup_1, wrkwayup_2, wrkwayup_3) %>%
     mutate(weight_2 = as.character(wtpannr12),
            weight_3 = as.character(wtpannr123),
@@ -50,6 +52,9 @@ clean_gss_08_12 <- function(path) {
            across(c(trust_1, trust_2, trust_3,
                     fair_1, fair_2, fair_3,
                     helpful_1, helpful_2, helpful_3), ~ifelse(.x == 3, 1.5, .x)),
+           across(c(workblks_1, workblks_2, workblks_3,
+                    workwhts_1, workwhts_2, workwhts_3),
+                  ~ifelse(.x %in% c(8,9), NA, .x)),
            across(c(tax_1, tax_2, tax_3), ~ifelse(.x %in% c(4), NA, .x)),
            across(partyid_1:wrkwayup_3,
                   ~as.character(.x))) %>%
@@ -71,9 +76,10 @@ clean_gss_08_12 <- function(path) {
                   ~(.x-1)/2*100), 
            across(c(cappun), 
                   ~(.x)/2*100), 
+           across(c(workblks, workwhts), ~(.x - 1)/6*100),
            across(c(letin1a, wrkwayup), 
                   ~(.x - 1)/4*100)) %>%
-    pivot_longer(c(cappun, fair:trust, wrkwayup)) %>%
+    pivot_longer(c(cappun, fair:trust, workblks:wrkwayup)) %>%
     filter(!is.na(age), !is.na(value), !is.na(date)) %>% arrange(id, name, date) %>%
     mutate(df = "2008-12 GSS")
   

### 2025-06-06 code/cleaning/clean_gss_10_14.R

@@ -32,6 +32,8 @@ clean_gss_10_14 <- function(path) {
            cappun_1, cappun_2, cappun_3, 
            letin1a_1, letin1a_2, letin1a_3,
            tax_1, tax_2, tax_3,
+           workblks_1, workblks_2, workblks_3,
+           workwhts_1, workwhts_2, workwhts_3,
            wrkwayup_1, wrkwayup_2, wrkwayup_3) %>%
     mutate(weight_2 = as.character(wtpannr12),
            weight_3 = as.character(WTPANNR123),
@@ -52,6 +54,9 @@ clean_gss_10_14 <- function(path) {
            across(c(trust_1, trust_2, trust_3,
                     fair_1, fair_2, fair_3,
                     helpful_1, helpful_2, helpful_3), ~ifelse(.x == 3, 1.5, .x)),
+           across(c(workblks_1, workblks_2, workblks_3,
+                    workwhts_1, workwhts_2, workwhts_3),
+                  ~ifelse(.x %in% c(8,9), NA, .x)),
            across(c(tax_1, tax_2, tax_3), ~ifelse(.x %in% c(4), NA, .x)),
            across(partyid_1:wrkwayup_3,
                   ~as.character(.x))) %>%
@@ -72,9 +77,10 @@ clean_gss_10_14 <- function(path) {
                   ~(.x-1)/2*100), 
            across(c(cappun), 
                   ~(.x)/2*100),
+           across(c(workblks, workwhts), ~(.x - 1)/6*100),
            across(c(letin1a, wrkwayup), 
                   ~(.x - 1)/4*100)) %>%
-    pivot_longer(c(cappun, fair:trust, wrkwayup)) %>%
+    pivot_longer(c(cappun, fair:trust, workblks:wrkwayup)) %>%
     filter(!is.na(age), !is.na(value), !is.na(date)) %>% arrange(id, name, date) %>%
     mutate(df = "2010-14 GSS")
   

### 2025-06-06 code/cleaning/clean_gss_16_20.R

@@ -30,6 +30,8 @@ clean_gss_16_20 <- function(path) {
            cappun_1a, cappun_1b, cappun_2, 
            letin1a_1a, letin1a_1b, letin1a_2,
            tax_1a, tax_1b, tax_2,
+           workblks_1a, workblks_1b, workblks_2,
+           workwhts_1a, workwhts_1b, workwhts_2,
            wrkwayup_1a, wrkwayup_1b, wrkwayup_2) %>%
     mutate(weight_2 = as.character(wtssnr_2)) %>%
     mutate(age_1a = ifelse(is.na(age_1a) & !is.na(age_2), age_2 - 4, age_1a),
@@ -67,6 +69,7 @@ clean_gss_16_20 <- function(path) {
                   ~as.numeric(.x)),
            partyid = (partyid/6)*100,
            polviews = (polviews-1)/6*100,
+           across(c(workblks, workwhts), ~(.x-1)/6*100),
            across(c(trust, fair, helpful),
                   ~(.x-1)*100),
            across(c(natenvir, nataid,
@@ -79,7 +82,7 @@ clean_gss_16_20 <- function(path) {
                   ~(.x)/2*100), 
            across(c(letin1a, wrkwayup), 
                   ~(.x - 1)/4*100)) %>%
-    pivot_longer(c(cappun, fair:trust, wrkwayup)) %>%
+    pivot_longer(c(cappun, fair:trust, workblks:wrkwayup)) %>%
     filter(!is.na(age), !is.na(value), !is.na(date)) %>% arrange(id, name, date) %>%
     mutate(df = "2016-20 GSS")
   
